---
title: "MOT143A: Business Analytics group 20"
author: "Hidde Segers and Lotte Feilzer"
date: "2026-03-04"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

# 1. Business context and analytics task

## 1.1 Business problem

**Context:** Airlines running domestic flights in the United States
(U.S.) must manage constantly resources such as gates, crew members,
ground handling staff, and customer service capacity. These decisions
have to be made quickly and often with limited certainty.

Flight delays and cancellations can cause a lot of trouble for those who
are trying to get on time. When things go wrong, airlines have to pay
for: crew overtime, gate conflicts, aircraft re-positioning,
compensation, re-booking logistics, and re-putational damage. At the
same time, customer satisfaction declines, which affects their name and
in the long-term their revenue.

**Problem:** Delays and cancellations are really expensive for airlines
and can be frustrating for passengers. A single delay can create high
costs for crew overtime, gate conflicts, re-booking workloads. Because
flights are connected in a daily schedule, one delay can also cause
follow-on delays later in the network. This increases operational costs
and directly reduces customer satisfaction.

The practical challenge is that a lot of operational decisions must be
made before the flights departures. At that time it is still not clear
whether the flight is running on time or not. If the airline can figure
out flights that are likely to be delayed early, it can allocate
resources more active (e.g., assign standby crew, prepare customer
service capacity or adjust gate plans) and reduce the impact when such
delays occurs.

This leads to our business question:

**Business question:** How can we estimate the probability that a
scheduled flight will be delayed (e.g. more than 15 min) before
departure?

The goal is not to "predict the truth" perfectly, but to provide a risk
signal that helps operations teams prioritize attention and resources
towards flights with the highest expected delay risk. From an
organizational perspective, the insights are meant for operations
control centers, network planners and day-of-operations managers. The
model would be used in real-time operational setting, where planners
monitor flights and allocate resources (that are most of the time
limited). In this context, the prediction functions as a
decision-support tool that can help prioritize attention and
interventions for flights with the highest risk of delay.

## 1.2 Analytics task

This analysis is focused on the perspective of the airline operations
control and the airport station management to identify high risk
flights. This should support the workers with allocation of staff, gates
and premature consumer communication. The model will be used for
pre-departure planning setting and will be based on the historical
operational data of US domestic flights. The business problem will be
translated to an analytic task by approaching it as a predictive
risk-scoring problem. Three possible targets were considered:

-   **Classification A (delay):**

    **Target:** `IS_DELAYED_15 = 1` als `ARR_DELAY >= 15`, anders 0.

    Here it is analysed whether or not a flight will be significantly
    delayed or not (this is for flights that are neither diverted or
    delayed).

-   **Classification B (cancellation):**

    **Target:** `CANCELLED == 1`

    Here it is analysed whether or not a flight will be cancelled.

-   **Regression:**

    **Target:** `ARR_DELAY`

    Here it is analysed how much too late a flight will likely be in
    minutes.

The primary task chosen is classification A, this is because it will be
directly usable to relocate resources if needed and where focus in
logistics needs to focused. It is important to press that the model
can't reason and does not rely on causality. It also will not be able to
guarantee outcomes. It will estimate the risk for significant delay and
consequently improve and support operational actions. Classification for
cancellation will be harder because of the rarity of this event. The
regression of the delay is also hard to predict and the type of delay
can but does not also translate to how much delay there will be.

## 1.3 Success criteria

We consider the success criteria for this report in two perspectives.
Firstly, model criteria that will be need to be met to answer our
business question. Secondly, the business criteria that will ensure
positive effects of the model in practice.

From an operational point of view, missing a real delay is usually worse
than raising a false alarm. If the model says a flight will be on time
but it is actually delayed, the airline is unprepared. This can quickly
lead to crew problems, gate conflicts, and unhappy passengers.

If the model predicts a delay that does not happen, it may lead to some
extra costs because resources were allocated unnecessarily. However,
this is less harmful than being unprepared for a real delay.

For this reason, we focus on high recall for the delayed flight s. We
want the model to catch as many real delays as possible, even if that
means accepting some false alarms. At the same time, we monitor
precision to make sure we do not allocate too many unnecessary
resources.

# 2. Setup

Here all packages are downloaded to make the code reproducible.

```{r setup}
# CRAN mirror
local({r <- getOption("repos")
r["CRAN"] <- "https://cran.rstudio.com/"
options(repos = r)})

N_READ <- -1

packages <- c("ggplot2", "Hmisc", "rsample", "gmodels", "C50", "randomForest",
  "factoextra", "lattice", "reshape2", "pROC"
)

for (i in packages) {
  if (!require(i, character.only = TRUE)) {
    install.packages(i, dependencies = TRUE)
    library(i, character.only = TRUE)
  }
}

set.seed(46748717)
```

# 3. Data loading

## 3.1 Data in `data/` folder

Here we read our data. In the code it is easy to change the year of data
we want to see, just change the variable `YEAR`. Because the data_files
are quite big we added the possibility to print only a specific amount
of rows as given in `nrows` a value of `-1` will give all rows.

```{r data_files}
data_dir <- "data"
YEAR <- 2018
csv_path <- file.path(data_dir, paste0(YEAR, ".csv"))
flights <- read.csv(csv_path, stringsAsFactors = FALSE, nrows = N_READ)
```

## 3.2 Read data

Now we read our actual data and print the dimensions and columns names
to get a first clue of what kind of data were dealing with.

```{r read_data}
dim(flights)
names(flights)
```

As seen the data file contains:

-   7,213,446 observations

-   28 variables

We see that each observation represents one commercial flight in 2018.
The variables are as follows and categorized in definitions that can be
used to refer to groups of variables further in the report:

-   **Flight identifiers:** FL_DATE, OP_CARRIER, OP_CARRIER_FL_NUM

-   **Route information:** ORIGIN, DEST

-   **Scheduled times:** CRS_DEP_TIME, CRS_ARR_TIME

-   **Actual flight performance:** DEP_TIME, ARR_TIME, DEP_DELAY,
    ARR_DELAY

-   **Operational metrics:** TAXI_OUT, TAXI_IN, AIR_TIME

-   **Delay causes:** CARRIER_DELAY, WEATHER_DELAY, NAS_DELAY,
    SECURITY_DELAY, LATE_AIRCRAFT_DELAY

-   **Flight status indicators:** CANCELLED, DIVERTED

The variables can be descriped into the following categories:

### Flight Identifiers

| Variable          | Description        | Variable type           |
|-------------------|--------------------|-------------------------|
| FL_DATE           | Date of the flight | character (date string) |
| OP_CARRIER        | Operating airline  | character               |
| OP_CARRIER_FL_NUM | Flight number      | integer / numeric       |

## Route information

| Variable | Description               | Variable type     |
|----------|---------------------------|-------------------|
| ORIGIN   | Departure airport         | character         |
| DEST     | Arrival airport           | character         |
| DISTANCE | Distance between airports | integer / numeric |

## Scheduled times

| Variable         | Description               | Variable type                   |
|------------------|---------------------------|---------------------------------|
| CRS_DEP_TIME     | Scheduled departure time  | integer / numeric (HHMM format) |
| CRS_ARR_TIME     | Scheduled arrival time    | integer / numeric (HHMM format) |
| CRS_ELAPSED_TIME | Scheduled flight duration | integer / numeric (minutes)     |

## Actual flight performance

| Variable  | Description           | Variable type                   |
|-----------|-----------------------|---------------------------------|
| DEP_TIME  | Actual departure time | integer / numeric (HHMM format) |
| ARR_TIME  | Actual arrival time   | integer / numeric (HHMM format) |
| DEP_DELAY | Departure delay       | integer / numeric (minutes)     |
| ARR_DELAY | Arrival delay         | integer / numeric (minutes)     |
| AIR_TIME  | Time spent in air     | integer / numeric (minutes)     |

## Operational metrics

| Variable   | Description              | Variable type                   |
|------------|--------------------------|---------------------------------|
| TAXI_OUT   | Taxi time before takeoff | integer / numeric (minutes)     |
| TAXI_IN    | Taxi time after landing  | integer / numeric (minutes)     |
| WHEELS_OFF | Takeoff time             | integer / numeric (HHMM format) |
| WHEELS_ON  | Landing time             | integer / numeric (HHMM format) |

## Flight status

| Variable  | Description                 | Variable type |
|-----------|-----------------------------|---------------|
| CANCELLED | Indicates cancelled flights | binary (0/1)  |
| DIVERTED  | Indicates diverted flights  | binary (0/1)  |

## Delay causes

| Variable | Description | Variable type |
|----|----|----|
| CARRIER_DELAY | Delay caused by airline | integer / numeric (minutes) |
| WEATHER_DELAY | Delay caused by weather | integer / numeric (minutes) |
| NAS_DELAY | Delay caused by air traffic system | integer / numeric (minutes) |
| SECURITY_DELAY | Delay caused by security | integer / numeric (minutes) |
| LATE_AIRCRAFT_DELAY | Delay caused by late incoming aircraft | integer / numeric (minutes) |

## 4.1 Data summary

Here we show a small summary of the data. The following things are
noticeable. Columns in this datafile that contain dates are characters
preferably we change this later to datetimes. It is also remarkable that
the times are given in military time but as integers, this will make it
hard to calculate certain metrics inveolving time such as "time passed"
therefor we will preferably also change these datatypes later on to
datetimes. For metrics such as `TAXI_OUT` and `DEP_DELAY` the integers
are suitable because it represents the amount of minutes passed. We see
that for the `DIVERTED` and `CANCELLED` columns `0` and `1` are used
instead of Booleans to respresent `FALSE` and `TRUE` respectively, this
is also workable if remarked. Let's further explore the data.

```{r explore_basic}
str(flights)
```

To align this notebook more closely with the workbook on exploration and
preparation, we explicitly separate three questions:

-   What is the type and scale of each variable?

-   Which variables show data quality issues or unusual structure?

-   Which of those issues require action before modelling?

The next sections answer these questions step by step. We focus on
variables that are most relevant for a pre-departure delay analysis,
rather than producing every possible plot for all 28 columns.

```{r selected_numeric_diagnostics}
numeric_vars <- c("DISTANCE", "CRS_ELAPSED_TIME", "DEP_DELAY", "ARR_DELAY")
numeric_vars <- intersect(numeric_vars, names(flights))

numeric_diagnostics <- data.frame(
  variable = numeric_vars,
  mean = sapply(flights[numeric_vars], function(x) round(mean(x, na.rm = TRUE), 2)),
  median = sapply(flights[numeric_vars], function(x) round(median(x, na.rm = TRUE), 2)),
  sd = sapply(flights[numeric_vars], function(x) round(sd(x, na.rm = TRUE), 2)),
  min = sapply(flights[numeric_vars], function(x) round(min(x, na.rm = TRUE), 2)),
  max = sapply(flights[numeric_vars], function(x) round(max(x, na.rm = TRUE), 2))
)

numeric_diagnostics
```

This table functions as our first univariate numeric exploration step.
In workbook terms, we compare central tendency and spread before we look
at plots. The main thing to watch for is a large gap between mean and
median, or an extreme maximum relative to the bulk of the data, because
that signals skewness or potentially influential observations.

```{r selected_numeric_plots}
set.seed(46748717)
idx_diag <- sample(seq_len(nrow(flights)), min(200000, nrow(flights)))
flights_diag <- flights[idx_diag, ]

par(mfrow = c(2, 2))
hist(flights_diag$DISTANCE, breaks = 40, main = "Distance", xlab = "Miles", col = "grey80")
boxplot(flights_diag$DISTANCE, horizontal = TRUE, main = "Distance boxplot", col = "grey80")
hist(flights_diag$ARR_DELAY, breaks = 40, main = "Arrival delay", xlab = "Minutes", col = "grey80")
boxplot(flights_diag$ARR_DELAY, horizontal = TRUE, main = "Arrival delay boxplot", col = "grey80")
par(mfrow = c(1, 1))
```

We keep this visual check selective. `DISTANCE` is a core pre-departure
predictor and should be inspected as a regular numeric feature.
`ARR_DELAY` is not a predictor in the pre-departure setting, but it is
the outcome we eventually simplify into a binary target. Inspecting its
raw distribution helps justify why a classification target is more
practical here than modelling raw delay minutes directly: the delay
distribution is highly skewed and dominated by a long right tail.

## 4.2 Missing values overview and discussion

Here we check whether or not we have missing values in our dataset, how
many there are and what are the possible reasons for these missing
values are. We use logical reasoning in order to determine what to do
with potential missing data. Here we show a table that shows the amount
of NA values for each column and also a percentage of how many values
are missing in regards to the total dataset.

```{r missingness}
na_count <- sapply(flights, function(x) sum(is.na(x)))
na_pct   <- round(100 * na_count / nrow(flights), 2)

na_table <- data.frame(
  na_count = as.integer(na_count),
  na_pct = na_pct
)

na_table <- na_table[order(-na_table$na_pct), ]
na_table
```

Firstly we see a column `Unnamed..27` which seems to have only NA values
making the column absolute and removable. We see for the operational
variables and flight performance variables such as `DEP_TIME`,
`ARR_TIME`, `AIR_TIME`, `WHEELS_ON`, `WHEELS_OFF` that 1-2% of all
values are missing. We should note that all arrival variables have about
the same amount of NA values missing and idem for the departure values.
This probably means that the NA values are probably/hopefully
explainable for the same reasons. First we check whether the NAs
occurring in the delay rows all occur in the same row. The cell below
tells us this is the case. This lets us know that these `DELAY` causes
variables are jointly observed or missing. This is good sign. These rows
are supposed to be filled in after a flight and give a reason for the
delay if it was significant delayed (15 minutes or more). In our check
this also seems to be the case. All flights that are 15+ minutes delayed
have a cause and vice versa.

```{r delay-na-same}
cause_cols <- intersect(c("CARRIER_DELAY","WEATHER_DELAY",
                    "NAS_DELAY","SECURITY_DELAY","LATE_AIRCRAFT_DELAY"), names(flights))
delay15 <- !is.na(flights$ARR_DELAY) & flights$ARR_DELAY >= 15
cause_present <- !is.na(flights[[cause_cols[1]]])

miss <- is.na(flights[, cause_cols, drop=FALSE])
same_na_rows <- all(apply(miss, 2, function(x) identical(which(x), which(miss[,1]))))

cat("All the delay causes with NA values are for the same row, is", same_na_rows, "\n")
cat("Causes are present but there was no 15 min delay:", sum(cause_present & !delay15), "\n")
cat("More than 15 min delay but cause is NA value:", sum(delay15 & !cause_present))
```

Now we check if it possible for a delay flight to have multiple reasons
for delay this might handy to know later on. This seems to be possible
and is the case for 51.56% of the delayed flights.

```{r cause-compact}
k <- rowSums(flights[cause_present, cause_cols, drop=FALSE] > 0)
tab <- table(k)

multi_n <- sum(k > 1)
total_n <- length(k)
multi_pct <- round(100 * multi_n / total_n, 2)

cat("Of the flights with delay causes,", multi_pct, "% have multiple causes for delay.")
```

We now know that our delay causes columns are in fact complete. We can
now check if for the flights that have taken off all the important
operational and performance variables are present (`ARR_DELAY`,
`ARR_TIME` ect.) or that some or all values are missing for certain
rows. The NA values for the these rows as presented above could be
explained if these are the rows of flights that were diverted or
cancelled. If not, crucial information might be missing. If so, we
analyse this here.

```{r}
flights_wo_cancelled <- which(flights$CANCELLED == 0 & flights$DIVERTED == 0)
flights_wo_cancelled_df <- flights[flights_wo_cancelled, ]

na_count_wo_cd <- sapply(flights_wo_cancelled_df, function(x) sum(is.na(x)))
na_pct_wo_cd   <- round(100 * na_count_wo_cd / nrow(flights_wo_cancelled_df), 2)

na_table_wo_cd <- data.frame(
  variable = names(flights_wo_cancelled_df),
  na_count_wo_cd = as.integer(na_count_wo_cd),
  na_pct_wo_cd = na_pct_wo_cd
)

na_table_wo_cd <- na_table_wo_cd[order(-na_table_wo_cd$na_pct_wo_cd), ]
na_table_wo_cd
```

We can see that a flight that was either cancelled or diverted, it
accounts for the reasons that rows contain NA values for performance and
operational variables. We can see that the only two performance
variables that have still NA values are `ARR_DELAY` and `DEP_DELAY`,
why? These NA values might cause some trouble when modelling because we
use `ARR_DELAY` as predictors in the pre-departure model. Let's check by
showing the rows were this data is missing.

```{r check-missing-mostly-delays}
dep_delays_missing <- which(is.na(flights_wo_cancelled_df$DEP_DELAY)) 
arr_delays_missing <- which(is.na(flights_wo_cancelled_df$ARR_DELAY))
missing_delays <- flights_wo_cancelled_df[c(dep_delays_missing, arr_delays_missing), ]
head(missing_delays)
```

First we need to look into what we want to do with the rows in which the
`DEP_DELAY` and/or `ARR_DELAY` variables are missing. It seems that
there are only NA values for these columns when the `CRS_DEP_TIME` and
`DEP_TIME` have the same value and idem for the arrival. This means that
in this dataset the NA values just mean that plane left exactly at the
intended time. If this is the case we can replace the NA values in these
rows for 0 values and these NA values are also explained than as well.
We perform a quick check to see this is a 100 percent the case.

```{r fill_zero_delays}
dep_na <- !is.na(missing_delays$CRS_DEP_TIME) & !is.na(missing_delays$DEP_TIME) &
                 is.na(missing_delays$DEP_DELAY) &
                 missing_delays$CRS_DEP_TIME == missing_delays$DEP_TIME

arr_na <- !is.na(missing_delays$CRS_ARR_TIME) & !is.na(missing_delays$ARR_TIME) &
                 is.na(missing_delays$ARR_DELAY) &
                 missing_delays$CRS_ARR_TIME == missing_delays$ARR_TIME

cat("DEP_DELAY NA but not CRS_DEP_TIME==DEP_TIME:", sum(dep_na), "\n")
cat("ARR_DELAY NA but not CRS_ARR_TIME==ARR_TIME:", sum(arr_na))
```

Now we have analysed all NA values and begin to clean our data properly.

The workbook logic is visible here: we do not immediately delete all
rows with missing values, because in operational data missingness is
often informative. We first distinguish between:

-   structurally missing values, for example delay causes that are only
    recorded after substantial delays;

-   outcome-related missing values, for example flights that were
    cancelled or diverted and therefore have no realised arrival
    metrics;

-   repairable missing values, for example cases where scheduled time
    and realised time are equal and the delay can logically be set to 0.

This distinction matters because it prevents us from throwing away valid
flights too early and helps us explain each cleaning choice.

## 4.3 Cleaning and datatypes

First, we do some final sanity checks for the values for variables. We
check whether the distances are positive as well as airtime. We also
check whether the departure and arrival delays are plausible which all
seems to be the case.

```{r sanity_checks}
cat("There are", sum(flights$DISTANCE <= 0), "negative distances.", "\n")
cat("There are", sum(flights$AIR_TIME <= 0, na.rm = TRUE) , "negative airtimes.", "\n")
print("Departure delay summary:")
summary(flights$DEP_DELAY) 
print("Arrival delay summary:")
summary(flights$ARR_DELAY) 
```

We create a df named `flights_clean` to not overwrite the `flights`
variable that can be cleaned without overwriting our original data.

```{r create_flights_clean}
flights_clean <- flights
```

We begin with adjusting in accordance with the remarks made in section
3.2. We the `FL_DATE` column a datatype datetime and drop the unuseful
column `Unnamed..27` and ensure that the columns are the correct
datatypes.

```{r clean_types}
drop_cols <- grep("^Unnamed", names(flights_clean), value = TRUE)
flights_clean[drop_cols] <- NULL

flights_clean$FL_DATE <- as.Date(flights_clean$FL_DATE)

head(flights_clean, 4)
```

Now we do some minimal cleaning make sure the datatypes are correct. We
make the integer time columns POSIXct datetime columns to be able to
treat the time correctly.

```{r datetime_fix_overwrite}
hhmm_to_hhmm_str <- function(x) {
  x <- as.integer(x)
  hh <- x %/% 100
  mm <- x %% 100
  out <- sprintf("%02d:%02d", hh, mm)
  out[is.na(x) | mm > 59 | hh < 0 | hh > 29] <- NA
  out
}

time_cols <- intersect(c("WHEELS_OFF","WHEELS_ON","DEP_TIME","ARR_TIME","CRS_DEP_TIME","CRS_ARR_TIME"), names(flights_clean))

# Replace columns to POSIXct
for (cn in time_cols) {
  tmp_str <- hhmm_to_hhmm_str(flights_clean[[cn]])
  flights_clean[[cn]] <- as.POSIXct(
    paste(as.character(flights_clean$FL_DATE), tmp_str),
    format = "%Y-%m-%d %H:%M",
    tz = "UTC"
  )
}

# Overnight: is ARR_TIME < DEP_TIME + one day
if (all(c("DEP_TIME","ARR_TIME") %in% names(flights_clean))) {
  overnight <- !is.na(flights_clean$DEP_TIME) &
               !is.na(flights_clean$ARR_TIME) &
               (flights_clean$ARR_TIME < flights_clean$DEP_TIME)

  arrival_side <- intersect(c("ARR_TIME","WHEELS_ON","CRS_ARR_TIME"), time_cols)
  for (cn in arrival_side) {
    flights_clean[[cn]][overnight] <- flights_clean[[cn]][overnight] + 24*3600
  }
}

head(flights_clean[, time_cols, drop=FALSE], 5)
```

We now clean the NA values for the delay columns when the `CRS_DEP_TIME`
and `DEP_TIME` are the same and idem for the arrival times.

```{r impute_delays_posixct}
idx_dep0 <- which(
  is.na(flights_clean$DEP_DELAY) &
  !is.na(flights_clean$DEP_TIME) & !is.na(flights_clean$CRS_DEP_TIME) &
  flights_clean$DEP_TIME == flights_clean$CRS_DEP_TIME
)

idx_arr0 <- which(
  is.na(flights_clean$ARR_DELAY) &
  !is.na(flights_clean$ARR_TIME) & !is.na(flights_clean$CRS_ARR_TIME) &
  flights_clean$ARR_TIME == flights_clean$CRS_ARR_TIME
)

# vullen
flights_clean$DEP_DELAY[idx_dep0] <- 0
flights_clean$ARR_DELAY[idx_arr0] <- 0

# check
head(flights_clean[idx_dep0, c("CRS_DEP_TIME","DEP_TIME","DEP_DELAY")])
head(flights_clean[idx_arr0, c("CRS_ARR_TIME","ARR_TIME","ARR_DELAY")])
```

Now we check if all the data is clean. First we delete all the cancelled
and diverted rows because theserows are not relevant for our analysis.

```{r}
keep <- (is.na(flights_clean$CANCELLED) | flights_clean$CANCELLED == 0) &
        (is.na(flights_clean$DIVERTED)  | flights_clean$DIVERTED  == 0)

flights_clean <- flights_clean[keep, , drop = FALSE]

cat("Removed:", sum(!keep), "\n")
cat("Rows after:", nrow(flights_clean))
```

We decide to add some columns that make our analysis and modelling
easier later on, such as `IS_DELAYED` day of the week column which might
be interesting in the analysis and a column which contains a Boolean on
whether a flight is significantly delayed (15 minutes or later).Now we
add a column in which we see if a plane was significantly delayed. We
also add a `MONTH` and `WEEKDAY` column.

```{r}
flights_clean$IS_DELAYED <- ifelse(flights_clean$ARR_DELAY >= 15, TRUE, FALSE)
flights_clean$WEEKDAY <- weekdays(flights_clean$FL_DATE)
flights_clean$MONTH <- as.integer(format(flights_clean$FL_DATE, "%m"))
flights_clean$SCHED_DEP_HOUR <- as.numeric(format(flights_clean$CRS_DEP_TIME, "%H")) +
  as.numeric(format(flights_clean$CRS_DEP_TIME, "%M")) / 60
flights_clean$SCHED_ARR_HOUR <- as.numeric(format(flights_clean$CRS_ARR_TIME, "%H")) +
  as.numeric(format(flights_clean$CRS_ARR_TIME, "%M")) / 60
```

The target is now based on `ARR_DELAY >= 15`. This is the cleanest
definition for the business question because it captures whether the
flight ultimately arrived with a meaningful disruption. We remove
cancelled and diverted flights before defining this target, so the class
labels are only assigned to completed flights.

```{r cleaning_summary}
cleaning_summary <- data.frame(
  step = c(
    "Original rows",
    "After removing cancelled/diverted flights",
    "Rows with defined delay target"
  ),
  n_rows = c(
    nrow(flights),
    nrow(flights_clean),
    sum(!is.na(flights_clean$IS_DELAYED))
  )
)

cleaning_summary$share_of_original <- round(
  100 * cleaning_summary$n_rows / cleaning_summary$n_rows[1], 2
)

cleaning_summary
```

The table above makes the cleaning impact transparent. This is useful in
the workbook context because it shows that our later analysis is based
on a clearly defined subset of operationally comparable flights, rather
than on an unclear mix of completed, cancelled and diverted flights.

## 4.4 Exploration log and preparation decisions

The workbook stresses that exploration should lead to explicit
preparation choices. We therefore summarize our main findings in a short
exploration log.

```{r exploration_log}
exploration_log <- data.frame(
  variable_or_group = c(
    "Unnamed columns",
    "Cancelled/diverted flights",
    "Departure/arrival times",
    "Delay fields",
    "Carrier/origin categories",
    "Distance and elapsed time"
  ),
  observation = c(
    "Unnamed column contains only missing values",
    "Operational outcome fields are often missing for these flights",
    "Stored as HHMM integers rather than usable times",
    "Some zero-delay cases stored as NA when scheduled and actual times match",
    "High-cardinality categories with uneven group sizes",
    "Numeric variables look plausible but can contain influential extremes"
  ),
  action = c(
    "Drop",
    "Exclude from completed-flight modelling sample",
    "Convert to POSIXct and derive scheduled hour features",
    "Impute those cases with 0 when the equality check supports it",
    "Use volume filters in visual ranking and factor encoding in models",
    "Keep, but inspect visually and avoid unnecessary transformation unless analysis needs it"
  ),
  stringsAsFactors = FALSE
)

exploration_log
```

```{r duplicate_check}
dup_key <- c("FL_DATE", "OP_CARRIER", "OP_CARRIER_FL_NUM", "ORIGIN", "DEST", "CRS_DEP_TIME")
dup_key <- intersect(dup_key, names(flights_clean))

duplicate_rows <- duplicated(flights_clean[, dup_key])

data.frame(
  key_columns_checked = paste(dup_key, collapse = ", "),
  n_duplicate_rows = sum(duplicate_rows),
  pct_duplicate_rows = round(100 * mean(duplicate_rows), 4)
)
```

This duplicate check follows the workbook logic that cleaning is not
only about missing values. In this data set, the practical question is
whether two rows appear to describe the same scheduled flight. If this
count is small or zero, that supports the assumption that the unit of
analysis is one row per flight. If it is not, the duplicates need manual
inspection before modelling.

Some workbook preparation steps are not needed here. Data integration is
not relevant yet because we currently work with one main flight file.
Likewise, we do not apply global standardization at this stage because
our current models are tree-based and therefore do not require all
predictors to be on the same scale. Where scaling does matter, such as
in PCA, we explicitly do it there.

# 5. Exploratory visualisation

First we have a look at the base rate, this is the amount of how many
flights are more than 15 minutes delayed

```{r mean_delay_rate}
delay_rate <- mean(flights_clean$IS_DELAYED)
delay_rate 
```

First we show a bar chart of our data that simply displays the amount of
flights that are delayed and on-time. We see that around 18% of flights
are significantly delayed.

```{r}
df_target <- data.frame(
  class = factor(
    ifelse(flights_clean$IS_DELAYED, "Delayed (15 min)", "On time"),
    levels = c("On time", "Delayed (15 min)")
  )
)

ggplot(df_target, aes(x = class)) +
  geom_bar() +
  labs(title = "Class distribution: delayed vs on-time", x = "", y = "amount of flights") +
  coord_flip()
```

This class balance matters for the rest of the notebook. A delay rate of
roughly 18% means the data is not perfectly balanced, but the delayed
class is still common enough to model without the extreme rarity issues
that would occur for cancellations.

```{r numeric_vs_target_boxplots}
set.seed(46748717)
idx_box <- sample(seq_len(nrow(flights_clean)), min(150000, nrow(flights_clean)))
flights_box <- flights_clean[idx_box, ]

par(mfrow = c(1, 2))
boxplot(SCHED_DEP_HOUR ~ IS_DELAYED,
        data = flights_box,
        outline = FALSE,
        names = c("On time", "Delayed"),
        main = "Scheduled departure hour by delay status",
        ylab = "Scheduled departure hour",
        col = c("grey80", "tomato"))

boxplot(CRS_ELAPSED_TIME ~ IS_DELAYED,
        data = flights_box,
        outline = FALSE,
        names = c("On time", "Delayed"),
        main = "Scheduled elapsed time by delay status",
        ylab = "Scheduled elapsed time (minutes)",
        col = c("grey80", "tomato"))
par(mfrow = c(1, 1))
```

This figure adds the workbook perspective of comparing a numerical
variable against a categorical one. It helps us see whether delayed and
on-time flights differ in their planned characteristics before we start
modelling. We suppress plotted outlier points here because the sample is
very large; otherwise the boxes become unreadable. The distribution
itself remains visible through the median and spread.

These two plots show delay rate per month and the number of flights per
month. The main thing we see is that delays go up in the summer months.
There are also a bit more flights in that period, but the rise in delay
rate is larger than the rise in volume, so seasonality seems to matter.

```{r}
df_month_rate <- aggregate(IS_DELAYED ~ MONTH, flights_clean, function(x) mean(x, na.rm = TRUE))
df_month_n    <- aggregate(IS_DELAYED ~ MONTH, flights_clean, length)
names(df_month_n)[2] <- "n"

df_month <- merge(df_month_rate, df_month_n, by = "MONTH")
names(df_month)[2] <- "delay_rate"

pct_label <- function(x) paste0(round(100*x, 1), "%")

ggplot(df_month, aes(x = MONTH, y = delay_rate)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(labels = pct_label) +
  coord_cartesian(ylim = c(0, 0.30)) +
  labs(title = "Delay rate per month", x = "Month", y = "Delay rate")

ggplot(df_month, aes(x = MONTH, y = n)) +
  geom_col() +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Amount of flights per month", x = "Month", y = "#")
```

The weekday pattern is fairly stable. Most days sit close together, with
slightly higher delay rates on Thursday and Friday and a lower rate on
Saturday. So weekday does matter a bit, but not as strongly as some
other variables. That suggests it makes sense to keep weekday in the
analysis, but not to expect it to explain delays on its own.

```{r}
weekday_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

df_wday_rate <- aggregate(IS_DELAYED ~ WEEKDAY, flights_clean, mean)
names(df_wday_rate)[2] <- "delay_rate"
df_wday_rate$WEEKDAY <- factor(df_wday_rate$WEEKDAY, levels = weekday_levels)
df_wday_rate <- df_wday_rate[order(df_wday_rate$WEEKDAY), ]

ggplot(df_wday_rate, aes(x = WEEKDAY, y = delay_rate)) +
  geom_col() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Delay rate per weekday", x = "Day of the week", y = "Delay rate")

```

Heat map: weekday x departure hour

```{r}
df_hw <- flights_clean[, c("WEEKDAY", "SCHED_DEP_HOUR", "IS_DELAYED")]
df_hw$DEP_HOUR_BIN <- floor(df_hw$SCHED_DEP_HOUR)
df_hw <- aggregate(IS_DELAYED ~ WEEKDAY + DEP_HOUR_BIN, df_hw, mean)
names(df_hw)[3] <- "delay_rate"
df_hw$WEEKDAY <- factor(df_hw$WEEKDAY, levels = weekday_levels)

ggplot(df_hw, aes(x = DEP_HOUR_BIN, y = WEEKDAY, fill = delay_rate)) +
  geom_tile(color = "white") +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  scale_fill_gradient(low = "white", high = "steelblue", labels = pct_label, limits = c(0, 0.4)) +
  labs(title = "Delay rate by weekday and departure hour",
       x = "Planned departure hour", y = "Day of the week", fill = "Delay rate")

```

This heatmap is more useful than the weekday bar chart on its own,
because it shows when delay risk builds up during the day. If later
flights are more often delayed, that can point to delays stacking up as
the day goes on.

Delay rate per airline

```{r}
MIN_N_CARRIER <- 5000

carrier_rate <- aggregate(IS_DELAYED ~ OP_CARRIER, flights_clean, mean)
carrier_n    <- aggregate(IS_DELAYED ~ OP_CARRIER, flights_clean, length)
names(carrier_n)[2] <- "n"

carrier_summary <- merge(carrier_rate, carrier_n, by = "OP_CARRIER")
names(carrier_summary)[2] <- "delay_rate"

carrier_summary <- subset(carrier_summary, n >= MIN_N_CARRIER)
carrier_summary <- carrier_summary[order(-carrier_summary$delay_rate), ]

top_carriers <- head(carrier_summary, 10)

ggplot(top_carriers, aes(x = reorder(OP_CARRIER, delay_rate), y = delay_rate)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Delay rate (>=15m) per airline (top 10, min volume filter)",
       x = "Airplane", y = "Delay rate")
```

We use a minimum-volume filter before ranking carriers and airports.
Without that, very small groups can end up at the top just because they
have only a few flights. For this analysis we want stable patterns, not
coincidental extremes.

Delay rate per airport

```{r}
MIN_N_AIRPORT <- 20000

origin_rate <- aggregate(IS_DELAYED ~ ORIGIN, flights_clean, mean)
origin_n    <- aggregate(IS_DELAYED ~ ORIGIN, flights_clean, length)
names(origin_n)[2] <- "n"

origin_summary <- merge(origin_rate, origin_n, by = "ORIGIN")
names(origin_summary)[2] <- "delay_rate"

origin_summary <- subset(origin_summary, n >= MIN_N_AIRPORT)
origin_summary <- origin_summary[order(-origin_summary$delay_rate), ]

top_origins <- head(origin_summary, 15)

ggplot(top_origins, aes(x = reorder(ORIGIN, delay_rate), y = delay_rate)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Delay rate (>=15m) per origin airport (top 15, min volume filter)",
       x = "Origin airport", y = "Delay rate")
```

Scatterplot

```{r}
origin_top50_vol <- origin_summary[order(-origin_summary$n), ][1:50, ]

ggplot(origin_top50_vol, aes(x = n, y = delay_rate)) +
  geom_point() +
  scale_x_log10() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Origin airports: volume vs delay rate (top 50 volume)",
       x = "Aantal vluchten (log schaal)", y = "Delay rate")
```

This scatterplot helps us separate volume from risk. A large airport
with a high delay rate matters more than a very small airport with a
high but unstable rate, because it affects many more flights.

Next we look at two simple relationships: distance and planned departure
time.

```{r}
set.seed(46748717)
N_VIS <- min(200000, nrow(flights_clean))
idx_vis <- sample(seq_len(nrow(flights_clean)), N_VIS)
flights_vis <- flights_clean[idx_vis, ]

distance_breaks <- pretty(flights_vis$DISTANCE, n = 12)
flights_vis$DISTANCE_BIN <- cut(
  flights_vis$DISTANCE,
  breaks = unique(distance_breaks),
  include.lowest = TRUE
)

distance_risk <- aggregate(IS_DELAYED ~ DISTANCE_BIN, flights_vis, mean)
distance_mid  <- aggregate(DISTANCE ~ DISTANCE_BIN, flights_vis, mean)
distance_n    <- aggregate(IS_DELAYED ~ DISTANCE_BIN, flights_vis, length)

names(distance_risk)[2] <- "delay_rate"
names(distance_mid)[2]  <- "distance_mid"
names(distance_n)[2]    <- "n"

distance_risk <- merge(distance_risk, distance_mid, by = "DISTANCE_BIN")
distance_risk <- merge(distance_risk, distance_n, by = "DISTANCE_BIN")
distance_risk <- subset(distance_risk, n >= 100)
distance_risk <- distance_risk[order(distance_risk$distance_mid), ]

print(
  ggplot(distance_risk, aes(x = distance_mid, y = delay_rate)) +
    geom_line() +
    geom_point() +
    scale_y_continuous(labels = pct_label, limits = c(0, 1)) +
    labs(title = "Delay risk vs distance",
         x = "Distance (miles)", y = "Delay risk")
)

flights_vis$DEP_HOUR_BIN <- floor(flights_vis$SCHED_DEP_HOUR)

hour_risk <- aggregate(IS_DELAYED ~ DEP_HOUR_BIN, flights_vis, mean)
hour_n    <- aggregate(IS_DELAYED ~ DEP_HOUR_BIN, flights_vis, length)

names(hour_risk)[2] <- "delay_rate"
names(hour_n)[2]    <- "n"

hour_risk <- merge(hour_risk, hour_n, by = "DEP_HOUR_BIN")
hour_risk <- subset(hour_risk, n >= 100)
hour_risk <- hour_risk[order(hour_risk$DEP_HOUR_BIN), ]

print(
  ggplot(hour_risk, aes(x = DEP_HOUR_BIN, y = delay_rate)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = 0:23) +
    scale_y_continuous(labels = pct_label, limits = c(0, 1)) +
    labs(title = "Delay risk vs planned departure time",
         x = "Scheduled departure hour", y = "Delay risk")
)
```

These plots are just meant to explore the pattern. They help us decide
which variables are worth keeping and how to code them. Scheduled
departure time clearly works better as an hour variable than as the raw
HHMM value.

```{r route_risk_summary}
route_summary <- aggregate(IS_DELAYED ~ ORIGIN + DEST, flights_clean, mean)
route_n <- aggregate(IS_DELAYED ~ ORIGIN + DEST, flights_clean, length)
names(route_summary)[3] <- "delay_rate"
names(route_n)[3] <- "n"

route_summary <- merge(route_summary, route_n, by = c("ORIGIN", "DEST"))
route_summary <- subset(route_summary, n >= 3000)
route_summary <- route_summary[order(-route_summary$delay_rate), ]

head(route_summary, 10)
```

This route table adds one more useful view. It shows whether high delay
rates mainly belong to certain airports in general, or to specific
origin-destination combinations.

# 6. Dimension reduction and clustering

PCA and clustering are exploratory steps. We only use them if we can
build a sensible feature set with variables that are known before
departure. If we put in raw times, airport labels, or post-event
variables, the output becomes much less useful.

## 6.1 PCA

We use PCA here to get a quick view of the structure in the data. The
first step is not choosing components yet, but deciding which variables
should even go into PCA and clustering.

```{r pca_cluster_screening}
structure_screen <- data.frame(
  variable = c(
    "FL_DATE",
    "OP_CARRIER",
    "OP_CARRIER_FL_NUM",
    "ORIGIN",
    "DEST",
    "DISTANCE",
    "CRS_DEP_TIME",
    "CRS_ARR_TIME",
    "CRS_ELAPSED_TIME",
    "DEP_TIME / ARR_TIME / WHEELS_*",
    "DEP_DELAY / ARR_DELAY",
    "Delay-cause columns",
    "CANCELLED / DIVERTED",
    "MONTH",
    "WEEKDAY"
  ),
  pca_role = c(
    "Transform",
    "Post hoc only",
    "Exclude",
    "Post hoc only",
    "Post hoc only",
    "Keep",
    "Transform",
    "Transform",
    "Keep",
    "Exclude",
    "Exclude",
    "Exclude",
    "Exclude",
    "Transform",
    "Transform"
  ),
  clustering_role = c(
    "Transform",
    "Post hoc only",
    "Exclude",
    "Post hoc only",
    "Post hoc only",
    "Keep",
    "Transform",
    "Transform",
    "Watch for redundancy",
    "Exclude",
    "Exclude",
    "Exclude",
    "Exclude",
    "Transform",
    "Transform"
  ),
  representation = c(
    "Do not use raw date; derive month/weekday information instead",
    "No dummy expansion in unsupervised step",
    "Identifier-like flight number; not analytically meaningful",
    "Retain for post hoc interpretation, not as dummy set",
    "Retain for post hoc interpretation, not as dummy set",
    "Use as numeric route-length feature",
    "Convert to cyclical sin/cos features",
    "Convert to cyclical sin/cos features",
    "Use as numeric schedule-length feature",
    "Known only after departure/arrival",
    "Leakage / target-adjacent outcome information",
    "Recorded after the event and therefore leakage",
    "Used only as sample filter, not as features",
    "Convert to cyclical sin/cos features",
    "Represent coarsely as weekend dummy"
  ),
  reason = c(
    "Raw dates create arbitrary scale differences",
    "One-hot encoding carriers would add dimensions and dominate structure",
    "Mostly identifies a flight record rather than a latent trait",
    "Airport dummies would mainly recover airport labels rather than shared structure",
    "Airport dummies would mainly recover airport labels rather than shared structure",
    "Operationally meaningful and known before departure",
    "Raw HHMM values are not valid linear distances",
    "Raw HHMM values are not valid linear distances",
    "Operationally meaningful and known before departure",
    "Post-event information must be excluded",
    "Would force PCA/clustering to recover realised delay instead of latent structure",
    "Pure leakage for pre-departure analysis",
    "These variables already define which rows remain in the analysis",
    "Month is cyclical rather than linear",
    "A low-dimensional weekend indicator is preferable to seven weekday dummies"
  ),
  stringsAsFactors = FALSE
)

structure_screen
```

This screening leads to two main choices. First, we do not put raw
carrier and airport labels into PCA or clustering, because they would
mostly make the analysis split on labels. Second, time variables need to
be transformed first, because time is cyclical and not linear.

```{r pca_cluster_preparation}
flights_structure <- flights_clean

flights_structure$DEP_HOUR_SIN <- sin(2 * pi * flights_structure$SCHED_DEP_HOUR / 24)
flights_structure$DEP_HOUR_COS <- cos(2 * pi * flights_structure$SCHED_DEP_HOUR / 24)
flights_structure$ARR_HOUR_SIN <- sin(2 * pi * flights_structure$SCHED_ARR_HOUR / 24)
flights_structure$ARR_HOUR_COS <- cos(2 * pi * flights_structure$SCHED_ARR_HOUR / 24)
flights_structure$MONTH_SIN <- sin(2 * pi * flights_structure$MONTH / 12)
flights_structure$MONTH_COS <- cos(2 * pi * flights_structure$MONTH / 12)
flights_structure$IS_WEEKEND <- ifelse(
  flights_structure$WEEKDAY %in% c("Saturday", "Sunday"), 1, 0
)

prep_decisions <- data.frame(
  feature = c(
    "DISTANCE",
    "CRS_ELAPSED_TIME",
    "SCHED_DEP_HOUR",
    "SCHED_ARR_HOUR",
    "MONTH",
    "WEEKDAY"
  ),
  preparation = c(
    "Z-score scaling only",
    "Z-score scaling only",
    "Cyclical encoding (sin/cos) + scaling",
    "Cyclical encoding (sin/cos) + scaling",
    "Cyclical encoding (sin/cos) + scaling",
    "Weekend dummy + scaling"
  ),
  rationale = c(
    "Meaningful numeric feature; no separate normalization needed once standardized",
    "Meaningful numeric feature; retained first to inspect overlap with distance",
    "Raw HHMM would create false distances, so cyclical encoding is required",
    "Raw HHMM would create false distances, so cyclical encoding is required",
    "Month 12 is adjacent to month 1, so linear coding is inappropriate",
    "A single dummy keeps dimensionality low and captures the main calendar contrast"
  ),
  stringsAsFactors = FALSE
)

prep_decisions
```

We use Z-score standardization here because PCA and distance-based
clustering are sensitive to scale. This puts the variables on a more
equal footing. We do not add an extra log transformation for `DISTANCE`
or `CRS_ELAPSED_TIME`, because we want to keep those variables easy to
interpret.

```{r pca_cluster_feature_space}
representation_vars <- c(
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "DEP_HOUR_SIN", "DEP_HOUR_COS",
  "ARR_HOUR_SIN", "ARR_HOUR_COS",
  "MONTH_SIN", "MONTH_COS",
  "IS_WEEKEND"
)

post_hoc_vars <- c("IS_DELAYED", "OP_CARRIER", "ORIGIN", "DEST")

structure_df <- flights_structure[, c(representation_vars, post_hoc_vars)]
structure_df <- na.omit(structure_df)

set.seed(46748717)
N_STRUCTURE <- min(50000, nrow(structure_df))
idx_structure <- sample(seq_len(nrow(structure_df)), N_STRUCTURE)

structure_sample <- structure_df[idx_structure, ]
structure_z <- as.data.frame(scale(structure_sample[, representation_vars]))

dim(structure_sample)
summary(structure_z)
```

## 6.2 Correlation heatmap

```{r correlation_heatmap}
cor_mat <- round(cor(structure_z), 2)
cor_mat

reshape2::melt(cor_mat) |>
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "firebrick", mid = "white", high = "steelblue", midpoint = 0) +
  labs(title = "Correlation heatmap of PCA/clustering feature space",
       x = "", y = "", fill = "Correlation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r correlation_pairs}
high_corr_idx <- which(abs(cor_mat) >= 0.75 & lower.tri(cor_mat), arr.ind = TRUE)

if (nrow(high_corr_idx) == 0) {
  high_corr_pairs <- data.frame(
    variable_1 = character(),
    variable_2 = character(),
    correlation = numeric()
  )
} else {
  high_corr_pairs <- data.frame(
    variable_1 = rownames(cor_mat)[high_corr_idx[, 1]],
    variable_2 = colnames(cor_mat)[high_corr_idx[, 2]],
    correlation = cor_mat[high_corr_idx],
    row.names = NULL
  )
}

high_corr_pairs
```

This heatmap is a quick check before PCA and clustering. If variables
are strongly correlated, they can capture the same signal twice. That is
especially relevant for clustering, where this can distort the
distances.

## 6.3 Principal component analysis

```{r pca_run}
pca_results <- prcomp(structure_z, center = FALSE, scale. = FALSE)

pca_var_explained <- (pca_results$sdev)^2 / sum((pca_results$sdev)^2)
pca_var_table <- data.frame(
  PC_num = seq_along(pca_var_explained),
  PC = paste0("PC", seq_along(pca_var_explained)),
  Variance_Explained = pca_var_explained,
  Cumulative = cumsum(pca_var_explained)
)

pca_var_table
```

```{r pca_variance_plots}
ggplot(pca_var_table, aes(x = PC, y = Variance_Explained)) +
  geom_col(fill = "steelblue", width = 0.7) +
  geom_text(aes(label = paste0(round(100 * Variance_Explained, 1), "%")),
            vjust = -0.4, size = 3) +
  ylim(0, max(pca_var_table$Variance_Explained) * 1.15) +
  labs(title = "Scree plot",
       x = "Principal component", y = "Variance explained")

ggplot(pca_var_table, aes(x = PC_num, y = Cumulative)) +
  geom_line(color = "steelblue") +
  geom_point(size = 2, color = "steelblue") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey50") +
  scale_x_continuous(breaks = pca_var_table$PC_num, labels = pca_var_table$PC) +
  scale_y_continuous(labels = pct_label, limits = c(0, 1)) +
  labs(title = "Cumulative explained variance",
       x = "Principal component", y = "Cumulative variance explained")
```

```{r pca_loadings}
top_loadings <- function(pca_obj, pc = 1, n = 5) {
  out <- data.frame(
    variable = names(pca_obj$rotation[, pc]),
    loading = pca_obj$rotation[, pc]
  )
  out$abs_loading <- abs(out$loading)
  out <- out[order(-out$abs_loading), ]
  head(out, n)
}

top_loadings(pca_results, pc = 1, n = 8)
top_loadings(pca_results, pc = 2, n = 8)
top_loadings(pca_results, pc = 3, n = 8)

plot_top_loadings <- function(pca_obj, pc = 1, n = 8) {
  out <- top_loadings(pca_obj, pc = pc, n = n)
  out$variable <- factor(out$variable, levels = rev(out$variable))

  ggplot(out, aes(x = variable, y = loading, fill = loading > 0)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    labs(title = paste("Main loadings for PC", pc),
         x = "Variable", y = "Loading")
}

print(plot_top_loadings(pca_results, pc = 1, n = 8))
print(plot_top_loadings(pca_results, pc = 2, n = 8))
```

The point of PCA here is not to replace the original variables straight
away. It is mainly a way to see whether a few components already capture
most of the structure, and which variables drive those components. The
scree plot shows how much variance each component explains, and the
loading plots show which variables matter most inside each component.

```{r pca_followup_decisions, include=FALSE}
cum4 <- pca_var_table$Cumulative[min(4, nrow(pca_var_table))]
distance_elapsed_cor <- cor(structure_z$DISTANCE, structure_z$CRS_ELAPSED_TIME)
```

The first four principal components explain `r round(100 * cum4, 1)`% of
the variance. That is helpful for understanding the data, but not enough
to drop the original variables. For the later models, the original
variables are still easier to explain.

The correlation between `DISTANCE` and `CRS_ELAPSED_TIME` is
`r round(distance_elapsed_cor, 2)`.
`r if (abs(distance_elapsed_cor) >= 0.85) "That overlap is quite strong, so we do not want both variables to weigh too heavily in the clustering step." else "That overlap is not extreme, so we can keep both for now, but we still read the clustering results with that overlap in mind."`

We also leave carrier and airport dummies out of PCA and clustering,
because they would mainly separate labels instead of showing broader
patterns.

## 6.4 Clustering

Clustering answers a different question than PCA. Here we use it to see
whether flights fall into a few recurring types based on route length,
timing and simple calendar structure. These clusters are not delay
classes. They are better seen as flight profiles.

```{r clustering_input}
cluster_vars <- c(
  "DISTANCE",
  "DEP_HOUR_SIN", "DEP_HOUR_COS",
  "ARR_HOUR_SIN", "ARR_HOUR_COS",
  "MONTH_SIN", "MONTH_COS",
  "IS_WEEKEND"
)

if (abs(distance_elapsed_cor) < 0.85) {
  cluster_vars <- c("DISTANCE", "CRS_ELAPSED_TIME", setdiff(cluster_vars, "DISTANCE"))
}

set.seed(46748717)
N_CLUSTER <- min(20000, nrow(structure_sample))
idx_cluster <- sample(seq_len(nrow(structure_sample)), N_CLUSTER)

cluster_data_z <- structure_z[idx_cluster, cluster_vars, drop = FALSE]
cluster_data_original <- flights_structure[idx_structure[idx_cluster], c(
  "DISTANCE", "CRS_ELAPSED_TIME", "SCHED_DEP_HOUR", "SCHED_ARR_HOUR",
  "IS_WEEKEND", "IS_DELAYED", "OP_CARRIER", "ORIGIN", "DEST"
)]
```

```{r elbow_plot}
wss <- sapply(2:6, function(k) {
  kmeans(cluster_data_z, centers = k, nstart = 20)$tot.withinss
})

elbow_df <- data.frame(k = 2:6, tot_withinss = wss)

ggplot(elbow_df, aes(x = k, y = tot_withinss)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 2:6) +
  labs(title = "Elbow plot for k-means", x = "Number of clusters", y = "Total within-cluster sum of squares")
```

We use the elbow plot only as a guide. There is no single correct number
of clusters, so we choose a value that is still easy to interpret.

```{r kmeans_clustering}
set.seed(46748717)
K_CHOSEN <- 4
kmeans_result <- kmeans(cluster_data_z, centers = K_CHOSEN, nstart = 25)

cluster_projection <- prcomp(cluster_data_z, center = FALSE, scale. = FALSE)$x[, 1:2]
cluster_plot_df <- data.frame(
  PC1 = cluster_projection[, 1],
  PC2 = cluster_projection[, 2],
  cluster = factor(kmeans_result$cluster)
)

set.seed(46748717)
idx_cluster_plot <- sample(seq_len(nrow(cluster_plot_df)), min(5000, nrow(cluster_plot_df)))

ggplot(cluster_plot_df[idx_cluster_plot, ], aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.2) +
  labs(title = "K-means clusters projected onto the first two dimensions")
```

```{r cluster_profiles}
cluster_profile_df <- cluster_data_original
cluster_profile_df$cluster <- factor(kmeans_result$cluster)
cluster_profile_df$IS_DELAYED <- as.numeric(cluster_profile_df$IS_DELAYED)

cluster_sizes <- data.frame(
  cluster = factor(names(table(cluster_profile_df$cluster))),
  n = as.integer(table(cluster_profile_df$cluster))
)

cluster_sizes

cluster_summary <- aggregate(
  cbind(DISTANCE, CRS_ELAPSED_TIME, SCHED_DEP_HOUR, SCHED_ARR_HOUR, IS_WEEKEND, IS_DELAYED) ~ cluster,
  data = cluster_profile_df,
  FUN = mean
)

cluster_summary
```

The cluster summaries are mainly descriptive. If the clusters differ a
lot in route length and timing, but only a little in delay rate, then
they are telling us more about flight types than about delay risk.

```{r cluster_density_plots}
cluster_density_df <- cluster_profile_df

density_vars <- c("DISTANCE", "CRS_ELAPSED_TIME", "SCHED_DEP_HOUR", "SCHED_ARR_HOUR")
density_vars <- intersect(density_vars, names(cluster_density_df))

cluster_density_long <- reshape2::melt(
  cluster_density_df[, c("cluster", density_vars)],
  id.vars = "cluster"
)

ggplot(cluster_density_long, aes(x = value, color = cluster)) +
  geom_density() +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Cluster distributions for key continuous variables",
       x = "Value", y = "Density", color = "Cluster")
```

These density plots help us check whether the clusters really differ in
their underlying distributions and not only in their averages. We only
use them for continuous variables, because they are not very useful for
binary ones like `IS_WEEKEND`.

```{r hierarchical_clustering}
set.seed(46748717)
N_HCLUST <- min(150, nrow(cluster_data_z))
idx_hclust <- sample(seq_len(nrow(cluster_data_z)), N_HCLUST)

hclust_input <- cluster_data_z[idx_hclust, , drop = FALSE]
hclust_dist <- dist(hclust_input, method = "euclidean")
hclust_result <- hclust(hclust_dist, method = "ward.D2")

plot(hclust_result, labels = FALSE, hang = -1,
     main = "Hierarchical clustering dendrogram (sample)",
     xlab = "", sub = "", ylab = "Height")
```

Hierarchical clustering is only used on a small sample. On a dataset
this size it is not practical as a main method, so here it is just a
simple extra check.

So our clustering conclusion is fairly simple. The clusters are useful
to describe flight types, but we do not use them directly in the main
prediction models. For that, the original engineered variables are still
the better choice.

# 7. Modelling

In this section we follow the classification workbook more closely. The
goal is not to try lots of complex models, but to compare a few clear
ones on our delay problem. We focus mainly on the delayed class, because
missing a real delay is worse than giving a false alarm.

## 7.1 Classification setup

We prepare one modelling data set and then split it into training and
testing data. The training data is used to estimate the classification
rules, while the testing data is held back to inspect how the model
behaves on unseen flights.

## 7.2 Preparing the modelling data

```{r}
flights_model <- flights_clean

flights_model$IS_DELAYED_15 <- ifelse(flights_model$ARR_DELAY >= 15, "Delayed", "On_time")
flights_model$IS_DELAYED_15 <- factor(flights_model$IS_DELAYED_15)

flights_model$DEP_HOUR <- as.numeric(format(flights_model$CRS_DEP_TIME, "%H")) +
                          as.numeric(format(flights_model$CRS_DEP_TIME, "%M")) / 60

flights_model$ARR_HOUR <- as.numeric(format(flights_model$CRS_ARR_TIME, "%H")) +
                          as.numeric(format(flights_model$CRS_ARR_TIME, "%M")) / 60

factor_vars <- c("OP_CARRIER", "ORIGIN", "DEST", "WEEKDAY")

for (v in factor_vars) {
  if (v %in% names(flights_model)) {
    flights_model[[v]] <- factor(flights_model[[v]])
  }
}

model_vars <- c(
  "IS_DELAYED_15",
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "DEP_HOUR",
  "ARR_HOUR",
  "MONTH",
  "WEEKDAY",
  "OP_CARRIER",
  "ORIGIN",
  "DEST"
)

model_df <- flights_model[, model_vars]
model_df <- na.omit(model_df)

str(model_df)
summary(model_df$IS_DELAYED_15)

set.seed(46748717)

N_MODEL <- min(300000, nrow(model_df))   # adjust if needed
idx <- sample(seq_len(nrow(model_df)), N_MODEL)

model_df_small <- model_df[idx, ]
dim(model_df_small)

set.seed(46748717)

split <- rsample::initial_split(model_df_small, prop = 0.7)
training <- training(split)
testing  <- testing(split)

dim(training)
dim(testing)
```

We keep the feature set simple and close to the workbook. All variables
used here are known before departure and are easy to explain.

```{r training_testing_comparison}
split_compare <- rbind(
  data.frame(split = "Training", training[, c("IS_DELAYED_15", "DISTANCE", "DEP_HOUR")]),
  data.frame(split = "Testing", testing[, c("IS_DELAYED_15", "DISTANCE", "DEP_HOUR")])
)

ggplot(split_compare, aes(x = IS_DELAYED_15, fill = split)) +
  geom_bar(position = "dodge") +
  labs(title = "Class balance in training and testing data",
       x = "Delay class", y = "Number of flights")

ggplot(split_compare, aes(x = DISTANCE, fill = split)) +
  geom_histogram(bins = 30, color = "white") +
  facet_wrap(~ split, ncol = 1) +
  labs(title = "Distance distribution: training vs testing",
       x = "Distance (miles)", y = "Count")

ggplot(split_compare, aes(x = DEP_HOUR, fill = split)) +
  geom_histogram(bins = 24, color = "white") +
  facet_wrap(~ split, ncol = 1) +
  labs(title = "Departure-hour distribution: training vs testing",
       x = "Scheduled departure hour", y = "Count")
```

These checks are simple but important. If training and testing look very
different, then the model comparison is less reliable. Here we mainly
want to see whether the split produced two similar samples.

```{r model_eval_helpers, include=FALSE}
metric_summary <- function(actual, pred) {
  actual <- factor(actual, levels = c("On_time", "Delayed"))
  pred   <- factor(pred, levels = c("On_time", "Delayed"))

  tp <- sum(actual == "Delayed" & pred == "Delayed")
  tn <- sum(actual == "On_time" & pred == "On_time")
  fp <- sum(actual == "On_time" & pred == "Delayed")
  fn <- sum(actual == "Delayed" & pred == "On_time")

  accuracy  <- (tp + tn) / (tp + tn + fp + fn)
  recall    <- ifelse(tp + fn == 0, NA, tp / (tp + fn))
  precision <- ifelse(tp + fp == 0, NA, tp / (tp + fp))
  f1        <- ifelse(
    is.na(recall) | is.na(precision) | (recall + precision) == 0,
    NA,
    2 * recall * precision / (recall + precision)
  )

  data.frame(
    accuracy = accuracy,
    recall_delayed = recall,
    precision_delayed = precision,
    f1_delayed = f1,
    tp = tp, tn = tn, fp = fp, fn = fn
  )
}

hard_vote <- function(pred_df) {
  delayed_votes <- rowSums(pred_df == "Delayed")
  factor(
    ifelse(delayed_votes >= ceiling(ncol(pred_df) / 2), "Delayed", "On_time"),
    levels = c("On_time", "Delayed")
  )
}
```

## 7.3 Model 1: C5.0 decision tree

```{r}

model_tree <- C50::C5.0(IS_DELAYED_15 ~ ., data = training)

summary(model_tree)
```

The decision tree is our first baseline because it is easy to explain.
We can see which variables are used first and how the data gets split.

```{r tree_conditions_plot}
tree_plot_model <- rpart::rpart(
  IS_DELAYED_15 ~ .,
  data = training,
  method = "class",
  control = rpart::rpart.control(maxdepth = 3, minsplit = 3000, cp = 0.001)
)

plot(tree_plot_model, uniform = TRUE, margin = 0.1)
text(tree_plot_model, use.n = TRUE, all = TRUE, cex = 0.7)
```

The C5.0 summary is still the main output, but this extra shallow tree
makes the first split rules easier to read.

## 7.4 Inspecting tree behaviour on unseen data

```{r}

pred_test <- predict(model_tree, testing)

gmodels::CrossTable(
  testing$IS_DELAYED_15,
  pred_test,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

# Accuracy
mean(pred_test == testing$IS_DELAYED_15)

actual <- testing$IS_DELAYED_15
pred   <- pred_test

tp <- sum(actual == "Delayed" & pred == "Delayed")
tn <- sum(actual == "On_time" & pred == "On_time")
fp <- sum(actual == "On_time" & pred == "Delayed")
fn <- sum(actual == "Delayed" & pred == "On_time")

accuracy  <- (tp + tn) / (tp + tn + fp + fn)
recall    <- tp / (tp + fn)
precision <- tp / (tp + fp)
f1_tree   <- 2 * recall * precision / (recall + precision)

data.frame(
  accuracy = accuracy,
  recall_delayed = recall,
  precision_delayed = precision,
  f1_delayed = f1_tree,
  tp = tp, tn = tn, fp = fp, fn = fn
)
```

The confusion matrix matters because accuracy alone is not enough. For
this problem, false negatives matter most: flights predicted as on time
that are actually delayed. So recall is especially important here.

## 7.5 Altering tree behaviour: boosting

```{r}
model_tree_boost <- C50::C5.0(IS_DELAYED_15 ~ ., data = training, trials = 10)

pred_boost <- predict(model_tree_boost, testing)

gmodels::CrossTable(
  testing$IS_DELAYED_15,
  pred_boost,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

tp_boost <- sum(actual == "Delayed" & pred_boost == "Delayed")
tn_boost <- sum(actual == "On_time" & pred_boost == "On_time")
fp_boost <- sum(actual == "On_time" & pred_boost == "Delayed")
fn_boost <- sum(actual == "Delayed" & pred_boost == "On_time")

data.frame(
  model = "Boosted C5.0",
  accuracy = (tp_boost + tn_boost) / (tp_boost + tn_boost + fp_boost + fn_boost),
  recall_delayed = tp_boost / (tp_boost + fn_boost),
  precision_delayed = tp_boost / (tp_boost + fp_boost),
  f1_delayed = 2 * (tp_boost / (tp_boost + fn_boost)) * (tp_boost / (tp_boost + fp_boost)) /
    ((tp_boost / (tp_boost + fn_boost)) + (tp_boost / (tp_boost + fp_boost)))
)
```

Boosting is useful here because it lets the model correct some mistakes
from earlier trees. We keep the number of trials low so the workflow
stays simple.

## 7.6 Altering tree behaviour: cost-sensitive classification

```{r}
cost_matrix <- matrix(c(NA, 1,
                        2, NA),
                      nrow = 2,
                      ncol = 2,
                      byrow = FALSE)
rownames(cost_matrix) <- colnames(cost_matrix) <- c("Delayed", "On_time")

model_tree_cost <- C50::C5.0(IS_DELAYED_15 ~ ., data = training, costs = cost_matrix)

pred_cost <- predict(model_tree_cost, testing)

gmodels::CrossTable(
  testing$IS_DELAYED_15,
  pred_cost,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

tp_cost <- sum(actual == "Delayed" & pred_cost == "Delayed")
tn_cost <- sum(actual == "On_time" & pred_cost == "On_time")
fp_cost <- sum(actual == "On_time" & pred_cost == "Delayed")
fn_cost <- sum(actual == "Delayed" & pred_cost == "On_time")

data.frame(
  model = "Cost-sensitive C5.0",
  accuracy = (tp_cost + tn_cost) / (tp_cost + tn_cost + fp_cost + fn_cost),
  recall_delayed = tp_cost / (tp_cost + fn_cost),
  precision_delayed = tp_cost / (tp_cost + fp_cost),
  f1_delayed = 2 * (tp_cost / (tp_cost + fn_cost)) * (tp_cost / (tp_cost + fp_cost)) /
    ((tp_cost / (tp_cost + fn_cost)) + (tp_cost / (tp_cost + fp_cost)))
)
```

This step fits the business problem well. We explicitly say that missing
a real delay is worse than a false alarm. That usually increases recall,
even if precision or accuracy drops a bit.

## 7.7 Model 2: Random Forest

```{r}
rf_vars <- c(
  "IS_DELAYED_15",
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "DEP_HOUR",
  "ARR_HOUR",
  "MONTH",
  "WEEKDAY",
  "OP_CARRIER"
)

training_rf <- training[, rf_vars]
testing_rf  <- testing[, rf_vars]

set.seed(46748717)
model_rf <- randomForest::randomForest(IS_DELAYED_15 ~ ., data = training_rf, ntree = 200)

pred_rf <- predict(model_rf, testing_rf)

tp_rf <- sum(actual == "Delayed" & pred_rf == "Delayed")
tn_rf <- sum(actual == "On_time" & pred_rf == "On_time")
fp_rf <- sum(actual == "On_time" & pred_rf == "Delayed")
fn_rf <- sum(actual == "Delayed" & pred_rf == "On_time")

data.frame(
  model = "Random forest",
  accuracy = (tp_rf + tn_rf) / (tp_rf + tn_rf + fp_rf + fn_rf),
  recall_delayed = tp_rf / (tp_rf + fn_rf),
  precision_delayed = tp_rf / (tp_rf + fp_rf),
  f1_delayed = 2 * (tp_rf / (tp_rf + fn_rf)) * (tp_rf / (tp_rf + fp_rf)) /
    ((tp_rf / (tp_rf + fn_rf)) + (tp_rf / (tp_rf + fp_rf)))
)

importance(model_rf)
varImpPlot(model_rf, main = "Random forest variable importance")
```

The random forest is a second benchmark next to the single tree. Here we
use a slightly simpler feature set, because this `randomForest`
implementation cannot handle predictors with too many categories, such
as `ORIGIN` and `DEST`.

## 7.8 Comparison

```{r}
comparison_table <- data.frame(
  model = c("C5.0 tree", "Boosted C5.0", "Cost-sensitive C5.0", "Random forest"),
  accuracy = c(
    accuracy,
    (tp_boost + tn_boost) / (tp_boost + tn_boost + fp_boost + fn_boost),
    (tp_cost + tn_cost) / (tp_cost + tn_cost + fp_cost + fn_cost),
    (tp_rf + tn_rf) / (tp_rf + tn_rf + fp_rf + fn_rf)
  ),
  recall_delayed = c(
    recall,
    tp_boost / (tp_boost + fn_boost),
    tp_cost / (tp_cost + fn_cost),
    tp_rf / (tp_rf + fn_rf)
  ),
  precision_delayed = c(
    precision,
    tp_boost / (tp_boost + fp_boost),
    tp_cost / (tp_cost + fp_cost),
    tp_rf / (tp_rf + fp_rf)
  ),
  f1_delayed = c(
    f1_tree,
    2 * (tp_boost / (tp_boost + fn_boost)) * (tp_boost / (tp_boost + fp_boost)) /
      ((tp_boost / (tp_boost + fn_boost)) + (tp_boost / (tp_boost + fp_boost))),
    2 * (tp_cost / (tp_cost + fn_cost)) * (tp_cost / (tp_cost + fp_cost)) /
      ((tp_cost / (tp_cost + fn_cost)) + (tp_cost / (tp_cost + fp_cost))),
    2 * (tp_rf / (tp_rf + fn_rf)) * (tp_rf / (tp_rf + fp_rf)) /
      ((tp_rf / (tp_rf + fn_rf)) + (tp_rf / (tp_rf + fp_rf)))
  )
)

comparison_table
```

For this case, recall on delayed flights is still the most important
metric. A model with slightly lower accuracy can still be better if it
finds more real delays.

The extra F1 column is useful because it summarizes the balance between
precision and recall in one number. We still do not treat it as the main
metric, but it helps us see when a model improves one of the two at the
expense of the other.

## 7.9 Why one split is not enough

One train-test split gives only one answer. To see whether our
conclusions are stable, we repeat the same evaluation several times on a
smaller fixed sample. We keep this step simple and limit it to three
different model behaviours: a plain tree, a cost-sensitive tree and the
random forest. We do not repeat every model from above, because that
would mainly add runtime without adding much new insight.

```{r repeated_evaluation}
set.seed(46748717)
N_EVAL <- min(60000, nrow(model_df_small))
idx_eval <- sample(seq_len(nrow(model_df_small)), N_EVAL)
eval_df <- model_df_small[idx_eval, ]

N_REPEATS <- 8
repeated_results <- data.frame()

for (i in seq_len(N_REPEATS)) {
  set.seed(46748717 + i)
  split_eval <- rsample::initial_split(eval_df, prop = 0.7, strata = IS_DELAYED_15)
  training_eval <- rsample::training(split_eval)
  testing_eval  <- rsample::testing(split_eval)

  tree_eval <- C50::C5.0(IS_DELAYED_15 ~ ., data = training_eval)
  pred_tree_eval <- predict(tree_eval, testing_eval)
  metrics_tree_eval <- metric_summary(testing_eval$IS_DELAYED_15, pred_tree_eval)
  metrics_tree_eval$model <- "C5.0 tree"
  metrics_tree_eval$iteration <- i

  cost_eval <- C50::C5.0(IS_DELAYED_15 ~ ., data = training_eval, costs = cost_matrix)
  pred_cost_eval <- predict(cost_eval, testing_eval)
  metrics_cost_eval <- metric_summary(testing_eval$IS_DELAYED_15, pred_cost_eval)
  metrics_cost_eval$model <- "Cost-sensitive C5.0"
  metrics_cost_eval$iteration <- i

  training_eval_rf <- training_eval[, rf_vars]
  testing_eval_rf  <- testing_eval[, rf_vars]
  rf_eval <- randomForest::randomForest(IS_DELAYED_15 ~ ., data = training_eval_rf, ntree = 100)
  pred_rf_eval <- predict(rf_eval, testing_eval_rf)
  metrics_rf_eval <- metric_summary(testing_eval_rf$IS_DELAYED_15, pred_rf_eval)
  metrics_rf_eval$model <- "Random forest"
  metrics_rf_eval$iteration <- i

  pred_ensemble_eval <- hard_vote(data.frame(
    tree = pred_tree_eval,
    cost_sensitive = pred_cost_eval,
    rf = pred_rf_eval
  ))
  metrics_ensemble_eval <- metric_summary(testing_eval$IS_DELAYED_15, pred_ensemble_eval)
  metrics_ensemble_eval$model <- "Hard vote ensemble"
  metrics_ensemble_eval$iteration <- i

  repeated_results <- rbind(
    repeated_results,
    metrics_tree_eval,
    metrics_cost_eval,
    metrics_rf_eval,
    metrics_ensemble_eval
  )
}

ggplot(repeated_results, aes(x = model, y = accuracy)) +
  geom_boxplot() +
  labs(title = "Accuracy across repeated train-test splits",
       x = "Model", y = "Accuracy") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

ggplot(repeated_results, aes(x = model, y = recall_delayed)) +
  geom_boxplot() +
  labs(title = "Recall across repeated train-test splits",
       x = "Model", y = "Recall for delayed flights") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

aggregate(
  cbind(accuracy, recall_delayed, precision_delayed, f1_delayed) ~ model,
  data = repeated_results,
  FUN = mean
)
```

This step matters because model performance is not just one number. Some
models look more stable across splits, while others move around more.
That is especially important here, because we want a model that behaves
reasonably well and not only on one lucky test set.

## 7.10 Looking beyond one threshold

So far we mostly looked at class labels such as "Delayed" or "On_time".
ROC curves and AUC add another view: they show how well the model ranks
flights from lower to higher delay risk across many possible thresholds.
We include this for the main models that give usable class
probabilities.

```{r roc_auc}
tree_prob <- predict(model_tree, testing, type = "prob")[, "Delayed"]
cost_prob <- predict(model_tree_cost, testing, type = "prob")[, "Delayed"]
rf_prob   <- predict(model_rf, testing_rf, type = "prob")[, "Delayed"]

roc_tree <- pROC::roc(testing$IS_DELAYED_15, tree_prob,
                      levels = c("On_time", "Delayed"), quiet = TRUE)
roc_cost <- pROC::roc(testing$IS_DELAYED_15, cost_prob,
                      levels = c("On_time", "Delayed"), quiet = TRUE)
roc_rf   <- pROC::roc(testing_rf$IS_DELAYED_15, rf_prob,
                      levels = c("On_time", "Delayed"), quiet = TRUE)

plot(roc_tree, col = "steelblue", main = "ROC curves on the testing data")
lines(roc_cost, col = "firebrick")
lines(roc_rf, col = "darkgreen")
legend("bottomright",
       legend = c("C5.0 tree", "Cost-sensitive C5.0", "Random forest"),
       col = c("steelblue", "firebrick", "darkgreen"),
       lwd = 2, bty = "n")

auc_table <- data.frame(
  model = c("C5.0 tree", "Cost-sensitive C5.0", "Random forest"),
  auc = c(
    as.numeric(pROC::auc(roc_tree)),
    as.numeric(pROC::auc(roc_cost)),
    as.numeric(pROC::auc(roc_rf))
  )
)

auc_table
```

AUC is useful here because it is not tied to one single threshold. We
still keep the confusion matrix and recall as our main practical view,
because the airline eventually needs a yes/no decision. But AUC helps us
check whether a model is at least ranking higher-risk flights above
lower-risk ones.

## 7.11 Model disagreement and simple voting

If models disagree on the same flights, that is useful information. It
usually means those flights are harder to classify. We therefore first
look at how often the main models agree, and then test one simple hard
voting ensemble.

```{r model_disagreement}
pred_patterns <- data.frame(
  tree = pred_test,
  cost_sensitive = pred_cost,
  rf = pred_rf
)

agreement_summary <- data.frame(
  outcome = c("All three agree", "At least one model disagrees"),
  share = c(
    mean(pred_patterns$tree == pred_patterns$cost_sensitive &
           pred_patterns$cost_sensitive == pred_patterns$rf),
    mean(!(pred_patterns$tree == pred_patterns$cost_sensitive &
             pred_patterns$cost_sensitive == pred_patterns$rf))
  )
)

ggplot(agreement_summary, aes(x = outcome, y = share)) +
  geom_col() +
  scale_y_continuous(labels = pct_label, limits = c(0, 1)) +
  labs(title = "How often do the main models agree?",
       x = "", y = "Share of testing flights")

pred_ensemble <- hard_vote(pred_patterns)
ensemble_metrics <- metric_summary(testing$IS_DELAYED_15, pred_ensemble)

ensemble_metrics

comparison_with_ensemble <- rbind(
  comparison_table,
  data.frame(
    model = "Hard vote ensemble",
    accuracy = ensemble_metrics$accuracy,
    recall_delayed = ensemble_metrics$recall_delayed,
    precision_delayed = ensemble_metrics$precision_delayed,
    f1_delayed = ensemble_metrics$f1_delayed
  )
)

comparison_with_ensemble
```

We keep the ensemble very simple on purpose. Hard voting is easy to
explain: each model gets one vote and the majority decides. That makes
it a good workbook-level ensemble.

We do not add weighted voting, stacking or another optimized ensemble
layer here. Those methods would add extra tuning and make it harder to
see whether any improvement comes from the ensemble idea itself or from
extra optimization. We also do not use soft voting, because then we
would need to treat the probabilities from different models as directly
comparable, which is a stronger assumption than we need here.

So the point of this section is not to build the most advanced ensemble.
It is just to check whether a very simple combination changes the
stability or the error balance in a useful way.

# 8. Conclusions and recommendations

## 8.1 Operational recommendations

Based on this first notebook, the clearest recommendation is to use the
cleaned pre-departure data for an early warning view. Delay risk is not
random. It changes by month, time of day, carrier, airport and route, so
those patterns can already help with planning and monitoring.

## 8.2 Limitations & next steps

This notebook still has clear limits. We only use historical operational
variables and a first set of engineered features. Weather, network
effects, inbound aircraft information and congestion measures could be
good next steps.

## 8.3 Assignment check

To make it easier to check the assignment, we summarize the main parts
that are already covered in the workbook:

-   Business context, analytics task and success criteria are stated
    explicitly.

-   Data exploration is documented with interpretation rather than plots
    only.

-   Data preparation choices are motivated, including missing values,
    datatype fixes and derived variables.

-   PCA and clustering are only used after variable screening and only
    on a clear pre-departure feature space.

-   Classification models are evaluated with a train-test split and a
    confusion matrix, with explicit attention to recall versus
    precision.

-   Model evaluation goes beyond one split by checking repeated
    evaluation, ROC/AUC and a simple voting ensemble.

-   Operational recommendations, limitations and next steps are stated
    separately.

So overall, the workbook now covers the main analytical parts that would
normally be expected. The last important step is still to knit the file
and check whether the written interpretation matches the final output.

# 9. Appendix

## 9.1 Variable screening for PCA and clustering

The main text already contains the screening and preparation decisions
used for PCA and clustering. We reproduce the key tables here so the
representation choices remain easy to review.

```{r appendix_pca_screening}
structure_screen
prep_decisions
```
