---
title: "MOT143A: Business Analytics group 18"
author: "Hidde Segers and Lotte Feilzer"
date: "2026-03-04"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

# 1. Business context and analytics task

## 1.1 Business problem

**Context:** Airlines running domestic flights in the United States
(U.S.) must manage constantly resources such as gates, crew members,
ground handling staff, and customer service capacity. These decisions
have to be made quickly and often with limited certainty.

Flight delays and cancellations can cause a lot of trouble for those who
are trying to get on time. When things go wrong, airlines have to pay
for: crew overtime, gate conflicts, aircraft re-positioning,
compensation, re-booking logistics, and re-putational damage. At the
same time, customer satisfaction declines, which affects their name and
in the long-term their revenue.

**Problem:** Delays and cancellations are really expensive for airlines
and can be frustrating for passengers. A single delay can create high
costs for crew overtime, gate conflicts, re-booking workloads. Because
flights are connected in a daily schedule, one delay can also cause
follow-on delays later in the network. This increases operational costs
and directly reduces customer satisfaction.

The practical challenge is that a lot of operational decisions must be
made before the flights departures. At that time it is still not clear
whether the flight is running on time or not. If the airline can figure
out flights that are likely to be delayed early, it can allocate
resources more active (e.g., assign standby crew, prepare customer
service capacity or adjust gate plans) and reduce the impact when such
delays occurs.

This leads to our business question:

**Business question:** How can we estimate the probability that a
scheduled flight will be delayed (e.g. more than 15 min) before
departure?

The goal is not to "predict the truth" perfectly, but to provide a risk
signal that helps operations teams prioritize attention and resources
towards flights with the highest expected delay risk. From an
organizational perspective, the insights are meant for operations
control centers, network planners and day-of-operations managers. The
model would be used in real-time operational setting, where planners
monitor flights and allocate resources (that are most of the time
limited). In this context, the prediction functions as a
decision-support tool that can help prioritize attention and
interventions for flights with the highest risk of delay.

## 1.2 Analytics task

This analysis is focused on the perspective of the airline operations
control and the airport station management to identify high risk
flights. This should support the workers with allocation of staff, gates
and premature consumer communication. The model will be used for
pre-departure planning setting and will be based on the historical
operational data of US domestic flights. The business problem will be
translated to an analytic task by approaching it as a predictive
risk-scoring problem. Three possible targets were considered:

-   **Classification A (delay):**

    **Target:** `IS_DELAYED_15 = 1` als `ARR_DELAY >= 15`, anders 0.

    Here it is analysed whether or not a flight will be significantly
    delayed or not (this is for flights that are neither diverted or
    delayed).

-   **Classification B (cancellation):**

    **Target:** `CANCELLED == 1`

    Here it is analysed whether or not a flight will be cancelled.

-   **Regression:**

    **Target:** `ARR_DELAY`

    Here it is analysed how much too late a flight will likely be in
    minutes.

The primary task chosen is classification A, this is because it will be
directly usable to relocate resources if needed and where focus in
logistics needs to focused. It is important to press that the model
can't reason and does not rely on causality. It also will not be able to
guarantee outcomes. It will estimate the risk for significant delay and
consequently improve and support operational actions. Classification for
cancellation will be harder because of the rarity of this event. The
regression of the delay is also hard to predict and the type of delay
can but does not also translate to how much delay there will be.

## 1.3 Success criteria

We consider the success criteria for this report in two perspectives.
Firstly, model criteria that will be need to be met to answer our
business question. Secondly, the business criteria that will ensure
positive effects of the model in practice.

From an operational point of view, missing a real delay is usually worse
than raising a false alarm. If the model says a flight will be on time
but it is actually delayed, the airline is unprepared. This can quickly
lead to crew problems, gate conflicts, and unhappy passengers.

If the model predicts a delay that does not happen, it may lead to some
extra costs because resources were allocated unnecessarily. However,
this is less harmful than being unprepared for a real delay.

For this reason, we focus on high recall for the delayed flight s. We
want the model to catch as many real delays as possible, even if that
means accepting some false alarms. At the same time, we monitor
precision to make sure we do not allocate too many unnecessary
resources.

# 2. Setup

Here all packages are downloaded to make the code reproducible.

```{r setup}
# CRAN mirror
local({r <- getOption("repos")
r["CRAN"] <- "https://cran.rstudio.com/"
options(repos = r)})

packages <- c("ggplot2", "Hmisc", "rsample", "gmodels", "C50", "randomForest",
  "factoextra", "lattice", "reshape2"
)

for (i in packages) {
  if (!require(i, character.only = TRUE)) {
    install.packages(i, dependencies = TRUE)
    library(i, character.only = TRUE)
  }
}

set.seed(46748717)
```

# 3. Data loading 

## 3.1 Data in `data/` folder

Here we read

```{r data_files}
data_dir <- "data"
YEAR <- 2018
csv_path <- file.path(data_dir, paste0(YEAR, ".csv"))
flights <- read.csv(csv_path, stringsAsFactors = FALSE, nrows = N_READ)
```

## 3.2 Read data

```{r read_data}
N_READ <- -1
dim(flights)
names(flights)
```

# 4. Data exploration & preparation

## 4.1 Reality check

```{r explore_basic}
str(flights)
summary(flights)
```

## 4.2 Missing values overview

```{r missingness}
na_count <- sapply(flights, function(x) sum(is.na(x)))
na_pct   <- round(100 * na_count / nrow(flights), 2)

na_table <- data.frame(
  variable = names(flights),
  na_count = as.integer(na_count),
  na_pct = na_pct
)

na_table <- na_table[order(-na_table$na_pct), ]
head(na_table, 15)
```

## 4.3 Minimal cleaning + datatypes

```{r clean_types}
# Drop unnamed columns if present
drop_cols <- grep("^Unnamed", names(flights), value = TRUE)
if (length(drop_cols) > 0) flights[drop_cols] <- NULL

# Parse FL_DATE if present
if ("FL_DATE" %in% names(flights)) flights$FL_DATE <- as.Date(flights$FL_DATE)

# Ensure integers
for (cname in c("CANCELLED","DIVERTED")) {
  if (cname %in% names(flights)) flights[[cname]] <- as.integer(flights[[cname]])
}
```

## 4.4 Feature engineering (pre-departure features)

```{r feature_engineering}
hhmm_to_minutes <- function(x) {
  x <- suppressWarnings(as.integer(x))
  hh <- x %/% 100
  mm <- x %% 100
  out <- hh*60 + mm
  out[is.na(x) | mm > 59 | hh < 0 | hh > 29] <- NA
  out
}

if ("FL_DATE" %in% names(flights)) {
  flights$MONTH <- as.integer(format(flights$FL_DATE, "%m"))
  flights$WEEKDAY <- weekdays(flights$FL_DATE)
}

if ("CRS_DEP_TIME" %in% names(flights)) {
  flights$CRS_DEP_MIN <- hhmm_to_minutes(flights$CRS_DEP_TIME)
  flights$CRS_DEP_HOUR <- floor(flights$CRS_DEP_MIN / 60)
}

if ("CRS_ARR_TIME" %in% names(flights)) {
  flights$CRS_ARR_MIN <- hhmm_to_minutes(flights$CRS_ARR_TIME)
  flights$CRS_ARR_HOUR <- floor(flights$CRS_ARR_MIN / 60)
}

if (all(c("ORIGIN","DEST") %in% names(flights))) {
  flights$ROUTE <- paste(flights$ORIGIN, flights$DEST, sep = "-")
}
```

## 4.5 Target definition (classification)

> **No leakage in modelling:** we will use only pre-departure features
> for the models. Targets come from outcome columns.

```{r targets}
# Delay target: ARR_DELAY >= 15 (only not-cancelled and not-diverted)
if (all(c("ARR_DELAY","CANCELLED","DIVERTED") %in% names(flights))) {
  flights$IS_DELAYED_15 <- ifelse(
    flights$CANCELLED == 0 & flights$DIVERTED == 0 & !is.na(flights$ARR_DELAY),
    as.integer(flights$ARR_DELAY >= 15),
    NA
  )
}

# Cancellation target
if ("CANCELLED" %in% names(flights)) {
  flights$IS_CANCELLED <- flights$CANCELLED
}

table(flights$IS_DELAYED_15, useNA="ifany")
table(flights$IS_CANCELLED, useNA="ifany")
```

# 5. Exploratory visualisation (ggplot2)

```{r eda_plots}
if (all(c("MONTH","IS_DELAYED_15") %in% names(flights))) {
  dfm <- aggregate(IS_DELAYED_15 ~ MONTH, flights, function(x) mean(x, na.rm = TRUE))
  ggplot(dfm, aes(x = MONTH, y = IS_DELAYED_15)) +
    geom_line() + geom_point() +
    scale_x_continuous(breaks = 1:12) +
    ylab("Delay rate (>=15 min)") + ggtitle("Delay rate per month")
}
```

```{r eda_airline_airport}
# Filter to flights where delay target is defined
if (all(c("IS_DELAYED_15","OP_CARRIER","ORIGIN") %in% names(flights))) {
  flights_clean <- flights[!is.na(flights$IS_DELAYED_15), ]

  airline_summary <- aggregate(IS_DELAYED_15 ~ OP_CARRIER, flights_clean,
                               function(x) mean(x, na.rm = TRUE))
  airline_summary <- airline_summary[order(-airline_summary$IS_DELAYED_15), ]

  ggplot(airline_summary[1:10, ],
         aes(x = reorder(OP_CARRIER, IS_DELAYED_15), y = IS_DELAYED_15)) +
    geom_col() +
    coord_flip() +
    labs(title = "Delay rate (>=15 min) per airline (top 10)",
         x = "Airline", y = "Delay rate")
}
```

**Interpretation (write):** - Seasonality?\
- Which carriers/airports stand out?\
- Class imbalance?

# 6. Dimension reduction & clustering (Workbook: Dimension reduction and clustering)

We cluster **airports** based on *pre-departure* aggregated profiles
(interpretability).

```{r airport_profiles}
needed <- c("ORIGIN","IS_DELAYED_15","IS_CANCELLED","DISTANCE","CRS_ELAPSED_TIME")
needed <- needed[needed %in% names(flights)]

if (all(c("ORIGIN","IS_DELAYED_15") %in% names(flights))) {
  base_df <- flights[!is.na(flights$IS_DELAYED_15), ]

  # aggregate per ORIGIN
  ap <- aggregate(
    cbind(IS_DELAYED_15, IS_CANCELLED, DISTANCE, CRS_ELAPSED_TIME) ~ ORIGIN,
    data = base_df,
    FUN = function(x) mean(x, na.rm = TRUE)
  )

  # add volume
  ap_n <- aggregate(IS_DELAYED_15 ~ ORIGIN, data = base_df, FUN = length)
  names(ap_n)[2] <- "n"
  ap <- merge(ap, ap_n, by = "ORIGIN")

  names(ap) <- c("ORIGIN","delay_rate","cancel_rate","avg_distance","avg_sched_time","n")

  # scale numeric columns
  ap_scaled <- scale(ap[, c("delay_rate","cancel_rate","avg_distance","avg_sched_time","n")])

  set.seed(1)
  km <- kmeans(ap_scaled, centers = 4, nstart = 20)
  ap$cluster <- factor(km$cluster)

  head(ap)
}
```

```{r cluster_viz}
if (exists("ap")) {
  ggplot(ap, aes(x = delay_rate, y = cancel_rate, color = cluster)) +
    geom_point(alpha = 0.8) +
    ggtitle("Airport clusters (delay & cancellation rates)") +
    xlab("Delay rate (>=15 min)") + ylab("Cancellation rate")
}
```

## 6.1 PCA

```{r pca}
if (exists("ap_scaled")) {
  pca_res <- prcomp(ap_scaled, center = TRUE, scale. = TRUE)
  summary(pca_res)

  factoextra::fviz_eig(pca_res, addlabels = TRUE)
  factoextra::fviz_pca_biplot(pca_res, repel = TRUE)
}
```

## 6.2 Correlation heatmap (replace GGally with reshape2 + ggplot2)

```{r corr_heatmap}
if (exists("ap")) {
  num_df <- ap[, c("delay_rate","cancel_rate","avg_distance","avg_sched_time","n")]
  cm <- cor(num_df, use = "complete.obs")
  cm_melt <- reshape2::melt(cm)

  ggplot(cm_melt, aes(x = Var1, y = Var2, fill = value)) +
    geom_tile() +
    labs(title = "Correlation matrix (airport profile features)",
         x = "", y = "", fill = "cor") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

**Interpretation (write):** - What do clusters represent?\
- Do PCA axes have meaning (loadings)?\
- Would you use a cluster label as a feature (yes/no and why)?

# 7. Modelling (Workbook: Classification)

We fit two models (C5.0 tree + Random Forest) using **pre-departure**
features only.

```{r modelling_data}
TARGET <- "IS_DELAYED_15"  # TODO: or "IS_CANCELLED"

candidate_features <- c(
  "OP_CARRIER","ORIGIN","DEST",
  "MONTH","WEEKDAY","CRS_DEP_HOUR",
  "CRS_ELAPSED_TIME","DISTANCE"
)
features <- intersect(candidate_features, names(flights))

stopifnot(TARGET %in% names(flights))
stopifnot(length(features) >= 3)

model_df <- flights[, c(TARGET, features)]
model_df <- model_df[complete.cases(model_df), ]

# collapse high-cardinality airports to top-N
topN_collapse <- function(x, N = 40) {
  tb <- sort(table(x), decreasing = TRUE)
  keep <- names(tb)[1:min(N, length(tb))]
  out <- ifelse(x %in% keep, x, "OTHER")
  factor(out)
}

if ("ORIGIN" %in% names(model_df)) model_df$ORIGIN <- topN_collapse(model_df$ORIGIN, 40)
if ("DEST" %in% names(model_df)) model_df$DEST <- topN_collapse(model_df$DEST, 40)
if ("OP_CARRIER" %in% names(model_df)) model_df$OP_CARRIER <- factor(model_df$OP_CARRIER)
if ("WEEKDAY" %in% names(model_df)) model_df$WEEKDAY <- factor(model_df$WEEKDAY)

split <- rsample::initial_split(model_df, prop = 0.7)
train <- rsample::training(split)
test  <- rsample::testing(split)

dim(train); dim(test)
table(train[[TARGET]])
```

## 7.1 Baseline

```{r baseline}
y_true <- test[[TARGET]]
y_pred0 <- rep(0, length(y_true))

gmodels::CrossTable(y_true, y_pred0,
                    prop.chisq = FALSE, prop.t = FALSE, prop.r = FALSE,
                    dnn = c("Actual", "Predicted (always 0)"))
```

## 7.2 Model 1 — C5.0 decision tree

```{r c50}
form <- as.formula(paste(TARGET, "~", paste(features, collapse = " + ")))

train_c50 <- train
test_c50 <- test
train_c50[[TARGET]] <- factor(train_c50[[TARGET]])
test_c50[[TARGET]]  <- factor(test_c50[[TARGET]])

tree <- C50::C5.0(form, data = train_c50)
tree

pred_tree <- predict(tree, newdata = test_c50)
gmodels::CrossTable(test_c50[[TARGET]], pred_tree,
                    prop.chisq = FALSE, prop.t = FALSE, prop.r = FALSE,
                    dnn = c("Actual", "Predicted (C5.0)"))
```

## 7.3 Model 2 — Random Forest

```{r rf}
train_rf <- train
test_rf <- test
train_rf[[TARGET]] <- factor(train_rf[[TARGET]])
test_rf[[TARGET]]  <- factor(test_rf[[TARGET]])

rf <- randomForest::randomForest(form, data = train_rf, ntree = 200)
rf

pred_rf <- predict(rf, newdata = test_rf)
gmodels::CrossTable(test_rf[[TARGET]], pred_rf,
                    prop.chisq = FALSE, prop.t = FALSE, prop.r = FALSE,
                    dnn = c("Actual", "Predicted (RF)"))

randomForest::varImpPlot(rf)
```

## 7.4 Comparison (write)

Compare: - performance (confusion matrices) - interpretability (tree
rules vs RF importance) - robustness / overfitting risk - operational
usability (how would you use scores in planning?)

**We recommend model \_\_\_ because \_\_\_ .**

# 8. Conclusions & recommendations

## 8.1 Operational recommendations (write)

Examples: - “High-risk flight list” per day/airport → allocate buffers
(gate agents/crew) - proactive passenger communication above a risk
threshold - airport/carrier monitoring dashboard (month/weekday/hour)

## 8.2 Limitations & next steps (write)

Limitations: - no weather / ATC constraints - concept drift across years
(policies, demand, post-COVID changes) - class imbalance (especially
cancellations)

Next steps (conceptual, workbooks 6–8): - when would representation
learning add value and what data is needed? - why text analytics is not
suitable here (no text fields) - how RL could apply if reformulated as
sequential staffing decisions

# 9. Reproducibility

```{r sessioninfo}
sessionInfo()
```
