---
title: "MOT143A: Business Analytics group 20"
author: "Hidde Segers and Lotte Feilzer"
date: "2026-03-04"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

# 1. Business context and analytics task

## 1.1 Business problem

**Context:** Airlines running domestic flights in the United States
(U.S.) must manage constantly resources such as gates, crew members,
ground handling staff, and customer service capacity. These decisions
have to be made quickly and often with limited certainty.

Flight delays and cancellations can cause a lot of trouble for those who
are trying to get on time. When things go wrong, airlines have to pay
for: crew overtime, gate conflicts, aircraft re-positioning,
compensation, re-booking logistics, and re-putational damage. At the
same time, customer satisfaction declines, which affects their name and
in the long-term their revenue.

**Problem:** Delays and cancellations are really expensive for airlines
and can be frustrating for passengers. A single delay can create high
costs for crew overtime, gate conflicts, re-booking workloads. Because
flights are connected in a daily schedule, one delay can also cause
follow-on delays later in the network. This increases operational costs
and directly reduces customer satisfaction.

The practical challenge is that a lot of operational decisions must be
made before the flights departures. At that time it is still not clear
whether the flight is running on time or not. If the airline can figure
out flights that are likely to be delayed early, it can allocate
resources more active (e.g., assign standby crew, prepare customer
service capacity or adjust gate plans) and reduce the impact when such
delays occurs.

This leads to our business question:

**Business question:** How can we estimate the probability that a
scheduled flight will be delayed (e.g. more than 15 min) before
departure?

The goal is not to "predict the truth" perfectly, but to provide a risk
signal that helps operations teams prioritize attention and resources
towards flights with the highest expected delay risk. From an
organizational perspective, the insights are meant for operations
control centers, network planners and day-of-operations managers. The
model would be used in real-time operational setting, where planners
monitor flights and allocate resources (that are most of the time
limited). In this context, the prediction functions as a
decision-support tool that can help prioritize attention and
interventions for flights with the highest risk of delay.

## 1.2 Analytics task

This analysis is focused on the perspective of the airline operations
control and the airport station management to identify high risk
flights. This should support the workers with allocation of staff, gates
and premature consumer communication. The model will be used for
pre-departure planning setting and will be based on the historical
operational data of US domestic flights. The business problem will be
translated to an analytic task by approaching it as a predictive
risk-scoring problem. Three possible targets were considered:

-   **Classification A (delay):**

    **Target:** `IS_DELAYED_15 = 1` als `ARR_DELAY >= 15`, anders 0.

    Here it is analysed whether or not a flight will be significantly
    delayed or not (this is for flights that are neither diverted or
    delayed).

-   **Classification B (cancellation):**

    **Target:** `CANCELLED == 1`

    Here it is analysed whether or not a flight will be cancelled.

-   **Regression:**

    **Target:** `ARR_DELAY`

    Here it is analysed how much too late a flight will likely be in
    minutes.

The primary task chosen is classification A, this is because it will be
directly usable to relocate resources if needed and where focus in
logistics needs to focused. It is important to press that the model
can't reason and does not rely on causality. It also will not be able to
guarantee outcomes. It will estimate the risk for significant delay and
consequently improve and support operational actions. Classification for
cancellation will be harder because of the rarity of this event. The
regression of the delay is also hard to predict and the type of delay
can but does not also translate to how much delay there will be.

## 1.3 Success criteria

We consider the success criteria for this report in two perspectives.
Firstly, model criteria that will be need to be met to answer our
business question. Secondly, the business criteria that will ensure
positive effects of the model in practice.

From an operational point of view, missing a real delay is usually worse
than raising a false alarm. If the model says a flight will be on time
but it is actually delayed, the airline is unprepared. This can quickly
lead to crew problems, gate conflicts, and unhappy passengers.

If the model predicts a delay that does not happen, it may lead to some
extra costs because resources were allocated unnecessarily. However,
this is less harmful than being unprepared for a real delay.

For this reason, we focus on high recall for the delayed flight s. We
want the model to catch as many real delays as possible, even if that
means accepting some false alarms. At the same time, we monitor
precision to make sure we do not allocate too many unnecessary
resources.

# 2. Setup

Here all packages are downloaded to make the code reproducible.

```{r setup}
# CRAN mirror
local({r <- getOption("repos")
r["CRAN"] <- "https://cran.rstudio.com/"
options(repos = r)})

N_READ <- -1

packages <- c("ggplot2", "Hmisc", "rsample", "gmodels", "C50", "randomForest",
  "factoextra", "lattice", "reshape2"
)

for (i in packages) {
  if (!require(i, character.only = TRUE)) {
    install.packages(i, dependencies = TRUE)
    library(i, character.only = TRUE)
  }
}

set.seed(46748717)
```

# 3. Data loading

## 3.1 Data in `data/` folder

Here we read our data. In the code it is easy to change the year of data
we want to see, just change the variable `YEAR`. Because the data_files
are quite big we added the possibility to print only a specific amount
of rows as given in `nrows` a value of `-1` will give all rows.

```{r data_files}
data_dir <- "data"
YEAR <- 2018
csv_path <- file.path(data_dir, paste0(YEAR, ".csv"))
flights <- read.csv(csv_path, stringsAsFactors = FALSE, nrows = N_READ)
```

## 3.2 Read data

Now we read our actual data and print the dimensions and columns names
to get a first clue of what kind of data were dealing with.

```{r read_data}
dim(flights)
names(flights)
```

As seen the data file contains:

-   7,213,446 observations

-   28 variables

We see that each observation represents one commercial flight in 2018.
The variables are as follows and categorized in definitions that can be
used to refer to groups of variables further in the report:

-   **Flight identifiers:** FL_DATE, OP_CARRIER, OP_CARRIER_FL_NUM

-   **Route information:** ORIGIN, DEST

-   **Scheduled times:** CRS_DEP_TIME, CRS_ARR_TIME

-   **Actual flight performance:** DEP_TIME, ARR_TIME, DEP_DELAY,
    ARR_DELAY

-   **Operational metrics:** TAXI_OUT, TAXI_IN, AIR_TIME

-   **Delay causes:** CARRIER_DELAY, WEATHER_DELAY, NAS_DELAY,
    SECURITY_DELAY, LATE_AIRCRAFT_DELAY

-   **Flight status indicators:** CANCELLED, DIVERTED

The variables can be descriped into the following categories:

### Flight Identifiers

| Variable          | Description        | Variable type           |
|-------------------|--------------------|-------------------------|
| FL_DATE           | Date of the flight | character (date string) |
| OP_CARRIER        | Operating airline  | character               |
| OP_CARRIER_FL_NUM | Flight number      | integer / numeric       |

## Route information

| Variable | Description               | Variable type     |
|----------|---------------------------|-------------------|
| ORIGIN   | Departure airport         | character         |
| DEST     | Arrival airport           | character         |
| DISTANCE | Distance between airports | integer / numeric |

## Scheduled times

| Variable         | Description               | Variable type                   |
|------------------|---------------------------|---------------------------------|
| CRS_DEP_TIME     | Scheduled departure time  | integer / numeric (HHMM format) |
| CRS_ARR_TIME     | Scheduled arrival time    | integer / numeric (HHMM format) |
| CRS_ELAPSED_TIME | Scheduled flight duration | integer / numeric (minutes)     |

## Actual flight performance

| Variable  | Description           | Variable type                   |
|-----------|-----------------------|---------------------------------|
| DEP_TIME  | Actual departure time | integer / numeric (HHMM format) |
| ARR_TIME  | Actual arrival time   | integer / numeric (HHMM format) |
| DEP_DELAY | Departure delay       | integer / numeric (minutes)     |
| ARR_DELAY | Arrival delay         | integer / numeric (minutes)     |
| AIR_TIME  | Time spent in air     | integer / numeric (minutes)     |

## Operational metrics

| Variable   | Description              | Variable type                   |
|------------|--------------------------|---------------------------------|
| TAXI_OUT   | Taxi time before takeoff | integer / numeric (minutes)     |
| TAXI_IN    | Taxi time after landing  | integer / numeric (minutes)     |
| WHEELS_OFF | Takeoff time             | integer / numeric (HHMM format) |
| WHEELS_ON  | Landing time             | integer / numeric (HHMM format) |

## Flight status

| Variable  | Description                 | Variable type |
|-----------|-----------------------------|---------------|
| CANCELLED | Indicates cancelled flights | binary (0/1)  |
| DIVERTED  | Indicates diverted flights  | binary (0/1)  |

## Delay causes

| Variable | Description | Variable type |
|----|----|----|
| CARRIER_DELAY | Delay caused by airline | integer / numeric (minutes) |
| WEATHER_DELAY | Delay caused by weather | integer / numeric (minutes) |
| NAS_DELAY | Delay caused by air traffic system | integer / numeric (minutes) |
| SECURITY_DELAY | Delay caused by security | integer / numeric (minutes) |
| LATE_AIRCRAFT_DELAY | Delay caused by late incoming aircraft | integer / numeric (minutes) |

## 4.1 Data summary

Here we show a small summary of the data. The following things are
noticeable. Columns in this datafile that contain dates are characters
preferably we change this later to datetimes. It is also remarkable that
the times are given in military time but as integers, this will make it
hard to calculate certain metrics inveolving time such as "time passed"
therefor we will preferably also change these datatypes later on to
datetimes. For metrics such as `TAXI_OUT` and `DEP_DELAY` the integers
are suitable because it represents the amount of minutes passed. We see
that for the `DIVERTED` and `CANCELLED` columns `0` and `1` are used
instead of Booleans to respresent `FALSE` and `TRUE` respectively, this
is also workable if remarked. Let's further explore the data.

```{r explore_basic}
str(flights)
```

To align this notebook more closely with the workbook on exploration and
preparation, we explicitly separate three questions:

-   What is the type and scale of each variable?

-   Which variables show data quality issues or unusual structure?

-   Which of those issues require action before modelling?

The next sections answer these questions step by step. We focus on
variables that are most relevant for a pre-departure delay analysis,
rather than producing every possible plot for all 28 columns.

```{r selected_numeric_diagnostics}
numeric_vars <- c("DISTANCE", "CRS_ELAPSED_TIME", "DEP_DELAY", "ARR_DELAY")
numeric_vars <- intersect(numeric_vars, names(flights))

numeric_diagnostics <- data.frame(
  variable = numeric_vars,
  mean = sapply(flights[numeric_vars], function(x) round(mean(x, na.rm = TRUE), 2)),
  median = sapply(flights[numeric_vars], function(x) round(median(x, na.rm = TRUE), 2)),
  sd = sapply(flights[numeric_vars], function(x) round(sd(x, na.rm = TRUE), 2)),
  min = sapply(flights[numeric_vars], function(x) round(min(x, na.rm = TRUE), 2)),
  max = sapply(flights[numeric_vars], function(x) round(max(x, na.rm = TRUE), 2))
)

numeric_diagnostics
```

This table functions as our first univariate numeric exploration step.
In workbook terms, we compare central tendency and spread before we look
at plots. The main thing to watch for is a large gap between mean and
median, or an extreme maximum relative to the bulk of the data, because
that signals skewness or potentially influential observations.

```{r selected_numeric_plots}
set.seed(46748717)
idx_diag <- sample(seq_len(nrow(flights)), min(200000, nrow(flights)))
flights_diag <- flights[idx_diag, ]

par(mfrow = c(2, 2))
hist(flights_diag$DISTANCE, breaks = 40, main = "Distance", xlab = "Miles", col = "grey80")
boxplot(flights_diag$DISTANCE, horizontal = TRUE, main = "Distance boxplot", col = "grey80")
hist(flights_diag$ARR_DELAY, breaks = 40, main = "Arrival delay", xlab = "Minutes", col = "grey80")
boxplot(flights_diag$ARR_DELAY, horizontal = TRUE, main = "Arrival delay boxplot", col = "grey80")
par(mfrow = c(1, 1))
```

We keep this visual check selective. `DISTANCE` is a core pre-departure
predictor and should be inspected as a regular numeric feature.
`ARR_DELAY` is not a predictor in the pre-departure setting, but it is
the outcome we eventually simplify into a binary target. Inspecting its
raw distribution helps justify why a classification target is more
practical here than modelling raw delay minutes directly: the delay
distribution is highly skewed and dominated by a long right tail.

## 4.2 Missing values overview and discussion

Here we check whether or not we have missing values in our dataset, how
many there are and what are the possible reasons for these missing
values are. We use logical reasoning in order to determine what to do
with potential missing data. Here we show a table that shows the amount
of NA values for each column and also a percentage of how many values
are missing in regards to the total dataset.

```{r missingness}
na_count <- sapply(flights, function(x) sum(is.na(x)))
na_pct   <- round(100 * na_count / nrow(flights), 2)

na_table <- data.frame(
  na_count = as.integer(na_count),
  na_pct = na_pct
)

na_table <- na_table[order(-na_table$na_pct), ]
na_table
```

Firstly we see a column `Unnamed..27` which seems to have only NA values
making the column absolute and removable. We see for the operational
variables and flight performance variables such as `DEP_TIME`,
`ARR_TIME`, `AIR_TIME`, `WHEELS_ON`, `WHEELS_OFF` that 1-2% of all
values are missing. We should note that all arrival variables have about
the same amount of NA values missing and idem for the departure values.
This probably means that the NA values are probably/hopefully
explainable for the same reasons. First we check whether the NAs
occurring in the delay rows all occur in the same row. The cell below
tells us this is the case. This lets us know that these `DELAY` causes
variables are jointly observed or missing. This is good sign. These rows
are supposed to be filled in after a flight and give a reason for the
delay if it was significant delayed (15 minutes or more). In our check
this also seems to be the case. All flights that are 15+ minutes delayed
have a cause and vice versa.

```{r delay-na-same}
cause_cols <- intersect(c("CARRIER_DELAY","WEATHER_DELAY",
                    "NAS_DELAY","SECURITY_DELAY","LATE_AIRCRAFT_DELAY"), names(flights))
delay15 <- !is.na(flights$ARR_DELAY) & flights$ARR_DELAY >= 15
cause_present <- !is.na(flights[[cause_cols[1]]])

miss <- is.na(flights[, cause_cols, drop=FALSE])
same_na_rows <- all(apply(miss, 2, function(x) identical(which(x), which(miss[,1]))))

cat("All the delay causes with NA values are for the same row, is", same_na_rows, "\n")
cat("Causes are present but there was no 15 min delay:", sum(cause_present & !delay15), "\n")
cat("More than 15 min delay but cause is NA value:", sum(delay15 & !cause_present))
```

Now we check if it possible for a delay flight to have multiple reasons
for delay this might handy to know later on. This seems to be possible
and is the case for 51.56% of the delayed flights.

```{r cause-compact}
k <- rowSums(flights[cause_present, cause_cols, drop=FALSE] > 0)
tab <- table(k)

multi_n <- sum(k > 1)
total_n <- length(k)
multi_pct <- round(100 * multi_n / total_n, 2)

cat("Of the flights with delay causes,", multi_pct, "% have multiple causes for delay.")
```

We now know that our delay causes columns are in fact complete. We can
now check if for the flights that have taken off all the important
operational and performance variables are present (`ARR_DELAY`,
`ARR_TIME` ect.) or that some or all values are missing for certain
rows. The NA values for the these rows as presented above could be
explained if these are the rows of flights that were diverted or
cancelled. If not, crucial information might be missing. If so, we
analyse this here.

```{r}
flights_wo_cancelled <- which(flights$CANCELLED == 0 & flights$DIVERTED == 0)
flights_wo_cancelled_df <- flights[flights_wo_cancelled, ]

na_count_wo_cd <- sapply(flights_wo_cancelled_df, function(x) sum(is.na(x)))
na_pct_wo_cd   <- round(100 * na_count_wo_cd / nrow(flights_wo_cancelled_df), 2)

na_table_wo_cd <- data.frame(
  variable = names(flights_wo_cancelled_df),
  na_count_wo_cd = as.integer(na_count_wo_cd),
  na_pct_wo_cd = na_pct_wo_cd
)

na_table_wo_cd <- na_table_wo_cd[order(-na_table_wo_cd$na_pct_wo_cd), ]
na_table_wo_cd
```

We can see that a flight that was either cancelled or diverted, it
accounts for the reasons that rows contain NA values for performance and
operational variables. We can see that the only two performance
variables that have still NA values are `ARR_DELAY` and `DEP_DELAY`,
why? These NA values might cause some trouble when modelling because we
use `ARR_DELAY` as predictors in the pre-departure model. Let's check by
showing the rows were this data is missing.

```{r check-missing-mostly-delays}
dep_delays_missing <- which(is.na(flights_wo_cancelled_df$DEP_DELAY)) 
arr_delays_missing <- which(is.na(flights_wo_cancelled_df$ARR_DELAY))
missing_delays <- flights_wo_cancelled_df[c(dep_delays_missing, arr_delays_missing), ]
head(missing_delays)
```

First we need to look into what we want to do with the rows in which the
`DEP_DELAY` and/or `ARR_DELAY` variables are missing. It seems that
there are only NA values for these columns when the `CRS_DEP_TIME` and
`DEP_TIME` have the same value and idem for the arrival. This means that
in this dataset the NA values just mean that plane left exactly at the
intended time. If this is the case we can replace the NA values in these
rows for 0 values and these NA values are also explained than as well.
We perform a quick check to see this is a 100 percent the case.

```{r fill_zero_delays}
dep_na <- !is.na(missing_delays$CRS_DEP_TIME) & !is.na(missing_delays$DEP_TIME) &
                 is.na(missing_delays$DEP_DELAY) &
                 missing_delays$CRS_DEP_TIME == missing_delays$DEP_TIME

arr_na <- !is.na(missing_delays$CRS_ARR_TIME) & !is.na(missing_delays$ARR_TIME) &
                 is.na(missing_delays$ARR_DELAY) &
                 missing_delays$CRS_ARR_TIME == missing_delays$ARR_TIME

cat("DEP_DELAY NA but not CRS_DEP_TIME==DEP_TIME:", sum(dep_na), "\n")
cat("ARR_DELAY NA but not CRS_ARR_TIME==ARR_TIME:", sum(arr_na))
```

Now we have analysed all NA values and begin to clean our data properly.

The workbook logic is visible here: we do not immediately delete all
rows with missing values, because in operational data missingness is
often informative. We first distinguish between:

-   structurally missing values, for example delay causes that are only
    recorded after substantial delays;

-   outcome-related missing values, for example flights that were
    cancelled or diverted and therefore have no realised arrival
    metrics;

-   repairable missing values, for example cases where scheduled time
    and realised time are equal and the delay can logically be set to 0.

This distinction matters because it prevents us from throwing away valid
flights too early and helps us explain each cleaning choice.

## 4.3 Cleaning and datatypes

First, we do some final sanity checks for the values for variables. We
check whether the distances are positive as well as airtime. We also
check whether the departure and arrival delays are plausible which all
seems to be the case.

```{r sanity_checks}
cat("There are", sum(flights$DISTANCE <= 0), "negative distances.", "\n")
cat("There are", sum(flights$AIR_TIME <= 0, na.rm = TRUE) , "negative airtimes.", "\n")
print("Departure delay summary:")
summary(flights$DEP_DELAY) 
print("Arrival delay summary:")
summary(flights$ARR_DELAY) 
```

We create a df named `flights_clean` to not overwrite the `flights`
variable that can be cleaned without overwriting our original data.

```{r create_flights_clean}
flights_clean <- flights
```

We begin with adjusting in accordance with the remarks made in section
3.2. We the `FL_DATE` column a datatype datetime and drop the unuseful
column `Unnamed..27` and ensure that the columns are the correct
datatypes.

```{r clean_types}
drop_cols <- grep("^Unnamed", names(flights_clean), value = TRUE)
flights_clean[drop_cols] <- NULL

flights_clean$FL_DATE <- as.Date(flights_clean$FL_DATE)

head(flights_clean, 4)
```

Now we do some minimal cleaning make sure the datatypes are correct. We
make the integer time columns POSIXct datetime columns to be able to
treat the time correctly.

```{r datetime_fix_overwrite}
hhmm_to_hhmm_str <- function(x) {
  x <- as.integer(x)
  hh <- x %/% 100
  mm <- x %% 100
  out <- sprintf("%02d:%02d", hh, mm)
  out[is.na(x) | mm > 59 | hh < 0 | hh > 29] <- NA
  out
}

time_cols <- intersect(c("WHEELS_OFF","WHEELS_ON","DEP_TIME","ARR_TIME","CRS_DEP_TIME","CRS_ARR_TIME"), names(flights_clean))

# Replace columns to POSIXct
for (cn in time_cols) {
  tmp_str <- hhmm_to_hhmm_str(flights_clean[[cn]])
  flights_clean[[cn]] <- as.POSIXct(
    paste(as.character(flights_clean$FL_DATE), tmp_str),
    format = "%Y-%m-%d %H:%M",
    tz = "UTC"
  )
}

# Overnight: is ARR_TIME < DEP_TIME + one day
if (all(c("DEP_TIME","ARR_TIME") %in% names(flights_clean))) {
  overnight <- !is.na(flights_clean$DEP_TIME) &
               !is.na(flights_clean$ARR_TIME) &
               (flights_clean$ARR_TIME < flights_clean$DEP_TIME)

  arrival_side <- intersect(c("ARR_TIME","WHEELS_ON","CRS_ARR_TIME"), time_cols)
  for (cn in arrival_side) {
    flights_clean[[cn]][overnight] <- flights_clean[[cn]][overnight] + 24*3600
  }
}

head(flights_clean[, time_cols, drop=FALSE], 5)
```

We now clean the NA values for the delay columns when the `CRS_DEP_TIME`
and `DEP_TIME` are the same and idem for the arrival times.

```{r impute_delays_posixct}
idx_dep0 <- which(
  is.na(flights_clean$DEP_DELAY) &
  !is.na(flights_clean$DEP_TIME) & !is.na(flights_clean$CRS_DEP_TIME) &
  flights_clean$DEP_TIME == flights_clean$CRS_DEP_TIME
)

idx_arr0 <- which(
  is.na(flights_clean$ARR_DELAY) &
  !is.na(flights_clean$ARR_TIME) & !is.na(flights_clean$CRS_ARR_TIME) &
  flights_clean$ARR_TIME == flights_clean$CRS_ARR_TIME
)

# vullen
flights_clean$DEP_DELAY[idx_dep0] <- 0
flights_clean$ARR_DELAY[idx_arr0] <- 0

# check
head(flights_clean[idx_dep0, c("CRS_DEP_TIME","DEP_TIME","DEP_DELAY")])
head(flights_clean[idx_arr0, c("CRS_ARR_TIME","ARR_TIME","ARR_DELAY")])
```

Now we check if all the data is clean. First we delete all the cancelled
and diverted rows because theserows are not relevant for our analysis.

```{r}
keep <- (is.na(flights_clean$CANCELLED) | flights_clean$CANCELLED == 0) &
        (is.na(flights_clean$DIVERTED)  | flights_clean$DIVERTED  == 0)

flights_clean <- flights_clean[keep, , drop = FALSE]

cat("Removed:", sum(!keep), "\n")
cat("Rows after:", nrow(flights_clean))
```

We decide to add some columns that make our analysis and modelling
easier later on, such as `IS_DELAYED` day of the week column which might
be interesting in the analysis and a column which contains a Boolean on
whether a flight is significantly delayed (15 minutes or later).Now we
add a column in which we see if a plane was significantly delayed. We
also add a `MONTH` and `WEEKDAY` column.

```{r}
flights_clean$IS_DELAYED <- ifelse(flights_clean$ARR_DELAY >= 15, TRUE, FALSE)
flights_clean$WEEKDAY <- weekdays(flights_clean$FL_DATE)
flights_clean$MONTH <- as.integer(format(flights_clean$FL_DATE, "%m"))
flights_clean$SCHED_DEP_HOUR <- as.numeric(format(flights_clean$CRS_DEP_TIME, "%H")) +
  as.numeric(format(flights_clean$CRS_DEP_TIME, "%M")) / 60
flights_clean$SCHED_ARR_HOUR <- as.numeric(format(flights_clean$CRS_ARR_TIME, "%H")) +
  as.numeric(format(flights_clean$CRS_ARR_TIME, "%M")) / 60
```

The target is now based on `ARR_DELAY >= 15`. This is the cleanest
definition for the business question because it captures whether the
flight ultimately arrived with a meaningful disruption. We remove
cancelled and diverted flights before defining this target, so the class
labels are only assigned to completed flights.

```{r cleaning_summary}
cleaning_summary <- data.frame(
  step = c(
    "Original rows",
    "After removing cancelled/diverted flights",
    "Rows with defined delay target"
  ),
  n_rows = c(
    nrow(flights),
    nrow(flights_clean),
    sum(!is.na(flights_clean$IS_DELAYED))
  )
)

cleaning_summary$share_of_original <- round(
  100 * cleaning_summary$n_rows / cleaning_summary$n_rows[1], 2
)

cleaning_summary
```

The table above makes the cleaning impact transparent. This is useful in
the workbook context because it shows that our later analysis is based
on a clearly defined subset of operationally comparable flights, rather
than on an unclear mix of completed, cancelled and diverted flights.

## 4.4 Exploration log and preparation decisions

The workbook stresses that exploration should lead to explicit
preparation choices. We therefore summarize our main findings in a short
exploration log.

```{r exploration_log}
exploration_log <- data.frame(
  variable_or_group = c(
    "Unnamed columns",
    "Cancelled/diverted flights",
    "Departure/arrival times",
    "Delay fields",
    "Carrier/origin categories",
    "Distance and elapsed time"
  ),
  observation = c(
    "Unnamed column contains only missing values",
    "Operational outcome fields are often missing for these flights",
    "Stored as HHMM integers rather than usable times",
    "Some zero-delay cases stored as NA when scheduled and actual times match",
    "High-cardinality categories with uneven group sizes",
    "Numeric variables look plausible but can contain influential extremes"
  ),
  action = c(
    "Drop",
    "Exclude from completed-flight modelling sample",
    "Convert to POSIXct and derive scheduled hour features",
    "Impute those cases with 0 when the equality check supports it",
    "Use volume filters in visual ranking and factor encoding in models",
    "Keep, but inspect visually and avoid unnecessary transformation unless analysis needs it"
  ),
  stringsAsFactors = FALSE
)

exploration_log
```

```{r duplicate_check}
dup_key <- c("FL_DATE", "OP_CARRIER", "OP_CARRIER_FL_NUM", "ORIGIN", "DEST", "CRS_DEP_TIME")
dup_key <- intersect(dup_key, names(flights_clean))

duplicate_rows <- duplicated(flights_clean[, dup_key])

data.frame(
  key_columns_checked = paste(dup_key, collapse = ", "),
  n_duplicate_rows = sum(duplicate_rows),
  pct_duplicate_rows = round(100 * mean(duplicate_rows), 4)
)
```

This duplicate check follows the workbook logic that cleaning is not
only about missing values. In this data set, the practical question is
whether two rows appear to describe the same scheduled flight. If this
count is small or zero, that supports the assumption that the unit of
analysis is one row per flight. If it is not, the duplicates need manual
inspection before modelling.

Some workbook preparation steps are not needed here. Data integration is
not relevant yet because we currently work with one main flight file.
Likewise, we do not apply global standardization at this stage because
our current models are tree-based and therefore do not require all
predictors to be on the same scale. Where scaling does matter, such as
in PCA, we explicitly do it there.

# 5. Exploratory visualisation

First we have a look at the base rate, this is the amount of how many
flights are more than 15 minutes delayed

```{r mean_delay_rate}
delay_rate <- mean(flights_clean$IS_DELAYED)
delay_rate 
```

First we show a bar chart of our data that simply displays the amount of
flights that are delayed and on-time. We see that around 18% of flights
are significantly delayed.

```{r}
df_target <- data.frame(
  class = factor(
    ifelse(flights_clean$IS_DELAYED, "Delayed (15 min)", "On time"),
    levels = c("On time", "Delayed (15 min)")
  )
)

ggplot(df_target, aes(x = class)) +
  geom_bar() +
  labs(title = "Class distribution: delayed vs on-time", x = "", y = "amount of flights") +
  coord_flip()
```

This class balance matters for the rest of the notebook. A delay rate of
roughly 18% means the data is not perfectly balanced, but the delayed
class is still common enough to model without the extreme rarity issues
that would occur for cancellations.

```{r numeric_vs_target_boxplots}
set.seed(46748717)
idx_box <- sample(seq_len(nrow(flights_clean)), min(150000, nrow(flights_clean)))
flights_box <- flights_clean[idx_box, ]

par(mfrow = c(1, 2))
boxplot(SCHED_DEP_HOUR ~ IS_DELAYED,
        data = flights_box,
        outline = FALSE,
        names = c("On time", "Delayed"),
        main = "Scheduled departure hour by delay status",
        ylab = "Scheduled departure hour",
        col = c("grey80", "tomato"))

boxplot(CRS_ELAPSED_TIME ~ IS_DELAYED,
        data = flights_box,
        outline = FALSE,
        names = c("On time", "Delayed"),
        main = "Scheduled elapsed time by delay status",
        ylab = "Scheduled elapsed time (minutes)",
        col = c("grey80", "tomato"))
par(mfrow = c(1, 1))
```

This figure adds the workbook perspective of comparing a numerical
variable against a categorical one. It helps us see whether delayed and
on-time flights differ in their planned characteristics before we start
modelling. We suppress plotted outlier points here because the sample is
very large; otherwise the boxes become unreadable. The distribution
itself remains visible through the median and spread.

Here we plot the delay rate per month and the amount of flights per
month. It is worth mentioning that the in the summer months the delay
amount significantly increasing proportional to the slight increase in
flights that we see in the graph.

```{r}
df_month_rate <- aggregate(IS_DELAYED ~ MONTH, flights_clean, function(x) mean(x, na.rm = TRUE))
df_month_n    <- aggregate(IS_DELAYED ~ MONTH, flights_clean, length)
names(df_month_n)[2] <- "n"

df_month <- merge(df_month_rate, df_month_n, by = "MONTH")
names(df_month)[2] <- "delay_rate"

pct_label <- function(x) paste0(round(100*x, 1), "%")

ggplot(df_month, aes(x = MONTH, y = delay_rate)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(labels = pct_label) +
  coord_cartesian(ylim = c(0, 0.30)) +
  labs(title = "Delay rate per month", x = "Month", y = "Delay rate")

ggplot(df_month, aes(x = MONTH, y = n)) +
  geom_col() +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Amount of flights per month", x = "Month", y = "#")
```

The delays per weekday shows that those are relatively stable during the
week, with values around 18-21%. On Fridays and Thursdays there are
slightly highiher percentagerates, while on Saterday their are clearly
lower delay rates. This pattern can be explained by the fact that there
are differneces in flightvolume and operational pressure: during the
week their are more commercial and business related flights that arwe
aslo connected, while on Saturday the operational pressure is lower,
what results in less delay. In general, it seems that the day of the
week does not have significantly influence on the change of delay, what
suggest that there are other factors, suchs as departure time, airport
or airplane that are maybe play ar ole in the explanination of the
delays.

```{r}
weekday_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

df_wday_rate <- aggregate(IS_DELAYED ~ WEEKDAY, flights_clean, mean)
names(df_wday_rate)[2] <- "delay_rate"
df_wday_rate$WEEKDAY <- factor(df_wday_rate$WEEKDAY, levels = weekday_levels)
df_wday_rate <- df_wday_rate[order(df_wday_rate$WEEKDAY), ]

ggplot(df_wday_rate, aes(x = WEEKDAY, y = delay_rate)) +
  geom_col() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Delay rate per weekday", x = "Day of the week", y = "Delay rate")

```

Heat map: weekday x departure hour

```{r}
df_hw <- flights_clean[, c("WEEKDAY", "SCHED_DEP_HOUR", "IS_DELAYED")]
df_hw$DEP_HOUR_BIN <- floor(df_hw$SCHED_DEP_HOUR)
df_hw <- aggregate(IS_DELAYED ~ WEEKDAY + DEP_HOUR_BIN, df_hw, mean)
names(df_hw)[3] <- "delay_rate"
df_hw$WEEKDAY <- factor(df_hw$WEEKDAY, levels = weekday_levels)

ggplot(df_hw, aes(x = DEP_HOUR_BIN, y = WEEKDAY, fill = delay_rate)) +
  geom_tile(color = "white") +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  scale_fill_gradient(low = "white", high = "steelblue", labels = pct_label, limits = c(0, 0.4)) +
  labs(title = "Delay rate by weekday and departure hour",
       x = "Planned departure hour", y = "Day of the week", fill = "Delay rate")

```

This heatmap is more informative than the weekday bar chart alone,
because average weekday effects can hide strong within-day patterns. If
delay risk rises later in the day, that suggests network accumulation:
once earlier flights run late, the rest of the day inherits part of that
disruption.

Delay rate per airplane. The first planes seem to be private jets.

```{r}
MIN_N_CARRIER <- 5000

carrier_rate <- aggregate(IS_DELAYED ~ OP_CARRIER, flights_clean, mean)
carrier_n    <- aggregate(IS_DELAYED ~ OP_CARRIER, flights_clean, length)
names(carrier_n)[2] <- "n"

carrier_summary <- merge(carrier_rate, carrier_n, by = "OP_CARRIER")
names(carrier_summary)[2] <- "delay_rate"

carrier_summary <- subset(carrier_summary, n >= MIN_N_CARRIER)
carrier_summary <- carrier_summary[order(-carrier_summary$delay_rate), ]

top_carriers <- head(carrier_summary, 10)

ggplot(top_carriers, aes(x = reorder(OP_CARRIER, delay_rate), y = delay_rate)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Delay rate (>=15m) per airline (top 10, min volume filter)",
       x = "Airplane", y = "Delay rate")
```

We apply a minimum-volume filter before ranking carriers and airports.
Without that filter, very small groups can appear extreme simply because
they have only a handful of flights. For an operational interpretation,
we prefer stable patterns over noisy ranking effects.

Delay rate per airport

```{r}
MIN_N_AIRPORT <- 20000

origin_rate <- aggregate(IS_DELAYED ~ ORIGIN, flights_clean, mean)
origin_n    <- aggregate(IS_DELAYED ~ ORIGIN, flights_clean, length)
names(origin_n)[2] <- "n"

origin_summary <- merge(origin_rate, origin_n, by = "ORIGIN")
names(origin_summary)[2] <- "delay_rate"

origin_summary <- subset(origin_summary, n >= MIN_N_AIRPORT)
origin_summary <- origin_summary[order(-origin_summary$delay_rate), ]

top_origins <- head(origin_summary, 15)

ggplot(top_origins, aes(x = reorder(ORIGIN, delay_rate), y = delay_rate)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Delay rate (>=15m) per origin airport (top 15, min volume filter)",
       x = "Origin airport", y = "Delay rate")
```

Scatterplot

```{r}
origin_top50_vol <- origin_summary[order(-origin_summary$n), ][1:50, ]

ggplot(origin_top50_vol, aes(x = n, y = delay_rate)) +
  geom_point() +
  scale_x_log10() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Origin airports: volume vs delay rate (top 50 volume)",
       x = "Aantal vluchten (log schaal)", y = "Delay rate")
```

The scatterplot helps separate two different effects: size and risk. A
high-volume airport with an above-average delay rate is operationally
more important than a tiny airport with a volatile rate, because the
large airport contributes more delayed flights in absolute numbers.

Multivariate relations and representation choices for features. In other
words distance and scheduled departure as risk factors

```{r}
set.seed(46748717)
N_VIS <- min(200000, nrow(flights_clean))
idx_vis <- sample(seq_len(nrow(flights_clean)), N_VIS)
flights_vis <- flights_clean[idx_vis, ]

distance_breaks <- pretty(flights_vis$DISTANCE, n = 12)
flights_vis$DISTANCE_BIN <- cut(
  flights_vis$DISTANCE,
  breaks = unique(distance_breaks),
  include.lowest = TRUE
)

distance_risk <- aggregate(IS_DELAYED ~ DISTANCE_BIN, flights_vis, mean)
distance_mid  <- aggregate(DISTANCE ~ DISTANCE_BIN, flights_vis, mean)
distance_n    <- aggregate(IS_DELAYED ~ DISTANCE_BIN, flights_vis, length)

names(distance_risk)[2] <- "delay_rate"
names(distance_mid)[2]  <- "distance_mid"
names(distance_n)[2]    <- "n"

distance_risk <- merge(distance_risk, distance_mid, by = "DISTANCE_BIN")
distance_risk <- merge(distance_risk, distance_n, by = "DISTANCE_BIN")
distance_risk <- subset(distance_risk, n >= 100)
distance_risk <- distance_risk[order(distance_risk$distance_mid), ]

print(
  ggplot(distance_risk, aes(x = distance_mid, y = delay_rate)) +
    geom_line() +
    geom_point() +
    scale_y_continuous(labels = pct_label, limits = c(0, 1)) +
    labs(title = "Delay risk vs distance",
         x = "Distance (miles)", y = "Delay risk")
)

flights_vis$DEP_HOUR_BIN <- floor(flights_vis$SCHED_DEP_HOUR)

hour_risk <- aggregate(IS_DELAYED ~ DEP_HOUR_BIN, flights_vis, mean)
hour_n    <- aggregate(IS_DELAYED ~ DEP_HOUR_BIN, flights_vis, length)

names(hour_risk)[2] <- "delay_rate"
names(hour_n)[2]    <- "n"

hour_risk <- merge(hour_risk, hour_n, by = "DEP_HOUR_BIN")
hour_risk <- subset(hour_risk, n >= 100)
hour_risk <- hour_risk[order(hour_risk$DEP_HOUR_BIN), ]

print(
  ggplot(hour_risk, aes(x = DEP_HOUR_BIN, y = delay_rate)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = 0:23) +
    scale_y_continuous(labels = pct_label, limits = c(0, 1)) +
    labs(title = "Delay risk vs planned departure time",
         x = "Scheduled departure hour", y = "Delay risk")
)
```

The two binned plots above are exploratory rather than causal. They help
us decide whether variables deserve a place in later models and how they
should be represented. In particular, scheduled departure time clearly
behaves as a continuous time-of-day feature and is therefore better used
as an engineered hour variable than as a raw HHMM code.

```{r route_risk_summary}
route_summary <- aggregate(IS_DELAYED ~ ORIGIN + DEST, flights_clean, mean)
route_n <- aggregate(IS_DELAYED ~ ORIGIN + DEST, flights_clean, length)
names(route_summary)[3] <- "delay_rate"
names(route_n)[3] <- "n"

route_summary <- merge(route_summary, route_n, by = c("ORIGIN", "DEST"))
route_summary <- subset(route_summary, n >= 3000)
route_summary <- route_summary[order(-route_summary$delay_rate), ]

head(route_summary, 10)
```

This extra route table adds one useful level of analysis beyond airline
and airport summaries. It helps us see whether elevated delay risk is
mainly tied to specific airports in general, or to specific airport-pair
combinations that may reflect congestion patterns, geography or network
timing.

# 6. Dimension reduction and clustering

PCA and clustering are exploratory tools, not mandatory steps. We only
use them if we can define a coherent pre-departure feature space. For
our flight data, this means that the representation may contain only
variables that are known before departure and that behave meaningfully
under variance- and distance-based methods.

This matters because PCA and clustering are much more sensitive to
representation choices than our tree-based predictive models. If we feed
them raw clock times, high-cardinality labels, or post-event variables,
the output will mostly reflect encoding artifacts or leakage rather than
useful operational structure.

## 6.1 PCA

PCA is included here as an exploratory tool, not as the main modelling
goal. The first question is therefore not "how many components do we
get?", but "which variables deserve to define the PCA/clustering
representation at all?"

```{r pca_cluster_screening}
structure_screen <- data.frame(
  variable = c(
    "FL_DATE",
    "OP_CARRIER",
    "OP_CARRIER_FL_NUM",
    "ORIGIN",
    "DEST",
    "DISTANCE",
    "CRS_DEP_TIME",
    "CRS_ARR_TIME",
    "CRS_ELAPSED_TIME",
    "DEP_TIME / ARR_TIME / WHEELS_*",
    "DEP_DELAY / ARR_DELAY",
    "Delay-cause columns",
    "CANCELLED / DIVERTED",
    "MONTH",
    "WEEKDAY"
  ),
  pca_role = c(
    "Transform",
    "Post hoc only",
    "Exclude",
    "Post hoc only",
    "Post hoc only",
    "Keep",
    "Transform",
    "Transform",
    "Keep",
    "Exclude",
    "Exclude",
    "Exclude",
    "Exclude",
    "Transform",
    "Transform"
  ),
  clustering_role = c(
    "Transform",
    "Post hoc only",
    "Exclude",
    "Post hoc only",
    "Post hoc only",
    "Keep",
    "Transform",
    "Transform",
    "Watch for redundancy",
    "Exclude",
    "Exclude",
    "Exclude",
    "Exclude",
    "Transform",
    "Transform"
  ),
  representation = c(
    "Do not use raw date; derive month/weekday information instead",
    "No dummy expansion in unsupervised step",
    "Identifier-like flight number; not analytically meaningful",
    "Retain for post hoc interpretation, not as dummy set",
    "Retain for post hoc interpretation, not as dummy set",
    "Use as numeric route-length feature",
    "Convert to cyclical sin/cos features",
    "Convert to cyclical sin/cos features",
    "Use as numeric schedule-length feature",
    "Known only after departure/arrival",
    "Leakage / target-adjacent outcome information",
    "Recorded after the event and therefore leakage",
    "Used only as sample filter, not as features",
    "Convert to cyclical sin/cos features",
    "Represent coarsely as weekend dummy"
  ),
  reason = c(
    "Raw dates create arbitrary scale differences",
    "One-hot encoding carriers would add dimensions and dominate structure",
    "Mostly identifies a flight record rather than a latent trait",
    "Airport dummies would mainly recover airport labels rather than shared structure",
    "Airport dummies would mainly recover airport labels rather than shared structure",
    "Operationally meaningful and known before departure",
    "Raw HHMM values are not valid linear distances",
    "Raw HHMM values are not valid linear distances",
    "Operationally meaningful and known before departure",
    "Post-event information must be excluded",
    "Would force PCA/clustering to recover realised delay instead of latent structure",
    "Pure leakage for pre-departure analysis",
    "These variables already define which rows remain in the analysis",
    "Month is cyclical rather than linear",
    "A low-dimensional weekend indicator is preferable to seven weekday dummies"
  ),
  stringsAsFactors = FALSE
)

structure_screen
```

The screening table leads to two important decisions. First, raw
categorical labels such as carrier and airport are not used to define
the PCA or clustering space, because one-hot encoding them would create
many sparse dimensions and make the analysis mostly recover those labels
themselves. Second, schedule variables must be re-encoded before use:
clock time and calendar position are cyclical, not linear.

```{r pca_cluster_preparation}
flights_structure <- flights_clean

flights_structure$DEP_HOUR_SIN <- sin(2 * pi * flights_structure$SCHED_DEP_HOUR / 24)
flights_structure$DEP_HOUR_COS <- cos(2 * pi * flights_structure$SCHED_DEP_HOUR / 24)
flights_structure$ARR_HOUR_SIN <- sin(2 * pi * flights_structure$SCHED_ARR_HOUR / 24)
flights_structure$ARR_HOUR_COS <- cos(2 * pi * flights_structure$SCHED_ARR_HOUR / 24)
flights_structure$MONTH_SIN <- sin(2 * pi * flights_structure$MONTH / 12)
flights_structure$MONTH_COS <- cos(2 * pi * flights_structure$MONTH / 12)
flights_structure$IS_WEEKEND <- ifelse(
  flights_structure$WEEKDAY %in% c("Saturday", "Sunday"), 1, 0
)

prep_decisions <- data.frame(
  feature = c(
    "DISTANCE",
    "CRS_ELAPSED_TIME",
    "SCHED_DEP_HOUR",
    "SCHED_ARR_HOUR",
    "MONTH",
    "WEEKDAY"
  ),
  preparation = c(
    "Z-score scaling only",
    "Z-score scaling only",
    "Cyclical encoding (sin/cos) + scaling",
    "Cyclical encoding (sin/cos) + scaling",
    "Cyclical encoding (sin/cos) + scaling",
    "Weekend dummy + scaling"
  ),
  rationale = c(
    "Meaningful numeric feature; no separate normalization needed once standardized",
    "Meaningful numeric feature; retained first to inspect overlap with distance",
    "Raw HHMM would create false distances, so cyclical encoding is required",
    "Raw HHMM would create false distances, so cyclical encoding is required",
    "Month 12 is adjacent to month 1, so linear coding is inappropriate",
    "A single dummy keeps dimensionality low and captures the main calendar contrast"
  ),
  stringsAsFactors = FALSE
)

prep_decisions
```

We deliberately use Z-score standardization rather than min-max
normalization. PCA is driven by variance and clustering here is based on
Euclidean distance, so it is more natural to put all features on a
common mean-0, sd-1 scale. We also do **not** force a log transformation
on `DISTANCE` or `CRS_ELAPSED_TIME`: these variables are already
operationally interpretable, and we first want to see whether shared
route structure appears without adding another transformation layer.

```{r pca_cluster_feature_space}
representation_vars <- c(
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "DEP_HOUR_SIN", "DEP_HOUR_COS",
  "ARR_HOUR_SIN", "ARR_HOUR_COS",
  "MONTH_SIN", "MONTH_COS",
  "IS_WEEKEND"
)

post_hoc_vars <- c("IS_DELAYED", "OP_CARRIER", "ORIGIN", "DEST")

structure_df <- flights_structure[, c(representation_vars, post_hoc_vars)]
structure_df <- na.omit(structure_df)

set.seed(46748717)
N_STRUCTURE <- min(50000, nrow(structure_df))
idx_structure <- sample(seq_len(nrow(structure_df)), N_STRUCTURE)

structure_sample <- structure_df[idx_structure, ]
structure_z <- as.data.frame(scale(structure_sample[, representation_vars]))

dim(structure_sample)
summary(structure_z)
```

## 6.2 Correlation heatmap

```{r correlation_heatmap}
cor_mat <- round(cor(structure_z), 2)
cor_mat

reshape2::melt(cor_mat) |>
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "firebrick", mid = "white", high = "steelblue", midpoint = 0) +
  labs(title = "Correlation heatmap of PCA/clustering feature space",
       x = "", y = "", fill = "Correlation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r correlation_pairs}
high_corr_idx <- which(abs(cor_mat) >= 0.75 & lower.tri(cor_mat), arr.ind = TRUE)

if (nrow(high_corr_idx) == 0) {
  high_corr_pairs <- data.frame(
    variable_1 = character(),
    variable_2 = character(),
    correlation = numeric()
  )
} else {
  high_corr_pairs <- data.frame(
    variable_1 = rownames(cor_mat)[high_corr_idx[, 1]],
    variable_2 = colnames(cor_mat)[high_corr_idx[, 2]],
    correlation = cor_mat[high_corr_idx],
    row.names = NULL
  )
}

high_corr_pairs
```

The heatmap is our checkpoint before PCA and clustering. If almost all
features were independent, PCA would add little value. If some pairs are
very strongly correlated, clustering would also double-count the same
underlying signal. In that case, we prefer to keep the richer set for
PCA diagnostics, but possibly use a slimmer set for clustering.

## 6.3 Principal component analysis

```{r pca_run}
pca_results <- prcomp(structure_z, center = FALSE, scale. = FALSE)

pca_var_explained <- (pca_results$sdev)^2 / sum((pca_results$sdev)^2)
pca_var_table <- data.frame(
  PC = paste0("PC", seq_along(pca_var_explained)),
  Variance_Explained = pca_var_explained,
  Cumulative = cumsum(pca_var_explained)
)

pca_var_table
```

```{r pca_variance_plots}
ggplot(pca_var_table, aes(x = PC, y = Variance_Explained)) +
  geom_col(fill = "steelblue") +
  labs(title = "Scree plot", x = "Principal component", y = "Variance explained")

ggplot(pca_var_table, aes(x = seq_along(Cumulative), y = Cumulative)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq_along(pca_var_table$PC), labels = pca_var_table$PC) +
  labs(title = "Cumulative explained variance",
       x = "Principal component", y = "Cumulative variance explained")
```

```{r pca_loadings}
top_loadings <- function(pca_obj, pc = 1, n = 5) {
  out <- data.frame(
    variable = names(pca_obj$rotation[, pc]),
    loading = pca_obj$rotation[, pc]
  )
  out$abs_loading <- abs(out$loading)
  out <- out[order(-out$abs_loading), ]
  head(out, n)
}

top_loadings(pca_results, pc = 1, n = 8)
top_loadings(pca_results, pc = 2, n = 8)
top_loadings(pca_results, pc = 3, n = 8)
```

The PCA here is not intended to manufacture a predictive feature set by
default. It is used to inspect whether a few components capture a large
share of the structure and whether those components are interpretable.
If the leading components are mainly combinations of route length and
schedule timing, then PCA still teaches us something useful: it tells us
which original variables overlap and where redundancy sits. We therefore
keep the interpretation at workbook level: the scree plot tells us how
much variance the first components capture, while the loading tables show
which original variables are driving those components.

```{r pca_followup_decisions}
cum4 <- pca_var_table$Cumulative[min(4, nrow(pca_var_table))]
distance_elapsed_cor <- cor(structure_z$DISTANCE, structure_z$CRS_ELAPSED_TIME)

pca_followup_decisions <- data.frame(
  decision_item = c(
    "Use principal components as direct replacement in later predictive models",
    "Keep both DISTANCE and CRS_ELAPSED_TIME in the clustering input",
    "Use carrier/airport dummies in PCA or clustering"
  ),
  decision = c(
    ifelse(
      cum4 >= 0.80,
      "Optional for later model comparison, but keep original engineered features as main representation",
      "No; keep PCA as exploratory diagnostic and retain original engineered features for modelling"
    ),
    ifelse(
      abs(distance_elapsed_cor) >= 0.85,
      "No; keep DISTANCE and drop CRS_ELAPSED_TIME from the cluster distance space",
      "Yes; overlap is not strong enough to justify dropping one upfront"
    ),
    "No; retain these variables only for post hoc interpretation"
  ),
  rationale = c(
    "Even when PCA compresses variance reasonably well, the original engineered variables are operationally easier to interpret",
    "If both variables capture nearly the same route-length signal, including both would overweight that structure in Euclidean clustering",
    "High-cardinality dummy sets would dominate the unsupervised representation and mainly recover label partitions"
  ),
  stringsAsFactors = FALSE
)

pca_followup_decisions
```

This table makes the follow-up decision explicit. In our project, PCA is
mainly useful as a screening and simplification diagnostic. Unless the
components turn out to be both compact **and** interpretable, we do not
replace the original engineered predictors in the main predictive
analysis.

## 6.4 Clustering

Clustering asks a different question than PCA: do flights form recurring
operational profiles under the chosen representation? Here that means
profiles of route length, scheduled timing and coarse calendar
structure. This is logically defensible, but it should be read
carefully: these clusters are not delay classes, and they are not
customer segments. At best, they describe schedule-and-route archetypes.

```{r clustering_input}
cluster_vars <- c(
  "DISTANCE",
  "DEP_HOUR_SIN", "DEP_HOUR_COS",
  "ARR_HOUR_SIN", "ARR_HOUR_COS",
  "MONTH_SIN", "MONTH_COS",
  "IS_WEEKEND"
)

if (abs(distance_elapsed_cor) < 0.85) {
  cluster_vars <- c("DISTANCE", "CRS_ELAPSED_TIME", setdiff(cluster_vars, "DISTANCE"))
}

set.seed(46748717)
N_CLUSTER <- min(20000, nrow(structure_sample))
idx_cluster <- sample(seq_len(nrow(structure_sample)), N_CLUSTER)

cluster_data_z <- structure_z[idx_cluster, cluster_vars, drop = FALSE]
cluster_data_original <- flights_structure[idx_structure[idx_cluster], c(
  "DISTANCE", "CRS_ELAPSED_TIME", "SCHED_DEP_HOUR", "SCHED_ARR_HOUR",
  "IS_WEEKEND", "IS_DELAYED", "OP_CARRIER", "ORIGIN", "DEST"
)]
```

```{r elbow_plot}
wss <- sapply(2:6, function(k) {
  kmeans(cluster_data_z, centers = k, nstart = 20)$tot.withinss
})

elbow_df <- data.frame(k = 2:6, tot_withinss = wss)

ggplot(elbow_df, aes(x = k, y = tot_withinss)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 2:6) +
  labs(title = "Elbow plot for k-means", x = "Number of clusters", y = "Total within-cluster sum of squares")
```

We use the elbow plot only as a heuristic. There is no single correct
number of clusters. We choose a moderate value that balances compactness
and interpretability rather than forcing very fine-grained segments.

```{r kmeans_clustering}
set.seed(46748717)
K_CHOSEN <- 4
kmeans_result <- kmeans(cluster_data_z, centers = K_CHOSEN, nstart = 25)

cluster_projection <- prcomp(cluster_data_z, center = FALSE, scale. = FALSE)$x[, 1:2]
cluster_plot_df <- data.frame(
  PC1 = cluster_projection[, 1],
  PC2 = cluster_projection[, 2],
  cluster = factor(kmeans_result$cluster)
)

set.seed(46748717)
idx_cluster_plot <- sample(seq_len(nrow(cluster_plot_df)), min(5000, nrow(cluster_plot_df)))

ggplot(cluster_plot_df[idx_cluster_plot, ], aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.2) +
  labs(title = "K-means clusters projected onto the first two dimensions")
```

```{r cluster_profiles}
cluster_profile_df <- cluster_data_original
cluster_profile_df$cluster <- factor(kmeans_result$cluster)
cluster_profile_df$IS_DELAYED <- as.numeric(cluster_profile_df$IS_DELAYED)

cluster_sizes <- data.frame(
  cluster = factor(names(table(cluster_profile_df$cluster))),
  n = as.integer(table(cluster_profile_df$cluster))
)

cluster_sizes

cluster_summary <- aggregate(
  cbind(DISTANCE, CRS_ELAPSED_TIME, SCHED_DEP_HOUR, SCHED_ARR_HOUR, IS_WEEKEND, IS_DELAYED) ~ cluster,
  data = cluster_profile_df,
  FUN = mean
)

cluster_summary
```

The cluster summaries should be interpreted as descriptive profiles. If
cluster delay rates differ only modestly while route length and schedule
features differ strongly, that means the clustering solution is mainly
recovering operational flight types rather than directly useful delay
risk segments. In that case, cluster labels may still be helpful for
communication or later experimentation, but they should not be treated
as essential modelling features.

```{r cluster_density_plots}
cluster_density_df <- cluster_profile_df

density_vars <- c("DISTANCE", "CRS_ELAPSED_TIME", "SCHED_DEP_HOUR", "SCHED_ARR_HOUR")
density_vars <- intersect(density_vars, names(cluster_density_df))

cluster_density_long <- reshape2::melt(
  cluster_density_df[, c("cluster", density_vars)],
  id.vars = "cluster"
)

ggplot(cluster_density_long, aes(x = value, color = cluster)) +
  geom_density() +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Cluster distributions for key continuous variables",
       x = "Value", y = "Density", color = "Cluster")
```

These density plots are useful here because they show whether the
clusters actually differ in the underlying continuous features, rather
than only in their mean values. We restrict them to continuous
variables. For binary variables such as `IS_WEEKEND`, density plots are
not very informative, so a table or mean per cluster remains clearer.

```{r hierarchical_clustering}
set.seed(46748717)
N_HCLUST <- min(150, nrow(cluster_data_z))
idx_hclust <- sample(seq_len(nrow(cluster_data_z)), N_HCLUST)

hclust_input <- cluster_data_z[idx_hclust, , drop = FALSE]
hclust_dist <- dist(hclust_input, method = "euclidean")
hclust_result <- hclust(hclust_dist, method = "ward.D2")

plot(hclust_result, labels = FALSE, hang = -1,
     main = "Hierarchical clustering dendrogram (sample)",
     xlab = "", sub = "", ylab = "Height")
```

Hierarchical clustering is included only on a small sample. For a data
set of this size it is not a realistic full-data segmentation tool,
because pairwise distances scale too poorly. We therefore use it as an
illustrative check on whether the schedule-and-route representation
shows nested similarity structure at all.

```{r clustering_followup_decisions}
clustering_followup_decisions <- data.frame(
  item = c(
    "Use k-means labels in the main predictive model right away",
    "Use hierarchical clustering as definitive segmentation",
    "Retain original engineered predictors after PCA/clustering"
  ),
  decision = c(
    "Not yet; treat clusters as exploratory profiles first",
    "No; use only as an illustrative small-sample check",
    "Yes; they remain the main representation for later supervised models"
  ),
  rationale = c(
    "These clusters describe route/time archetypes and must first prove added value before entering prediction",
    "The full flight data is too large and the result would still depend strongly on representation and sample choice",
    "The original engineered features stay closest to the business process and remain easier to interpret operationally"
  ),
  stringsAsFactors = FALSE
)

clustering_followup_decisions
```

# 7. Modelling

In this section we align more closely with the classification workbook.
The goal is not to try many complicated models, but to understand how a
small set of classification models behaves on our flight-delay task.
Throughout this section, the target class of interest is a delayed
flight, because from an operational perspective missing a real delay is
more costly than issuing a false alarm.

## 7.1 Classification setup

We prepare one modelling data set and then split it into training and
testing data. The training data is used to estimate the classification
rules, while the testing data is held back to inspect how the model
behaves on unseen flights.

## 7.2 Preparing the modelling data

```{r}
flights_model <- flights_clean

flights_model$IS_DELAYED_15 <- ifelse(flights_model$ARR_DELAY >= 15, "Delayed", "On_time")
flights_model$IS_DELAYED_15 <- factor(flights_model$IS_DELAYED_15)

flights_model$DEP_HOUR <- as.numeric(format(flights_model$CRS_DEP_TIME, "%H")) +
                          as.numeric(format(flights_model$CRS_DEP_TIME, "%M")) / 60

flights_model$ARR_HOUR <- as.numeric(format(flights_model$CRS_ARR_TIME, "%H")) +
                          as.numeric(format(flights_model$CRS_ARR_TIME, "%M")) / 60

factor_vars <- c("OP_CARRIER", "ORIGIN", "DEST", "WEEKDAY")

for (v in factor_vars) {
  if (v %in% names(flights_model)) {
    flights_model[[v]] <- factor(flights_model[[v]])
  }
}

model_vars <- c(
  "IS_DELAYED_15",
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "DEP_HOUR",
  "ARR_HOUR",
  "MONTH",
  "WEEKDAY",
  "OP_CARRIER",
  "ORIGIN",
  "DEST"
)

model_df <- flights_model[, model_vars]
model_df <- na.omit(model_df)

str(model_df)
summary(model_df$IS_DELAYED_15)

set.seed(46748717)

N_MODEL <- min(300000, nrow(model_df))   # adjust if needed
idx <- sample(seq_len(nrow(model_df)), N_MODEL)

model_df_small <- model_df[idx, ]
dim(model_df_small)

set.seed(46748717)

split <- rsample::initial_split(model_df_small, prop = 0.7)
training <- training(split)
testing  <- testing(split)

dim(training)
dim(testing)
```

The feature set stays deliberately simple and close to the workbook
style. We use scheduled timing, route length, carrier and coarse
calendar information, because these are known before departure and are
easy to interpret in operational terms.

```{r training_testing_comparison}
split_compare <- rbind(
  data.frame(split = "Training", training[, c("IS_DELAYED_15", "DISTANCE", "DEP_HOUR")]),
  data.frame(split = "Testing", testing[, c("IS_DELAYED_15", "DISTANCE", "DEP_HOUR")])
)

ggplot(split_compare, aes(x = IS_DELAYED_15, fill = split)) +
  geom_bar(position = "dodge") +
  labs(title = "Class balance in training and testing data",
       x = "Delay class", y = "Number of flights")

ggplot(split_compare, aes(x = DISTANCE, fill = split)) +
  geom_histogram(bins = 30, color = "white") +
  facet_wrap(~ split, ncol = 1) +
  labs(title = "Distance distribution: training vs testing",
       x = "Distance (miles)", y = "Count")

ggplot(split_compare, aes(x = DEP_HOUR, fill = split)) +
  geom_histogram(bins = 24, color = "white") +
  facet_wrap(~ split, ncol = 1) +
  labs(title = "Departure-hour distribution: training vs testing",
       x = "Scheduled departure hour", y = "Count")
```

These checks are simple but important. If the training and testing data
show a very different class balance or very different feature
distributions, then model evaluation becomes less trustworthy. Here we
mainly want to confirm that the random split produced two comparable
samples of the same flight population.

## 7.3 Model 1: C5.0 decision tree

```{r}

model_tree <- C50::C5.0(IS_DELAYED_15 ~ ., data = training)

summary(model_tree)
```

The decision tree is our first baseline because it is transparent. It
shows which variables are used to split the data and in what order. That
makes it easier to explain to non-technical readers than a more opaque
model.

```{r tree_conditions_plot}
tree_plot_model <- rpart::rpart(
  IS_DELAYED_15 ~ .,
  data = training,
  method = "class",
  control = rpart::rpart.control(maxdepth = 3, minsplit = 3000, cp = 0.001)
)

plot(tree_plot_model, uniform = TRUE, margin = 0.1)
text(tree_plot_model, use.n = TRUE, all = TRUE, cex = 0.7)
```

The C5.0 summary remains our main model output, but the extra shallow
tree above makes the split conditions easier to read visually. We keep
this plot intentionally small and shallow so the first decision rules are
still interpretable at workbook level.

## 7.4 Inspecting tree behaviour on unseen data

```{r}

pred_test <- predict(model_tree, testing)

gmodels::CrossTable(
  testing$IS_DELAYED_15,
  pred_test,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

# Accuracy
mean(pred_test == testing$IS_DELAYED_15)

actual <- testing$IS_DELAYED_15
pred   <- pred_test

tp <- sum(actual == "Delayed" & pred == "Delayed")
tn <- sum(actual == "On_time" & pred == "On_time")
fp <- sum(actual == "On_time" & pred == "Delayed")
fn <- sum(actual == "Delayed" & pred == "On_time")

accuracy  <- (tp + tn) / (tp + tn + fp + fn)
recall    <- tp / (tp + fn)
precision <- tp / (tp + fp)

data.frame(
  accuracy = accuracy,
  recall_delayed = recall,
  precision_delayed = precision,
  tp = tp, tn = tn, fp = fp, fn = fn
)
```

The confusion matrix is important because accuracy alone is not enough.
For this business problem, the most costly error is a false negative: a
flight predicted as on time that is actually delayed. Recall therefore
deserves more attention than overall accuracy.

## 7.5 Altering tree behaviour: boosting

```{r}
model_tree_boost <- C50::C5.0(IS_DELAYED_15 ~ ., data = training, trials = 10)

pred_boost <- predict(model_tree_boost, testing)

gmodels::CrossTable(
  testing$IS_DELAYED_15,
  pred_boost,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

tp_boost <- sum(actual == "Delayed" & pred_boost == "Delayed")
tn_boost <- sum(actual == "On_time" & pred_boost == "On_time")
fp_boost <- sum(actual == "On_time" & pred_boost == "Delayed")
fn_boost <- sum(actual == "Delayed" & pred_boost == "On_time")

data.frame(
  model = "Boosted C5.0",
  accuracy = (tp_boost + tn_boost) / (tp_boost + tn_boost + fp_boost + fn_boost),
  recall_delayed = tp_boost / (tp_boost + fn_boost),
  precision_delayed = tp_boost / (tp_boost + fp_boost)
)
```

Boosting is relevant here because it allows the model to correct some of
the mistakes made by earlier trees. We keep the number of trials modest
so the workflow remains simple and close to the workbook.

## 7.6 Altering tree behaviour: cost-sensitive classification

```{r}
cost_matrix <- matrix(c(NA, 1,
                        2, NA),
                      nrow = 2,
                      ncol = 2,
                      byrow = FALSE)
rownames(cost_matrix) <- colnames(cost_matrix) <- c("Delayed", "On_time")

model_tree_cost <- C50::C5.0(IS_DELAYED_15 ~ ., data = training, costs = cost_matrix)

pred_cost <- predict(model_tree_cost, testing)

gmodels::CrossTable(
  testing$IS_DELAYED_15,
  pred_cost,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

tp_cost <- sum(actual == "Delayed" & pred_cost == "Delayed")
tn_cost <- sum(actual == "On_time" & pred_cost == "On_time")
fp_cost <- sum(actual == "On_time" & pred_cost == "Delayed")
fn_cost <- sum(actual == "Delayed" & pred_cost == "On_time")

data.frame(
  model = "Cost-sensitive C5.0",
  accuracy = (tp_cost + tn_cost) / (tp_cost + tn_cost + fp_cost + fn_cost),
  recall_delayed = tp_cost / (tp_cost + fn_cost),
  precision_delayed = tp_cost / (tp_cost + fp_cost)
)
```

This step fits the business problem particularly well. We explicitly say
that missing a real delay should count as more costly than a false
alarm. That usually shifts the tree toward higher recall for delayed
flights, even if precision or accuracy decreases somewhat.

## 7.7 Model 2: Random Forest

```{r}
rf_vars <- c(
  "IS_DELAYED_15",
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "DEP_HOUR",
  "ARR_HOUR",
  "MONTH",
  "WEEKDAY",
  "OP_CARRIER"
)

training_rf <- training[, rf_vars]
testing_rf  <- testing[, rf_vars]

set.seed(46748717)
model_rf <- randomForest::randomForest(IS_DELAYED_15 ~ ., data = training_rf, ntree = 200)

pred_rf <- predict(model_rf, testing_rf)

tp_rf <- sum(actual == "Delayed" & pred_rf == "Delayed")
tn_rf <- sum(actual == "On_time" & pred_rf == "On_time")
fp_rf <- sum(actual == "On_time" & pred_rf == "Delayed")
fn_rf <- sum(actual == "Delayed" & pred_rf == "On_time")

data.frame(
  model = "Random forest",
  accuracy = (tp_rf + tn_rf) / (tp_rf + tn_rf + fp_rf + fn_rf),
  recall_delayed = tp_rf / (tp_rf + fn_rf),
  precision_delayed = tp_rf / (tp_rf + fp_rf)
)

importance(model_rf)
varImpPlot(model_rf, main = "Random forest variable importance")
```

The random forest adds a second benchmark with more flexibility than a
single tree. For this model we use a slightly simpler feature set than
for the decision tree, because the `randomForest` package in base usage
cannot handle predictors with very many factor levels, such as `ORIGIN`
and `DEST`. Excluding those variables is acceptable here because this
model is meant as a workbook-style benchmark rather than as the final
best possible model.

## 7.8 Comparison

```{r}
comparison_table <- data.frame(
  model = c("C5.0 tree", "Boosted C5.0", "Cost-sensitive C5.0", "Random forest"),
  accuracy = c(
    accuracy,
    (tp_boost + tn_boost) / (tp_boost + tn_boost + fp_boost + fn_boost),
    (tp_cost + tn_cost) / (tp_cost + tn_cost + fp_cost + fn_cost),
    (tp_rf + tn_rf) / (tp_rf + tn_rf + fp_rf + fn_rf)
  ),
  recall_delayed = c(
    recall,
    tp_boost / (tp_boost + fn_boost),
    tp_cost / (tp_cost + fn_cost),
    tp_rf / (tp_rf + fn_rf)
  ),
  precision_delayed = c(
    precision,
    tp_boost / (tp_boost + fp_boost),
    tp_cost / (tp_cost + fp_cost),
    tp_rf / (tp_rf + fp_rf)
  )
)

comparison_table
```

For the business case, recall on delayed flights remains the most
important metric. A model with slightly lower accuracy can still be
preferable if it catches more truly delayed flights, because missing a
real disruption is more costly operationally than raising some false
alarms.

# 8. Conclusions and recommendations

## 8.1 Operational recommendations

Based on this first notebook, the strongest practical recommendation is
to use the cleaned pre-departure data to create an early-warning view
for operations teams. The exploratory plots suggest that delay risk is
not random: it varies by month, time of day, carrier, airport and route.
That means an airline could already prioritize monitoring and standby
resources for flights that match historically riskier patterns.

## 8.2 Limitations & next steps

This notebook is still limited to historical operational variables and a
first set of engineered features. We have not yet added weather, lagged
network effects, inbound aircraft information or airport congestion
proxies. Those are relevant next steps because they may improve recall
without violating the pre-departure information constraint.

## 8.3 Assignment check

To make the workbook easier to assess against the assignment, we
summarize the main deliverables that are covered in the report:

-   Business context, analytics task and success criteria are stated
    explicitly.

-   Data exploration is documented with interpretation rather than plots
    only.

-   Data preparation choices are motivated, including missing values,
    datatype fixes and derived variables.

-   PCA and clustering are only used after variable screening and only
    on a coherent pre-departure feature space.

-   Classification models are evaluated with a train-test split and a
    confusion matrix, with explicit attention to recall versus
    precision.

-   Operational recommendations, limitations and next steps are stated
    separately.

This means the workbook now covers the main analytical elements that are
normally expected in the group assignment. What still matters most for
the final submission is that you verify the outputs after knitting and
that the written interpretations still match the actual numbers and
plots produced on your machine.

# 9. Appendix

## 9.1 Variable screening for PCA and clustering

The main text already contains the screening and preparation decisions
used for PCA and clustering. We reproduce the key tables here so the
representation choices remain easy to review.

```{r appendix_pca_screening}
structure_screen
prep_decisions
pca_followup_decisions
clustering_followup_decisions
```
