---
title: "MOT143A: Business Analytics group 20"
author: "Hidde Segers and Lotte Feilzer"
date: "2026-03-04"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

# 1. Business context and analytics task

## 1.1 Business problem

**Context:** Airlines running domestic flights in the United States
(U.S.) must manage constantly resources such as gates, crew members,
ground handling staff, and customer service capacity. These decisions
have to be made quickly and often with limited certainty.

Flight delays and cancellations can cause a lot of trouble for those who
are trying to get on time. When things go wrong, airlines have to pay
for: crew overtime, gate conflicts, aircraft re-positioning,
compensation, re-booking logistics, and re-putational damage. At the
same time, customer satisfaction declines, which affects their name and
in the long-term their revenue.

**Problem:** Delays and cancellations are really expensive for airlines
and can be frustrating for passengers. A single delay can create high
costs for crew overtime, gate conflicts, re-booking workloads. Because
flights are connected in a daily schedule, one delay can also cause
follow-on delays later in the network. This increases operational costs
and directly reduces customer satisfaction.

The practical challenge is that a lot of operational decisions must be
made before the flights departures. At that time it is still not clear
whether the flight is running on time or not. If the airline can figure
out flights that are likely to be delayed early, it can allocate
resources more active (e.g., assign standby crew, prepare customer
service capacity or adjust gate plans) and reduce the impact when such
delays occurs.

This leads to our business question:

**Business question:** How can we estimate the probability that a
scheduled flight will be delayed (e.g. more than 15 min) before
departure?

The goal is not to "predict the truth" perfectly, but to provide a risk
signal that helps operations teams prioritize attention and resources
towards flights with the highest expected delay risk. From an
organizational perspective, the insights are meant for operations
control centers, network planners and day-of-operations managers. The
model would be used in real-time operational setting, where planners
monitor flights and allocate resources (that are most of the time
limited). In this context, the prediction functions as a
decision-support tool that can help prioritize attention and
interventions for flights with the highest risk of delay.

## 1.2 Analytics task

This analysis is focused on the perspective of the airline operations
control and the airport station management to identify high risk
flights. This should support the workers with allocation of staff, gates
and premature consumer communication. The model will be used for
pre-departure planning setting and will be based on the historical
operational data of US domestic flights. The business problem will be
translated to an analytic task by approaching it as a predictive
risk-scoring problem. Three possible targets were considered:

-   **Classification A (delay):**

    **Target:** `IS_DELAYED_15 = 1` als `ARR_DELAY >= 15`, anders 0.

    Here it is analysed whether or not a flight will be significantly
    delayed or not (this is for flights that are neither diverted or
    delayed).

-   **Classification B (cancellation):**

    **Target:** `CANCELLED == 1`

    Here it is analysed whether or not a flight will be cancelled.

-   **Regression:**

    **Target:** `ARR_DELAY`

    Here it is analysed how much too late a flight will likely be in
    minutes.

The primary task chosen is classification A, this is because it will be
directly usable to relocate resources if needed and where focus in
logistics needs to focused. It is important to press that the model
can't reason and does not rely on causality. It also will not be able to
guarantee outcomes. It will estimate the risk for significant delay and
consequently improve and support operational actions. Classification for
cancellation will be harder because of the rarity of this event. The
regression of the delay is also hard to predict and the type of delay
can but does not also translate to how much delay there will be.

## 1.3 Success criteria

We consider the success criteria for this report in two perspectives.
Firstly, model criteria that will be need to be met to answer our
business question. Secondly, the business criteria that will ensure
positive effects of the model in practice.

From an operational point of view, missing a real delay is usually worse
than raising a false alarm. If the model says a flight will be on time
but it is actually delayed, the airline is unprepared. This can quickly
lead to crew problems, gate conflicts, and unhappy passengers.

If the model predicts a delay that does not happen, it may lead to some
extra costs because resources were allocated unnecessarily. However,
this is less harmful than being unprepared for a real delay.

For this reason, we focus on high recall for the delayed flight s. We
want the model to catch as many real delays as possible, even if that
means accepting some false alarms. At the same time, we monitor
precision to make sure we do not allocate too many unnecessary
resources.

# 2. Setup

Here all packages are downloaded to make the code reproducible.

```{r setup}
# CRAN mirror
local({r <- getOption("repos")
r["CRAN"] <- "https://cran.rstudio.com/"
options(repos = r)})

packages <- c("ggplot2", "Hmisc", "rsample", "gmodels", "C50", "randomForest",
  "factoextra", "lattice", "reshape2"
)

for (i in packages) {
  if (!require(i, character.only = TRUE)) {
    install.packages(i, dependencies = TRUE)
    library(i, character.only = TRUE)
  }
}

set.seed(46748717)
```

# 3. Data loading 

## 3.1 Data in `data/` folder

Here we read our data. In the code it is easy to change the year of data
we want to see, just change the variable `YEAR`. Because the data_files
are quite big we added the possibility to print only a specific amount
of rows as given in `nrows` a value of `-1` will give all rows.

```{r data_files}
data_dir <- "data"
YEAR <- 2018
csv_path <- file.path(data_dir, paste0(YEAR, ".csv"))
flights <- read.csv(csv_path, stringsAsFactors = FALSE, nrows = N_READ)
```

## 3.2 Read data

Now we read our actual data and print the dimensions and columns names
to get a first clue of what kind of data were dealing with.

```{r read_data}
N_READ <- -1
dim(flights)
names(flights)
```

As seen the datafile contains:

-   7,213,446 observations

-   28 variables

We see that each observation represents one commercial flight in 2018.
The variables are as follows and categorized in definitions that can be
used to refer to groups of variables further in the report:

-   **Flight identifiers:** FL_DATE, OP_CARRIER, OP_CARRIER_FL_NUM

-   **Route information:** ORIGIN, DEST

-   **Scheduled times:** CRS_DEP_TIME, CRS_ARR_TIME

-   **Actual flight performance:** DEP_TIME, ARR_TIME, DEP_DELAY,
    ARR_DELAY

-   **Operational metrics:** TAXI_OUT, TAXI_IN, AIR_TIME

-   **Delay causes:** CARRIER_DELAY, WEATHER_DELAY, NAS_DELAY,
    SECURITY_DELAY, LATE_AIRCRAFT_DELAY

-   **Flight status indicators:** CANCELLED, DIVERTED

# 4. Data exploration & preparation

## 4.1 Data summary

Here we show a small summary of the data. The following things are
noticeable. Columns in this datafile that contain dates are characters
preferably we change this later to datetimes. It is also remarkable that
the times are given in military time but as integers, this will make it
hard to calculate certain metrics inveolving time such as "time passed"
therefor we will preferably also change these datatypes later on to
datetimes. For metrics such as `TAXI_OUT` and `DEP_DELAY` the integers
are suitable because it represents the amount of minutes passed. We see
that for the `DIVERTED` and `CANCELLED` columns `0` and `1` are used
instead of Booleans to respresent `FALSE` and `TRUE` respectively, this
is also workable if remarked. Let's further explore the data.

```{r explore_basic}
str(flights)
```

## 4.2 Missing values overview and discussion

Here we check whether or not we have missing values in our dataset, how
many there are and what are the possible reasons for these missing
values are. We use logical reasoning in order to determine what to do
with potential missing data. Here we show a table that shows the amount
of NA values for each column and also a percentage of how many values
are missing in regards to the total dataset.

```{r missingness}
na_count <- sapply(flights, function(x) sum(is.na(x)))
na_pct   <- round(100 * na_count / nrow(flights), 2)

na_table <- data.frame(
  na_count = as.integer(na_count),
  na_pct = na_pct
)

na_table <- na_table[order(-na_table$na_pct), ]
na_table
```

Firstly we see a column `Unnamed..27` which seems to have only NA values
making the column absolute and removable. We see for the operational
variables and flight performance variables such as `DEP_TIME`,
`ARR_TIME`, `AIR_TIME`, `WHEELS_ON`, `WHEELS_OFF` that 1-2% of all
values are missing. We should note that all arrival variables have about
the same amount of NA values missing and idem for the departure values.
This probably means that the NA values are probably/hopefully
explainable for the same reasons. First we check whether the NAs
occurring in the delay rows all occur in the same row. The cell below
tells us this is the case. This lets us know that these `DELAY` causes
variables are jointly observed or missing. This is good sign. These rows
are supposed to be filled in after a flight and give a reason for the
delay if it was significant delayed (15 minutes or more). In our check
this also seems to be the case. All flights that are 15+ minutes delayed
have a cause and vice versa.

```{r delay-na-same}
cause_cols <- intersect(c("CARRIER_DELAY","WEATHER_DELAY",
                    "NAS_DELAY","SECURITY_DELAY","LATE_AIRCRAFT_DELAY"), names(flights))
delay15 <- !is.na(flights$ARR_DELAY) & flights$ARR_DELAY >= 15
cause_present <- !is.na(flights[[cause_cols[1]]])

miss <- is.na(flights[, cause_cols, drop=FALSE])
same_na_rows <- all(apply(miss, 2, function(x) identical(which(x), which(miss[,1]))))

cat("All the delay causes with NA values are for the same row, is", same_na_rows, "\n")
cat("Causes are present but there was no 15 min delay:", sum(cause_present & !delay15), "\n")
cat("More than 15 min delay but cause is NA value:", sum(delay15 & !cause_present))
```

Now we check if it possible for a delay flight to have multiple reasons
for delay this might handy to know later on. This seems to be possible
and is the case for 51.56% of the delayed flights.

```{r cause-compact}
k <- rowSums(flights[cause_present, cause_cols, drop=FALSE] > 0)
tab <- table(k)

multi_n <- sum(k > 1)
total_n <- length(k)
multi_pct <- round(100 * multi_n / total_n, 2)

cat("Of the flights with delay causes,", multi_pct, "% have multiple causes for delay.")
```

We now know that our delay causes columns are in fact complete. We can
now check if for the flights that have taken off all the important
operational and performance variables are present (`ARR_DELAY`,
`ARR_TIME` ect.) or that some or all values are missing for certain
rows. The NA values for the these rows as presented above could be
explained if these are the rows of flights that were diverted or
cancelled. If not, crucial information might be missing. If so, we
analyse this here.

```{r}
flights_wo_cancelled <- which(flights$CANCELLED == 0 & flights$DIVERTED == 0)
flights_wo_cancelled_df <- flights[flights_wo_cancelled, ]

na_count_wo_cd <- sapply(flights_wo_cancelled_df, function(x) sum(is.na(x)))
na_pct_wo_cd   <- round(100 * na_count_wo_cd / nrow(flights_wo_cancelled_df), 2)

na_table_wo_cd <- data.frame(
  variable = names(flights_wo_cancelled_df),
  na_count_wo_cd = as.integer(na_count_wo_cd),
  na_pct_wo_cd = na_pct_wo_cd
)

na_table_wo_cd <- na_table_wo_cd[order(-na_table_wo_cd$na_pct_wo_cd), ]
na_table_wo_cd
```

We can see that a flight that was either cancelled or diverted, it
accounts for the reasons that rows contain NA values for performance and
operational variables. We can see that the only two performance
variables that have still NA values are `ARR_DELAY` and `DEP_DELAY`,
why? These NA values might cause some trouble when modelling because we
use `ARR_DELAY` as predictors in the pre-departure model. Let's check by
showing the rows were this data is missing.

```{r check-missing-mostly-delays}
dep_delays_missing <- which(is.na(flights_wo_cancelled_df$DEP_DELAY)) 
arr_delays_missing <- which(is.na(flights_wo_cancelled_df$ARR_DELAY))
missing_delays <- flights_wo_cancelled_df[c(dep_delays_missing, arr_delays_missing), ]
head(missing_delays)
```

First we need to look into what we want to do with the rows in which the
`DEP_DELAY` and/or `ARR_DELAY` variables are missing. It seems that
there are only NA values for these columns when the `CRS_DEP_TIME` and
`DEP_TIME` have the same value and idem for the arrival. This means that
in this dataset the NA values just mean that plane left exactly at the
intended time. If this is the case we can replace the NA values in these
rows for 0 values and these NA values are also explained than as well.
We perform a quick check to see this is a 100 percent the case.

```{r fill_zero_delays}
dep_na <- !is.na(missing_delays$CRS_DEP_TIME) & !is.na(missing_delays$DEP_TIME) &
                 is.na(missing_delays$DEP_DELAY) &
                 missing_delays$CRS_DEP_TIME == missing_delays$DEP_TIME

arr_na <- !is.na(missing_delays$CRS_ARR_TIME) & !is.na(missing_delays$ARR_TIME) &
                 is.na(missing_delays$ARR_DELAY) &
                 missing_delays$CRS_ARR_TIME == missing_delays$ARR_TIME

cat("DEP_DELAY NA but not CRS_DEP_TIME==DEP_TIME:", sum(dep_na_other), "\n")
cat("ARR_DELAY NA but not CRS_ARR_TIME==ARR_TIME:", sum(arr_na_other))
```

Now we have analysed all NA values and begin to clean our data properly.

## 4.3 Cleaning and datatypes

First, we do some final sanity checks for the values for variables. We
check whether the distances are positive as well as airtime. We also
check whether the departure and arrival delays are plausible which all
seems to be the case.

```{r sanity_checks}
cat("There are", sum(flights$DISTANCE <= 0), "negative distances.", "\n")
cat("There are", sum(flights$AIR_TIME <= 0, na.rm = TRUE) , "negative airtimes.", "\n")
print("Departure delay summary:")
summary(flights$DEP_DELAY) 
print("Arrival delay summary:")
summary(flights$ARR_DELAY) 
```

We create a df named `flights_clean` to not overwrite the `flights`
variable that can be cleaned without overwriting our original data.

```{r create_flights_clean}
flights_clean <- flights
```

We begin with adjusting in accordance with the remarks made in section
3.2. We the `FL_DATE` column a datatype datetime and drop the unuseful
column `Unnamed..27` and ensure that the columns are the correct
datatypes.

```{r clean_types}
drop_cols <- grep("^Unnamed", names(flights_clean), value = TRUE)
flights_clean[drop_cols] <- NULL

flights_clean$FL_DATE <- as.Date(flights_clean$FL_DATE)

head(flights_clean, 4)
```

Now we do some minimal cleaning make sure the datatypes are correct. We
make the integer time columns POSIXct datetime columns to be able to
treat the time correctly.

```{r datetime_fix_overwrite}
hhmm_to_hhmm_str <- function(x) {
  x <- as.integer(x)
  hh <- x %/% 100
  mm <- x %% 100
  out <- sprintf("%02d:%02d", hh, mm)
  out[is.na(x) | mm > 59 | hh < 0 | hh > 29] <- NA
  out
}

time_cols <- intersect(c("WHEELS_OFF","WHEELS_ON","DEP_TIME","ARR_TIME","CRS_DEP_TIME","CRS_ARR_TIME"), names(flights_clean))

# Replace columns to POSIXct
for (cn in time_cols) {
  tmp_str <- hhmm_to_hhmm_str(flights_clean[[cn]])
  flights_clean[[cn]] <- as.POSIXct(
    paste(as.character(flights_clean$FL_DATE), tmp_str),
    format = "%Y-%m-%d %H:%M",
    tz = "UTC"
  )
}

# Overnight: is ARR_TIME < DEP_TIME + one day
if (all(c("DEP_TIME","ARR_TIME") %in% names(flights_clean))) {
  overnight <- !is.na(flights_clean$DEP_TIME) &
               !is.na(flights_clean$ARR_TIME) &
               (flights_clean$ARR_TIME < flights_clean$DEP_TIME)

  arrival_side <- intersect(c("ARR_TIME","WHEELS_ON","CRS_ARR_TIME"), time_cols)
  for (cn in arrival_side) {
    flights_clean[[cn]][overnight] <- flights_clean[[cn]][overnight] + 24*3600
  }
}

head(flights_clean[, time_cols, drop=FALSE], 5)
```

We now clean the NA values for the delay columns when the `CRS_DEP_TIME`
and `DEP_TIME` are the same and idem for the arrival times.

```{r impute_delays_posixct}
idx_dep0 <- which(
  is.na(flights_clean$DEP_DELAY) &
  !is.na(flights_clean$DEP_TIME) & !is.na(flights_clean$CRS_DEP_TIME) &
  flights_clean$DEP_TIME == flights_clean$CRS_DEP_TIME
)

idx_arr0 <- which(
  is.na(flights_clean$ARR_DELAY) &
  !is.na(flights_clean$ARR_TIME) & !is.na(flights_clean$CRS_ARR_TIME) &
  flights_clean$ARR_TIME == flights_clean$CRS_ARR_TIME
)

# vullen
flights_clean$DEP_DELAY[idx_dep0] <- 0
flights_clean$ARR_DELAY[idx_arr0] <- 0

# check
head(flights_clean[idx_dep0, c("CRS_DEP_TIME","DEP_TIME","DEP_DELAY")])
head(flights_clean[idx_arr0, c("CRS_ARR_TIME","ARR_TIME","ARR_DELAY")])
```

Now we check if all the data is clean. First we delete all the cancelled
and diverted rows because theserows are not relevant for our analysis.

```{r}
keep <- (is.na(flights_clean$CANCELLED) | flights_clean$CANCELLED == 0) &
        (is.na(flights_clean$DIVERTED)  | flights_clean$DIVERTED  == 0)

flights_clean <- flights_clean[keep, , drop = FALSE]

cat("Removed:", sum(!keep), "\n")
cat("Rows after:", nrow(flights_clean))
```

We decide to add some columns that make our analysis and modelling
easier later on, such as `IS_DELAYED` day of the week column which might
be interesting in the analysis and a column which contains a Boolean on
whether a flight is significantly delayed (15 minutes or later).Now we
add a column in which we see if a plane was significantly delayed.

```{r}
flights_clean$IS_DELAYED <- ifelse(flights_clean$DEP_DELAY > 15, TRUE, FALSE) 
```

## 4.4 Target definition (classification)

> **No leakage in modelling:** we will use only pre-departure features
> for the models. Targets come from outcome columns.

```{r targets}
# Delay target: ARR_DELAY >= 15 (only not-cancelled and not-diverted)
if (all(c("ARR_DELAY","CANCELLED","DIVERTED") %in% names(flights))) {
  flights$IS_DELAYED_15 <- ifelse(
    flights$CANCELLED == 0 & flights$DIVERTED == 0 & !is.na(flights$ARR_DELAY),
    as.integer(flights$ARR_DELAY >= 15), 
    NA
  )
}

# Cancellation target
if ("CANCELLED" %in% names(flights)) {
  flights$IS_CANCELLED <- flights$CANCELLED
}

table(flights$IS_DELAYED_15, useNA="ifany")
table(flights$IS_CANCELLED, useNA="ifany")
```

# 5. Exploratory visualisation (ggplot2)

```{r eda_plots}
if (all(c("MONTH","IS_DELAYED_15") %in% names(flights))) {
  dfm <- aggregate(IS_DELAYED_15 ~ MONTH, flights, function(x) mean(x, na.rm = TRUE))
  ggplot(dfm, aes(x = MONTH, y = IS_DELAYED_15)) +
    geom_line() + geom_point() +
    scale_x_continuous(breaks = 1:12) +
    ylab("Delay rate (>=15 min)") + ggtitle("Delay rate per month")
}
```

```{r eda_airline_airport}
# Filter to flights where delay target is defined
if (all(c("IS_DELAYED_15","OP_CARRIER","ORIGIN") %in% names(flights))) {
  flights_clean <- flights[!is.na(flights$IS_DELAYED_15), ]

  airline_summary <- aggregate(IS_DELAYED_15 ~ OP_CARRIER, flights_clean,
                               function(x) mean(x, na.rm = TRUE))
  airline_summary <- airline_summary[order(-airline_summary$IS_DELAYED_15), ]

  ggplot(airline_summary[1:10, ],
         aes(x = reorder(OP_CARRIER, IS_DELAYED_15), y = IS_DELAYED_15)) +
    geom_col() +
    coord_flip() +
    labs(title = "Delay rate (>=15 min) per airline (top 10)",
         x = "Airline", y = "Delay rate")
}
```

**Interpretation (write):** - Seasonality?\
- Which carriers/airports stand out?\
- Class imbalance?

# 6. Dimension reduction & clustering (Workbook: Dimension reduction and clustering)

We cluster **airports** based on *pre-departure* aggregated profiles
(interpretability).

```{r airport_profiles}
needed <- c("ORIGIN","IS_DELAYED_15","IS_CANCELLED","DISTANCE","CRS_ELAPSED_TIME")
needed <- needed[needed %in% names(flights)]

if (all(c("ORIGIN","IS_DELAYED_15") %in% names(flights))) {
  base_df <- flights[!is.na(flights$IS_DELAYED_15), ]

  # aggregate per ORIGIN
  ap <- aggregate(
    cbind(IS_DELAYED_15, IS_CANCELLED, DISTANCE, CRS_ELAPSED_TIME) ~ ORIGIN,
    data = base_df,
    FUN = function(x) mean(x, na.rm = TRUE)
  )

  # add volume
  ap_n <- aggregate(IS_DELAYED_15 ~ ORIGIN, data = base_df, FUN = length)
  names(ap_n)[2] <- "n"
  ap <- merge(ap, ap_n, by = "ORIGIN")

  names(ap) <- c("ORIGIN","delay_rate","cancel_rate","avg_distance","avg_sched_time","n")

  # scale numeric columns
  ap_scaled <- scale(ap[, c("delay_rate","cancel_rate","avg_distance","avg_sched_time","n")])

  set.seed(1)
  km <- kmeans(ap_scaled, centers = 4, nstart = 20)
  ap$cluster <- factor(km$cluster)

  head(ap)
}
```

```{r cluster_viz}
if (exists("ap")) {
  ggplot(ap, aes(x = delay_rate, y = cancel_rate, color = cluster)) +
    geom_point(alpha = 0.8) +
    ggtitle("Airport clusters (delay & cancellation rates)") +
    xlab("Delay rate (>=15 min)") + ylab("Cancellation rate")
}
```

## 6.1 PCA

```{r pca}
if (exists("ap_scaled")) {
  pca_res <- prcomp(ap_scaled, center = TRUE, scale. = TRUE)
  summary(pca_res)

  factoextra::fviz_eig(pca_res, addlabels = TRUE)
  factoextra::fviz_pca_biplot(pca_res, repel = TRUE)
}
```

## 6.2 Correlation heatmap (replace GGally with reshape2 + ggplot2)

```{r corr_heatmap}
if (exists("ap")) {
  num_df <- ap[, c("delay_rate","cancel_rate","avg_distance","avg_sched_time","n")]
  cm <- cor(num_df, use = "complete.obs")
  cm_melt <- reshape2::melt(cm)

  ggplot(cm_melt, aes(x = Var1, y = Var2, fill = value)) +
    geom_tile() +
    labs(title = "Correlation matrix (airport profile features)",
         x = "", y = "", fill = "cor") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

**Interpretation (write):** - What do clusters represent?\
- Do PCA axes have meaning (loadings)?\
- Would you use a cluster label as a feature (yes/no and why)?

# 7. Modelling (Workbook: Classification)

We fit two models (C5.0 tree + Random Forest) using **pre-departure**
features only.

```{r modelling_data}
TARGET <- "IS_DELAYED_15"  # TODO: or "IS_CANCELLED"

candidate_features <- c(
  "OP_CARRIER","ORIGIN","DEST",
  "MONTH","WEEKDAY","CRS_DEP_HOUR",
  "CRS_ELAPSED_TIME","DISTANCE"
)
features <- intersect(candidate_features, names(flights))

stopifnot(TARGET %in% names(flights))
stopifnot(length(features) >= 3)

model_df <- flights[, c(TARGET, features)]
model_df <- model_df[complete.cases(model_df), ]

# collapse high-cardinality airports to top-N
topN_collapse <- function(x, N = 40) {
  tb <- sort(table(x), decreasing = TRUE)
  keep <- names(tb)[1:min(N, length(tb))]
  out <- ifelse(x %in% keep, x, "OTHER")
  factor(out)
}

if ("ORIGIN" %in% names(model_df)) model_df$ORIGIN <- topN_collapse(model_df$ORIGIN, 40)
if ("DEST" %in% names(model_df)) model_df$DEST <- topN_collapse(model_df$DEST, 40)
if ("OP_CARRIER" %in% names(model_df)) model_df$OP_CARRIER <- factor(model_df$OP_CARRIER)
if ("WEEKDAY" %in% names(model_df)) model_df$WEEKDAY <- factor(model_df$WEEKDAY)

split <- rsample::initial_split(model_df, prop = 0.7)
train <- rsample::training(split)
test  <- rsample::testing(split)

dim(train); dim(test)
table(train[[TARGET]])
```

## 7.1 Baseline

```{r baseline}
y_true <- test[[TARGET]]
y_pred0 <- rep(0, length(y_true))

gmodels::CrossTable(y_true, y_pred0,
                    prop.chisq = FALSE, prop.t = FALSE, prop.r = FALSE,
                    dnn = c("Actual", "Predicted (always 0)"))
```

## 7.2 Model 1 — C5.0 decision tree

```{r c50}
form <- as.formula(paste(TARGET, "~", paste(features, collapse = " + ")))

train_c50 <- train
test_c50 <- test
train_c50[[TARGET]] <- factor(train_c50[[TARGET]])
test_c50[[TARGET]]  <- factor(test_c50[[TARGET]])

tree <- C50::C5.0(form, data = train_c50)
tree

pred_tree <- predict(tree, newdata = test_c50)
gmodels::CrossTable(test_c50[[TARGET]], pred_tree,
                    prop.chisq = FALSE, prop.t = FALSE, prop.r = FALSE,
                    dnn = c("Actual", "Predicted (C5.0)"))
```

## 7.3 Model 2 — Random Forest

```{r rf}
train_rf <- train
test_rf <- test
train_rf[[TARGET]] <- factor(train_rf[[TARGET]])
test_rf[[TARGET]]  <- factor(test_rf[[TARGET]])

rf <- randomForest::randomForest(form, data = train_rf, ntree = 200)
rf

pred_rf <- predict(rf, newdata = test_rf)
gmodels::CrossTable(test_rf[[TARGET]], pred_rf,
                    prop.chisq = FALSE, prop.t = FALSE, prop.r = FALSE,
                    dnn = c("Actual", "Predicted (RF)"))

randomForest::varImpPlot(rf)
```

## 7.4 Comparison (write)

Compare: - performance (confusion matrices) - interpretability (tree
rules vs RF importance) - robustness / overfitting risk - operational
usability (how would you use scores in planning?)

**We recommend model \_\_\_ because \_\_\_ .**

# 8. Conclusions & recommendations

## 8.1 Operational recommendations (write)

Examples: - “High-risk flight list” per day/airport → allocate buffers
(gate agents/crew) - proactive passenger communication above a risk
threshold - airport/carrier monitoring dashboard (month/weekday/hour)

## 8.2 Limitations & next steps (write)

Limitations: - no weather / ATC constraints - concept drift across years
(policies, demand, post-COVID changes) - class imbalance (especially
cancellations)

Next steps (conceptual, workbooks 6–8): - when would representation
learning add value and what data is needed? - why text analytics is not
suitable here (no text fields) - how RL could apply if reformulated as
sequential staffing decisions

# 9. Reproducibility

```{r sessioninfo}
sessionInfo()
```
