---
title: 'MOT143A: Business Analytics group 20'
author: "Nora Donders, Stijn, Brian, Hidde Segers and Lotte Feilzer"
date: "2026-03-04"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    highlight: tango
editor_options:
  markdown:
    wrap: 72
---

# 1. Business context and analytics task

## 1.1 Business problem

**Context:** Airlines operating domestic flights in the United States
need to manage gates, crews, ground handling staff and customer service
capacity. These decisions often have to be made quickly and without
knowing the whole of the situation.

Flight delays and cancellations are expensive for the airlines and
stressful for the passengers. Airlines have to deal with numerous
problems including crew overtime, gate conflicts, aircraft
repositioning, compensation, rebooking work and reputational damage.

**Problem:** A single delay usually has repercussions for other flights
as well. Flights are using the same planes and are therefor intertwined
with their schedule. If one flight is delayed, it creates more delays
later in the day raising operational costs and reducing customer
satisfaction.

The most challenging part of operations is that many decisions need to
be made before a flight departs. At that point, not all setbacks are
known, and therefore neither is the final outcome of the departure. If
an airline were able to identify flights with a high delay risk far
before departure, an appropiate solution can be thought beforehand.
Preparing standby crew, adjusting gate plans, or increasing customer
service capacity are examples of operations that the airline could
impose beforehand so further delays are avoided or reduced. Delays would
still occur, but the impact of a single delay on further delays, costs,
and customer dissatisfaction would be reduced.

**Business question:** How can we estimate the probability that a
scheduled flight will be delayed (more than 15 min) before departure?

The purpose of this analysis is not to predict the future perfectly. The
goal is to be able to make a risk signal that aids operations teams with
the division of focus attention and resources on flights with the
highest expected delay risk. From an organizational perspective, the
main users would be operations control centres, network planners, and
day-of-operations managers. In that setting, the model is best seen as a
decision-support tool rather than an automatic decision-maker.

## 1.2 Analytics task

This analysis takes the perspective of airline operations control and
airport station management. The aim is to identify higher-risk flights
early enough to support staffing, gate planning, and earlier passenger
communication. Because the decision moment is before departure, we only
want to use information that would realistically be known at that time.

This leads naturally to a predictive risk-scoring task based on
historical U.S. domestic flight data. Before choosing the final task, we
considered three possible targets:

-   **Classification I (delay):**

    **Target:** `IS_DELAYED_15 = 1` if `ARR_DELAY >= 15`, otherwise 0.

    This asks whether a completed flight arrives with a meaningful
    delay.

-   **Classification II (cancellation):**

    **Target:** `CANCELLED == 1`

    This asks whether a flight will be cancelled.

-   **Regression:**

    **Target:** `ARR_DELAY`

    This asks how many minutes of arrival delay we should expect.

This asks how many minutes of arrival delay we should expect.

We choose Classification I as the main task. That choice follows from
the business question: operations teams mainly need an early warning
about whether a flight is likely to become meaningfully delayed, because
that is directly useful for reallocating attention and resources. This
is also an assumption we make explicit in the report: a clear delayed vs
on-time signal is more actionable here than a raw minute prediction.

We do not choose cancellation as the main task because cancellations are
much rarer and therefore harder to model well. We also do not choose
regression as the main task because predicting the exact number of delay
minutes is harder to explain and less directly useful for the first
operational decision. So, the reasoning pipeline is that the business
problem is pre-departure resource allocation, which requires a
pre-departure risk signal, and therefore the main analytics task becomes
the binary classification of meaningful arrival delay.

## 1.3 Success criteria

We define success from two angles. First, the model should answer the
analytics question in a useable way. Second, the model should support
the business problem in practice.

From an operational point of view, missing a real delay is usually worse
than raising a false alarm. If the model says a flight will be on time
but it is actually delayed, the airline is unprepared. This can quickly
lead to crew problems, gate conflicts, and unhappy passengers.

If the model predicts a delay that does not happen, it may lead to some
extra costs because resources were allocated unnecessarily. However,
this is less harmful than being unprepared for a real delay.

For this reason, we focus on high recall for delayed flights. We want
the model to catch as many real delays as possible, even if that means
accepting some false alarms. At the same time, we monitor precision to
make sure we do not allocate too many unnecessary resources.

# 2. Setup

This section loads the packages used throughout the notebook so the
report stays reproducible for all readers.

```{r setup}
# CRAN mirror
local({r <- getOption("repos")
r["CRAN"] <- "https://cran.rstudio.com/"
options(repos = r)})

# Main packages 
packages <- c("ggplot2", "Hmisc", "rsample", "gmodels", "C50", "randomForest",
  "factoextra", "lattice", "reshape2", "pROC","e1071"
)

# Install missing packages and load everything
for (i in packages) {
  if (!require(i, character.only = TRUE)) {
    install.packages(i, dependencies = TRUE)
    library(i, character.only = TRUE)
  }
}

set.seed(46748717)
```

# 3. Data loading

## 3.1 Data in `data/` folder

Here, we read the flight data. The year can be changed through `YEAR`,
because the files are large. For the analysis in this workbook we use
the data of 2018.

```{r data_files}
# get data for year 2018
data_dir <- "data"
YEAR <- 2018
csv_path <- file.path(data_dir, paste0(YEAR, ".csv"))
flights <- read.csv(csv_path, stringsAsFactors = FALSE)
```

## 3.2 Read data

We first inspect the dimensions and column names to understand what kind
of data we are dealing with and what the scope and units of the analysis
are.

```{r read_data}
# size and columns dataset
dim(flights)
names(flights)
```

The dataset contains:

-   7,213,446 observations

-   28 variables

Each observation represents one commercial flight in 2018 in the United
States of America. The variables represent information about these
flights. Here, we print a concise summary.

```{r summary }
# print summary 
str(flights)
```

From this we can see the variable types and the first values. By
analysing we can write down the variable descriptions.

| Variable | Description | Variable type |
|------------------------|------------------------|------------------------|
| FL_DATE | Scheduled date of the flight | character (date string `YYYY-MM-DD` format) |
| OP_CARRIER | Airline operating the flight | character |
| OP\_ CARRIER_FL_NUM | Identifier number of the flight operated by the carrier | integer |
| ORIGIN | Airport from which the flight departs | character |
| DEST | Airport at which the flight arrives | character |
| CRS_DEP_TIME | Scheduled departure time | integer (`HHMM` format) |
| DEP_TIME | Actual departure time | numeric (`HHMM` format) |
| DEP_DELAY | Difference in minutes between actual and scheduled departure time | numeric (minutes) |
| TAXI_OUT | Time from leaving the departure gate until takeoff | numeric (minutes) |
| WHEELS_OFF | Actual time at which the aircraft leaves the ground | numeric (`HHMM` format) |
| WHEELS_ON | Actual time at which the aircraft touches down | numeric (`HHMM` format) |
| TAXI_IN | Time from landing until arrival at the gate | numeric (minutes) |
| CRS_ARR_TIME | Scheduled arrival time | integer (`HHMM` format) |
| ARR_TIME | Actual arrival time | numeric (`HHMM` format) |
| ARR_DELAY | Difference in minutes between actual and scheduled arrival time | numeric (minutes) |
| CANCELLED | Indicator showing whether the flight was cancelled | numeric (binary) |
| CAN CELLATION_CODE | Reason code explaining why the flight was cancelled | character |
| DIVERTED | Indicator showing whether the flight was diverted to another airport | numeric (binary) |
| CR S_ELASPED_TIME | Scheduled total travel time from departure to arrival | numeric (minutes) |
| ACTUA L_ELAPSED_TIME | Actual total travel time from departure to arrival | numeric (minutes) |
| AIR_TIME | Total time the aircraft is in the air | numeric (minutes) |
| DISTANCE | Flight distance between origin and destination airports | numeric (miles) |
| CARRIER_DELAY | Delay attributed to the airline carrier | numeric (minutes) |
| WEATHER_DELAY | Delay attributed to weather conditions | numeric (minutes) |
| NAS_DELAY | Delay attributed to the National Airspace System | numeric (minutes) |
| SECURITY_DELAY | Delay attributed to security-related issues | numeric (minutes) |
| LATE\_ AIRCRAFT_DELAY | Delay caused by the late arrival of the incoming aircraft | numeric (minutes) |
| Unnamed..27 | Unnamed column | logical |

Now we group the variables into a few practical categories so we can
refer back to them more easily later in the report.

| Category | Variables |
|------------------------------------|------------------------------------|
| Flight identifiers | FL_DATE, OP_CARRIER, OP_CARRIER_FL_NUM |
| Route information | ORIGIN, DEST |
| Scheduled times | CRS_DEP_TIME, CRS_ARR_TIME |
| Actual flight performance | DEP_TIME, ARR_TIME, DEP_DELAY, ARR_DELAY |
| Operational metrics | TAXI_OUT, TAXI_IN, AIR_TIME |
| Delay causes | CARRIER_DELAY, WEATHER_DELAY, NAS_DELAY, SECURITY_DELAY, LATE_AIRCRAFT_DELAY |
| Flight status indicators | CANCELLED, DIVERTED |

# 4. Data cleaning

## 4.1 Data summary

We start with a basic structural check of the data. This is important
because before we decide how to clean or model the data, we first need
to understand how the variables are stored. We inspect structure and
missing data points. Then we decide which datatype changes and cleaning
steps are actually needed for the pre-departure analysis.

To better align this notebook with the workbook on exploration and
preparation, we explicitly separate the analysis around three questions:

-   What are the type and scale of each variable?

-   Which variables show data quality issues or unusual structure?

-   Which of those issues require action before modelling?

The next sections answer these questions step by step. We focus on
variables that are most relevant for a pre-departure delay analysis,
rather than producing every possible plot for all 28 columns.

## 4.2 Missing values overview and discussion

We now inspect missing values. The point is not only to count `NA`s, but
also to understand why they occur. In operational data, it is known that
missingness can be structural rather than accidental. This makes it that
in the next cleaning steps we should prioritize the reason for the
missing value and not just on the amount. We start with counting the
missing values per column so we can get a good overview of the missing
values.

```{r missingness}
# calculate NA amount and the NA percentage per column
na_count <- sapply(flights, function(x) sum(is.na(x)))
na_pct   <- round(100 * na_count / nrow(flights), 2)

na_table <- data.frame(
  na_count = as.integer(na_count),
  na_pct = na_pct
)

na_table <- na_table[order(-na_table$na_pct), ]
na_table
```

This table suggests three obvious missingness patterns at first glance.
First, `Unnamed..27` contains only missing values, does not carry useful
information and can be dropped. Second, several operational variables
such as `DEP_TIME`, `ARR_TIME`, `AIR_TIME`, `WHEELS_ON`, and
`WHEELS_OFF` are missing in roughly the same small amount of rows. This
leads us to think that there is a shared operational reason for these
missing values rather than random data entry errors. Third, the
delay-cause columns also seem to be jointly missing or jointly present.

That leads to the next check. Before removing anything, we verify
whether the delay causes columns have structurally missing fields.
Logical thinking gives the impression that the missingness is because
these columns only are filled when a substantial realised delay exists.
If that is true, we should not treat those `NA`s as a data-quality
problem.

```{r delay-na-same}
# check delay-cause missingness 
cause_cols <- intersect(c("CARRIER_DELAY","WEATHER_DELAY",
                    "NAS_DELAY","SECURITY_DELAY","LATE_AIRCRAFT_DELAY"), names(flights))
delay15 <- !is.na(flights$ARR_DELAY) & flights$ARR_DELAY >= 15
cause_present <- !is.na(flights[[cause_cols[1]]])

miss <- is.na(flights[, cause_cols, drop=FALSE])
same_na_rows <- all(apply(miss, 2, function(x) identical(which(x), which(miss[,1]))))

# print concluson
cat("All the delay causes with NA values are for the same row, is", same_na_rows, "\n")
cat("Causes are present but there was no 15 min delay:", sum(cause_present & !delay15), "\n")
cat("More than 15 min delay but cause is NA value:", sum(delay15 & !cause_present))
```

As you can see in the summary above, all delay-cause variables with
missing values occur in the same row and there are no cases where causes
are present without a 15 minute delay, nor cases where a delay exceeds
15 minutes while the cause is missing. This indicates no structural data
quality problem regarding the data quality.

Next we check whether a delayed flight can have multiple recorded delay
causes. This is useful context for interpretation later on, because it
tells us whether delay causes behave like a single-label explanation or
whether multiple operational problems can happen on the same flight.

```{r cause-compact}
# select multiple delay causes
k <- rowSums(flights[cause_present, cause_cols, drop=FALSE] > 0)
tab <- table(k)

# calculate percentages
multi_n <- sum(k > 1)
total_n <- length(k)
multi_pct <- round(100 * multi_n / total_n, 2)

# print values 
cat("Of the flights with delay causes,", multi_pct, "% have multiple causes for delay.")
```

This summary above indicates that delays are often caused by multiple
factors rather than a single one. We now know that the delay-cause
columns are structurally complete for their purpose.

The next step is to check the missing values in the operational fields
such as `ARR_DELAY` and `ARR_TIME`. Here the hypothesis is that many of
those missing values belong to cancelled or diverted flights. If that is
true, the missingness is again explained by the flight outcome rather
than by poor data quality.

```{r non_cancelled_missingness_check}
# remove cancelled and diverted flights
completed_flight_rows <- which(flights$CANCELLED == 0 & flights$DIVERTED == 0)
completed_flights_df <- flights[completed_flight_rows, ]

completed_na_count <- sapply(completed_flights_df, function(x) sum(is.na(x)))
completed_na_pct   <- round(100 * completed_na_count / nrow(completed_flights_df), 2)

#check remaining missing values 
completed_missingness_table <- data.frame(
  na_count_completed = as.integer(completed_na_count),
  na_pct_completed = completed_na_pct
)

completed_missingness_table <- completed_missingness_table[order(-completed_missingness_table$na_pct_completed), ]
completed_missingness_table
```

This table shows that cancelled or diverted flights explain almost all
of the missingness in realised performance fields. After removing those
rows, the only important remaining gaps are in `ARR_DELAY` and
`DEP_DELAY` columns. That matters because `ARR_DELAY` is used to define
our target, so we need a clear reason for these remaining missing values
before moving on. We start by printing the rows where these `NA` values
are still remaining so that we can begin to evaluate the reason.

```{r check-missing-mostly-delays}
# remaining rows with missing delay values 
dep_delay_missing_rows <- which(is.na(completed_flights_df$DEP_DELAY)) 
arr_delay_missing_rows <- which(is.na(completed_flights_df$ARR_DELAY))
missing_delay_rows <- completed_flights_df[c(dep_delay_missing_rows, arr_delay_missing_rows), ]
head(missing_delay_rows)
```

We now inspect those remaining missing `ARR_DELAY` and `DEP_DELAY`
values. Looking at this table it is seen that in this dataset an `NA`
means that the scheduled and realised times are exactly equal, so the
delay should effectively be `0`. If that is true for all rows, these are
repairable missing values rather than rows we need to drop. We test that
assumption before imputing anything.

```{r fill_zero_delays}
# rows where scheduled and realised times are not the same but has NA value 
dep_delay_is_zero_case <- !is.na(missing_delay_rows$CRS_DEP_TIME) & !is.na(missing_delay_rows$DEP_TIME) &
                 is.na(missing_delay_rows$DEP_DELAY) &
                 !missing_delay_rows$CRS_DEP_TIME == missing_delay_rows$DEP_TIME

arr_delay_is_zero_case <- !is.na(missing_delay_rows$CRS_ARR_TIME) & !is.na(missing_delay_rows$ARR_TIME) &
                 is.na(missing_delay_rows$ARR_DELAY) &
                 !missing_delay_rows$CRS_ARR_TIME == missing_delay_rows$ARR_TIME

# print conclusion
cat("DEP_DELAY is a NA value but CRS_DEP_TIME is not equal to DEP_TIME:", sum(dep_delay_is_zero_case), "\n")
cat("ARR_DELAY is a NA value but CRS_ARR_TIME is not equal to ARR_TIME:", sum(arr_delay_is_zero_case))
```

This supports our assumptions that missing delay values occur when
scheduled and actual times are equal, and can thus be interpreted as
zero minutes.

At this point, we have explained the main missing value patterns. We can
conclude that all the data quality of this dataset is actually really
high. We can move on to cleaning with a clearer understanding and
knowing that the data is complete.

We shall not delete rows with missing values, because in the data the
missingness seems to be informative. We first distinguish between:

-   **structurally missing values:** for example delay causes that are
    only recorded after substantial delays

-   **outcome-related missing values:** for example flights that were
    cancelled or diverted and therefore have no actual arrival metrics

-   **repairable missing values:** for example cases where scheduled
    time and realised time are equal and the delay can logically be set
    to 0

We take these different observations with us to the next section where
we can clean the data and prepare it for our analysis.

## 4.3 Cleaning NA values and datatypes

At this moment a few things stand out from the concise summary printed
in 3.2 about the datatypes that occur in this dataset that might need to
be changed for convenience.

Firstly, `FL_DATE` is stored as date string, so we will later convert it
to a datetime variable to handle and plot the dates more easily. Several
scheduled and actual time variables are stored as integers in `HHMM`
format as shown in the table in 3.2 rather than as usable time values,
which makes direct time calculations easier. Variables such as
`TAXI_OUT` and `DEP_DELAY` are already stored as numeric minutes, which
are suitable for analysis. We also see that `CANCELLED` and `DIVERTED`
use bineary `0` and `1` coding rather than logical `TRUE` and `FALSE`,
which is fine as long as we handle it consistently and keep track of the
meaning. These observations will directly motivate the next steps.

We create a separate `flights_clean` object so the raw import remains
available for reference.

```{r create_flights_clean}
# make a copy of raw data 
flights_clean <- flights
```

The first cleaning step follows directly from the structural inspection
above. We convert `FL_DATE` to a real date and drop the unnamed column
that contains no information.

```{r clean_types}
# drop empty unnamed column and convert the flight date into datetime 
drop_cols <- grep("^Unnamed", names(flights_clean), value = TRUE)
flights_clean[drop_cols] <- NULL

flights_clean$FL_DATE <- as.Date(flights_clean$FL_DATE)

head(flights_clean, 4)
```

Next we deal with the time variables that are stored as `HHMM` integers,
we convert them to proper datetime values so that time-based comparisons
and derived hour variables become more reliable and more easy to plot.

```{r datetime_fix_overwrite}
# make a function that changes the hhmm to a hhmm string
hhmm_to_hhmm_str <- function(x) {
  x <- as.integer(x)
  hh <- x %/% 100
  mm <- x %% 100
  out <- sprintf("%02d:%02d", hh, mm)
  out[is.na(x) | mm > 59 | hh < 0 | hh > 29] <- NA
  out
}

# select the appropiate columns
time_cols <- intersect(c("WHEELS_OFF","WHEELS_ON","DEP_TIME","ARR_TIME","CRS_DEP_TIME","CRS_ARR_TIME"), names(flights_clean))

# replace columns to POSIXct
for (cn in time_cols) {
  tmp_str <- hhmm_to_hhmm_str(flights_clean[[cn]])
  flights_clean[[cn]] <- as.POSIXct(
    paste(as.character(flights_clean$FL_DATE), tmp_str),
    format = "%Y-%m-%d %H:%M",
    tz = "UTC"
  )
}

# keep overnight flights in mind: is ARR_TIME < DEP_TIME + one day
if (all(c("DEP_TIME","ARR_TIME") %in% names(flights_clean))) {
  overnight <- !is.na(flights_clean$DEP_TIME) &
               !is.na(flights_clean$ARR_TIME) &
               (flights_clean$ARR_TIME < flights_clean$DEP_TIME)

  arrival_side <- intersect(c("ARR_TIME","WHEELS_ON","CRS_ARR_TIME"), time_cols)
  for (cn in arrival_side) {
    flights_clean[[cn]][overnight] <- flights_clean[[cn]][overnight] + 24*3600
  }
}

head(flights_clean[, time_cols, drop=FALSE], 5)
```

We now fix the `NA` values occurrence in `ARR_DELAY` and `DEP_DELAY`
columns that are present when the scheduled and actual times are
identical and the delay is missing, we set the delay to `0`.

```{r impute_delays_posixct}
# select the rows where crs and actual time are the same 
idx_dep0 <- which(
  is.na(flights_clean$DEP_DELAY) &
  !is.na(flights_clean$DEP_TIME) & !is.na(flights_clean$CRS_DEP_TIME) &
  flights_clean$DEP_TIME == flights_clean$CRS_DEP_TIME
)

idx_arr0 <- which(
  is.na(flights_clean$ARR_DELAY) &
  !is.na(flights_clean$ARR_TIME) & !is.na(flights_clean$CRS_ARR_TIME) &
  flights_clean$ARR_TIME == flights_clean$CRS_ARR_TIME
)

# delay values are 0 if equal crs as actual time 
flights_clean$DEP_DELAY[idx_dep0] <- 0
flights_clean$ARR_DELAY[idx_arr0] <- 0

# print check
head(flights_clean[idx_dep0, c("CRS_DEP_TIME","DEP_TIME","DEP_DELAY")])
head(flights_clean[idx_arr0, c("CRS_ARR_TIME","ARR_TIME","ARR_DELAY")])
```

Now the needed datatype fixes are implemented and the repairable missing
values are handled. We still need to remove cancelled and diverted
flights because the main task is to compare completed flights on
meaningful arrival delay as our problem statement stated in the
beginning. The flights that are diverted and cancelled are not needed in
our scope and cause `NA` values in the data that corrupt our data
quality.

```{r filter_cancelled_diverted}
# delete diverted and cancelled flights to get completed-flight
keep <- (is.na(flights_clean$CANCELLED) | flights_clean$CANCELLED == 0) &
        (is.na(flights_clean$DIVERTED)  | flights_clean$DIVERTED  == 0)

flights_clean <- flights_clean[keep, , drop = FALSE]

cat("Removed:", sum(!keep), "\n")
cat("Rows after:", nrow(flights_clean))
```

Now the appropriate sample is selected, the next step is feature
creation for modeling. We insert a binary delay indicator for the main
target of significant delay of 15 minutes The target is
`ARR_DELAY >= 15`. This is the best definition for the business question
because it illustrates whether the flight arrived with a significant
delay. We have removed cancelled and diverted flights before defining
this target, so the class labels are only assigned to completed flights.
We also add calendar and scheduled-hour variables that are still known
before departure and therefore fit the pre-departure condition.

```{r create_analysis_features}
# create feature variables
flights_clean$IS_DELAYED <- ifelse(flights_clean$ARR_DELAY >= 15, TRUE, FALSE)
flights_clean$WEEKDAY <- weekdays(flights_clean$FL_DATE)
flights_clean$MONTH <- as.integer(format(flights_clean$FL_DATE, "%m"))
flights_clean$SCHED_DEP_HOUR <- as.numeric(format(flights_clean$CRS_DEP_TIME, "%H")) +
  as.numeric(format(flights_clean$CRS_DEP_TIME, "%M")) / 60
flights_clean$SCHED_ARR_HOUR <- as.numeric(format(flights_clean$CRS_ARR_TIME, "%H")) +
  as.numeric(format(flights_clean$CRS_ARR_TIME, "%M")) / 60
```

## 4.4 Extreme values

Before finishing up the cleaning, we do a few final sanity checks by
printing a detailed summary about the variables.

```{r summaries}
summary(flights_clean)
```

The overall summary indicates that most variables fall within plausible
ranges. For example, `DISTANCE` contains no negative values and appears
realistic. Negative values in `DEP_DELAY` and `ARR_DELAY` are also not
problematic, since flights may depart or arrive earlier than scheduled.

We consider the columns that contain data that will probably be key in
our modelling: `DISTANCE`, `CRS_ELAPSED_TIME`, `DEP_DELAY` and
`ARR_DELAY` and will be predictors for understanding the target or are
the target. We look at the values head on and decide if the values are
realistic.

```{r selected_numeric_diagnostics}
# select set of numeric variables
numeric_vars <- c("DISTANCE", "CRS_ELAPSED_TIME", "DEP_DELAY", "ARR_DELAY")
numeric_vars <- intersect(numeric_vars, names(flights_clean))

# make a center, spread and extreme values before plots 
numeric_diagnostics <- data.frame(
  variable = numeric_vars,
  mean = sapply(flights_clean[numeric_vars], function(x) round(mean(x, na.rm = TRUE), 2)),
  median = sapply(flights_clean[numeric_vars], function(x) round(median(x, na.rm = TRUE), 2)),
  sd = sapply(flights_clean[numeric_vars], function(x) round(sd(x, na.rm = TRUE), 2)),
  min = sapply(flights_clean[numeric_vars], function(x) round(min(x, na.rm = TRUE), 2)),
  max = sapply(flights_clean[numeric_vars], function(x) round(max(x, na.rm = TRUE), 2))
)

numeric_diagnostics
```

Comparing measures of central tendency and spread shows that the
distributions are not fully symmetric. In particular, `DEP_DELAY` and
`ARR_DELAY` are clearly right-skewed. Their means are positive (`9.89`
and `5.05`), while their medians are negative (`-2` and `-6`),
indicating that many flights depart or arrive on time or slightly early,
whereas a smaller number of severe delays pull the average upward.
`DISTANCE` and `CRS_ELAPSED_TIME` also have means above their medians,
which suggests some positive skewness there as well.

The large standard deviations and extreme maximum values confirm
substantial spread and the presence of outliers. For the cleaned data,
no invalid values remain for `DISTANCE` or `CRS_ELAPSED_TIME`. However,
extreme delay values are still present: 61 observations for `DEP_DELAY`
and 66 observations for `ARR_DELAY` exceed 1440 minutes in absolute
value. These observations are rare and may still be valid, so they are
flagged for inspection rather than removed automatically.

```{r clean_numeric_values}
# flag impossible values
flights_clean$flag_distance_invalid <- !is.na(flights_clean$DISTANCE) & flights_clean$DISTANCE <= 0
flights_clean$flag_crs_elapsed_invalid <- !is.na(flights_clean$CRS_ELAPSED_TIME) & flights_clean$CRS_ELAPSED_TIME <= 0

# flag extreme but plausible delay values
flights_clean$flag_dep_delay_extreme <- !is.na(flights_clean$DEP_DELAY) & abs(flights_clean$DEP_DELAY) > 1440
flights_clean$flag_arr_delay_extreme <- !is.na(flights_clean$ARR_DELAY) & abs(flights_clean$ARR_DELAY) > 1440

# summary of flagged observations
data_quality_summary <- data.frame(
  issue = c(
    "Invalid DISTANCE (<= 0)",
    "Invalid CRS_ELAPSED_TIME (<= 0)",
    "Extreme DEP_DELAY (abs > 1440)",
    "Extreme ARR_DELAY (abs > 1440)"
  ),
  n = c(
    sum(flights_clean$flag_distance_invalid, na.rm = TRUE),
    sum(flights_clean$flag_crs_elapsed_invalid, na.rm = TRUE),
    sum(flights_clean$flag_dep_delay_extreme, na.rm = TRUE),
    sum(flights_clean$flag_arr_delay_extreme, na.rm = TRUE)
  )
)

data_quality_summary

# replace impossible values with NA
flights_clean$DISTANCE[flights_clean$flag_distance_invalid] <- NA
flights_clean$CRS_ELAPSED_TIME[flights_clean$flag_crs_elapsed_invalid] <- NA
```

Overall, the cleaned data appears to be plausible and largely free of
impossible numeric values, while a small number of extreme but
potentially valid delay observations remain.

## 4.5 Exploration and preparation decisions

The exploration of data that was conducted until now should lead to
explicit preparation choices. We here summarize the main findings in a
short table.

| Case | Observation | Action |
|------------------------|------------------------|------------------------|
| Unnamed columns | Unnamed column contains only missing values | Drop |
| Cancelled or diverted flights | Operational outcome fields are often missing for these flights | Exclude from flight modelling sample |
| Departure and arrival times | Stored as HHMM numerics rather than usable times | Convert to POSIXct and derive scheduled hour features |
| Delay fields | Zero-delay cases are stored as `NA` | Impute with `0` when the equality check calls for it |
| Carrier and origin categories | These are categories with uneven group sizes | Use volume filters in visual ranking and factor encoding in models |
| Distance and elapsed time | Numeric variables look plausible | Keep, but inspect visually and avoid unnecessary transformation |

```{r duplicate_check}
# duplicate check for flights 
dup_key <- c("FL_DATE", "OP_CARRIER", "OP_CARRIER_FL_NUM", "ORIGIN", "DEST", "CRS_DEP_TIME")
dup_key <- intersect(dup_key, names(flights_clean))

duplicate_rows <- duplicated(flights_clean[, dup_key])

data.frame(
  key_columns_checked = paste(dup_key, collapse = ", "),
  n_duplicate_rows = sum(duplicate_rows),
  pct_duplicate_rows = round(100 * mean(duplicate_rows), 4)
)
```

This check looks for duplicate flights and thus two rows describe the
same scheduled flight. If there are a few or no duplicates, this
supports the idea that each row represents one flight. If not, the
duplicates need to be checked before we move on to the modelling part.

Data integration is not relevant yet because we currently work with one
main flight file. Likewise, we do not apply global standardisation at
this stage because our planned models are tree-based and therefore do
not require all predictors to be on the same scale. Where scaling does
matter, such as in PCA, we apply it at that stage in specific.

# 5. Exploratory visualisation

We start with the exploratory visualisation with the base rate: the
share of completed flights that arrive 15 minutes late or more. This is
the natural first plot, because it tells us how common the target event
is before we interpret any subgroup patterns.

```{r mean_delay_rate}
# base rate of delayed flights
delay_rate <- mean(flights_clean$IS_DELAYED)
delay_rate 
```

This shows that about 19% of the flights are delayed by 15 minutes or
more, providing a baseline to which we can compare all subgroup
patterns. We first show a simple bar chart with delayed versus on-time
flights. This gives a direct visual check of the class balance.

```{r class_distribution_plot}
# target into class labels bar plot
df_target <- data.frame(
  class = factor(
    ifelse(flights_clean$IS_DELAYED, "Delayed (15 min)", "On time"),
    levels = c("On time", "Delayed (15 min)")
  )
)

ggplot(df_target, aes(x = class)) +
  geom_bar() +
  labs(title = "Class distribution: delayed versus on time", x = "", y = "Number of flights") +
  coord_flip()
```

This class balance matters for the rest of the notebook. A delay rate of
roughly 19% means the data is not perfectly balanced, but the delayed
class is still common enough to model without the extreme rarity issues
that would occur for cancellations. At the same time, the graph makes
clear that "On time" is still the dominant class. That matters when we
evaluate models later, because a model can look acceptable on accuracy
while still doing a poor job on the delayed flights we care about most.

We shall now plot some exploratory plots to get some insights into the
skewness of the observations and potentially unrealistic influential
observations.

```{r selected_numeric_plots}
# diagnostic numeric plots 
par(mfrow = c(2, 2))
hist(flights_clean$DISTANCE, breaks = 40, main = "Distance", xlab = "Miles", col = "grey")
boxplot(flights_clean$DISTANCE, horizontal = TRUE, main = "Distance boxplot", col = "grey")
hist(flights_clean$ARR_DELAY, breaks = 40, main = "Arrival delay", xlab = "Minutes", col = "grey")
boxplot(flights_clean$ARR_DELAY, horizontal = TRUE, main = "Arrival delay boxplot", col = "grey")
par(mfrow = c(1, 1))
```

As we can see a few conclusions can be made from these plots. The plots
confirm that both `DISTANCE` and `ARR_DELAY` are strongly right-skewed.
For `DISTANCE`, most observations are concentrated among shorter and
medium distance flights, while a smaller number of long distance flights
that create a long right tail. The boxplot shows theese many
high-distance observations as outliers, but these appear to be plausible
values rather than data errors.

The distribution of `ARR_DELAY` is even more strongly skewed as shown by
the few bins in the histogram plots. Most flights are clustered around
zero delay, with some slightly negative values, meaning that many
flights arrive on time or a little early. At the same time, there is a
long right tail with a relatively small number of flights experiencing
very large delays. This indicates that raw arrival delay is dominated by
extreme observations and is not a symmetric distribution.

These plots show that the decision to model delay as a binary
classification target rather than using raw arrival delay minutes
directly is a good idea. The binary target will make interpretation much
easier. They also show that extreme values are part of the natural
structure of the data.

```{r numeric_vs_target_boxplots}
# delayed and on-time flights the boxplots
par(mfrow = c(1, 2))
boxplot(SCHED_DEP_HOUR ~ IS_DELAYED,
        data = flights_clean,
        outline = FALSE,
        names = c("On time", "Delayed"),
        main = "Scheduled departure hour by delay status",
        ylab = "Scheduled departure hour",
        col = c("grey", "red"))

boxplot(CRS_ELAPSED_TIME ~ IS_DELAYED,
        data = flights_clean,
        outline = FALSE,
        names = c("On time", "Delayed"),
        main = "Scheduled elapsed time by delay status",
        ylab = "Scheduled elapsed time (min)",
        col = c("grey", "red"))
par(mfrow = c(1, 1))
```

In these two boxplots, we compare a numerical variable with a
categorical one. This helps us analyse whether delayed and on-time
flights differ in their planned characteristics before moving on to
modelling. The main structure of the distributions remains visible
through the medians, interquartile ranges, and whiskers.

What we mainly want to examine is whether the centre and spread differ
between the two groups. For scheduled departure hour, the median is
clearly higher for delayed flights than for on-time flights, which
suggests that flights scheduled later in the day are more likely to be
delayed. The entire box for delayed flights also lies somewhat higher,
indicating that this difference is not only due to a few extreme
observations. For scheduled elapsed time, the difference between the two
groups is much smaller. Although delayed flights appear to have a
slightly higher median scheduled elapsed time, the two distributions
however seem to overlap mostly, suggesting that this variable may
contain only limited predictive signal on its own. The purpose here is
not to establish causality, but to check whether delayed and on-time
flights already show visible differences in simple planned
characteristics before departure.

The following two plots below show delay rate per month and the number
of flights per month. The plots show a clear seasonal pattern in flight
delays. The delay rate increases gradually in the spring and reaches its
highest levels in June, July and August and then drops sharply again in
September and October. Although the number of flights is somewhat higher
in the summer months, the variation in flight volume is much less than
the variation in delay rate. This shows that the monthly pattern is not
only driven by traffic volume. This is supported by the fact that in
September there is still a relatively high number of flights, while the
delay rate falls rapidly. This indicates seasonality as an independent
explanatory factor. The pattern also appears non-linear, suggesting that
`MONTH` may be more informative when treated as a categorical variable
rather than a purely numeric one. It is worth noting that these plots
are based on a single calendar year. The suggestion of seasonality is
there but do not by themselves prove that the same monthly pattern would
repeat identically in other years.

```{r monthly_delay_patterns}
# monthly delay rate and monthly flight volume 
df_month_rate <- aggregate(IS_DELAYED ~ MONTH, flights_clean, function(x) mean(x, na.rm = TRUE))
df_month_n    <- aggregate(IS_DELAYED ~ MONTH, flights_clean, length)
names(df_month_n)[2] <- "n"

df_month <- merge(df_month_rate, df_month_n, by = "MONTH")
names(df_month)[2] <- "delay_rate"

pct_label <- function(x) paste0(round(100*x, 1), "%")

ggplot(df_month, aes(x = MONTH, y = delay_rate)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(labels = pct_label) +
  coord_cartesian(ylim = c(0, 0.30)) +
  labs(title = "Delay rate per month", x = "Month", y = "Delay rate")

ggplot(df_month, aes(x = MONTH, y = n)) +
  geom_col() +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Amount of flights per month", x = "Month", y = "#")
```

The weekday pattern is relatively stable as shown below in the graph.
Most days sit close together, with slightly higher delay rates on
Thursday and Friday and a lower rate on Saturday. So weekday does matter
a bit, but not as strongly as some other variables. That suggests it is
useful to keep `WEEKDAY` in the analysis, but not to expect it to
explain delays on its own. In other words, weekday seems more like a
supporting variable than one of the strongest drivers in the analysis.

```{r weekday_delay_rates}
# weekday as delay risk.
weekday_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

df_wday_rate <- aggregate(IS_DELAYED ~ WEEKDAY, flights_clean, mean)
names(df_wday_rate)[2] <- "delay_rate"
df_wday_rate$WEEKDAY <- factor(df_wday_rate$WEEKDAY, levels = weekday_levels)
df_wday_rate <- df_wday_rate[order(df_wday_rate$WEEKDAY), ]

ggplot(df_wday_rate, aes(x = WEEKDAY, y = delay_rate)) +
  geom_col() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Delay rate per weekday", x = "Day of the week", y = "Delay rate")

```

Now we plot a heat map to get some more information on top of the
weekday bar chart alone because it shows the joint pattern of weekday
and departure hour. It shows delay risk build up during the day rather
than averaging that pattern away in the weekday graph. The darker cells
later in the day suggest that flights scheduled in the afternoon and
evening are more likely to be delayed, which is consistent with delays
causing more delay later in the day. Overall this plot shows that
departure time carries stronger predictive signal than weekday on its
own, while the combination of both variables may still add useful
information to later models.

```{r weekday_hour_heatmap}
# weekday and departure hour joint pattern
df_hw <- flights_clean[, c("WEEKDAY", "SCHED_DEP_HOUR", "IS_DELAYED")]
df_hw$DEP_HOUR_BIN <- floor(df_hw$SCHED_DEP_HOUR)
df_hw <- aggregate(IS_DELAYED ~ WEEKDAY + DEP_HOUR_BIN, df_hw, mean)
names(df_hw)[3] <- "delay_rate"
df_hw$WEEKDAY <- factor(df_hw$WEEKDAY, levels = weekday_levels)

ggplot(df_hw, aes(x = DEP_HOUR_BIN, y = WEEKDAY, fill = delay_rate)) +
  geom_tile(color = "white") +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  scale_fill_gradient(low = "white", high = "blue", labels = pct_label, limits = c(0, 0.4)) +
  labs(title = "Delay rate by weekday and departure hour",
       x = "Planned departure hour", y = "Day of the week", fill = "Delay rate")

```

So far, the results suggest that delay risk changes over the day. The
next plot shows delay rate by operating airline. The differences between
carriers are very big, which suggests that airline contains useful
predictive information for delay status. At the same time, these
differences should not immediately be interpreted as pure carrier
performance, because they may also reflect differences in route
networks, airports, departure times, and operating conditions.

```{r carrier_delay_rates}
# minimum number of flights per carrier
MIN_N_CARRIER <- 50000

# delay rate and number of flights per carrier
carrier_rate <- aggregate(IS_DELAYED ~ OP_CARRIER, flights_clean, mean)
carrier_n    <- aggregate(IS_DELAYED ~ OP_CARRIER, flights_clean, length)
names(carrier_n)[2] <- "n"

carrier_summary <- merge(carrier_rate, carrier_n, by = "OP_CARRIER")
names(carrier_summary)[2] <- "delay_rate"

# keep only carriers with enough volume
carrier_summary_filtered <- subset(carrier_summary, n >= MIN_N_CARRIER)

# sort by delay rate
carrier_summary_filtered <- carrier_summary_filtered[order(-carrier_summary_filtered$delay_rate), ]

# top 10 after volume filter
top_carriers <- head(carrier_summary_filtered, 10)

ggplot(top_carriers, aes(x = reorder(OP_CARRIER, delay_rate), y = delay_rate)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = pct_label) +
  labs(
    title = paste0("Delay rate (>= 15 min) per airline, min ", MIN_N_CARRIER, " flights"),
    x = "Airline",
    y = "Delay rate"
  )
```

The plot above shows clear differences in delay rates across airlines,
with some carriers having much higher delay rates than others. Before
ranking airlines and airports, we apply a minimum volume filter. This
makes sure that very small groups do not appear at the top or bottom of
the ranking just because they contain only a few flights. We are
interested in stable patterns rather than random extremes. We assume
that groups with at least 50000 flights provide a more reliable picture
than very small groups. The cutoff is chosen for stability and
interpretability, not because this amount represents a universal
threshold. This is completely arbitrary.

In addition to airline differences, delay risk may also depend on the
difference in airport. We therefore plot delay rates by origin airports.
We applied a volume filter of 20000 this time around.

```{r origin_delay_rates}
# origin airports plots 
MIN_N_AIRPORT <- 20000

origin_rate <- aggregate(IS_DELAYED ~ ORIGIN, flights_clean, mean)
origin_n    <- aggregate(IS_DELAYED ~ ORIGIN, flights_clean, length)
names(origin_n)[2] <- "n"

origin_summary <- merge(origin_rate, origin_n, by = "ORIGIN")
names(origin_summary)[2] <- "delay_rate"

origin_summary <- subset(origin_summary, n >= MIN_N_AIRPORT)
origin_summary <- origin_summary[order(-origin_summary$delay_rate), ]

top_origins <- head(origin_summary, 15)

ggplot(top_origins, aes(x = reorder(ORIGIN, delay_rate), y = delay_rate)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Delay rate (>=15m) per origin airport",
       x = "Origin airport", y = "Delay rate")
```

The plot shows us that differences exist in delay rates across origin
airports. Some airports, such as EWR and ORD, have clearly higher delay
rates than others. This suggests that departure airport contains useful
predictive information for delay status. At the same time, these
differences should not be interpreted as a purely due to the airport.
They may also reflect differences in airlines, route structure, weather
exposure and scheduling patterns. The figure therefore shows that
`ORIGIN` as a potentially informative variable in the models.

However, delay rates alone do not show whether busier airports behave
different. We therefore want to compare airport volume and delay rate
directly.

```{r origin_volume_vs_delay_final}
library(ggrepel)

origin_top50_vol <- origin_summary[order(-origin_summary$n), ][1:50, ]

avg_delay <- mean(origin_top50_vol$delay_rate, na.rm = TRUE)
med_volume <- median(origin_top50_vol$n, na.rm = TRUE)

label_points <- subset(origin_top50_vol, delay_rate > avg_delay | n > quantile(n, 0.9))

ggplot(origin_top50_vol, aes(x = n, y = delay_rate)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_hline(yintercept = avg_delay, linetype = "dashed") +
  geom_vline(xintercept = med_volume, linetype = "dashed") +
  geom_text_repel(
    data = label_points,
    aes(label = ORIGIN),
    size = 3,
    max.overlaps = 20
  ) +
  scale_x_log10() +
  scale_y_continuous(labels = pct_label) +
  labs(
    title = "Origin airports: volume vs delay rate (top 50 by volume)",
    x = "Number of flights (log scale)",
    y = "Delay rate"
  )
```

This scatterplot helps separate airport scale from delay risk. Airports
in the upper right of the figure are especially important, because they
combine high traffic with high delay rates and therefore affect many
flights overall. The plot shows that some large airports, such as ORD
and EWR, also have relatively high delay rates, while others, such as
ATL, have high volume but more moderate delay rates. This suggests that
airport size alone does not explain delay risk. The most relevant
airports are those that combine both high traffic and high delay rates.

Next we look at two simple relationships: distance and planned departure
time.

```{r distance_hour_delay_risk}
set.seed(46748717)
N_VIS <- min(200000, nrow(flights_clean))
idx_vis <- sample(seq_len(nrow(flights_clean)), N_VIS)
flights_vis <- flights_clean[idx_vis, ]

# make bin distance 
distance_breaks <- pretty(flights_vis$DISTANCE, n = 12)
flights_vis$DISTANCE_BIN <- cut(
  flights_vis$DISTANCE,
  breaks = unique(distance_breaks),
  include.lowest = TRUE
)

# calculate both the average delay rate and the number of flights per bin 
distance_risk <- aggregate(IS_DELAYED ~ DISTANCE_BIN, flights_vis, mean)
distance_mid  <- aggregate(DISTANCE ~ DISTANCE_BIN, flights_vis, mean)
distance_n    <- aggregate(IS_DELAYED ~ DISTANCE_BIN, flights_vis, length)

names(distance_risk)[2] <- "delay_rate"
names(distance_mid)[2]  <- "distance_mid"
names(distance_n)[2]    <- "n"

distance_risk <- merge(distance_risk, distance_mid, by = "DISTANCE_BIN")
distance_risk <- merge(distance_risk, distance_n, by = "DISTANCE_BIN")
distance_risk <- subset(distance_risk, n >= 100)
distance_risk <- distance_risk[order(distance_risk$distance_mid), ]

print(
  ggplot(distance_risk, aes(x = distance_mid, y = delay_rate)) +
    geom_line() +
    geom_point() +
    scale_y_continuous(labels = pct_label, limits = c(0, 1)) +
    labs(title = "Delay risk vs distance",
         x = "Distance (miles)", y = "Delay risk")
)

# planned departure hour bins 
flights_vis$DEP_HOUR_BIN <- floor(flights_vis$SCHED_DEP_HOUR)

hour_risk <- aggregate(IS_DELAYED ~ DEP_HOUR_BIN, flights_vis, mean)
hour_n    <- aggregate(IS_DELAYED ~ DEP_HOUR_BIN, flights_vis, length)

names(hour_risk)[2] <- "delay_rate"
names(hour_n)[2]    <- "n"

hour_risk <- merge(hour_risk, hour_n, by = "DEP_HOUR_BIN")
hour_risk <- subset(hour_risk, n >= 100)
hour_risk <- hour_risk[order(hour_risk$DEP_HOUR_BIN), ]

print(
  ggplot(hour_risk, aes(x = DEP_HOUR_BIN, y = delay_rate)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = 0:23) +
    scale_y_continuous(labels = pct_label, limits = c(0, 1)) +
    labs(title = "Delay risk vs planned departure time",
         x = "Scheduled departure hour", y = "Delay risk")
)
```

The two plots suggest that planned departure hour is a much stronger
predictor of delay risk than distance. For distance, the delay rate
changes little across the bins, indicating a weak and slightly
non-linear relationship rather than a clear linear trend. In contrast,
scheduled departure hour shows a pronounced pattern: delay risk is
lowest in the early morning, rises gradually during the day. It reaches
highest levels in the late afternoon and evening before falling slightly
again at the end of the day. The plots suggest that both variables
capture some variation in delay risk, but that departure time is likely
to be substantially more informative than route length.

These plots are meant to explore patterns and help decide which
variables to keep and how to code them. Scheduled departure time works
better as an hour variable than as the raw HHMM value.

There is also a small plotting assumption here. We only keep bins with
at least 100 flights, because otherwise the average rate per bin becomes
too unstable to read. That choice is meant to make the plot more
reliable, not to change the underlying pattern.

Finally we move from individual variables to route patterns, where both
origin and destination are considered together.

```{r route_risk_summary}
# route level high-volume combinations
route_summary <- aggregate(IS_DELAYED ~ ORIGIN + DEST, flights_clean, mean)
route_n <- aggregate(IS_DELAYED ~ ORIGIN + DEST, flights_clean, length)
names(route_summary)[3] <- "delay_rate"
names(route_n)[3] <- "n"

route_summary <- merge(route_summary, route_n, by = c("ORIGIN", "DEST"))
route_summary <- subset(route_summary, n >= 3000)
route_summary <- route_summary[order(-route_summary$delay_rate), ]

head(route_summary, 10)
```

This route table adds insight by showing whether high delay rates are
associated mainly with airports in general or with specific origin and
destination combinations. If certain routes remain risky for highly
delay even after observing airport level differences, this suggests that
route structure itself may carry predictive information.

The results indeed indicate that some route combinations stand out
clearly. In particular, many of the highest-risk routes in the table
involve `EWR` as destination, suggesting that congestion or operating
conditions on specific route networks do probably matter. The delay
rates for these routes are above the base rate, which suggests that
route-level structure is relevant rather than negligible.

As before, we apply a minimum-volume rule and retain only routes with at
least 3000 flights. With these route patterns in mind, we now move to
more exploratory modelling techniques. We can now start with dimension
reduction and clustering.

# 6. Dimension reduction and clustering

PCA and clustering are exploratory steps. We only use them if we can
build a sensible feature set with variables that are known before
departure. If we put in raw times, airport labels, or post-event
variables, the output becomes much less useful.

## 6.1 Preparing data for PCA and clustering

We use PCA here as an exploratory tool to understand the structure of
the pre-departure data. PCA helps us assess whether some variables move
together. But also whether the feature space contains redundancy and
whether part of the variation can be summarized in a smaller number of
dimensions.

Before deciding how many components to keep, we first need to determine
which variables should be allowed to shape the PCA and clustering space.
This is an important modelling choice, because PCA is only as useful as
the feature representation it starts from. If we include raw time
stamps, airport labels or post-event variables, the resulting structure
would not contain meaningful pre-departure differences.

| Variable | Decision | Treatment | Reason |
|------------------|------------------|------------------|-------------------|
| `DISTANCE` | Keep | Standardize | Numeric pre-departure variable; potentially informative for route structure |
| `CRS_E LAPSED_TIME` | Keep | Standardize | Planned flight duration, pre-departure and useful as a compact route characteristic |
| `SCH ED_DEP_HOUR` | T ransform | Standardize | Seems better interpretable than raw scheduled departure time from exploratory analysis |
| `MONTH` | T ransform | Convert to cyclical sine and cosine features and standardize | Month 12 is adjacent to month 1, so simple linear coding would be inappropriate |
| `WEEKDAY` | T ransform | Represent as a weekend dummy | Low-dimensional representation for exploratory PCA/clustering |
| `FL_DATE` | Exclude | \- | A full calendar date is too specific for PCA or clustering |
| `OP_CARRIER` | Exclude | \- | Highly diverse label; useful for supervised models, but less suitable for a compact and interpretable PCA and/or clustering feature set |
| `OP_CAR RIER_FL_NUM` | Exclude | \- | Flight number is an identifier rather than a meaningful numeric feature |
| `ORIGIN` `DEST` | Exclude | \- | Airport code is a categorical label and thus not suitable for PCA and/or clustering |
| `C RS_DEP_TIME` `C RS_ARR_TIME` | Exclude | Replace by `SCHE D_DEP_HOUR` and `SCHE D_ARR_HOUR` | Raw timestamp is not ideal for PCA. If arrival timing is used, a simplified hour feature is preferable |
| `SCH ED_ARR_HOUR` | Exclude | \- | Redundant with scheduled departure hour and planned elapsed time, excluding keeps the feature set more compact |
| `DEP_DELAY` `ARR_DELAY` `ARR_TIME` `DEP_TIME` | Exclude | \- | Not known before departure and including it would introduce data leakage |
| `IS_DELAYED` | Exclude | \- | Is the target variable |
| `TAXI_OUT` `WHEELS_OFF` `WHEELS_ON` `TAXI_IN` | Exclude | \- | Observed after departure |
| ``` ACTUAL _ELAPSED_TIM E``AIR_TIME ``` | Exclude | \- | Not valid in a pre-departure setting |
| `SEC URITY_DELAY` `LATE_AI RCRAFT_DELAY NAS_DELAY WE ATHER_DELAY` `CA RRIER_DELAY` | Exclude | \- | Post-event delay attribution, not known before departure |
| `CANCELLED` ``` DIVE RTED``CANCEL LATION_CODE ``` | Exclude | \- | Not in completed flight sample and therefore not used as a feature |

For PCA and clustering we need to only keep variables that are available
before departure and that can be represented in a compact numeric form.
This leads to a small feature set consisting of planned numeric
characteristics such as distance, scheduled elapsed time, scheduled
departure hour and simple calendar features. Variables that are
identifiers, raw labels, target variables or post-departure are
excluded. This keeps the exploratory structure easier to interpret and
aligned with the pre-departure prediction setting.

In this step, we prepare the variables for PCA and clustering. Cyclical
variables such as scheduled time and month are converted to sine/cosine
representations so that values at the boundaries remain close together.
Raw clock times do not work well as simple numeric inputs. The numeric
inputs in HHMM format are also not a good representation of time that is
why it is converted. We also simplify `WEEKDAY` to a weekend indicator
in order to keep the feature space compact.

```{r pca_cluster_preparation, echo=FALSE}
flights_structure <- flights_clean

# cyclical encoding for time and month
flights_structure$DEP_HOUR_SIN <- sin(2 * pi * flights_structure$SCHED_DEP_HOUR / 24)
flights_structure$DEP_HOUR_COS <- cos(2 * pi * flights_structure$SCHED_DEP_HOUR / 24)
flights_structure$MONTH_SIN    <- sin(2 * pi * flights_structure$MONTH / 12)
flights_structure$MONTH_COS    <- cos(2 * pi * flights_structure$MONTH / 12)

# weekend indicator 
flights_structure$IS_WEEKEND <- ifelse(
  flights_structure$WEEKDAY %in% c("Saturday", "Sunday"), 1, 0
)
```

The table below then summarizes what we did for each variable and why.

| Feature | Preparation | Why |
|-------------------|-------------------|----------------------------------|
| `DISTANCE` | Standardize | Already meaningful as a numeric feature and scaling makes it comparable to the other variables |
| `CRS_EL APSED_TIME` | Standardize | Planned numeric feature and scaling is sufficient, though some overlap with distance is expected |
| `SCHE D_DEP_HOUR` | Encode cyclically, then standardize | Time of day is circular, so sine/cosine encoding avoids jumps at midnight |
| `MONTH` | Encode cyclically, then standardize | The calendar is circular, so December and January should remain close in feature space |
| `WEEKDAY` | Convert to weekend dummy | Keeps dimensions low and easy to interpret |

Now we define the feature space used for PCA and clustering as described
in the table above. Variables such as delay status, carrier, origin, and
destination are retained only for post-hoc interpretation of the
resulting components or clusters. We standardize the feature space so
that no variable dominates purely because of its scale.

```{r pca_cluster_feature_space}
# define the variables used
features_retain <- c(
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "DEP_HOUR_SIN", "DEP_HOUR_COS",
  "MONTH_SIN", "MONTH_COS",
  "IS_WEEKEND"
)

# define variables retained only for post hoc interpretation
features_post_hoc <- c("IS_DELAYED", "OP_CARRIER", "ORIGIN", "DEST")

# construct the analysis frame
flights_structure_analysis <- flights_structure[, c(features_retain, features_post_hoc)]

# remove incomplete observations
flights_structure_analysis <- na.omit(flights_structure_analysis)

# standardize the representation variables
flights_structure_analysis_z <- as.data.frame(
  scale(flights_structure_analysis[, features_retain])
)

# inspect result
dim(flights_structure_analysis)
apply(flights_structure_analysis_z, 2, sd)
```

We use Z-score standardization because both PCA and distance-based
clustering are sensitive to scale. Without scaling, variables with
larger numeric ranges would dominate the analysis even if they do not
contain more information. Standardization places the variables on a more
comparable footing.

The feature representation is kept simple. We do not add extra
transformations such as log transformations for `DISTANCE` or
`CRS_ELAPSED_TIME`, because for this exploratory step it is more
important that the variables remain understandable than that every
distribution is made statistically neater.

Thus, first we defined a coherent feature representation and transform
non-linear or categorical information into a compact numeric form. Then
we separated variables used only for post-hoc interpretation and
standardize the resulting feature space before PCA and clustering. We
kept calendar related structures through cyclical month encoding and a
weekend indicator, because these patterns appear relevant in the flight
delay risk analysis after our exploration.

Before running the PCA itself, we first inspect the correlation
structure among the retained variables. This is useful because PCA is
especially informative when some variables overlap or move together. A
correlation heat map therefore gives a first indication of whether
dimensionality reduction is likely to add value.

## 6.2 Correlation heatmap

We inspect the correlation structure of the kept variables. This is a
useful step, because PCA is most informative when at least part of the
feature space contains overlap. The heat map therefore helps us assess
whether some variables capture similar information and whether the
dimensionality reduction is meaningful.

This step is also relevant for clustering. If strongly correlated
variables are kept together, they can effectively count the same signal
more than once in the distance calculation influencing the accurateness.
Clustering may place too much weight on one part of the structure simply
because it is represented multiple times. The heat map is a diagnostic
tool for evaluating whether the features are reasonably balanced for the
next steps.

```{r correlation_heatmap}
# heatmap
cor_mat <- round(cor(flights_structure_analysis_z), 2)
cor_mat
```

```{r correlation_heatmap_plot}
cor_long <- reshape2::melt(cor_mat)

label_map <- c(
  DISTANCE = "Distance",
  CRS_ELAPSED_TIME = "Sched elapsed",
  DEP_HOUR_SIN = "Dep sin",
  DEP_HOUR_COS = "Dep cos",
  MONTH_SIN = "Month sin",
  MONTH_COS = "Month cos",
  IS_WEEKEND = "Weekend"
)

cor_long$Var1 <- factor(cor_long$Var1, levels = names(label_map), labels = label_map)
cor_long$Var2 <- factor(cor_long$Var2, levels = names(label_map), labels = label_map)

ggplot(cor_long, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "red",
    mid = "white",
    high = "blue",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation heatmap of retained features",
    x = "",
    y = "",
    fill = "Correlation"
  ) +
  coord_fixed() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14)
  )
```

```{r correlation_pairs}
# show overlaps explicitly
high_corr_idx <- which(abs(cor_mat) >= 0.75 & lower.tri(cor_mat), arr.ind = TRUE)

if (nrow(high_corr_idx) == 0) {
  high_corr_pairs <- data.frame(
    variable_1 = character(),
    variable_2 = character(),
    correlation = numeric()
  )
} else {
  high_corr_pairs <- data.frame(
    variable_1 = rownames(cor_mat)[high_corr_idx[, 1]],
    variable_2 = colnames(cor_mat)[high_corr_idx[, 2]],
    correlation = cor_mat[high_corr_idx],
    row.names = NULL
  )
}

high_corr_pairs
```

The clearest pattern in the heat map is the very high correlation
between `DISTANCE` and `CRS_ELAPSED_TIME` (`0.98`). This is not
surprising since longer routes usually also have longer scheduled flight
times. For our analysis, this means that these two variables are close
to carrying the same information. This does not necessarily mean that
one of them must be removed, but it does mean that they should be
interpreted as a strongly overlapping pair. This is exactly the type of
redundancy that PCA can summarize into a shared component.

Apart from this pair, the remaining correlations are small. In
particular, the cyclical month variables and the weekend indicator show
little correlation with most of the other retained features. In other
words, the feature space contains some clear redundancy, but it still
also contains dimensions that add distinct information.

The table of pairs with correlation above `0.75` makes this easier to
summarize. Under this cutoff, the main source of overlap in the retained
feature space is the pair `DISTANCE` and `CRS_ELAPSED_TIME`.

The heat map supports PCA as a logical and sensible next step. The
retained variables are not independent and some of them move together.
Dimensionality reduction can help summarize shared structure. At the
same time, the heat map also shows that the feature space is not
dominated entirely by redundancy.

This is both a warning and a justification for clustering. The warning
is that overlapping variables may influences distances heavily if this
standardized feature space is used. The justification is that PCA may
help stabilize the structure first by reducing some of this double
counting. We therefore use the heat map as a diagnostic step before
moving on to the PCA itself.

## 6.3 Principal component analysis

After inspecting the correlation structure, we can now run the PCA on
the standardized feature space. The goal is not to replace the original
variables, but to see whether a smaller number of components captures a
meaningful share of the structure and what those components represent.
This helps us assess how compressible the feature space is and whether
PCA adds something useful before clustering.

```{r pca_run}
# PCA
pca_results <- prcomp(flights_structure_analysis_z, center = FALSE, scale. = FALSE)

pca_var_explained <- (pca_results$sdev)^2 / sum((pca_results$sdev)^2)
pca_var_table <- data.frame(
  PC_num = seq_along(pca_var_explained),
  PC = paste0("PC", seq_along(pca_var_explained)),
  Variance_explained = pca_var_explained,
  Cumulative = cumsum(pca_var_explained)
)

pca_var_table
```

This table gives an overview of how much structure is captured by the
different principal components. The first component explains the largest
share of variance, but the feature space is not dominated by a single
PC. Instead, the variance is spread across several components. This
shows that the variables contain multiple kinds of structure rather than
having one or two very simple patterns.

```{r pca_variance_plots}
# PCA plots
ggplot(pca_var_table, aes(x = PC, y = Variance_explained)) +
  geom_col(fill = "blue", width = 0.7) +
  geom_text(
    aes(label = paste0(round(100 * Variance_explained, 1), "%")),
    vjust = -0.4,
    size = 3
  ) +
  ylim(0, max(pca_var_table$Variance_explained) * 1.15) +
  labs(
    title = "Scree plot",
    x = "Principal component",
    y = "Variance explained"
  )

ggplot(pca_var_table, aes(x = PC_num, y = Cumulative)) +
  geom_line(color = "blue") +
  geom_point(size = 2, color = "blue") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey") +
  scale_x_continuous(breaks = pca_var_table$PC_num, labels = pca_var_table$PC) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  labs(
    title = "Cumulative explained variance",
    x = "Principal component",
    y = "Cumulative variance explained"
  )
```

The scree plot and cumulative variance plot show us whether PCA give a
summary of the retained feature space. In our case, the first component
explains about 28.4% of the variance. The first two components together
explain about 46.6%. The first four components explain about 75.3%. Five
components are needed to exceed the 80% benchmark, reaching a cumulative
of 89.5%.

We can conclude that the feature space is structured, but not extremely
compressible. The first two or three components do not capture enough of
the total variation to replace the original variables. However, the
first few components still summarize important parts of the structure
and can be useful for clustering.

To understand what the main components represent, we inspect the
variable loadings. These show which feature variables contribute most
strongly to each principal component.

```{r pca_loadings}
# extract the strongest loadings for each PC
top_loadings <- function(pca_obj, pc = 1, n = 7) {
  out <- data.frame(
    variable = names(pca_obj$rotation[, pc]),
    loading = pca_obj$rotation[, pc]
  )
  out$abs_loading <- abs(out$loading)
  out <- out[order(-out$abs_loading), ]
  head(out, n)
}

top_loadings(pca_results, pc = 1, n = 7)
top_loadings(pca_results, pc = 2, n = 7)
top_loadings(pca_results, pc = 3, n = 7)
```

```{r pca_loading_plots}
#plot top loadings 
plot_top_loadings <- function(pca_obj, pc = 1, n = 7) {
  out <- top_loadings(pca_obj, pc = pc, n = n)
  out$variable <- factor(out$variable, levels = rev(out$variable))

  ggplot(out, aes(x = variable, y = loading, fill = loading > 0)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    labs(
      title = paste("Main loadings for PC", pc),
      x = "Variable",
      y = "Loading"
    )
}

print(plot_top_loadings(pca_results, pc = 1, n = 7))
print(plot_top_loadings(pca_results, pc = 2, n = 7))
print(plot_top_loadings(pca_results, pc = 3, n = 7))
```

The loading table shows that the first principal component is dominated
almost entirely by `DISTANCE` and `CRS_ELAPSED_TIME` with very similar
absolute loadings. This indicates that PC1 mainly captures a shared
route-length or schedule-length dimension. This is in line with the
conclusion drawn from the earlier correlation heat map, where these two
variables showed a very strong correlation. PCA here summarizes the
common information into one main direction of variation. This also makes
sense when thinking logically the time that a flight is in the air is
ofcourse highly correlated with how far the flight is.

The second principal component is dominated by `DEP_HOUR_SIN` and
`DEP_HOUR_COS`. This suggests that PC2 mainly captures time-of-day
structure. The third principal component is driven mainly by
`MONTH_COS`, `MONTH_SIN`, and `IS_WEEKEND`. This shows that calendar
structure is another distinct dimension. It shows us that the cyclical
month encoding and weekend indicator are contributing real structure
rather than merely adding noise.

This means that the PCA is indeed useful for understanding the structure
of the data we have here. The scree plot shows no extremely sharp elbow.
A reasonably compact summary still requires several components. PCA here
is a diagnostic and exploratory tool and will not be used as a
replacement for the original explanatory variables.

The dashed line at 80% cumulative explained variance is used as a
practical benchmark rather than a hard rule. If a relatively small
number of components explains around 80% of the variance, that is often
considered a reasonably compact summary. In this case, this requires
five components, which is informative but not compact enough that the
original variables become unnecessary.

```{r pca_followup_decisions}
# summary values for interpretation
cum4 <- pca_var_table$Cumulative[min(4, nrow(pca_var_table))]
cum5 <- pca_var_table$Cumulative[min(5, nrow(pca_var_table))]
distance_elapsed_cor <- cor(
  flights_structure_analysis_z$DISTANCE,
  flights_structure_analysis_z$CRS_ELAPSED_TIME
)

data.frame(
  first_4_components = round(100 * cum4, 1),
  first_5_components = round(100 * cum5, 1),
  distance_elapsed_correlation = round(distance_elapsed_cor, 2)
)
```

Overall, we see that t he correlation between `DISTANCE` and
`CRS_ELAPSED_TIME` is very strong, so we do not want both variables to
weigh too heavily in the clustering step. PCA combines the shared
variation into the first principal component. The analysis confirms that
the retained variables are not independent, that some overlap can be
summarized efficiently. The feature space contains several dimensions.
In the next step we start with clustering and from the PCA we can
conclude that it will be logical to cluster on a small set of principal
component scores than on the full standardized feature space, because
PCA reduces some of the double counting caused by strongly overlapping
variables.

## 6.4 Clustering

Clustering answers a different question than PCA. PCA summarizes
variation into a smaller number of dimensions. Clustering groups flights
into a small number of profiles based on their characteristics. The
clusters are not be interpreted as delay classes, but as flight profiles
defined by route length, departure timing and calendar structure.

For the clustering step, we only use variables that are known
pre-departure. We also use the earlier feature-screening results. We
keep in mind the strong correlation between `DISTANCE` and
`CRS_ELAPSED_TIME` because we do not want both variables to dominate the
distance calculation. The overlap is strong enough to justify a simpler
representation.

We run k-means on standardized clustering variables because of the scale
sensitivity. We also work with a random sample, which keeps the
computation stable while still finding broad patterns.

```{r clustering_input}
# Final clustering feature set
if (!exists("distance_elapsed_cor")) {
  distance_elapsed_cor <- cor(
    flights_structure_analysis_z$DISTANCE,
    flights_structure_analysis_z$CRS_ELAPSED_TIME
  )
}

cluster_vars <- c(
  "DISTANCE",
  "DEP_HOUR_SIN", "DEP_HOUR_COS",
  "MONTH_SIN", "MONTH_COS",
  "IS_WEEKEND"
)

if (abs(distance_elapsed_cor) < 0.85) {
  cluster_vars <- c(
    "DISTANCE",
    "CRS_ELAPSED_TIME",
    setdiff(cluster_vars, "DISTANCE")
  )
}

# Sample used for clustering
set.seed(46748717)
N_CLUSTER <- min(20000, nrow(flights_structure_analysis))
idx_cluster <- sample(seq_len(nrow(flights_structure_analysis)), N_CLUSTER)

cluster_data_z <- flights_structure_analysis_z[idx_cluster, cluster_vars, drop = FALSE]

# Safe mapping back to original rows
cluster_rows <- as.integer(rownames(flights_structure_analysis)[idx_cluster])

interpretation_vars <- c(
    "DISTANCE",
    "CRS_ELAPSED_TIME",
    "SCHED_DEP_HOUR",
    "SCHED_ARR_HOUR",
    "MONTH",
    "IS_WEEKEND",
    "IS_DELAYED",
    "OP_CARRIER",
    "ORIGIN",
    "DEST"
  )

cluster_data_original <- flights_structure[
  cluster_rows,
  interpretation_vars,
  drop = FALSE
]

rownames(cluster_data_original) <- NULL

head(cluster_data_original)
```

To choose the number of clusters, we first use the elbow plot. This plot
shows the total cluster sum of squares for different values of `k`. The
point of using it is not to find one uniquely correct number of
clusters, but to assess at what amount of clusters the plot begins to
flatten out. When that happens we can choose to not add any extra
clusters because it would cause less extra structure and mainly adds
complexity.

```{r elbow_plot}
k_grid <- 2:6

set.seed(46748717)
wss <- sapply(k_grid, function(k) {
  kmeans(cluster_data_z, centers = k, nstart = 20)$tot.withinss
})

elbow_df <- data.frame(
  k = k_grid,
  tot_withinss = wss
)

elbow_df

ggplot(elbow_df, aes(x = k, y = tot_withinss)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = k_grid) +
  labs(
    title = "Elbow plot for k-means",
    x = "Number of clusters",
    y = "Total within-cluster sum of squares"
  )
```

The elbow plot does not show a sharp break, but the strongest flattening
occurs from `k = 2` to `k = 4` afterwards the flattening becomes more
gradual. We therefore choose `k = 4`. This should be seen as a modelling
decision, four is not the one true number of flight types in the data.

```{r kmeans_clustering}
# fit one interpretable k-means solution
set.seed(46748717)
K_CHOSEN <- 4

kmeans_result <- kmeans(cluster_data_z, centers = K_CHOSEN, nstart = 25)

cluster_profile_df <- cluster_data_original
cluster_profile_df$cluster <- factor(kmeans_result$cluster)
cluster_profile_df$IS_DELAYED <- as.numeric(cluster_profile_df$IS_DELAYED)

head(cluster_profile_df)
```

We first inspect cluster sizes, because a useful clustering solution
should not be driven only by very small niche groups.

```{r cluster_sizes}
# Cluster sizes
cluster_sizes <- data.frame(
  cluster = factor(names(table(cluster_profile_df$cluster))),
  n = as.integer(table(cluster_profile_df$cluster))
)

cluster_sizes

ggplot(cluster_sizes, aes(x = cluster, y = n)) +
  geom_col() +
  labs(
    title = "Number of flights per cluster",
    x = "Cluster",
    y = "Number of flights"
  )
```

The cluster size distribution shows that the solution is fairly stable
and not driven only by niche and small groups. Cluster 1 is the largest
with 7,731 flights, followed by clusters 3 and 4 with 5,245 and 4,861
flights respectively. Cluster 2 is smaller with 2,163 flights, which
suggests a more specific profile.

Now we look at the cluster centers in the standardized feature space,
because these are the variables on which k-means actually clustered.

```{r cluster_centers}
cluster_centers <- as.data.frame(kmeans_result$centers)
cluster_centers$cluster <- factor(seq_len(nrow(cluster_centers)))

cluster_centers
```

The cluster centers give the impression of some structure existing in
the data, but it is needed to state that the differences are not
extreme. Cluster 2 stands out most clearly as a long distance profile,
since its standardized `DISTANCE` is much higher than in the other
clusters. Cluster 4 is characterized most strongly by a higher
`IS_WEEKEND`, which makes it a weekend-oriented profile. Clusters 1 and
3 differ in the cyclical departure-time columns, which suggests that
they capture different time-of-day patterns. These differences are
however easier to detect in the transformed feature space than in cyclic
form. The month features lie close to zero in all cluster centres, this
shows that seasonality is probably not a major driver of the differences
in this analysis.

We next summarise the clusters in the original variables. We do this
because the cyclical time structure is not always perfectly visible in
raw average hours.

```{r cluster_summary}
summary_vars <- c(
    "DISTANCE",
    "CRS_ELAPSED_TIME",
    "SCHED_DEP_HOUR",
    "SCHED_ARR_HOUR",
    "MONTH",
    "IS_WEEKEND",
    "IS_DELAYED"
  )

cluster_summary <- aggregate(
  cluster_profile_df[, summary_vars, drop = FALSE],
  by = list(cluster = cluster_profile_df$cluster),
  FUN = function(x) mean(x, na.rm = TRUE)
)

cluster_summary

```

The raw cluster summaries show that the differences between clusters are
fairly small. The mean values for distance, scheduled departure hour,
scheduled arrival hour, month, and weekend are all very close together.
Indicating that the clustering does not show four sharply separated
groups in the original variables. Instead, it seems to capture weaker
multivariate differences that are clearer in the transformed feature
space than in the raw data summaries. For example, the departure-time
differences between clusters 1 and 3 is easier to see in the
standardized cyclical cluster centers than in the raw average of
`SCHED_DEP_HOUR`. This is actually logical because k-means grouped the
flights in sine/cosine space rather than on a single linear hour scale.

Next we inspect some boxplots to add more insights. These show not only
the average, but also the spread and overlap within each cluster.

```{r cluster_profiles}
ggplot(cluster_profile_df, aes(x = cluster, y = DISTANCE, fill = cluster)) +
  geom_boxplot(show.legend = FALSE, outlier.shape = NA, na.rm = TRUE) +
  labs(
    title = "Distance per cluster",
    x = "Cluster",
    y = "Distance (miles)"
  )

ggplot(cluster_profile_df, aes(x = cluster, y = SCHED_DEP_HOUR, fill = cluster)) +
  geom_boxplot(show.legend = FALSE, outlier.shape = NA, na.rm = TRUE) +
  labs(
    title = "Scheduled departure hour per cluster",
    x = "Cluster",
    y = "Scheduled departure hour"
  )

ggplot(cluster_profile_df, aes(x = cluster, y = MONTH, fill = cluster)) +
  geom_boxplot(show.legend = FALSE, outlier.shape = NA, na.rm = TRUE) +
  labs(
    title = "Month per cluster",
    x = "Cluster",
    y = "Month"
  )
```

```{r cluster_arrival_plot}
if ("SCHED_ARR_HOUR" %in% names(cluster_profile_df)) {
  print(
    ggplot(cluster_profile_df, aes(x = cluster, y = SCHED_ARR_HOUR, fill = cluster)) +
      geom_boxplot(show.legend = FALSE, outlier.shape = NA, na.rm = TRUE) +
      labs(
        title = "Scheduled arrival hour per cluster",
        x = "Cluster",
        y = "Scheduled arrival hour"
      )
  )
}
```

The boxplots confirm our earlier conclusion and show that the clusters
are indeed not clearly separated on any single raw variable. There is
substantial overlap in distance, scheduled departure hour, scheduled
arrival hour and month. This means that the cluster solution should not
be interpreted as a set of strongly distinct natural categories. It is
better understood as a partition into broader and only partially
separated profiles. The strongest cluster interpretation therefore comes
from combining the cluster centers values with the raw data summaries:

-   **Cluster 2** is the clearest **longer-distance profile**.

-   **Cluster 4** is the clearest **weekend-oriented profile**.

-   **Clusters 1 and 3** appear to differ somewhat in **time-of-day
    structure**, but this distinction is more visible in the cyclical
    feature space than in the raw hour summaries.

As an additional descriptive step, we inspect the most common carrier,
origin, and destination within each cluster. These variables are not
used to form the clusters, but they may help with interpretation.

```{r cluster_top_labels}
top_label <- function(x) names(sort(table(x), decreasing = TRUE))[1]

cluster_top_carrier <- aggregate(OP_CARRIER ~ cluster, data = cluster_profile_df, FUN = top_label)
cluster_top_origin  <- aggregate(ORIGIN ~ cluster, data = cluster_profile_df, FUN = top_label)
cluster_top_dest    <- aggregate(DEST ~ cluster, data = cluster_profile_df, FUN = top_label)

cluster_top_carrier
cluster_top_origin
cluster_top_dest
```

The top labels also do not show a clear distinct in the clusters,
because the carrier and airport labels appear across several clusters.
This suggests that the divide is not mainly carrier-specific or
airport-specific, which is consistent with the decision to leave those
labels out of the clustering feature space itself.

Now we run hierarchical clustering on a smaller sample as a robustness
check. We do not use it as the main clustering method, because on a
dataset of this size it becomes computationally heavy and is harder
interpret. Here it serves only as a supplementary diagnostic.

```{r hierarchical_clustering}
set.seed(46748717)
N_HCLUST <- min(150, nrow(cluster_data_z))
idx_hclust <- sample(seq_len(nrow(cluster_data_z)), N_HCLUST)

hclust_input <- cluster_data_z[idx_hclust, , drop = FALSE]
hclust_dist <- dist(hclust_input, method = "euclidean")
hclust_result <- hclust(hclust_dist, method = "ward.D2")

plot(
  hclust_result,
  labels = FALSE,
  hang = -1,
  main = "Hierarchical clustering dendrogram (sample)",
  xlab = "",
  sub = "",
  ylab = "Height"
)

```

Overall, the clustering results are useful as an exploratory step in
line with the previous sections. Section 6.2 showed that route length
and elapsed time overlap are strong. In section 6.3 we showed that the
main directions of variation in the data are related to route length,
time of day and calendar structure. The clustering analysis shows
similar groups but in weaker and more overlapping way. That is way it is
important to state that the clustering results should not be
overinterpreted. The groups are only moderately distinct and some of the
structure is easier to see in the transformed feature space than in raw
summaries. They help reveal some flight profiles and show how the PCA
dimensions combine in a non-linear grouping of flights. For later
modelling, the original engineered variables remain the stronger basis.

# 7. Modelling

In this section, we move from exploration to prediction. Whereas Chapter
6 focused on understanding the structure of the data through
unsupervised methods, the goal here is different: we now examine whether
flight delay risk can be predicted before departure in a way that is
useful for decision making. In particular, we compare classification
models that use only information available pre-departure. This keeps the
analysis operationally relevant and avoids information leakage from
variables that are only observed after the flight has started.

The target in this chapter is the delay classification problem defined
earlier: a flight is delayed by at least 15 minutes. The main interest
is the delayed class. In practice, failing to identify a flight that
will be delayed is often more costly than incorrectly flagging a flight
that ultimately departs on time. Missed delay risks reduce the
opportunity to prepare gates, crew or other operational buffers. At the
same time, false delay risk alerts are not costless either. This is due
to unnecessary preparation that also creates operational costs. The data
does not contain the exact cost trade-off between these two types of
errors. That's why we do not impose an artificial cost model. Instead,
we use an evaluation framework that places somewhat more emphasis on
identifying delayed flights, while still penalizing false positives.

The PCA and clustering results from Chapter 6 remain useful in this
chapter because they helped us understand the structure of the feature
space and guided the selection of important predictors. We move to a
supervised setting in which models are trained on observed outcomes and
evaluated on unseen flights, so that we can assess predictive value
rather than structure alone.

## 7.1 Classification setup

We prepare a modelling data set by splitting our data set into training
and testing data. The training data is as always used to estimate the
classification rules, while the testing data is held back to inspect how
the model behaves on unseen flights. We do this because a model that
performs well on the same data it was trained on is not automatically
useful in practice. The real question is whether it still performs in a
proper way when applied to new flights. We use only variables that are
known before departure and that are easy to explain. This is important
both for fairness in the comparison and for business relevance. An
accurate model would not be useful if it relied on information that is
only observed after the flight has already started.

## 7.2 Preparing the modelling data

We now prepare the data set for the classification models. We continue
to work in a pre-departure setting, so only variables that are known
before departure are allowed to enter the modelling table. This is still
essential, because the goal is to support earlier operational decisions
on staffing, gates, and attention allocation. The predictor choice is in
line with Chapters 5 and 6. In the exploratory analysis, route length
and schedule timing emerged as two of the clearest dimensions in the
data. We therefore retain distance, scheduled elapsed time and scheduled
departure and arrival hour. We also keep month and weekday to represent
calendar structure. On top of that, we now include carrier, origin, and
destination. These variables were excluded from PCA and clustering
because those methods required a compact numeric feature space, but they
are useful in supervised classification because operational differences
across airlines and airports may help explain delay risk.\
\
There is one more difference with Chapter 6. In the PCA and clustering
section, cyclical encodings were useful because those methods depend
strongly on geometric representation. Here the first models are
tree-based, so we keep scheduled departure and arrival time in their
directly interpretable hour form. This makes the resulting splits easier
to read while still preserving the pre-departure timing information. In
the same spirit, we do not standardize the predictors here. As noted
earlier in the report, tree-based models do not require all variables to
be on the same scale.

```{r prepare_modelling_data}
# construct modelling table and create one train-test split
flights_model <- flights_clean

# target variable in class-label form
flights_model$IS_DELAYED_15 <- factor(
  ifelse(flights_model$IS_DELAYED, "Delayed", "On_time"),
  levels = c("On_time", "Delayed")
)

# convert categorical predictors to factors
factor_vars <- c("MONTH", "WEEKDAY", "OP_CARRIER", "ORIGIN", "DEST")

for (v in factor_vars) {
  if (v %in% names(flights_model)) {
    flights_model[[v]] <- factor(flights_model[[v]])
  }
}

# variables used in modelling
model_vars <- c(
  "IS_DELAYED_15",
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "SCHED_DEP_HOUR",
  "SCHED_ARR_HOUR",
  "MONTH",
  "WEEKDAY",
  "OP_CARRIER",
  "ORIGIN",
  "DEST"
)

model_df <- flights_model[, model_vars]
model_df <- na.omit(model_df)

# overview of the modelling data
str(model_df)

# class balance in full modelling table
class_counts_full <- table(model_df$IS_DELAYED_15)
class_props_full  <- prop.table(class_counts_full)

class_counts_full
round(100 * class_props_full, 2)

# sample for computational manageability
set.seed(46748717)

N_MODEL <- min(300000, nrow(model_df))
idx <- sample(seq_len(nrow(model_df)), N_MODEL)

model_df_small <- model_df[idx, ]
dim(model_df_small)

# class balance in sampled modelling data
class_counts_sample <- table(model_df_small$IS_DELAYED_15)
class_props_sample  <- prop.table(class_counts_sample)

class_counts_sample
round(100 * class_props_sample, 2)

# train-test split while preserving class proportions
set.seed(46748717)

split <- rsample::initial_split(
  model_df_small,
  prop = 0.7,
  strata = IS_DELAYED_15
)

training <- training(split)
testing  <- testing(split)

dim(training)
dim(testing)

# class balance in each split
class_counts_train <- table(training$IS_DELAYED_15)
class_props_train  <- prop.table(class_counts_train)

class_counts_test <- table(testing$IS_DELAYED_15)
class_props_test  <- prop.table(class_counts_test)

class_counts_train
round(100 * class_props_train, 2)

class_counts_test
round(100 * class_props_test, 2)
```

In Chapter 5 we found that meaningful delays are common enough to model,
but the delayed class is still clearly smaller than the on-time class.
To keep the analysis computationally manageable, we draw a sample of
300,000 flights. In that sample, 57,620 flights are delayed and 242,380
flights are on time, corresponding to 19.21% delayed flights and 80.79%
on-time flights. The sampled data therefore preserve the same overall
class pattern as the full modelling table. We then divide the sampled
data into a training set and a testing set using a 70/30 split. Because
delayed flights form the smaller class, we preserve the class
proportions across both subsets. This is useful because it keeps the
later model comparison easier to interpret: the models are trained and
evaluated under essentially the same target distribution. The training
set and testing have the same percentages with 19.21% for delayed
flights and 80.79% for on-time flights. This imbalance matters when we
evaluate the models. A classifier can look strong on overall accuracy
simply by predicting the majority class most of the time. That would
still be unsatisfactory in our setting, because the business problem
from Chapter 1 is precisely about identifying flights with elevated
delay risk early enough to act. For that reason, later model evaluation
will not rely on accuracy alone.

```{r training_testing_comparison}
# compare training and testing distributions
split_compare <- rbind(
  data.frame(split = "Training", training[, c("IS_DELAYED_15", "DISTANCE", "SCHED_DEP_HOUR")]),
  data.frame(split = "Testing",  testing[,  c("IS_DELAYED_15", "DISTANCE", "SCHED_DEP_HOUR")])
)

# class proportions by split
class_share <- as.data.frame(prop.table(
  table(split_compare$split, split_compare$IS_DELAYED_15),
  margin = 1
))
names(class_share) <- c("split", "IS_DELAYED_15", "share")

ggplot(class_share, aes(x = IS_DELAYED_15, y = share, fill = split)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Class balance in training and testing data",
    x = "Delay class",
    y = "Share of flights"
  )

ggplot(split_compare, aes(x = DISTANCE, y = after_stat(density), fill = split)) +
  geom_histogram(bins = 30, color = "white") +
  facet_wrap(~ split, ncol = 1) +
  labs(
    title = "Distance distribution: training vs testing",
    x = "Distance (miles)",
    y = "Density"
  )

ggplot(split_compare, aes(x = SCHED_DEP_HOUR, y = after_stat(density), fill = split)) +
  geom_histogram(bins = 24, color = "white") +
  facet_wrap(~ split, ncol = 1) +
  labs(
    title = "Scheduled departure hour: training vs testing",
    x = "Scheduled departure hour",
    y = "Density"
  )
```

Before fitting the models, we compare the training and testing sets to
check whether they still look similar after the split. This is a useful
diagnostic, because model comparison is easier to interpret when the
test data resemble the data on which the models were trained.\
\
The first plot shows that the proportion of delayed and on-time flights
is effectively the same in both sets. This is what we wanted, since it
means that the split did not distort the target distribution. The second
plot compares the distribution of distance. In both sets, most flights
are concentrated in the short- and medium-distance range, together with
a long right tail toward longer routes. This fits the earlier findings
from Chapters 5 and 6, where route length was one of the clearest
dimensions of variation across flights. The third plot compares the
distribution of scheduled departure hour. Again, the training and
testing data look very similar. Most departures are concentrated between
the morning and evening hours, with relatively few scheduled in the
late-night period. This also matches the earlier exploratory findings,
where schedule timing appeared as another important dimension in the
data. Overall, these checks suggest that the train-test split produced
two comparable samples. The target balance is preserved and two
important predictors retain the same overall shape.\
\
To make the later comparison as fair as possible, we also define one
shared evaluation function that will be used for every model. The
function reports the same set of measures for every model: accuracy,
recall, precision and F1 for the delayed class, together with the
underlying confusion-matrix counts. We do not judge the modelling step
by one number alone, but by whether it helps identify delayed flights in
a way that is meaningful for the business problem.\
\
The second helper function implements a simple hard-voting rule that
will be used later when we combine model predictions in an ensemble.

```{r model_eval_helpers, include=FALSE}
# shared evaluation function for all models
metric_summary <- function(actual, pred) {
  actual <- factor(actual, levels = c("On_time", "Delayed"))
  pred   <- factor(pred,   levels = c("On_time", "Delayed"))

  tp <- sum(actual == "Delayed" & pred == "Delayed")
  tn <- sum(actual == "On_time" & pred == "On_time")
  fp <- sum(actual == "On_time" & pred == "Delayed")
  fn <- sum(actual == "Delayed" & pred == "On_time")

  accuracy  <- (tp + tn) / (tp + tn + fp + fn)
  recall    <- ifelse(tp + fn == 0, NA, tp / (tp + fn))
  precision <- ifelse(tp + fp == 0, NA, tp / (tp + fp))
  f1        <- ifelse(
    is.na(recall) | is.na(precision) | (recall + precision) == 0,
    NA,
    2 * recall * precision / (recall + precision)
  )

  data.frame(
    accuracy = accuracy,
    recall_delayed = recall,
    precision_delayed = precision,
    f1_delayed = f1,
    tp = tp,
    tn = tn,
    fp = fp,
    fn = fn
  )
}

# hard-voting rule for later ensemble comparison
hard_vote <- function(pred_df) {
  delayed_votes <- rowSums(pred_df == "Delayed")
  factor(
    ifelse(delayed_votes >= ceiling(ncol(pred_df) / 2), "Delayed", "On_time"),
    levels = c("On_time", "Delayed")
  )
}
```

## 7.3 Model 1: C5.0 decision tree

We start with a simple decision tree as our first classification model.
This is a natural baseline because a tree is easy to inspect and easy to
explain. If delay risk could already be separated by a few simple split
rules, then we would have a model that is transparent and easy to
interpret.

```{r c50_tree_model}
# C5.0 decision tree
model_tree <- C50::C5.0(
  IS_DELAYED_15 ~ .,
  data = training,
  trials = 1
)

summary(model_tree)
```

The C5.0 output shows that the fitted tree stays at the root node and
predicts only the majority class `On_time`. In other words, the model
creates no actual split and classifies all flights as on time. The
training error is 40,334 out of 210,000 flights or 19.2%, which is
exactly the share of delayed flights in the training data. This means
that the model reduces error simply by following the class imbalance.
This is an important baseline result. It shows that, under the current
class distribution and with a plain tree structure, the model does not
move beyond the majority-class rule. From the business perspective of
this project, that is not sufficient. A model that never flags delayed
flights is not useful for pre-departure decision support, even if its
overall error rate appears acceptable at first sight.\
\
To evaluate the model, we also compute predictions for both the training
and testing data and summarize the results with the shared evaluation
function from Section 7.2.

```{r c50_tree_predictions}
# Predictions on training and testing data
pred_tree_train <- predict(model_tree, training)
pred_tree_test  <- predict(model_tree, testing)

# Evaluation with the shared metrics
tree_train_metrics <- metric_summary(training$IS_DELAYED_15, pred_tree_train)
tree_test_metrics  <- metric_summary(testing$IS_DELAYED_15, pred_tree_test)

cat("Training performance:\n")
print(tree_train_metrics)

cat("\nTesting performance:\n")
print(tree_test_metrics)
```

These results confirm the same pattern on the holdout data. Because the
tree predicts only `On_time`, it achieves a test accuracy that is still
relatively high simply because most flights are on time. At the same
time, recall for the delayed class is zero, because the model never
identifies a delayed flight. This is exactly why accuracy alone is not
enough in this setting. The confusion matrices make this behaviour
directly visible.

```{r c50_tree_confusion}
# confusion matrices on training and testing data
cat("Confusion matrix on training data:\n")
print(table(Actual = training$IS_DELAYED_15, Predicted = pred_tree_train))

cat("\nConfusion matrix on testing data:\n")
print(table(Actual = testing$IS_DELAYED_15, Predicted = pred_tree_test))
```

The confusion matrices make this behaviour directly visible. On the
training data, the model correctly classifies the 169,666 on-time
flights, but misses all 40,334 delayed flights. On the testing data, it
correctly classifies the 72,714 on-time flights, but again misses all
17,286 delayed flights. So although the overall test accuracy still
looks reasonably high, the model is operationally useless for the actual
problem of identifying flights with elevated delay risk.\
\
As an additional visual check, we also fit a shallow `rpart` tree. This
is not meant to replace the C5.0 model, but only to see whether a small
tree reveals any simple split conditions that are easy to display.

```{r tree_conditions_plot}
# extra visual check: shallow rpart tree
tree_plot_model <- rpart::rpart(
  IS_DELAYED_15 ~ .,
  data = training,
  method = "class",
  control = rpart::rpart.control(
    maxdepth = 3,
    minsplit = 3000,
    cp = 0.001
  )
)

if (nrow(tree_plot_model$frame) > 1 &&
    any(tree_plot_model$frame$var != "<leaf>")) {
  plot(tree_plot_model, uniform = TRUE, margin = 0.1)
  text(tree_plot_model, use.n = TRUE, all = TRUE, cex = 0.7)
} else {
  cat("The shallow rpart check also stayed at the root node, so there are no split conditions to plot here.\n")
}
```

The shallow `rpart` check leads to the same conclusion. It also stays at
the root node, so there are no split conditions to display. This
reinforces the idea that a simple tree structure is too weak for this
problem under the present setup.\
\
This result should not be interpreted as proof that the predictors
contain no useful information. It only shows that a plain decision tree,
under the current class imbalance and without further adjustments, is
too conservative to identify the delayed class. That is exactly why this
model is a useful baseline: it shows what a very simple and
interpretable classifier fails to do, and it motivates the move to more
flexible models in the next sections.

```{r c50_tree_store_results}
# Store test performance for later model comparison
tree_test_metrics$model <- "C5.0 tree"

model_results <- tree_test_metrics[, c(
  "model", "accuracy", "recall_delayed", "precision_delayed",
  "f1_delayed", "tp", "tn", "fp", "fn"
)]

model_results
```

## 7.4 Inspecting tree behaviour on unseen data

The previous section showed that the C5.0 tree stays at the root node on
the training data and predicts only the majority class. In this section,
we check whether that same behaviour also appears on unseen flights.
This matters because a model should not be judged only on the data on
which it was fitted. The relevant question is whether it remains useful
when applied to new observations.

```{r c50_tree_test_evaluation}
# evaluate the baseline tree on unseen test data
pred_test <- predict(model_tree, testing)

gmodels::CrossTable(
  testing$IS_DELAYED_15,
  pred_test,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

# accuracy on unseen data
mean(pred_test == testing$IS_DELAYED_15)

# shared evaluation metrics from Section 7.2
tree_test_metrics <- metric_summary(testing$IS_DELAYED_15, pred_test)
tree_test_metrics
```

The confusion matrix shows that the tree again predicts every flight as
`On_time`. On the testing data, this means that all 72,714 on-time
flights are classified correctly, while all 17,286 delayed flights are
missed. The resulting test accuracy is 0.8079, or about 80.8%. At first
sight, that accuracy may seem acceptable. However, it is entirely driven
by the class imbalance in the data. Because most flights are on time, a
classifier can obtain a relatively high accuracy without identifying any
delayed flights at all. That is exactly what happens here. The recall
for the delayed class is 0, meaning that the model fails completely on
the class that matters most for our decision problem. Precision and the
F1-score for the delayed class are not defined here, because the tree
never predicts a single delayed flight. In practical terms, this
reinforces the same conclusion: the model offers no useful delay
warnings on unseen data.\
\
This makes the baseline result very clear. A plain decision tree is too
conservative under the current setup and simply follows the majority
class. So even though the model appears acceptable on overall accuracy,
it is not useful for pre-departure delay detection. This motivates the
next modelling steps, where we consider approaches that may achieve a
better balance between identifying delayed flights and avoiding too many
false alarms.

## 7.5 Altering tree behaviour: boosting

The plain C5.0 tree in the previous section stayed at the root node and
simply predicted the majority class. A natural next step is therefore to
make the tree model more flexible. We do this by using boosting. The
idea of boosting is that later trees focus more strongly on observations
that were handled poorly earlier on, so that the combined model may
capture structure that a single tree misses.

```{r boosted_c50_evaluation}
# boosted C5.0 model: let multiple trees work sequentially
model_tree_boost <- C50::C5.0(
  IS_DELAYED_15 ~ .,
  data = training,
  trials = 10
)

# predictions on unseen test data
pred_boost <- predict(model_tree_boost, testing)

# confusion matrix on test data
gmodels::CrossTable(
  testing$IS_DELAYED_15,
  pred_boost,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

# shared evaluation metrics from Section 7.2
boost_test_metrics <- metric_summary(testing$IS_DELAYED_15, pred_boost)
boost_test_metrics$model <- "Boosted C5.0"

boost_test_metrics <- boost_test_metrics[, c(
  "model", "accuracy", "recall_delayed",
  "precision_delayed", "f1_delayed",
  "tp", "tn", "fp", "fn"
)]

boost_test_metrics
```

The confusion matrix shows that the boosted model still predicts every
flight as `On_time` again. On the testing data, this means that all
72,714 on-time flights are classified correctly, while all 17,286
delayed flights are missed. So even after boosting, the model produces
no true positives at all. The resulting accuracy is again 0.8079, which
is about 80.8%. On its own, that number may look acceptable, but here it
is misleading. The model reaches this accuracy only because on-time
flights are the majority class. In practical terms, the boosted tree
still behaves like a majority-class classifier rather than a useful
delay predictor.\
\
This is also reflected in the other evaluation measures. Recall for the
delayed class is 0, because the model fails to identify any delayed
flight. Precision and the F1-score for the delayed class are not defined
here, because the model never predicts a single delayed case. In
practical terms, that again means that the model provides no useful
delay warnings.\
\
This is an informative result. Boosting makes the tree model more
flexible in principle, but under the current setup it does not change
the basic behaviour of the classifier. The model remains too
conservative and still follows the majority class. So although boosting
is a logical next step after the plain tree, it does not solve the core
problem here. This motivates moving further away from simple tree-based
majority-class behaviour in the next models.

## 7.6 Altering tree behaviour: cost-sensitive classification

A second way to change tree behaviour is to use cost-sensitive
classification. This fits the business setting of the project well,
because missing a truly delayed flight is more problematic than raising
a false alarm for a flight that later turns out to be on time as
explained in the introduction of this chapter. We therefore assign a
higher cost to missed delays than to false positives and check whether
that changes the behaviour of the tree.

```{r cost_sensitive_c50_evaluation}
# cost-sensitive C5.0: make missed delays more expensive than false alarms
cost_matrix <- matrix(
  c(NA, 1,
    2,  NA),
  nrow = 2,
  ncol = 2,
  byrow = FALSE
)
rownames(cost_matrix) <- colnames(cost_matrix) <- c("Delayed", "On_time")

model_tree_cost <- C50::C5.0(
  IS_DELAYED_15 ~ .,
  data = training,
  costs = cost_matrix
)

# predictions on unseen test data
pred_cost <- predict(model_tree_cost, testing)

# confusion matrix on test data
gmodels::CrossTable(
  testing$IS_DELAYED_15,
  pred_cost,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

# shared evaluation metrics from Section 7.2
cost_test_metrics <- metric_summary(testing$IS_DELAYED_15, pred_cost)
cost_test_metrics$model <- "Cost-sensitive C5.0"

cost_test_metrics <- cost_test_metrics[, c(
  "model", "accuracy", "recall_delayed",
  "precision_delayed", "f1_delayed",
  "tp", "tn", "fp", "fn"
)]

cost_test_metrics
```

The confusion matrix shows that the cost-sensitive tree still predicts
every flight as On_time. So even with a higher penalty on missed delays,
the model produces no true positives at all. The model still behaves
like a majority-class classifier rather than a useful delay predictor.
Cost-sensitive classification is conceptually well aligned with the
business problem, since it allows us to penalize missed delays more
heavily than false alarms. However, under the current specification, it
does not change the basic behaviour of the tree. The classifier remains
too conservative and still follows the majority class. So although this
is a logical next step after the plain and boosted trees, it does not
solve the core problem in our data.

## 7.7 Model 2: Random Forest

The next model is a random forest. This is a natural step after the
earlier tree-based models, because a random forest keeps the tree logic
but reduces the instability of a single tree by averaging across many
trees. In that way, it can capture more complex patterns than a plain
decision tree while still remaining relatively interpretable through
variable-importance measures.\
\
In this implementation, however, we cannot use predictors with very many
categories. In particular, `ORIGIN` and `DEST` have too many levels for
the `randomForest` package. We therefore work with a reduced feature set
that keeps the main route, timing, calendar and carrier information, but
excludes the highest variety airport variables. We also set the forest
to 200 trees as a practical choice: large enough to stabilize the model,
but still workable in this notebook.

```{r random_forest_evaluation}
# Random forest exclude ORIGIN and DEST
rf_vars <- c(
  "IS_DELAYED_15",
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "SCHED_DEP_HOUR",
  "MONTH",
  "WEEKDAY",
  "OP_CARRIER"
)

training_rf <- training[, rf_vars]
testing_rf  <- testing[, rf_vars]

set.seed(46748717)
model_rf <- randomForest::randomForest(
  IS_DELAYED_15 ~ .,
  data = training_rf,
  ntree = 200,
  importance = TRUE
)

# predictions on unseen data
pred_rf <- predict(model_rf, testing_rf)

# confusion matrix on test data
gmodels::CrossTable(
  testing_rf$IS_DELAYED_15,
  pred_rf,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

# shared evaluation metrics from Section 7.2
rf_test_metrics <- metric_summary(testing_rf$IS_DELAYED_15, pred_rf)
rf_test_metrics$model <- "Random forest"

rf_test_metrics <- rf_test_metrics[, c(
  "model", "accuracy", "recall_delayed",
  "precision_delayed", "f1_delayed",
  "tp", "tn", "fp", "fn"
)]

rf_test_metrics
```

The random forest does improve on the earlier tree-based baselines a
little. On the testing data, it correctly identifies 1,048 delayed
flights, while still missing 16,238 delayed flights. It also incorrectly
flags 1,983 on-time flights as delayed. This gives an accuracy of
0.7975, a delayed-class recall of 0.0606, a precision of 0.3458, and an
F1-score of 0.1032. This means that the model no longer behaves as a
pure majority-class classifier. Unlike the earlier C5.0 variants, the
random forest does identify some delayed flights. That is an important
result, because it suggests that the pre-departure predictors do contain
at least some useful predictive signal. At the same time, recall remains
very low. The model still misses the large majority of actual delays, so
the improvement over the earlier tree models is real but limited.\
\
The confusion matrix makes that trade-off visible. Most flights are
still predicted as On_time, which keeps the number of false positives
relatively low and supports a fairly high overall accuracy. However, the
model remains too conservative for the business objective of identifying
flights with elevated delay risk before departure.

```{r random_forest_importance}
importance(model_rf)
varImpPlot(model_rf, main = "Random forest variable importance")
```

The variable-importance results help explain what the random forest is
using. Across both importance measures, the strongest predictors are
scheduled departure hour, distance and scheduled elapsed time. This
suggests that the forest relies mainly on schedule timing and route
characteristics when distinguishing delayed from on-time flights.
`MONTH` still contributes some information, but clearly less than the
three main predictors, while `WEEKDAY` and `OP_CARRIER` are the least
important variables in this reduced feature set.

This pattern fits well with the earlier exploratory analysis. In
Chapters 5 and 6, timing and route-length variables already emerged as
the clearest sources of structure in the data. The random forest
importance plot points in the same direction: even in a more flexible
supervised model, the main predictive information still comes from when
a flight is scheduled and from the type of route it operates. At the
same time, the earlier performance results show that this information is
not strong enough for the forest to identify delayed flights well. This
plot confirms which variables drive the model, not because it changes
the overall conclusion that predictive performance remains limited.

Overall, the random forest is a better benchmark than the earlier tree
variants, because it moves beyond trivial majority-class prediction.
However, its recall for delayed flights remains low, so it is still not
strong enough for the business goal. Before comparing all models side by
side, we therefore consider one more benchmark from a different model
family: Naive Bayes.

## 7.8 Model 3: Naive Bayes classifier

After the earlier tree-based models, we now consider a Naive Bayes
classifier. This is a useful next step because it introduces a different
modelling logic. Instead of separating the classes through recursive
splits, Naive Bayes estimates how likely a flight belongs to the delayed
or on-time class based on the observed predictor values. It is therefore
a simple probabilistic benchmark rather than another tree variant.

This also fits our business question. The purpose of the analysis is not
only to assign a hard delayed versus on-time label, but more
specifically to estimate delay risk before departure. A probabilistic
model is therefore useful, because it gives direct class probabilities
that can later be inspected under different thresholds. That is relevant
in our case, because from the business perspective missing a real delay
is worse than raising some false alarms.

Naive Bayes also fits well because it remains relatively transparent and
computationally light. At the same time, it makes a strong simplifying
assumption: conditional on the class label, the predictors are treated
like they contribute independently. In practice, that assumption is not
fully realistic here, since several predictors are clearly related, such
as route length, scheduled elapsed time and timing variables. The model
is therefore not expected to be literally correct in a statistical
sense. However, it can still be useful as a benchmark, because it shows
whether a simple probabilistic classifier can extract more signal from
the same pre-departure variables than the earlier tree models.

We fit the Naive Bayes model on the same training data and evaluate it
on the same testing data as before. This keeps the comparison fair and
allows us to judge whether the model improves delayed-flight detection
relative to the earlier approaches.

```{r naive_bayes_evaluation}
nb_vars <- c(
  "IS_DELAYED_15",
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "SCHED_DEP_HOUR",
  "SCHED_ARR_HOUR",
  "MONTH",
  "WEEKDAY",
  "OP_CARRIER"
)

training_nb <- training[, nb_vars]
testing_nb  <- testing[, nb_vars]

# Naive Bayes as a probabilistic benchmark on the reduced feature set
model_nb <- e1071::naiveBayes(
  IS_DELAYED_15 ~ .,
  data = training_nb
)

# class predictions on unseen data
pred_nb <- predict(model_nb, testing_nb)

# confusion matrix
gmodels::CrossTable(
  testing_nb$IS_DELAYED_15,
  pred_nb,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

# shared evaluation metrics from Section 7.2
nb_test_metrics <- metric_summary(testing_nb$IS_DELAYED_15, pred_nb)
nb_test_metrics$model <- "Naive Bayes"

nb_test_metrics <- nb_test_metrics[, c(
  "model", "accuracy", "recall_delayed",
  "precision_delayed", "f1_delayed",
  "tp", "tn", "fp", "fn"
)]

nb_test_metrics
```

In the table we can see the following. On the testing data, the model
correctly identifies 620 delayed flights, while missing 16,666 delayed
flights. It also incorrectly flags 994 on-time flights as delayed. This
gives an accuracy of 0.8038, a delayed-class recall of 0.0359, a
precision of 0.3841, and an F1-score of 0.0656.

These results show that Naive Bayes does move beyond trivial
majority-class prediction. Unlike the normal, boosted and cost-sensitive
C5.0 trees, it identifies at least some delayed flights. This indicates
that the pre-departure predictors contain some class information that
can also be used by a simple probabilistic model.

At the same time, the model remains weak for the delayed class. Recall
is still very low, which means that most actual delays are still missed.
Compared with the Random forest, Naive Bayes is more conservative: it
achieves a slightly higher accuracy and a higher precision, but it
identifies fewer delayed flights and therefore has a lower recall and
F1-score. In other words, the model produces fewer false alarms, but at
the cost of missing even more true delays.

This makes Naive Bayes a useful intermediate benchmark. It confirms that
the data contain some predictive signal beyond majority-class structure,
but it also reinforces the broader conclusion: with the current feature
set, even models that move beyond trivial prediction still perform too
weakly on the delayed class to support strong operational decision
making.

## 7.9 Comparison

To compare the models consistently, we place the main evaluation
measures of all tested classifiers side by side. This includes the
earlier C5.0 variants, the random forest and the Naive Bayes model. This
matters because overall accuracy alone is misleading in this setting.
Since most flights are on time, a model can achieve a fairly high
accuracy while still failing completely on the delayed class. For that
reason, recall, precision and the F1-score for delayed flights are more
informative than accuracy alone.

```{r model_comparison_table}
# combine the stored metric summaries from the earlier sections
comparison_table <- rbind(
  data.frame(
    model = "C5.0 tree",
    accuracy = tree_test_metrics$accuracy,
    recall_delayed = tree_test_metrics$recall_delayed,
    precision_delayed = tree_test_metrics$precision_delayed,
    f1_delayed = tree_test_metrics$f1_delayed
  ),
  data.frame(
    model = "Boosted C5.0",
    accuracy = boost_test_metrics$accuracy,
    recall_delayed = boost_test_metrics$recall_delayed,
    precision_delayed = boost_test_metrics$precision_delayed,
    f1_delayed = boost_test_metrics$f1_delayed
  ),
  data.frame(
    model = "Cost-sensitive C5.0",
    accuracy = cost_test_metrics$accuracy,
    recall_delayed = cost_test_metrics$recall_delayed,
    precision_delayed = cost_test_metrics$precision_delayed,
    f1_delayed = cost_test_metrics$f1_delayed
  ),
  data.frame(
    model = "Random forest",
    accuracy = rf_test_metrics$accuracy,
    recall_delayed = rf_test_metrics$recall_delayed,
    precision_delayed = rf_test_metrics$precision_delayed,
    f1_delayed = rf_test_metrics$f1_delayed
  ),
  data.frame(
    model = "Naive Bayes",
    accuracy = nb_test_metrics$accuracy,
    recall_delayed = nb_test_metrics$recall_delayed,
    precision_delayed = nb_test_metrics$precision_delayed,
    f1_delayed = nb_test_metrics$f1_delayed
  )
)

comparison_table
```

The comparison table shows a clear pattern. The `C5.0 tree`,
`Boosted C5.0`, and `Cost-sensitive C5.0` all lead to exactly the same
practical outcome: an accuracy of 0.8079, but a delayed-class recall of
0. In other words, all three behave like majority-class classifiers and
fail to identify any delayed flights at all. Their precision and
F1-score for the delayed class are therefore not defined, because they
never predict a single delayed case.

The`Random forest` and `Naive Bayes` both improve on this majority-class
behavior. The `Random forest` reaches an accuracy of 0.7975, a
delayed-class recall of 0.0606, a precision of 0.3458, and an F1-score
of 0.1032. The `Naive Bayes` model is slightly more conservative: it
achieves a higher accuracy of 0.8038 and a higher precision of 0.3841,
but its delayed-class recall is lower at 0.0359, with an F1-score of
0.0656. This means that Naive Bayes produces fewer false alarms, but
also misses more actual delays.

So the comparison reveals an important trade-off. The tree-based C5.0
variants achieve the highest raw accuracy, but only because they default
to the majority class. Naive Bayes and Random forest are both more
informative because they identify at least some delayed flights. Of
these two, the `Random forest` performs best on the criterion that
matters most in this project: detecting delayed flights early enough to
be useful. However, its recall remains low, so even the best-performing
model is still too weak to serve as a final operational solution.

## 7.10 Multiple splits

The comparison in the previous section was based on one particular
train-test split. That gives a useful first answer, but it does not show
how sensitive the results are to the exact split that happened to be
drawn. To make this visible, we repeat the same evaluation several times
on a smaller fixed sample. We focus on three distinct model behaviors: a
plain C5.0 tree, a cost-sensitive C5.0 tree, the random forest and Naive
Bayes. In addition, we include a simple hard-vote ensemble based on
these models. The goal is to examine whether the main conclusions remain
stable across repeated train-test splits. We repeat the evaluation 8
times on a subset of 60,000 observations. This is sufficient to make
instability visible without making the notebook too computationally
heavy.

```{r repeated_evaluation}
set.seed(46748717)
N_EVAL <- min(60000, nrow(model_df_small))
idx_eval <- sample(seq_len(nrow(model_df_small)), N_EVAL)
eval_df <- model_df_small[idx_eval, ]

# Repeated evaluation across multiple train-test splits
N_REPEATS <- 8
repeated_results <- data.frame()

for (i in seq_len(N_REPEATS)) {
  set.seed(46748717 + i)
  split_eval <- rsample::initial_split(eval_df, prop = 0.7, strata = IS_DELAYED_15)
  training_eval <- rsample::training(split_eval)
  testing_eval  <- rsample::testing(split_eval)

  # Plain decision tree
  tree_eval <- C50::C5.0(IS_DELAYED_15 ~ ., data = training_eval)
  pred_tree_eval <- predict(tree_eval, testing_eval)
  metrics_tree_eval <- metric_summary(testing_eval$IS_DELAYED_15, pred_tree_eval)
  metrics_tree_eval$model <- "C5.0 tree"
  metrics_tree_eval$iteration <- i

  # Cost-sensitive tree
  cost_eval <- C50::C5.0(IS_DELAYED_15 ~ ., data = training_eval, costs = cost_matrix)
  pred_cost_eval <- predict(cost_eval, testing_eval)
  metrics_cost_eval <- metric_summary(testing_eval$IS_DELAYED_15, pred_cost_eval)
  metrics_cost_eval$model <- "Cost-sensitive C5.0"
  metrics_cost_eval$iteration <- i

  # Random forest on reduced feature set
  training_eval_rf <- training_eval[, rf_vars]
  testing_eval_rf  <- testing_eval[, rf_vars]

  rf_eval <- randomForest::randomForest(
    IS_DELAYED_15 ~ .,
    data = training_eval_rf,
    ntree = 100
  )

  pred_rf_eval <- predict(rf_eval, testing_eval_rf)
  metrics_rf_eval <- metric_summary(testing_eval_rf$IS_DELAYED_15, pred_rf_eval)
  metrics_rf_eval$model <- "Random forest"
  metrics_rf_eval$iteration <- i

  # Naive Bayes on reduced feature set
  training_eval_nb <- training_eval[, nb_vars]
  testing_eval_nb  <- testing_eval[, nb_vars]

  nb_eval <- e1071::naiveBayes(
    IS_DELAYED_15 ~ .,
    data = training_eval_nb
  )

  pred_nb_eval <- predict(nb_eval, testing_eval_nb)
  metrics_nb_eval <- metric_summary(testing_eval_nb$IS_DELAYED_15, pred_nb_eval)
  metrics_nb_eval$model <- "Naive Bayes"
  metrics_nb_eval$iteration <- i

  # Hard-vote ensemble: simple tree + RF + NB
  pred_ensemble_eval <- hard_vote(data.frame(
    tree = pred_tree_eval,
    rf = pred_rf_eval,
    nb = pred_nb_eval
  ))
  metrics_ensemble_eval <- metric_summary(testing_eval$IS_DELAYED_15, pred_ensemble_eval)
  metrics_ensemble_eval$model <- "Hard vote ensemble"
  metrics_ensemble_eval$iteration <- i

  repeated_results <- rbind(
    repeated_results,
    metrics_tree_eval,
    metrics_cost_eval,
    metrics_rf_eval,
    metrics_nb_eval,
    metrics_ensemble_eval
  )
}

# Accuracy across repeated splits
ggplot(repeated_results, aes(x = model, y = accuracy)) +
  geom_boxplot() +
  labs(
    title = "Accuracy across repeated train-test splits",
    x = "Model",
    y = "Accuracy"
  ) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

# Recall across repeated splits
ggplot(repeated_results, aes(x = model, y = recall_delayed)) +
  geom_boxplot() +
  labs(
    title = "Recall across repeated train-test splits",
    x = "Model",
    y = "Recall for delayed flights"
  ) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

# Mean performance across repeats
repeated_summary <- dplyr::group_by(repeated_results, model) |>
  dplyr::summarise(
    mean_accuracy = mean(accuracy, na.rm = TRUE),
    mean_recall_delayed = mean(recall_delayed, na.rm = TRUE),
    mean_precision_delayed = mean(precision_delayed, na.rm = TRUE),
    mean_f1_delayed = mean(f1_delayed, na.rm = TRUE),
    sd_accuracy = sd(accuracy, na.rm = TRUE),
    sd_recall_delayed = sd(recall_delayed, na.rm = TRUE),
    .groups = "drop"
  )

repeated_summary
```

The boxplots and summary table show a clear and more nuanced pattern.
The `C5.0 tree` and the`Cost-sensitive C5.0` again show no variation
across repeated splits trapped in majority-class prediction and achieve
a delayed-class recall of 0 in every repeated split. The
`Hard-vote ensemble` no longer behaves exactly the same as these two
tree models, but its improvement is extremely small. On average, it
reaches an accuracy of about 0.8079 and a delayed-class recall of only
0.0023. Its precision is relatively high at 0.4460, but this mainly
reflects the fact that it predicts very few delayed flights at all. In
practical terms, the ensemble remains far too conservative to be useful
for delayed-flight detection.

The `Naive Bayes` model performs better than the `Hard-vote ensemble`
and the C5.0 variants, but it is still limited. Across the repeated
splits, it reaches an average accuracy of about 0.8071, an average
delayed-class recall of 0.0077, a precision of 0.3947 and an F1-score of
0.0152. So it does move beyond trivial majority-class prediction, but
only to a small extent. The `Random forest` remains the strongest model
in the repeated-split analysis. It has the lowest average accuracy at
about 0.7994, but it also achieves by far the highest delayed-class
recall, at 0.0473 on average, together with an average precision of
0.3396 and an F1-score of 0.0830. This means that it consistently
identifies more delayed flights than the other models, even though it
does so at the cost of slightly lower overall accuracy and somewhat more
variation across splits.

Overall, the repeated-split results strengthen the earlier conclusion
rather than changing it. The simplest tree-based models remain
ineffective for the delayed class, even when the split is changed.
`Naive Bayes` and the hard-vote ensemble move only slightly beyond that
behavior, while the `Random forest` remains the only model that produces
a clearly non-zero delayed-flight recall on a consistent basis. So the
main limitation is not one unlucky train-test split, but the underlying
weakness of the simpler models for this task.

## 7.11 Multiple thresholds

So far, we mainly evaluated models through final class labels such as
`Delayed` and `On_time`. That is useful for operational interpretation,
because in practice the airline ultimately needs a yes-or-no decision.
However, this approach fixes one classification threshold at the outset.
ROC curves and the AUC provide a different perspective: they show how
well a model ranks flights from lower to higher delay risk across many
possible threshol]\
\
We only include this analysis for models that return usable class
probabilities. That is an important restriction here. The
`cost-sensitive C5.0` model was designed to change the classification
decision under asymmetric costs, so we do not treat its output as an
ordinary probability ranking for ROC/AUC purposes. For that reason, we
focus on the plain `C5.0` tree, the `Random forest` and `Naive Bayes`.

```{r roc_auc}
# ROC/AUC: predicted probabilities rather than final class labels
tree_prob <- predict(model_tree, testing, type = "prob")[, "Delayed"]
rf_prob   <- predict(model_rf, testing_rf, type = "prob")[, "Delayed"]
nb_prob   <- predict(model_nb, testing_nb, type = "raw")[, "Delayed"]

# ROC curves
roc_tree <- pROC::roc(
  testing$IS_DELAYED_15,
  tree_prob,
  levels = c("On_time", "Delayed"),
  quiet = TRUE
)

roc_rf <- pROC::roc(
  testing_rf$IS_DELAYED_15,
  rf_prob,
  levels = c("On_time", "Delayed"),
  quiet = TRUE
)

roc_nb <- pROC::roc(
  testing_nb$IS_DELAYED_15,
  nb_prob,
  levels = c("On_time", "Delayed"),
  quiet = TRUE
)

plot(roc_tree, col = "blue", main = "ROC curves on the testing data")
lines(roc_rf, col = "green")
lines(roc_nb, col = "orange")
legend(
  "bottomright",
  legend = c("C5.0 tree", "Random forest", "Naive Bayes"),
  col = c("blue", "green", "orange"),
  lwd = 2,
  bty = "n"
)

# AUC summary table
auc_table <- data.frame(
  model = c("C5.0 tree", "Random forest", "Naive Bayes"),
  auc = c(
    as.numeric(pROC::auc(roc_tree)),
    as.numeric(pROC::auc(roc_rf)),
    as.numeric(pROC::auc(roc_nb))
  )
)

auc_table
```

The ROC plot shows a clear difference between the three models. The
`C5.0 tree` lies exactly on the diagonal, which means that it cannot
rank flights better than chance. This is fully consistent with the
earlier results, where the tree behaved like a majority-class classifier
and failed to identify delayed flights. Both the `Random forest`
and`Naive Bayes` lie above the diagonal, which means that both models
contain at least some ability to rank higher-risk flights above
lower-risk flights.

The AUC table summarizes this in one number. The `C5.0 tree` has an AUC
of 0.500, which is exactly chance level. The`Random forest` reaches an
AUC of 0.623, while `Naive Bayes` performs slightly better with an AUC
of 0.626. This adds an important nuance to the earlier results. At the
chosen classification threshold, the `Random forest` performed better
than`Naive Bayes` in terms of delayed-class recall and F1, so it
remained the strongest classifier for the main decision problem. But in
threshold-independent ranking terms, `Naive Bayes` performs at least as
well as, and here slightly better than, the `Random forest`. In other
words, `Naive Bayes` appears somewhat better at ordering flights by
relative delay risk, even though it is more conservative once a fixed
classification threshold is applied.

So the ROC and AUC results do not overturn the earlier conclusion, but
they refine it. The `C5.0 tree` is uninformative as both a classifier
and a ranking model. The `Random forest` remains the best model for
actually detecting some delayed flights at the chosen threshold.
`Naive Bayes`, however, shows that a model can have weak classification
performance while still containing modest ranking value. From a
decision-support perspective, both the `Random forest` and `Naive Bayes`
seem more useful as weak risk-ranking tools than as strong stand-alone
classifiers.

## 7.12 Model disagreement and simple voting

If models disagree on the same flights, that can be useful information.
In general, disagreement may indicate that some flights are harder to
classify than others. We therefore compare the predictions of three
models with clearly different behavior: the plain `C5.0 tree`, the
`Random forest` and `Naive Bayes`.

This section should still be interpreted carefully. The three models do
not contribute equally. The plain `C5.0 tree` remains highly
conservative and predicts the majority class, while the `Random forest`
and `Naive Bayes` are the only models that sometimes predict delays.
That makes disagreement more informative here than in the earlier
tree-only comparisons, because it reflects actual differences between
model types rather than only repeated majority-class behavior.

```{r model_disagreement}
# put the three main model predictions next to each other
pred_patterns <- data.frame(
  tree = pred_tree_test,
  rf = pred_rf,
  nb = pred_nb
)

# check how often all three models agree
all_agree <- pred_patterns$tree == pred_patterns$rf &
  pred_patterns$rf == pred_patterns$nb

agreement_summary <- data.frame(
  outcome = c("All three agree", "At least one model disagrees"),
  share = c(mean(all_agree), mean(!all_agree))
)

agreement_summary_print <- agreement_summary
agreement_summary_print$share <- round(100 * agreement_summary_print$share, 2)
agreement_summary_print

ggplot(agreement_summary, aes(x = outcome, y = share)) +
  geom_col(fill = "grey") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  labs(
    title = "How often do the main models agree?",
    x = "",
    y = "Share of testing flights"
  )

# Hard-vote ensemble
pred_ensemble <- hard_vote(pred_patterns)
ensemble_metrics <- metric_summary(testing$IS_DELAYED_15, pred_ensemble)

ensemble_metrics
```

The agreement plot shows that the three models give the same prediction
for the large majority of testing flights: about 96.45% of cases receive
the same label from all three models, while disagreement occurs for only
about 3.55% of flights. This means that the models behave similarly most
of the time, and that only a small subset of flights is treated
differently. The ensemble results show what this disagreement means in
practice. The hard-vote model does move slightly beyond majority-class
prediction, but only to a very limited extent. On the testing data, it
correctly identifies 38 delayed flights and incorrectly flags 39 on-time
flights as delayed. This gives an accuracy of 0.8079, a delayed-class
recall of only about 0.0022, a precision of 0.4935 and an F1-score of
about 0.0044. This pattern is important. The ensemble is no longer
identical to the majority-class C5.0 models, but it remains far too
conservative to be useful for delayed-flight detection. Its relatively
high precision does not indicate strong performance; it mainly reflects
the fact that the model predicts `Delayed` only in a very small number
of cases. In other words, when the ensemble does flag a delay, it is
often correct, but it almost never flags one at all.

```{r comparison_with_ensemble}
comparison_with_ensemble <- rbind(
  comparison_table,
  data.frame(
    model = "Hard vote ensemble",
    accuracy = ensemble_metrics$accuracy,
    recall_delayed = ensemble_metrics$recall_delayed,
    precision_delayed = ensemble_metrics$precision_delayed,
    f1_delayed = ensemble_metrics$f1_delayed
  )
)

comparison_with_ensemble
```

The hard-vote rule does not provide a useful compromise between the
individual models. Although it does move slightly beyond trivial
majority-class prediction, its delayed-class recall remains extremely
low at 0.0022, far below both the`Random forest` (0.0606) and
`Naive Bayes` (0.0058). Its relatively high precision (0.4935) should
also be interpreted carefully: the ensemble predicts very few delayed
flights at all, so it avoids false alarms mainly by remaining highly
conservative.

This means that the ensemble suppresses most of the delay-detection
signal coming from the `Random forest` and, to a lesser extent, from
`Naive Bayes`. As a result, it ends up much closer to the majority-class
baseline than to the stronger non-trivial models. The ensemble is an
informative negative result: in the present setup, a simple unweighted
voting rule is too conservative to improve the practical detection of
delayed flights.

This does not mean that ensemble methods are useless in general. Rather,
it shows that this particular ensemble is poorly matched to the models
being combined. When one strong majority-class baseline is combined with
two only mildly non-trivial classifiers, an unweighted majority vote can
still be dominated by conservative predictions. So the problem here is
not the ensemble idea itself, but the fact that this simple voting rule
does not preserve the limited delayed-flight signal that is present in
the better individual models.

# 8. Conclusions and recommendations

This project examined whether flight delays can be predicted before
departure using variables that are operationally available in advance.
Across the full pipeline one main conclusion emerges: the data does
contain useful structure, but that structure is not strong enough to
produce a reliable classifier for delayed flights with the current
feature set. The problem is therefore substantially harder than overall
accuracy alone would suggest.

The exploratory analysis already pointed in this direction. Timing and
route-related variables consistently carried most of the usable
structure in the data. In both the PCA and clustering steps, scheduled
timing, distance, and scheduled elapsed time emerged as the main
dimensions along which flights differed. This means the representation
choices were meaningful: the later models were learning from variables
that reflect real operational differences between flights, rather than
from arbitrary coding effects. At the same time, the unsupervised
analysis also showed that the structure was not cleanly separable.
Flights did not fall into a few sharply distinct groups, and the feature
space remained spread over multiple dimensions. That already suggested
that prediction would be challenging.

The supervised modelling results confirm this. The plain `C5.0 tree`,
the `boosted C5.0 tree` and the `cost-sensitive C5.0 tree` all defaulted
to the majority class and predicted every flight as `On_time`. As a
result, each of these models achieved an accuracy of 0.8079, but a
delayed-class recall of 0. They looked acceptable on overall accuracy
only because most flights are on time, while completely failing on the
class that matters for the decision problem. The `hard-vote ensemble`
moved only slightly beyond this trivial behavior. Its delayed-class
recall was 0.0022, with a precision of 0.4935 and an F1-score of 0.0044.
So although it no longer behaved exactly like a pure majority-class
classifier, it remained far too conservative to be useful in practice.
The `Random forest` and `Naive Bayes` were the only models that moved
more clearly beyond trivial majority-class prediction. The
`Random forest` performed best for delayed-flight detection on the main
test split, reaching an accuracy of 0.7975, a delayed-class recall of
0.0606, a precision of 0.3458 and an F1-score of 0.1032. This is still
weak, but it shows that the pre-departure variables do contain at least
some predictive signal. The random forest variable-importance results
also fit the earlier exploratory findings: scheduled departure hour,
distance, and scheduled elapsed time were the most influential
predictors, while month, weekday and carrier contributed much less.
`Naive Bayes` was more conservative: it achieved an accuracy of 0.8075,
a delayed-class recall of 0.0058, a precision of 0.4226, and an F1-score
of 0.0115. This means that `Naive Bayes` produced fewer false alarms,
but also missed substantially more actual delays.

The repeated-split evaluation strengthens rather than changes this
conclusion. Across repeated train-test splits, the `C5.0 tree` and
`cost-sensitive C5.0` kept reproducing the same trivial majority-class
result. The `hard-vote ensemble` moved only marginally beyond this, with
an average delayed-class recall of about 0.0023. `Naive Bayes` performed
somewhat better, with average performance across repeated splits of
approximately 0.8071 for accuracy, 0.0077 for delayed-class recall,
0.3947 for precision and 0.0152 for F1. The `Random forest` again
remained the strongest model, with average performance of approximately
0.7994 for accuracy, 0.0473 for delayed-class recall, 0.3396 for
precision and 0.0830 for F1. So the Random forest is consistently better
than the simpler models, but its performance is still too limited for
strong operational use.

The ROC and AUC results add an important nuance. The `C5.0 tree` had an
AUC of 0.500, which is exactly chance level. The `Random forest` reached
an AUC of 0.623, while `Naive Bayes` performed slightly better at 0.626.
This shows that both non-trivial models contain some useful ranking
signal, even though their classification performance at the chosen
threshold remained weak. So while the `Random forest` remains the
strongest classifier for delayed-flight detection, `Naive Bayes`
performs at least as well (and here slightly better) as a
threshold-independent ranking model.

This leads to an important business conclusion. The current models are
not strong enough to support full automation of pre-departure
intervention decisions. In particular, the recall for delayed flights is
too low, which means that too many real delays would still be missed. So
the main value of this analysis is not that it produced a
deployment-ready prediction system, but that it clarified where useful
signal exists, where it remains too weak, and which variables matter
most. In that sense, the project succeeds analytically even though the
predictive performance remains limited.

### Practical recommendations

Based on the current results, we would not recommend using the tested
models as a stand-alone system to automatically decide which flights
should trigger strong operational intervention. The delayed-class recall
is too low for that purpose.

A more realistic recommendation is to use the current modelling output
as a weak prioritization or ranking signal rather than as a hard
decision rule. In practice, this could mean using the stronger
non-trivial models, such as the `Random forest` for delayed-flight
detection and possibly `Naive Bayes` for broader ranking support, to
place flights on an operations watchlist for additional human review,
rather than automatically triggering expensive interventions. In
practice, this means the airline could use the model to support
light-touch pre-departure monitoring. Flights flagged as relatively
higher risk could be sent to an operations watchlist for additional
human review, rather than automatically triggering expensive
interventions such as extra crew allocation or major gate changes. This
is a more defensible use case because it treats the model as an input
into decision-making, not as a replacement for operational judgment.

A second recommendation is to improve the data before trying to improve
the models further. The current feature set is limited to structured
pre-departure variables such as distance, schedule information, carrier,
and airport identifiers. That is appropriate for a fair predictive
setup, but it also means that the models do not observe many of the
mechanisms that actually generate delays. For practical improvement,
richer data are likely to matter more than simply trying more algorithms
on the same inputs.

The most useful additional data would probably include upstream aircraft
rotation information, inbound aircraft delay, weather conditions,
airport congestion indicators, runway or air traffic control pressure,
maintenance-related signals, and possibly crew or turnaround
information. These variables are much closer to the operational
processes that actually create delay risk. Without that kind of
information, the models mainly learn broad structural patterns rather
than the most important disruption drivers.

### Limitations of the current approach

This project has several limitations that should be acknowledged
clearly.

First, the models were trained on a relatively simple structured feature
space. That was a deliberate choice to keep the problem aligned with
pre-departure decision-making and to maintain interpretability, but it
also limits predictive power.

Second, the target itself is simplified. A binary label such as
`Delayed` versus `On_time` compresses a more nuanced operational reality
into one threshold-based outcome. In practice, the consequences of a
16-minute delay and a 90-minute delay are not the same and different
delay magnitudes may require different actions.

Third, evaluation remains conditional on design choices. A single
train-test split does not reveal “the truth,” and model performance
always depends partly on how the data are partitioned and which metric
is emphasized. We addressed this partly through repeated splits and by
looking beyond accuracy to recall, precision, F1 and AUC, but
uncertainty remains.

Fourth, the project does not directly include deployment constraints.
Even if a model performed somewhat better, real-world implementation
would still depend on whether predictions can be integrated into
workflows, whether the input data is timely and reliable, and whether
operations teams trust and understand the output.

### How more advanced analytics could help

Looking beyond the implemented models, there are several directions that
could improve the problem in principle, but only if the data and
decision framing become richer.

A first direction is representation learning. If much richer and more
detailed operational data were available, more flexible models could
potentially learn interactions between airport, carrier, schedule,
disruption, and network effects that the current tree-based models
cannot capture well. The potential value is greater modelling
flexibility, but the trade-off would be lower transparency and higher
demands on validation.

A second direction is text analytics. Useful text sources might include
maintenance logs, dispatcher notes, incident descriptions, weather
bulletins, or pilot reports. These could contain early warning
information that does not appear in structured schedule data. In that
case, NLP methods could transform text into additional pre-departure
risk signals. The challenge is that such data are noisy,
context-dependent, and harder to standardize.

A third direction is to reformulate the problem toward decision support
rather than pure prediction. Instead of asking only “Will this flight be
delayed?”, the airline may eventually care more about “Which flights
should be monitored or acted on first, given limited operational
capacity?” That is a different task, and one that may be better served
by ranking, prioritization, or more explicit cost-based decision
modelling than by a simple binary classifier alone.

### Final recommendation

Overall, our recommendation is to treat the current analysis as a useful
first predictive screening exercise, not as a deployment-ready delay
decision system. The immediate next step should not be “use a more
complex model,” but rather “improve the data and sharpen the decision
problem.”

More specifically, the most valuable follow-up would be to:

-   add operationally richer pre-departure data, especially upstream and
    disruption-related signals

-   evaluate models within a clearer decision framework, for example
    ranking flights for human review instead of forcing direct
    automation

-   continue using stronger validation designs, so that evaluation is
    seen as a distribution rather than as one fixed number

-   only consider substantially more advanced methods once the data
    richness and decision framing justify them

So the final conclusion is not that the problem is unsolvable, but that
the current structured pre-departure variables are not sufficient for a
strong operational classifier on their own. The project still succeeds
analytically because it identifies the main sources of signal, shows
clearly why some modelling approaches fail, and translates those
findings into realistic next steps for decision support.
