---
title: "MOT143A: Business Analytics group 20"
author: "Hidde Segers and Lotte Feilzer"
date: "2026-03-04"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

# 1. Business context and analytics task

## 1.1 Business problem

**Context:** Airlines running domestic flights in the United States
(U.S.) must manage constantly resources such as gates, crew members,
ground handling staff, and customer service capacity. These decisions
have to be made quickly and often with limited certainty.

Flight delays and cancellations can cause a lot of trouble for those who
are trying to get on time. When things go wrong, airlines have to pay
for: crew overtime, gate conflicts, aircraft re-positioning,
compensation, re-booking logistics, and re-putational damage. At the
same time, customer satisfaction declines, which affects their name and
in the long-term their revenue.

**Problem:** Delays and cancellations are really expensive for airlines
and can be frustrating for passengers. A single delay can create high
costs for crew overtime, gate conflicts, re-booking workloads. Because
flights are connected in a daily schedule, one delay can also cause
follow-on delays later in the network. This increases operational costs
and directly reduces customer satisfaction.

The practical challenge is that a lot of operational decisions must be
made before the flights departures. At that time it is still not clear
whether the flight is running on time or not. If the airline can figure
out flights that are likely to be delayed early, it can allocate
resources more active (e.g., assign standby crew, prepare customer
service capacity or adjust gate plans) and reduce the impact when such
delays occurs.

This leads to our business question:

**Business question:** How can we estimate the probability that a
scheduled flight will be delayed (e.g. more than 15 min) before
departure?

The goal is not to "predict the truth" perfectly, but to provide a risk
signal that helps operations teams prioritize attention and resources
towards flights with the highest expected delay risk. From an
organizational perspective, the insights are meant for operations
control centers, network planners and day-of-operations managers. The
model would be used in real-time operational setting, where planners
monitor flights and allocate resources (that are most of the time
limited). In this context, the prediction functions as a
decision-support tool that can help prioritize attention and
interventions for flights with the highest risk of delay.

## 1.2 Analytics task

This analysis is focused on the perspective of the airline operations
control and the airport station management to identify high risk
flights. This should support the workers with allocation of staff, gates
and premature consumer communication. The model will be used for
pre-departure planning setting and will be based on the historical
operational data of US domestic flights. The business problem will be
translated to an analytic task by approaching it as a predictive
risk-scoring problem. Three possible targets were considered:

-   **Classification A (delay):**

    **Target:** `IS_DELAYED_15 = 1` als `ARR_DELAY >= 15`, anders 0.

    Here it is analysed whether or not a flight will be significantly
    delayed or not (this is for flights that are neither diverted or
    delayed).

-   **Classification B (cancellation):**

    **Target:** `CANCELLED == 1`

    Here it is analysed whether or not a flight will be cancelled.

-   **Regression:**

    **Target:** `ARR_DELAY`

    Here it is analysed how much too late a flight will likely be in
    minutes.

The primary task chosen is classification A, this is because it will be
directly usable to relocate resources if needed and where focus in
logistics needs to focused. It is important to press that the model
can't reason and does not rely on causality. It also will not be able to
guarantee outcomes. It will estimate the risk for significant delay and
consequently improve and support operational actions. Classification for
cancellation will be harder because of the rarity of this event. The
regression of the delay is also hard to predict and the type of delay
can but does not also translate to how much delay there will be.

## 1.3 Success criteria

We consider the success criteria for this report in two perspectives.
Firstly, model criteria that will be need to be met to answer our
business question. Secondly, the business criteria that will ensure
positive effects of the model in practice.

From an operational point of view, missing a real delay is usually worse
than raising a false alarm. If the model says a flight will be on time
but it is actually delayed, the airline is unprepared. This can quickly
lead to crew problems, gate conflicts, and unhappy passengers.

If the model predicts a delay that does not happen, it may lead to some
extra costs because resources were allocated unnecessarily. However,
this is less harmful than being unprepared for a real delay.

For this reason, we focus on high recall for the delayed flight s. We
want the model to catch as many real delays as possible, even if that
means accepting some false alarms. At the same time, we monitor
precision to make sure we do not allocate too many unnecessary
resources.

# 2. Setup

Here all packages are downloaded to make the code reproducible.

```{r setup}
# CRAN mirror
local({r <- getOption("repos")
r["CRAN"] <- "https://cran.rstudio.com/"
options(repos = r)})

packages <- c("ggplot2", "Hmisc", "rsample", "gmodels", "C50", "randomForest",
  "factoextra", "lattice", "reshape2"
)

for (i in packages) {
  if (!require(i, character.only = TRUE)) {
    install.packages(i, dependencies = TRUE)
    library(i, character.only = TRUE)
  }
}

set.seed(46748717)
```

# 3. Data loading

## 3.1 Data in `data/` folder

Here we read our data. In the code it is easy to change the year of data
we want to see, just change the variable `YEAR`. Because the data_files
are quite big we added the possibility to print only a specific amount
of rows as given in `nrows` a value of `-1` will give all rows.

```{r data_files}
data_dir <- "data"
YEAR <- 2018
csv_path <- file.path(data_dir, paste0(YEAR, ".csv"))
flights <- read.csv(csv_path, stringsAsFactors = FALSE, nrows = N_READ)
```

## 3.2 Read data

Now we read our actual data and print the dimensions and columns names
to get a first clue of what kind of data were dealing with.

```{r read_data}
N_READ <- -1
dim(flights)
names(flights)
```

As seen the datafile contains:

-   7,213,446 observations

-   28 variables

We see that each observation represents one commercial flight in 2018.
The variables are as follows and categorized in definitions that can be
used to refer to groups of variables further in the report:

-   **Flight identifiers:** FL_DATE, OP_CARRIER, OP_CARRIER_FL_NUM

-   **Route information:** ORIGIN, DEST

-   **Scheduled times:** CRS_DEP_TIME, CRS_ARR_TIME

-   **Actual flight performance:** DEP_TIME, ARR_TIME, DEP_DELAY,
    ARR_DELAY

-   **Operational metrics:** TAXI_OUT, TAXI_IN, AIR_TIME

-   **Delay causes:** CARRIER_DELAY, WEATHER_DELAY, NAS_DELAY,
    SECURITY_DELAY, LATE_AIRCRAFT_DELAY

-   **Flight status indicators:** CANCELLED, DIVERTED

The variables can be sorted into the following categories:

### Flight identifiers

| Variable          | Description        |
|-------------------|--------------------|
| FL_DATE           | Date of the flight |
| OP_CARRIER        | Operating airline  |
| OP_CARRIER_FL_NUM | Flight number      |

### Route information

| Variable | Description               |
|----------|---------------------------|
| ORIGIN   | Departure airport         |
| DEST     | Arrival airport           |
| DISTANCE | Distance between airports |

### Scheduled times

| Variable         | Description               |
|------------------|---------------------------|
| CRS_DEP_TIME     | Scheduled departure time  |
| CRS_ARR_TIME     | Scheduled arrival time    |
| CRS_ELAPSED_TIME | Scheduled flight duration |

### Actual flight performance

| Variable  | Description           |
|-----------|-----------------------|
| DEP_TIME  | Actual departure time |
| ARR_TIME  | Actual arrival time   |
| DEP_DELAY | Departure delay       |
| ARR_DELAY | Arrival delay         |
| AIR_TIME  | Time spent in air     |

### Operational metrics

| Variable   | Description              |
|------------|--------------------------|
| TAXI_OUT   | Taxi time before takeoff |
| TAXI_IN    | Taxi time after landing  |
| WHEELS_OFF | Takeoff time             |
| WHEELS_ON  | Landing time             |

### Flight status

| Variable  | Description                 |
|-----------|-----------------------------|
| CANCELLED | Indicates cancelled flights |
| DIVERTED  | Indicates diverted flights  |

### Delay causes

| Variable            | Description                            |
|---------------------|----------------------------------------|
| CARRIER_DELAY       | Delay caused by airline                |
| WEATHER_DELAY       | Delay caused by weather                |
| NAS_DELAY           | Delay caused by air traffic system     |
| SECURITY_DELAY      | Delay caused by security               |
| LATE_AIRCRAFT_DELAY | Delay caused by late incoming aircraft |

# 4. Data exploration & preparation

## 4.1 Data summary

Here we show a small summary of the data. The following things are
noticeable. Columns in this datafile that contain dates are characters
preferably we change this later to datetimes. It is also remarkable that
the times are given in military time but as integers, this will make it
hard to calculate certain metrics inveolving time such as "time passed"
therefor we will preferably also change these datatypes later on to
datetimes. For metrics such as `TAXI_OUT` and `DEP_DELAY` the integers
are suitable because it represents the amount of minutes passed. We see
that for the `DIVERTED` and `CANCELLED` columns `0` and `1` are used
instead of Booleans to respresent `FALSE` and `TRUE` respectively, this
is also workable if remarked. Let's further explore the data.

```{r explore_basic}
str(flights)
```

## 4.2 Missing values overview and discussion

Here we check whether or not we have missing values in our dataset, how
many there are and what are the possible reasons for these missing
values are. We use logical reasoning in order to determine what to do
with potential missing data. Here we show a table that shows the amount
of NA values for each column and also a percentage of how many values
are missing in regards to the total dataset.

```{r missingness}
na_count <- sapply(flights, function(x) sum(is.na(x)))
na_pct   <- round(100 * na_count / nrow(flights), 2)

na_table <- data.frame(
  na_count = as.integer(na_count),
  na_pct = na_pct
)

na_table <- na_table[order(-na_table$na_pct), ]
na_table
```

Firstly we see a column `Unnamed..27` which seems to have only NA values
making the column absolute and removable. We see for the operational
variables and flight performance variables such as `DEP_TIME`,
`ARR_TIME`, `AIR_TIME`, `WHEELS_ON`, `WHEELS_OFF` that 1-2% of all
values are missing. We should note that all arrival variables have about
the same amount of NA values missing and idem for the departure values.
This probably means that the NA values are probably/hopefully
explainable for the same reasons. First we check whether the NAs
occurring in the delay rows all occur in the same row. The cell below
tells us this is the case. This lets us know that these `DELAY` causes
variables are jointly observed or missing. This is good sign. These rows
are supposed to be filled in after a flight and give a reason for the
delay if it was significant delayed (15 minutes or more). In our check
this also seems to be the case. All flights that are 15+ minutes delayed
have a cause and vice versa.

```{r delay-na-same}
cause_cols <- intersect(c("CARRIER_DELAY","WEATHER_DELAY",
                    "NAS_DELAY","SECURITY_DELAY","LATE_AIRCRAFT_DELAY"), names(flights))
delay15 <- !is.na(flights$ARR_DELAY) & flights$ARR_DELAY >= 15
cause_present <- !is.na(flights[[cause_cols[1]]])

miss <- is.na(flights[, cause_cols, drop=FALSE])
same_na_rows <- all(apply(miss, 2, function(x) identical(which(x), which(miss[,1]))))

cat("All the delay causes with NA values are for the same row, is", same_na_rows, "\n")
cat("Causes are present but there was no 15 min delay:", sum(cause_present & !delay15), "\n")
cat("More than 15 min delay but cause is NA value:", sum(delay15 & !cause_present))
```

Now we check if it possible for a delay flight to have multiple reasons
for delay this might handy to know later on. This seems to be possible
and is the case for 51.56% of the delayed flights.

```{r cause-compact}
k <- rowSums(flights[cause_present, cause_cols, drop=FALSE] > 0)
tab <- table(k)

multi_n <- sum(k > 1)
total_n <- length(k)
multi_pct <- round(100 * multi_n / total_n, 2)

cat("Of the flights with delay causes,", multi_pct, "% have multiple causes for delay.")
```

We now know that our delay causes columns are in fact complete. We can
now check if for the flights that have taken off all the important
operational and performance variables are present (`ARR_DELAY`,
`ARR_TIME` ect.) or that some or all values are missing for certain
rows. The NA values for the these rows as presented above could be
explained if these are the rows of flights that were diverted or
cancelled. If not, crucial information might be missing. If so, we
analyse this here.

```{r}
flights_wo_cancelled <- which(flights$CANCELLED == 0 & flights$DIVERTED == 0)
flights_wo_cancelled_df <- flights[flights_wo_cancelled, ]

na_count_wo_cd <- sapply(flights_wo_cancelled_df, function(x) sum(is.na(x)))
na_pct_wo_cd   <- round(100 * na_count_wo_cd / nrow(flights_wo_cancelled_df), 2)

na_table_wo_cd <- data.frame(
  variable = names(flights_wo_cancelled_df),
  na_count_wo_cd = as.integer(na_count_wo_cd),
  na_pct_wo_cd = na_pct_wo_cd
)

na_table_wo_cd <- na_table_wo_cd[order(-na_table_wo_cd$na_pct_wo_cd), ]
na_table_wo_cd
```

We can see that a flight that was either cancelled or diverted, it
accounts for the reasons that rows contain NA values for performance and
operational variables. We can see that the only two performance
variables that have still NA values are `ARR_DELAY` and `DEP_DELAY`,
why? These NA values might cause some trouble when modelling because we
use `ARR_DELAY` as predictors in the pre-departure model. Let's check by
showing the rows were this data is missing.

```{r check-missing-mostly-delays}
dep_delays_missing <- which(is.na(flights_wo_cancelled_df$DEP_DELAY)) 
arr_delays_missing <- which(is.na(flights_wo_cancelled_df$ARR_DELAY))
missing_delays <- flights_wo_cancelled_df[c(dep_delays_missing, arr_delays_missing), ]
head(missing_delays)
```

First we need to look into what we want to do with the rows in which the
`DEP_DELAY` and/or `ARR_DELAY` variables are missing. It seems that
there are only NA values for these columns when the `CRS_DEP_TIME` and
`DEP_TIME` have the same value and idem for the arrival. This means that
in this dataset the NA values just mean that plane left exactly at the
intended time. If this is the case we can replace the NA values in these
rows for 0 values and these NA values are also explained than as well.
We perform a quick check to see this is a 100 percent the case.

```{r fill_zero_delays}
dep_na <- !is.na(missing_delays$CRS_DEP_TIME) & !is.na(missing_delays$DEP_TIME) &
                 is.na(missing_delays$DEP_DELAY) &
                 missing_delays$CRS_DEP_TIME == missing_delays$DEP_TIME

arr_na <- !is.na(missing_delays$CRS_ARR_TIME) & !is.na(missing_delays$ARR_TIME) &
                 is.na(missing_delays$ARR_DELAY) &
                 missing_delays$CRS_ARR_TIME == missing_delays$ARR_TIME

cat("DEP_DELAY NA but not CRS_DEP_TIME==DEP_TIME:", sum(dep_na), "\n")
cat("ARR_DELAY NA but not CRS_ARR_TIME==ARR_TIME:", sum(arr_na))
```

Now we have analysed all NA values and begin to clean our data properly.

## 4.3 Cleaning and datatypes

First, we do some final sanity checks for the values for variables. We
check whether the distances are positive as well as airtime. We also
check whether the departure and arrival delays are plausible which all
seems to be the case.

```{r sanity_checks}
cat("There are", sum(flights$DISTANCE <= 0), "negative distances.", "\n")
cat("There are", sum(flights$AIR_TIME <= 0, na.rm = TRUE) , "negative airtimes.", "\n")
print("Departure delay summary:")
summary(flights$DEP_DELAY) 
print("Arrival delay summary:")
summary(flights$ARR_DELAY) 
```

We create a df named `flights_clean` to not overwrite the `flights`
variable that can be cleaned without overwriting our original data.

```{r create_flights_clean}
flights_clean <- flights
```

We begin with adjusting in accordance with the remarks made in section
3.2. We the `FL_DATE` column a datatype datetime and drop the unuseful
column `Unnamed..27` and ensure that the columns are the correct
datatypes.

```{r clean_types}
drop_cols <- grep("^Unnamed", names(flights_clean), value = TRUE)
flights_clean[drop_cols] <- NULL

flights_clean$FL_DATE <- as.Date(flights_clean$FL_DATE)

head(flights_clean, 4)
```

Now we do some minimal cleaning make sure the datatypes are correct. We
make the integer time columns POSIXct datetime columns to be able to
treat the time correctly.

```{r datetime_fix_overwrite}
hhmm_to_hhmm_str <- function(x) {
  x <- as.integer(x)
  hh <- x %/% 100
  mm <- x %% 100
  out <- sprintf("%02d:%02d", hh, mm)
  out[is.na(x) | mm > 59 | hh < 0 | hh > 29] <- NA
  out
}

time_cols <- intersect(c("WHEELS_OFF","WHEELS_ON","DEP_TIME","ARR_TIME","CRS_DEP_TIME","CRS_ARR_TIME"), names(flights_clean))

# Replace columns to POSIXct
for (cn in time_cols) {
  tmp_str <- hhmm_to_hhmm_str(flights_clean[[cn]])
  flights_clean[[cn]] <- as.POSIXct(
    paste(as.character(flights_clean$FL_DATE), tmp_str),
    format = "%Y-%m-%d %H:%M",
    tz = "UTC"
  )
}

# Overnight: is ARR_TIME < DEP_TIME + one day
if (all(c("DEP_TIME","ARR_TIME") %in% names(flights_clean))) {
  overnight <- !is.na(flights_clean$DEP_TIME) &
               !is.na(flights_clean$ARR_TIME) &
               (flights_clean$ARR_TIME < flights_clean$DEP_TIME)

  arrival_side <- intersect(c("ARR_TIME","WHEELS_ON","CRS_ARR_TIME"), time_cols)
  for (cn in arrival_side) {
    flights_clean[[cn]][overnight] <- flights_clean[[cn]][overnight] + 24*3600
  }
}

head(flights_clean[, time_cols, drop=FALSE], 5)
```

We now clean the NA values for the delay columns when the `CRS_DEP_TIME`
and `DEP_TIME` are the same and idem for the arrival times.

```{r impute_delays_posixct}
idx_dep0 <- which(
  is.na(flights_clean$DEP_DELAY) &
  !is.na(flights_clean$DEP_TIME) & !is.na(flights_clean$CRS_DEP_TIME) &
  flights_clean$DEP_TIME == flights_clean$CRS_DEP_TIME
)

idx_arr0 <- which(
  is.na(flights_clean$ARR_DELAY) &
  !is.na(flights_clean$ARR_TIME) & !is.na(flights_clean$CRS_ARR_TIME) &
  flights_clean$ARR_TIME == flights_clean$CRS_ARR_TIME
)

# vullen
flights_clean$DEP_DELAY[idx_dep0] <- 0
flights_clean$ARR_DELAY[idx_arr0] <- 0

# check
head(flights_clean[idx_dep0, c("CRS_DEP_TIME","DEP_TIME","DEP_DELAY")])
head(flights_clean[idx_arr0, c("CRS_ARR_TIME","ARR_TIME","ARR_DELAY")])
```

Now we check if all the data is clean. First we delete all the cancelled
and diverted rows because theserows are not relevant for our analysis.

```{r}
keep <- (is.na(flights_clean$CANCELLED) | flights_clean$CANCELLED == 0) &
        (is.na(flights_clean$DIVERTED)  | flights_clean$DIVERTED  == 0)

flights_clean <- flights_clean[keep, , drop = FALSE]

cat("Removed:", sum(!keep), "\n")
cat("Rows after:", nrow(flights_clean))
```

We decide to add some columns that make our analysis and modelling
easier later on, such as `IS_DELAYED` day of the week column which might
be interesting in the analysis and a column which contains a Boolean on
whether a flight is significantly delayed (15 minutes or later).Now we
add a column in which we see if a plane was significantly delayed. We
also add a `MONTH` and `WEEKDAY` column.

```{r}
flights_clean$IS_DELAYED <- ifelse(flights_clean$DEP_DELAY > 15, TRUE, FALSE) 
flights_clean$WEEKDAY <- weekdays(flights_clean$FL_DATE)
flights_clean$MONTH <- as.integer(format(flights_clean$FL_DATE, "%m"))
```

# 5. Exploratory visualisation

First we have a look at the base rate, this is the amount of how many
flights are more than 15 minutes delayed

```{r mean_delay_rate}
delay_rate <- mean(flights_clean$IS_DELAYED)
delay_rate 
```

First we show a bar chart of our data that simply displays the amount of
flights that are delayed and on-time. We see that around 18% of flights
are significantly delayed.

```{r}
df_target <- data.frame(
  class = factor(flights_clean$IS_DELAYED, levels = c(0,1), labels = c("On time", "Delayed (15 min)"))
)

ggplot(df_target, aes(x = class)) +
  geom_bar() +
  labs(title = "Class distribution: delayed vs on-time", x = "", y = "amount of flights") +
  coord_flip()
```

Here we plot the delay rate per month and the amount of flights per
month. It is worth mentioning that the in the summer months the delay
amount significantly increasing proportional to the slight increase in
flights that we see in the graph.

```{r}
df_month_rate <- aggregate(IS_DELAYED ~ MONTH, flights_clean, function(x) mean(x, na.rm = TRUE))
df_month_n    <- aggregate(IS_DELAYED ~ MONTH, flights_clean, length)
names(df_month_n)[2] <- "n"

df_month <- merge(df_month_rate, df_month_n, by = "MONTH")
names(df_month)[2] <- "delay_rate"

pct_label <- function(x) paste0(round(100*x, 1), "%")

ggplot(df_month, aes(x = MONTH, y = delay_rate)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(labels = pct_label) +
  coord_cartesian(ylim = c(0, 0.30)) +
  labs(title = "Delay rate per month", x = "Month", y = "Delay rate")

ggplot(df_month, aes(x = MONTH, y = n)) +
  geom_col() +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Amount of flights per month", x = "Month", y = "#")
```

The delays per weekday shows that those are relatively stable during the
week, with values around 18-21%. On Fridays and Thursdays there are
slightly highiher percentagerates, while on Saterday their are clearly
lower delay rates. This pattern can be explained by the fact that there
are differneces in flightvolume and operational pressure: during the
week their are more commercial and business related flights that arwe
aslo connected, while on Saturday the operational pressure is lower,
what results in less delay. In general, it seems that the day of the
week does not have significantly influence on the change of delay, what
suggest that there are other factors, suchs as departure time, airport
or airplane that are maybe play ar ole in the explanination of the
delays.

```{r}
df_wday_rate <- aggregate(IS_DELAYED ~ WEEKDAY, flights_clean, mean)
names(df_wday_rate)[2] <- "delay_rate"

ggplot(df_wday_rate, aes(x = WEEKDAY, y = delay_rate)) +
  geom_col() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Delay rate per weekday", x = "Day of the week", y = "Delay rate")

```

Heat map: weekday x departure hour

```{r}
df_hw <- aggregate(IS_DELAYED ~ WEEKDAY + CRS_DEP_TIME, flights_clean, mean)
names(df_hw)[3] <- "delay_rate"

ggplot(df_hw, aes(x = CRS_DEP_TIME, y = WEEKDAY, fill = delay_rate)) +
  geom_tile() +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  scale_fill_gradient(labels = pct_label, limits = c(0, 1)) +
  labs(title = "Heatmap delay rate: weekday × vertrekuur",
       x = "Planned departure hour", y = "Day of the week", fill = "Delay rate")

```

Delay rate per airplane. The first planes seem to be private jets.

```{r}
MIN_N_CARRIER <- 5000

carrier_rate <- aggregate(IS_DELAYED ~ OP_CARRIER, flights_clean, mean)
carrier_n    <- aggregate(IS_DELAYED ~ OP_CARRIER, flights_clean, length)
names(carrier_n)[2] <- "n"

carrier_summary <- merge(carrier_rate, carrier_n, by = "OP_CARRIER")
names(carrier_summary)[2] <- "delay_rate"

carrier_summary <- subset(carrier_summary, n >= MIN_N_CARRIER)
carrier_summary <- carrier_summary[order(-carrier_summary$delay_rate), ]

top_carriers <- head(carrier_summary, 10)

ggplot(top_carriers, aes(x = reorder(OP_CARRIER, delay_rate), y = delay_rate)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Delay rate (>=15m) per airline (top 10, min volume filter)",
       x = "Airplane", y = "Delay rate")
```

Delay rate per airport

```{r}
MIN_N_AIRPORT <- 20000

origin_rate <- aggregate(IS_DELAYED ~ ORIGIN, flights_clean, mean)
origin_n    <- aggregate(IS_DELAYED ~ ORIGIN, flights_clean, length)
names(origin_n)[2] <- "n"

origin_summary <- merge(origin_rate, origin_n, by = "ORIGIN")
names(origin_summary)[2] <- "delay_rate"

origin_summary <- subset(origin_summary, n >= MIN_N_AIRPORT)
origin_summary <- origin_summary[order(-origin_summary$delay_rate), ]

top_origins <- head(origin_summary, 15)

ggplot(top_origins, aes(x = reorder(ORIGIN, delay_rate), y = delay_rate)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Delay rate (>=15m) per origin airport (top 15, min volume filter)",
       x = "Origin airport", y = "Delay rate")
```

Scatterplot

```{r}
origin_top50_vol <- origin_summary[order(-origin_summary$n), ][1:50, ]

ggplot(origin_top50_vol, aes(x = n, y = delay_rate)) +
  geom_point() +
  scale_x_log10() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Origin airports: volume vs delay rate (top 50 volume)",
       x = "Aantal vluchten (log schaal)", y = "Delay rate")
```

Multivariate relations and representation choices for features. In other
words distance and scheduled departure as risk factors

```{r}
set.seed(46748717)
N_VIS <- min(200000, nrow(flights_clean))
idx_vis <- sample(seq_len(nrow(flights_clean)), N_VIS)
flights_vis <- flights_clean[idx_vis, ]
ggplot(flights_vis, aes(x = DISTANCE, y = IS_DELAYED)) +
  stat_summary_bin(fun = mean, bins = 60, geom = "line") +
  scale_y_continuous(labels = pct_label, limits = c(0, 1)) +
  labs(title = "Delay risk vs distance (binned mean)", x = "Distance (miles)", y = "Delay risk")

flights_vis <- flights_clean[idx_vis, ]

ggplot(flights_vis, aes(x = CRS_DEP_TIME, y = IS_DELAYED)) +
  stat_summary_bin(fun = mean, bins = 48, geom = "line") +
  scale_y_continuous(labels = pct_label, limits = c(0, 1)) +
  labs(title = "Delay risk vs planned departure time (minutes)", x = "CRS_DEP_TIME", y = "Delay risk")
```

```{r eda_airline_airport}
# Filter to flights where delay target is defined
if (all(c("IS_DELAYED_15","OP_CARRIER","ORIGIN") %in% names(flights))) {
  flights_clean <- flights[!is.na(flights$IS_DELAYED_15), ]

  airline_summary <- aggregate(IS_DELAYED_15 ~ OP_CARRIER, flights_clean,
                               function(x) mean(x, na.rm = TRUE))
  airline_summary <- airline_summary[order(-airline_summary$IS_DELAYED_15), ]

  ggplot(airline_summary[1:10, ],
         aes(x = reorder(OP_CARRIER, IS_DELAYED_15), y = IS_DELAYED_15)) +
    geom_col() +
    coord_flip() +
    labs(title = "Delay rate (>=15 min) per airline (top 10)",
         x = "Airline", y = "Delay rate")
}
```

# 6. Dimension reduction and clustering

Before we can explore the PCA approach, we first have to anaylse which
variables are suitable for a PCA and which representation we choose. In
appendix X we compose a feature PCA-ready set. This in order to have
numerical variables which together ma contain a certain latent structure
without putting in target or post-event information. In this way the PCA
performs the best. We do not use PCA "because we have to, but because we
want to solve our business problem at the end. In this case the PCA have
three possible goals: 1. discover structyre in operational features (for
exmaple are schedulded departure time, rute distnace, elapsed time,
month, weekday, airport, pressure indicators related?) e.g. is there a
compact pattern underneath? 2. Reduce redunancy, for example: some time
variables or route/operational variables contain partly overlapping
information. PCA can summarize that overlap in fewer components. 3.
create more compact input for a model, if we later create a lot of
numerical derivative features and which are strongly related then PCA
can help to reduce it to a smaller number of components. Our first
concpetual quesgion here is: Do we actually have a numerical feature
space in which PCA makes sense?

## 6.1 PCA

## 6.2 Correlation heatmap

# 7. Modelling

## 7.1 Model 1: C5.0 decision tree

```{r}
flights_model <- flights_clean

flights_model$IS_DELAYED_15 <- ifelse(flights_model$ARR_DELAY >= 15, "Delayed", "On_time")
flights_model$IS_DELAYED_15 <- factor(flights_model$IS_DELAYED_15)

flights_model$DEP_HOUR <- as.numeric(format(flights_model$CRS_DEP_TIME, "%H")) +
                          as.numeric(format(flights_model$CRS_DEP_TIME, "%M")) / 60

flights_model$ARR_HOUR <- as.numeric(format(flights_model$CRS_ARR_TIME, "%H")) +
                          as.numeric(format(flights_model$CRS_ARR_TIME, "%M")) / 60

factor_vars <- c("OP_CARRIER", "ORIGIN", "DEST", "WEEKDAY")

for (v in factor_vars) {
  if (v %in% names(flights_model)) {
    flights_model[[v]] <- factor(flights_model[[v]])
  }
}

model_vars <- c(
  "IS_DELAYED_15",
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "DEP_HOUR",
  "ARR_HOUR",
  "MONTH",
  "WEEKDAY",
  "OP_CARRIER",
  "ORIGIN",
  "DEST"
)

model_df <- flights_model[, model_vars]
model_df <- na.omit(model_df)

str(model_df)
summary(model_df$IS_DELAYED_15)

set.seed(46748717)

N_MODEL <- min(300000, nrow(model_df))   # adjust if needed
idx <- sample(seq_len(nrow(model_df)), N_MODEL)

model_df_small <- model_df[idx, ]
dim(model_df_small)

set.seed(46748717)

split <- rsample::initial_split(model_df_small, prop = 0.7)
training <- training(split)
testing  <- testing(split)

dim(training)
dim(testing)

model_tree <- C50::C5.0(IS_DELAYED_15 ~ ., data = training)

summary(model_tree)

pred_test <- predict(model_tree, testing)

gmodels::CrossTable(
  testing$IS_DELAYED_15,
  pred_test,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

#Accuracy 
mean(pred_test == testing$IS_DELAYED_15)

actual <- testing$IS_DELAYED_15
pred   <- pred_test

tp <- sum(actual == "Delayed" & pred == "Delayed")
tn <- sum(actual == "On_time" & pred == "On_time")
fp <- sum(actual == "On_time" & pred == "Delayed")
fn <- sum(actual == "Delayed" & pred == "On_time")

accuracy  <- (tp + tn) / (tp + tn + fp + fn)
recall    <- tp / (tp + fn)
precision <- tp / (tp + fp)

data.frame(
  accuracy = accuracy,
  recall_delayed = recall,
  precision_delayed = precision,
  tp = tp, tn = tn, fp = fp, fn = fn
)
```

## 7.2 Model 2: Random Forest

## 7.3 Comparison

# 8. Conclusions and recommendations

## 8.1 Operational recommendations

## 8.2 Limitations & next steps

# 9. Appendix

## 9.1 Variable screening for PCA

```         
For our project, PCA only makes sense on variables that meet a few conditions.  A variable is especially interesting for PCA in your setting if it:  is available before departure, contains no leakage, is numerical, or can be logically made numerical,  can contribute something substantive to a joint structure, is not simply the outcome itself or a post-event consequence.  So this is not a purely technical selection, but above all an analytical selection. Because the business question is formulated in a sense that we may only rely on pre-departutre information. So everything that is newly known: during flight, after departure, after arrival, or after the delay has already occurred, should be excluded for PCA in this setting. That is the most important criterion.
```

```{r}
#Omdat we nog niet echt feature ready data hebben om de PCA uit te voeren, converteren we even de data in useable data zodat we die verder kunnen gebruiken in de PCA

pca_screen <- data.frame(
  variable = c(
    "FL_DATE", "OP_CARRIER", "OP_CARRIER_FL_NUM",
    "ORIGIN", "DEST", "DISTANCE",
    "CRS_DEP_TIME", "CRS_ARR_TIME", "CRS_ELAPSED_TIME",
    "DEP_TIME", "ARR_TIME", "DEP_DELAY", "ARR_DELAY", "AIR_TIME",
    "TAXI_OUT", "TAXI_IN", "WHEELS_OFF", "WHEELS_ON",
    "CANCELLED", "DIVERTED",
    "CARRIER_DELAY", "WEATHER_DELAY", "NAS_DELAY", "SECURITY_DELAY", "LATE_AIRCRAFT_DELAY",
    "IS_DELAYED", "WEEKDAY", "MONTH"
  ),
  type = c(
    "date", "categorical", "identifier",
    "categorical", "categorical", "numeric",
    "time", "time", "numeric",
    "time_actual", "time_actual", "post_event", "post_event", "post_event",
    "post_event", "post_event", "post_event", "post_event",
    "status", "status",
    "post_event_cause", "post_event_cause", "post_event_cause", "post_event_cause", "post_event_cause",
    "target", "categorical", "calendar_numeric"
  ),
  pre_departure_available = c(
    TRUE, TRUE, TRUE,
    TRUE, TRUE, TRUE,
    TRUE, TRUE, TRUE,
    FALSE, FALSE, FALSE, FALSE, FALSE,
    FALSE, FALSE, FALSE, FALSE,
    FALSE, FALSE,
    FALSE, FALSE, FALSE, FALSE, FALSE,
    FALSE, TRUE, TRUE
  ),
  pca_label = c(
    "watchlist", "watchlist", "ignore",
    "watchlist", "watchlist", "usable",
    "watchlist", "watchlist", "usable",
    "ignore", "ignore", "ignore", "ignore", "ignore",
    "ignore", "ignore", "ignore", "ignore",
    "ignore", "ignore",
    "ignore", "ignore", "ignore", "ignore", "ignore",
    "ignore", "watchlist", "watchlist"
  ),
  reason = c(
    "Raw date not suitable; use derived calendar features instead",
    "Raw carrier code not numeric; maybe useful through historical aggregates",
    "Identifier-like variable; no meaningful numeric PCA scale",
    "Raw airport code not numeric; maybe useful through aggregates",
    "Raw airport code not numeric; maybe useful through aggregates",
    "Numeric, pre-departure, operationally meaningful",
    "Raw clock time is cyclical; re-encode before PCA",
    "Raw clock time is cyclical; re-encode before PCA",
    "Numeric, pre-departure, operationally meaningful",
    "Actual departure time is post-event",
    "Actual arrival time is post-event",
    "Realized departure delay is post-event / leakage",
    "Realized arrival delay is post-event / target-adjacent",
    "Realized air time is post-event",
    "Realized taxi-out is post-event",
    "Realized taxi-in is post-event",
    "Realized takeoff time is post-event",
    "Realized landing time is post-event",
    "Outcome/status variable, not predictor",
    "Outcome/status variable, not predictor",
    "Delay cause recorded after event",
    "Delay cause recorded after event",
    "Delay cause recorded after event",
    "Delay cause recorded after event",
    "Delay cause recorded after event",
    "Target variable, must not enter PCA",
    "Useful only after encoding / transformation",
    "Potentially useful, but cyclical/calendar representation preferred"
  ),
  stringsAsFactors = FALSE
)


# =========================
# Variables to inspect visually for PCA screening
# =========================

screen_vars <- subset(pca_screen, pca_label %in% c("usable", "watchlist"))$variable
screen_vars <- intersect(screen_vars, names(flights_clean))

screen_vars

# =========================
# Appendix plotting function
# =========================

plot_screen_variable <- function(df, varname, max_levels = 20) {
  
  x <- df[[varname]]
  
  # Datetime / POSIXct
  if (inherits(x, "POSIXct") || inherits(x, "Date")) {
    message("Variable ", varname, " is date/datetime. Plot derived summary instead.")
    
    # For Date: show counts over time if not too many dates
    tmp <- data.frame(value = x)
    p <- ggplot(tmp, aes(x = value)) +
      geom_bar() +
      labs(title = paste("Distribution of", varname),
           x = varname, y = "Count")
    print(p)
    
  # Numeric / integer / logical
  } else if (is.numeric(x) || is.integer(x)) {
    tmp <- data.frame(value = x)
    p <- ggplot(tmp, aes(x = value)) +
      geom_histogram(bins = 40, fill = "steelblue", color = "white") +
      labs(title = paste("Histogram of", varname),
           x = varname, y = "Count")
    print(p)
    
    p2 <- ggplot(tmp, aes(y = value)) +
      geom_boxplot(fill = "tomato", alpha = 0.7) +
      labs(title = paste("Boxplot of", varname),
           y = varname, x = "")
    print(p2)
    
  # Character / factor
  } else {
    tab <- sort(table(x), decreasing = TRUE)
    tab <- head(tab, max_levels)
    tmp <- data.frame(level = names(tab), n = as.integer(tab))
    
    p <- ggplot(tmp, aes(x = reorder(level, n), y = n)) +
      geom_col(fill = "darkseagreen4") +
      coord_flip() +
      labs(title = paste("Top categories of", varname),
           x = varname, y = "Count")
    print(p)
  }
}

# =========================
# Loop through variables for appendix plots
# =========================

for (v in screen_vars) {
  cat("\n\n###", v, "\n")
  plot_screen_variable(flights_clean, v)
}

# =========================
# Add empirical data checks to screening table
# =========================

pca_screen$exists_in_data <- pca_screen$variable %in% names(flights_clean)

pca_screen$n_missing <- sapply(pca_screen$variable, function(v) {
  if (v %in% names(flights_clean)) sum(is.na(flights_clean[[v]])) else NA
})

pca_screen$pct_missing <- sapply(pca_screen$variable, function(v) {
  if (v %in% names(flights_clean)) round(100 * mean(is.na(flights_clean[[v]])), 2) else NA
})

pca_screen$n_distinct <- sapply(pca_screen$variable, function(v) {
  if (v %in% names(flights_clean)) length(unique(flights_clean[[v]])) else NA
})

pca_screen$is_numeric_in_data <- sapply(pca_screen$variable, function(v) {
  if (v %in% names(flights_clean)) is.numeric(flights_clean[[v]]) else NA
})

pca_screen

# =========================
# Feature engineering for PCA-ready variables
# =========================

pca_df <- flights_clean
# Departure hour and arrival hour from scheduled times
pca_df$DEP_HOUR <- as.numeric(format(pca_df$CRS_DEP_TIME, "%H")) +
                   as.numeric(format(pca_df$CRS_DEP_TIME, "%M")) / 60

pca_df$ARR_HOUR <- as.numeric(format(pca_df$CRS_ARR_TIME, "%H")) +
                   as.numeric(format(pca_df$CRS_ARR_TIME, "%M")) / 60

# Cyclical encoding for departure hour and arrival hour
pca_df$DEP_HOUR_SIN <- sin(2 * pi * pca_df$DEP_HOUR / 24)
pca_df$DEP_HOUR_COS <- cos(2 * pi * pca_df$DEP_HOUR / 24)

pca_df$ARR_HOUR_SIN <- sin(2 * pi * pca_df$ARR_HOUR / 24)
pca_df$ARR_HOUR_COS <- cos(2 * pi * pca_df$ARR_HOUR / 24)

# Cyclical encoding for month
#pca_df$MONTH_SIN <- sin(2 * pi * pca_df$MONTH / 12)
#pca_df$MONTH_COS <- cos(2 * pi * pca_df$MONTH / 12)

# Ensure weekday is ordered properly
weekday_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

pca_df$WEEKDAY_F <- factor(pca_df$WEEKDAY, levels = weekday_levels, ordered = TRUE)
pca_df$WEEKDAY_NUM <- as.numeric(pca_df$WEEKDAY_F)

# Cyclical encoding for weekday
pca_df$WEEKDAY_SIN <- sin(2 * pi * pca_df$WEEKDAY_NUM / 7)
pca_df$WEEKDAY_COS <- cos(2 * pi * pca_df$WEEKDAY_NUM / 7)

# =========================
# First PCA-ready feature set
# =========================

pca_vars <- c(
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "DEP_HOUR_SIN", "DEP_HOUR_COS",
  "ARR_HOUR_SIN", "ARR_HOUR_COS",
  "MONTH_SIN", "MONTH_COS",
  "WEEKDAY_SIN", "WEEKDAY_COS"
)

pca_input <- pca_df[, pca_vars]

pca_input_complete <- na.omit(pca_input)

dim(pca_input_complete)
summary(pca_input_complete)


# =========================
# Histograms for engineered PCA features
# =========================

for (v in names(pca_input_complete)) {
  tmp <- data.frame(value = pca_input_complete[[v]])
  
  p <- ggplot(tmp, aes(x = value)) +
    geom_histogram(bins = 40, fill = "steelblue", color = "white") +
    labs(title = paste("Histogram of engineered feature:", v),
         x = v, y = "Count")
  print(p)
}

# =========================
# Run PCA
# =========================

pca_model <- prcomp(pca_input_complete, center = TRUE, scale. = TRUE)

summary(pca_model)

# =========================
# Explained variance
# =========================

pca_var_explained <- (pca_model$sdev)^2 / sum((pca_model$sdev)^2)
pca_var_table <- data.frame(
  PC = paste0("PC", seq_along(pca_var_explained)),
  Variance_Explained = pca_var_explained,
  Cumulative = cumsum(pca_var_explained)
)

pca_var_table
# Scree plot
ggplot(pca_var_table, aes(x = PC, y = Variance_Explained)) +
  geom_col(fill = "steelblue") +
  labs(title = "Scree plot", x = "Principal Component", y = "Proportion of variance explained")
# Cumulative explained variance
ggplot(pca_var_table, aes(x = seq_along(Cumulative), y = Cumulative)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq_along(pca_var_table$PC), labels = pca_var_table$PC) +
  labs(title = "Cumulative explained variance",
       x = "Principal Component", y = "Cumulative variance explained")

# =========================
# PCA loadings
# =========================

loadings <- pca_model$rotation
round(loadings, 3)
round(loadings[, 1:4], 3)

# =========================
# PCA scores
# =========================

scores <- as.data.frame(pca_model$x)
head(scores)
ggplot(scores, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.15) +
  labs(title = "PCA plot: observations in PC1-PC2 space")


# =========================
# Add target back only for interpretation
# =========================

pca_complete_rows <- complete.cases(pca_input)
scores$IS_DELAYED <- pca_df$IS_DELAYED[pca_complete_rows]
ggplot(scores, aes(x = PC1, y = PC2, color = IS_DELAYED)) +
  geom_point(alpha = 0.2) +
  labs(title = "PCA plot colored by delay status")

# =========================
# Top absolute loadings per component
# =========================

top_loadings <- function(pca_model, pc = 1, n = 5) {
  ld <- pca_model$rotation[, pc]
  out <- data.frame(
    variable = names(ld),
    loading = ld,
    abs_loading = abs(ld)
  )
  out <- out[order(-out$abs_loading), ]
  head(out, n)
}

top_loadings(pca_model, pc = 1, n = 10)
top_loadings(pca_model, pc = 2, n = 10)
top_loadings(pca_model, pc = 3, n = 10)
```
