---
title: "MOT143A Business Analytics: Group Report"
subtitle: "Airline Delay and Cancellation (2009–2018)"
author: "Hidde Segers and Lotte Feilzer"
date: "2026-03-04"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

# 1. Business context and analytics task

## 1.1 Business problem

**Context.** Airline operations for domestic flights in the US must plan
resources (gates, crew, customer service capacity).\
**Problem.** Delays and cancellations create high costs (crew overtime,
gate conflicts, rebooking) and reduce customer satisfaction.

**Business question.**\
*How can we predict the risk of flight delay/cancellation
**pre-departure** to optimize the use of operational resources?*

The model will not state the “truth”; it is a decision-support tool to
identify *high risk* flights and allocate resources proactively.

## 1.2 Analytics task

We consider two possible targets:

-   **Classification A (delay):** `ARR_DELAY >= 15` minutes\
    (only for flights that are not cancelled and not diverted).
-   **Classification B (cancellation):** `CANCELLED == 1`.

**Primary task (choose and write):** \_\_\_\
**Why:** \_\_\_ (business impact / class imbalance / interpretability)

## 1.3 Success criteria (metrics & business trade-offs)

-   **Model metrics:** confusion matrix + accuracy + precision/recall
    for the positive class.
-   **Business trade-off:**
    -   false negatives (miss real delays) → too few buffers, higher
        customer impact\
    -   false positives (predict delay but on-time) → over-allocation,
        higher costs

**Which error is more costly?** \_\_\_

# 2. Setup (same style as workbooks)

```{r setup}
# CRAN mirror
local({r <- getOption("repos")
r["CRAN"] <- "https://cran.rstudio.com/"
options(repos = r)})

packages <- c(
  # EDA + modelling (workbooks)
  "ggplot2", "Hmisc", "rsample", "gmodels", "C50", "randomForest",
  # dimension reduction & clustering workbook
  "factoextra", "lattice", "reshape2"
)

for (i in packages) {
  if (!require(i, character.only = TRUE)) {
    install.packages(i, dependencies = TRUE)
    library(i, character.only = TRUE)
  }
}

set.seed(46748717)

# folders
if (!dir.exists("output")) dir.create("output")
```

# 3. Data loading (use **one year**)

## 3.1 Data in `data/` folder

Put **one** year CSV in `data/`, for example: `data/2009.csv`.

> We do **not** push the big CSVs to GitHub (GitHub limit 100MB).\
> Each team member downloads the CSV locally and places it in `data/`.

```{r data_files}
data_dir <- "data"
YEAR <- 2009                 # TODO: change to 2017, 2018, etc.
csv_path <- file.path(data_dir, paste0(YEAR, ".csv"))

if (!file.exists(csv_path)) {
  stop(paste0(
    "File not found: ", csv_path, "\n",
    "Download it from Kaggle and place it in the data/ folder."
  ))
}

csv_path
```

## 3.2 Read data (lightweight development option)

These CSVs are huge. For development we can read only the first *N*
rows. Set `N_READ <- -1` to read the full file (may take long).

```{r read_data}
N_READ <- 300000   # TODO: set to -1 for full data

if (N_READ > 0) {
  flights <- read.csv(csv_path, stringsAsFactors = FALSE, nrows = N_READ)
} else {
  flights <- read.csv(csv_path, stringsAsFactors = FALSE)
}

dim(flights)
names(flights)[1:20]
```

# 4. Data exploration & preparation (Workbook: EDA/Prep/Visualization)

## 4.1 Reality check

```{r explore_basic}
str(flights)
summary(flights)
```

## 4.2 Missing values overview

```{r missingness}
na_count <- sapply(flights, function(x) sum(is.na(x)))
na_pct   <- round(100 * na_count / nrow(flights), 2)

na_table <- data.frame(
  variable = names(flights),
  na_count = as.integer(na_count),
  na_pct = na_pct
)

na_table <- na_table[order(-na_table$na_pct), ]
head(na_table, 15)
```

## 4.3 Minimal cleaning + datatypes

```{r clean_types}
# Drop unnamed columns if present
drop_cols <- grep("^Unnamed", names(flights), value = TRUE)
if (length(drop_cols) > 0) flights[drop_cols] <- NULL

# Parse FL_DATE if present
if ("FL_DATE" %in% names(flights)) flights$FL_DATE <- as.Date(flights$FL_DATE)

# Ensure integers
for (cname in c("CANCELLED","DIVERTED")) {
  if (cname %in% names(flights)) flights[[cname]] <- as.integer(flights[[cname]])
}
```

## 4.4 Feature engineering (pre-departure features)

```{r feature_engineering}
hhmm_to_minutes <- function(x) {
  x <- suppressWarnings(as.integer(x))
  hh <- x %/% 100
  mm <- x %% 100
  out <- hh*60 + mm
  out[is.na(x) | mm > 59 | hh < 0 | hh > 29] <- NA
  out
}

if ("FL_DATE" %in% names(flights)) {
  flights$MONTH <- as.integer(format(flights$FL_DATE, "%m"))
  flights$WEEKDAY <- weekdays(flights$FL_DATE)
}

if ("CRS_DEP_TIME" %in% names(flights)) {
  flights$CRS_DEP_MIN <- hhmm_to_minutes(flights$CRS_DEP_TIME)
  flights$CRS_DEP_HOUR <- floor(flights$CRS_DEP_MIN / 60)
}

if ("CRS_ARR_TIME" %in% names(flights)) {
  flights$CRS_ARR_MIN <- hhmm_to_minutes(flights$CRS_ARR_TIME)
  flights$CRS_ARR_HOUR <- floor(flights$CRS_ARR_MIN / 60)
}

if (all(c("ORIGIN","DEST") %in% names(flights))) {
  flights$ROUTE <- paste(flights$ORIGIN, flights$DEST, sep = "-")
}
```

## 4.5 Target definition (classification)

> **No leakage in modelling:** we will use only pre-departure features
> for the models. Targets come from outcome columns.

```{r targets}
# Delay target: ARR_DELAY >= 15 (only not-cancelled and not-diverted)
if (all(c("ARR_DELAY","CANCELLED","DIVERTED") %in% names(flights))) {
  flights$IS_DELAYED_15 <- ifelse(
    flights$CANCELLED == 0 & flights$DIVERTED == 0 & !is.na(flights$ARR_DELAY),
    as.integer(flights$ARR_DELAY >= 15),
    NA
  )
}

# Cancellation target
if ("CANCELLED" %in% names(flights)) {
  flights$IS_CANCELLED <- flights$CANCELLED
}

table(flights$IS_DELAYED_15, useNA="ifany")
table(flights$IS_CANCELLED, useNA="ifany")
```

# 5. Exploratory visualisation (ggplot2)

```{r eda_plots}
if (all(c("MONTH","IS_DELAYED_15") %in% names(flights))) {
  dfm <- aggregate(IS_DELAYED_15 ~ MONTH, flights, function(x) mean(x, na.rm = TRUE))
  ggplot(dfm, aes(x = MONTH, y = IS_DELAYED_15)) +
    geom_line() + geom_point() +
    scale_x_continuous(breaks = 1:12) +
    ylab("Delay rate (>=15 min)") + ggtitle("Delay rate per month")
}
```

```{r eda_airline_airport}
# Filter to flights where delay target is defined
if (all(c("IS_DELAYED_15","OP_CARRIER","ORIGIN") %in% names(flights))) {
  flights_clean <- flights[!is.na(flights$IS_DELAYED_15), ]

  airline_summary <- aggregate(IS_DELAYED_15 ~ OP_CARRIER, flights_clean,
                               function(x) mean(x, na.rm = TRUE))
  airline_summary <- airline_summary[order(-airline_summary$IS_DELAYED_15), ]

  ggplot(airline_summary[1:10, ],
         aes(x = reorder(OP_CARRIER, IS_DELAYED_15), y = IS_DELAYED_15)) +
    geom_col() +
    coord_flip() +
    labs(title = "Delay rate (>=15 min) per airline (top 10)",
         x = "Airline", y = "Delay rate")
}
```

**Interpretation (write):** - Seasonality?\
- Which carriers/airports stand out?\
- Class imbalance?

# 6. Dimension reduction & clustering (Workbook: Dimension reduction and clustering)

We cluster **airports** based on *pre-departure* aggregated profiles
(interpretability).

```{r airport_profiles}
needed <- c("ORIGIN","IS_DELAYED_15","IS_CANCELLED","DISTANCE","CRS_ELAPSED_TIME")
needed <- needed[needed %in% names(flights)]

if (all(c("ORIGIN","IS_DELAYED_15") %in% names(flights))) {
  base_df <- flights[!is.na(flights$IS_DELAYED_15), ]

  # aggregate per ORIGIN
  ap <- aggregate(
    cbind(IS_DELAYED_15, IS_CANCELLED, DISTANCE, CRS_ELAPSED_TIME) ~ ORIGIN,
    data = base_df,
    FUN = function(x) mean(x, na.rm = TRUE)
  )

  # add volume
  ap_n <- aggregate(IS_DELAYED_15 ~ ORIGIN, data = base_df, FUN = length)
  names(ap_n)[2] <- "n"
  ap <- merge(ap, ap_n, by = "ORIGIN")

  names(ap) <- c("ORIGIN","delay_rate","cancel_rate","avg_distance","avg_sched_time","n")

  # scale numeric columns
  ap_scaled <- scale(ap[, c("delay_rate","cancel_rate","avg_distance","avg_sched_time","n")])

  set.seed(1)
  km <- kmeans(ap_scaled, centers = 4, nstart = 20)
  ap$cluster <- factor(km$cluster)

  head(ap)
}
```

```{r cluster_viz}
if (exists("ap")) {
  ggplot(ap, aes(x = delay_rate, y = cancel_rate, color = cluster)) +
    geom_point(alpha = 0.8) +
    ggtitle("Airport clusters (delay & cancellation rates)") +
    xlab("Delay rate (>=15 min)") + ylab("Cancellation rate")
}
```

## 6.1 PCA

```{r pca}
if (exists("ap_scaled")) {
  pca_res <- prcomp(ap_scaled, center = TRUE, scale. = TRUE)
  summary(pca_res)

  factoextra::fviz_eig(pca_res, addlabels = TRUE)
  factoextra::fviz_pca_biplot(pca_res, repel = TRUE)
}
```

## 6.2 Correlation heatmap (replace GGally with reshape2 + ggplot2)

```{r corr_heatmap}
if (exists("ap")) {
  num_df <- ap[, c("delay_rate","cancel_rate","avg_distance","avg_sched_time","n")]
  cm <- cor(num_df, use = "complete.obs")
  cm_melt <- reshape2::melt(cm)

  ggplot(cm_melt, aes(x = Var1, y = Var2, fill = value)) +
    geom_tile() +
    labs(title = "Correlation matrix (airport profile features)",
         x = "", y = "", fill = "cor") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

**Interpretation (write):** - What do clusters represent?\
- Do PCA axes have meaning (loadings)?\
- Would you use a cluster label as a feature (yes/no and why)?

# 7. Modelling (Workbook: Classification)

We fit two models (C5.0 tree + Random Forest) using **pre-departure**
features only.

```{r modelling_data}
TARGET <- "IS_DELAYED_15"  # TODO: or "IS_CANCELLED"

candidate_features <- c(
  "OP_CARRIER","ORIGIN","DEST",
  "MONTH","WEEKDAY","CRS_DEP_HOUR",
  "CRS_ELAPSED_TIME","DISTANCE"
)
features <- intersect(candidate_features, names(flights))

stopifnot(TARGET %in% names(flights))
stopifnot(length(features) >= 3)

model_df <- flights[, c(TARGET, features)]
model_df <- model_df[complete.cases(model_df), ]

# collapse high-cardinality airports to top-N
topN_collapse <- function(x, N = 40) {
  tb <- sort(table(x), decreasing = TRUE)
  keep <- names(tb)[1:min(N, length(tb))]
  out <- ifelse(x %in% keep, x, "OTHER")
  factor(out)
}

if ("ORIGIN" %in% names(model_df)) model_df$ORIGIN <- topN_collapse(model_df$ORIGIN, 40)
if ("DEST" %in% names(model_df)) model_df$DEST <- topN_collapse(model_df$DEST, 40)
if ("OP_CARRIER" %in% names(model_df)) model_df$OP_CARRIER <- factor(model_df$OP_CARRIER)
if ("WEEKDAY" %in% names(model_df)) model_df$WEEKDAY <- factor(model_df$WEEKDAY)

split <- rsample::initial_split(model_df, prop = 0.7)
train <- rsample::training(split)
test  <- rsample::testing(split)

dim(train); dim(test)
table(train[[TARGET]])
```

## 7.1 Baseline

```{r baseline}
y_true <- test[[TARGET]]
y_pred0 <- rep(0, length(y_true))

gmodels::CrossTable(y_true, y_pred0,
                    prop.chisq = FALSE, prop.t = FALSE, prop.r = FALSE,
                    dnn = c("Actual", "Predicted (always 0)"))
```

## 7.2 Model 1 — C5.0 decision tree

```{r c50}
form <- as.formula(paste(TARGET, "~", paste(features, collapse = " + ")))

train_c50 <- train
test_c50 <- test
train_c50[[TARGET]] <- factor(train_c50[[TARGET]])
test_c50[[TARGET]]  <- factor(test_c50[[TARGET]])

tree <- C50::C5.0(form, data = train_c50)
tree

pred_tree <- predict(tree, newdata = test_c50)
gmodels::CrossTable(test_c50[[TARGET]], pred_tree,
                    prop.chisq = FALSE, prop.t = FALSE, prop.r = FALSE,
                    dnn = c("Actual", "Predicted (C5.0)"))
```

## 7.3 Model 2 — Random Forest

```{r rf}
train_rf <- train
test_rf <- test
train_rf[[TARGET]] <- factor(train_rf[[TARGET]])
test_rf[[TARGET]]  <- factor(test_rf[[TARGET]])

rf <- randomForest::randomForest(form, data = train_rf, ntree = 200)
rf

pred_rf <- predict(rf, newdata = test_rf)
gmodels::CrossTable(test_rf[[TARGET]], pred_rf,
                    prop.chisq = FALSE, prop.t = FALSE, prop.r = FALSE,
                    dnn = c("Actual", "Predicted (RF)"))

randomForest::varImpPlot(rf)
```

## 7.4 Comparison (write)

Compare: - performance (confusion matrices) - interpretability (tree
rules vs RF importance) - robustness / overfitting risk - operational
usability (how would you use scores in planning?)

**We recommend model \_\_\_ because \_\_\_ .**

# 8. Conclusions & recommendations

## 8.1 Operational recommendations (write)

Examples: - “High-risk flight list” per day/airport → allocate buffers
(gate agents/crew) - proactive passenger communication above a risk
threshold - airport/carrier monitoring dashboard (month/weekday/hour)

## 8.2 Limitations & next steps (write)

Limitations: - no weather / ATC constraints - concept drift across years
(policies, demand, post-COVID changes) - class imbalance (especially
cancellations)

Next steps (conceptual, workbooks 6–8): - when would representation
learning add value and what data is needed? - why text analytics is not
suitable here (no text fields) - how RL could apply if reformulated as
sequential staffing decisions

# 9. Reproducibility

```{r sessioninfo}
sessionInfo()
```
