---
title: "MOT143A: Business Analytics group 20"
author: "Nora Donders, Stijn, Brian, Hidde Segers and Lotte Feilzer"
date: "2026-03-04"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    highlight: tango
editor_options:
  markdown:
    wrap: 72
---

# 1. Business context and analytics task

## 1.1 Business problem

**Context:** Airlines operating domestic flights in the United States
need to manage gates, crews, ground handling staff and customer service
capacity. These decisions often have to be made quickly and without
knowing the whole of the situation.

Flight delays and cancellations are expensive for the airlines and
stressful for the passengers. Airlines have to deal with numerous
problems including crew overtime, gate conflicts, aircraft
repositioning, compensation, rebooking work and reputational damage.

**Problem:** A single delay usually has repercussions for other flights
as well. Flights are using the same planes and are therefor intertwined
with their schedule. If one flight is delayed, it creates more delays
later in the day raising operational costs and reducing customer
satisfaction.

The most challenging part of operations is that many decisions need to
be made before a flight departs. At that point, not all setbacks are
known, and therefore neither is the final outcome of the departure. If
an airline were able to identify flights with a high delay risk far
before departure, an appropiate solution can be thought beforehand.
Preparing standby crew, adjusting gate plans, or increasing customer
service capacity are examples of operations that the airline could
impose beforehand so further delays are avoided or reduced. Delays would
still occur, but the impact of a single delay on further delays, costs,
and customer dissatisfaction would be reduced.

**Business question:** How can we estimate the probability that a
scheduled flight will be delayed (more than 15 min) before departure?

The purpose of this analysis is not to predict the future perfectly. The
goal is to be able to make a risk signal that aids operations teams with
the division of focus attention and resources on flights with the
highest expected delay risk. From an organizational perspective, the
main users would be operations control centres, network planners, and
day-of-operations managers. In that setting, the model is best seen as a
decision-support tool rather than an automatic decision-maker.

## 1.2 Analytics task

This analysis takes the perspective of airline operations control and
airport station management. The aim is to identify higher-risk flights
early enough to support staffing, gate planning, and earlier passenger
communication. Because the decision moment is before departure, we only
want to use information that would realistically be known at that time.

This leads naturally to a predictive risk-scoring task based on
historical U.S. domestic flight data. Before choosing the final task, we
considered three possible targets:

-   **Classification I (delay):**

    **Target:** `IS_DELAYED_15 = 1` if `ARR_DELAY >= 15`, otherwise 0.

    This asks whether a completed flight arrives with a meaningful
    delay.

-   **Classification II (cancellation):**

    **Target:** `CANCELLED == 1`

    This asks whether a flight will be cancelled.

-   **Regression:**

    **Target:** `ARR_DELAY`

    This asks how many minutes of arrival delay we should expect.

This asks how many minutes of arrival delay we should expect.

We choose Classification I as the main task. That choice follows from
the business question: operations teams mainly need an early warning
about whether a flight is likely to become meaningfully delayed, because
that is directly useful for reallocating attention and resources. This
is also an assumption we make explicit in the report: a clear delayed vs
on-time signal is more actionable here than a raw minute prediction.

We do not choose cancellation as the main task because cancellations are
much rarer and therefore harder to model well. We also do not choose
regression as the main task because predicting the exact number of delay
minutes is harder to explain and less directly useful for the first
operational decision. So, the reasoning pipeline is that the business
problem is pre-departure resource allocation, which requires a
pre-departure risk signal, and therefore the main analytics task becomes
the binary classification of meaningful arrival delay.

## 1.3 Success criteria

We define success from two angles. First, the model should answer the
analytics question in a useable way. Second, the model should support
the business problem in practice.

From an operational point of view, missing a real delay is usually worse
than raising a false alarm. If the model says a flight will be on time
but it is actually delayed, the airline is unprepared. This can quickly
lead to crew problems, gate conflicts, and unhappy passengers.

If the model predicts a delay that does not happen, it may lead to some
extra costs because resources were allocated unnecessarily. However,
this is less harmful than being unprepared for a real delay.

For this reason, we focus on high recall for delayed flights. We want
the model to catch as many real delays as possible, even if that means
accepting some false alarms. At the same time, we monitor precision to
make sure we do not allocate too many unnecessary resources.

# 2. Setup

This section loads the packages used throughout the notebook so the
report stays reproducible for all readers.

```{r setup}
# CRAN mirror
local({r <- getOption("repos")
r["CRAN"] <- "https://cran.rstudio.com/"
options(repos = r)})

# These are the main packages we use throughout the workbook
packages <- c("ggplot2", "Hmisc", "rsample", "gmodels", "C50", "randomForest",
  "factoextra", "lattice", "reshape2", "pROC"
)

# Install missing packages and load everything
for (i in packages) {
  if (!require(i, character.only = TRUE)) {
    install.packages(i, dependencies = TRUE)
    library(i, character.only = TRUE)
  }
}

set.seed(46748717)
```

# 3. Data loading

## 3.1 Data in `data/` folder

Here, we read the flight data. The year can be changed through `YEAR`,
because the files are large. For the analysis in this workbook we use
the data of 2018.

```{r data_files}
data_dir <- "data"
YEAR <- 2018
csv_path <- file.path(data_dir, paste0(YEAR, ".csv"))
flights <- read.csv(csv_path, stringsAsFactors = FALSE)
```

## 3.2 Read data

We first inspect the dimensions and column names to understand what kind
of data we are dealing with and what the scope and units of the analysis
are.

```{r read_data}
dim(flights)
names(flights)
```

The dataset contains:

-   7,213,446 observations

-   28 variables

Each observation represents one commercial flight in 2018 in the United
States of America. The variables represent information about these
flights. Here, we print a concise summary.

```{r summary }
str(flights)
```

From this we can see the variable types and the first values. By
analysing we can write down the variable descriptions.

| Variable | Description | Variable type |
|----|----|----|
| FL_DATE | Scheduled date of the flight | character (date string `YYYY-MM-DD` format) |
| OP_CARRIER | Airline operating the flight | character |
| OP_CARRIER_FL_NUM | Identifier number of the flight operated by the carrier | integer |
| ORIGIN | Airport from which the flight departs | character |
| DEST | Airport at which the flight arrives | character |
| CRS_DEP_TIME | Scheduled departure time | integer (`HHMM` format) |
| DEP_TIME | Actual departure time | numeric (`HHMM` format) |
| DEP_DELAY | Difference in minutes between actual and scheduled departure time | numeric (minutes) |
| TAXI_OUT | Time from leaving the departure gate until takeoff | numeric (minutes) |
| WHEELS_OFF | Actual time at which the aircraft leaves the ground | numeric (`HHMM` format) |
| WHEELS_ON | Actual time at which the aircraft touches down | numeric (`HHMM` format) |
| TAXI_IN | Time from landing until arrival at the gate | numeric (minutes) |
| CRS_ARR_TIME | Scheduled arrival time | integer (`HHMM` format) |
| ARR_TIME | Actual arrival time | numeric (`HHMM` format) |
| ARR_DELAY | Difference in minutes between actual and scheduled arrival time | numeric (minutes) |
| CANCELLED | Indicator showing whether the flight was cancelled | numeric (binary) |
| CANCELLATION_CODE | Reason code explaining why the flight was cancelled | character |
| DIVERTED | Indicator showing whether the flight was diverted to another airport | numeric (binary) |
| CRS_ELASPED_TIME | Scheduled total travel time from departure to arrival | numeric (minutes) |
| ACTUAL_ELAPSED_TIME | Actual total travel time from departure to arrival | numeric (minutes) |
| AIR_TIME | Total time the aircraft is in the air | numeric (minutes) |
| DISTANCE | Flight distance between origin and destination airports | numeric (miles) |
| CARRIER_DELAY | Delay attributed to the airline carrier | numeric (minutes) |
| WEATHER_DELAY | Delay attributed to weather conditions | numeric (minutes) |
| NAS_DELAY | Delay attributed to the National Airspace System | numeric (minutes) |
| SECURITY_DELAY | Delay attributed to security-related issues | numeric (minutes) |
| LATE_AIRCRAFT_DELAY | Delay caused by the late arrival of the incoming aircraft | numeric (minutes) |
| Unnamed..27 | Unnamed column | logical |

Now we group the variables into a few practical categories so we can
refer back to them more easily later in the report.

| Category | Variables |
|----|----|
| Flight identifiers | FL_DATE, OP_CARRIER, OP_CARRIER_FL_NUM |
| Route information | ORIGIN, DEST |
| Scheduled times | CRS_DEP_TIME, CRS_ARR_TIME |
| Actual flight performance | DEP_TIME, ARR_TIME, DEP_DELAY, ARR_DELAY |
| Operational metrics | TAXI_OUT, TAXI_IN, AIR_TIME |
| Delay causes | CARRIER_DELAY, WEATHER_DELAY, NAS_DELAY, SECURITY_DELAY, LATE_AIRCRAFT_DELAY |
| Flight status indicators | CANCELLED, DIVERTED |

# 4. Data cleaning

## 4.1 Data summary

We start with a basic structural check of the data. This is important
because before we decide how to clean or model anything, we first need
to understand how the variables are stored.

At this moment a few things stand out. `FL_DATE` is stored as text, so
we will later convert it to a real date. Several scheduled and actual
time variables are stored as integers in `HHMM` format rather than as
usable time values, which makes direct time calculations easier.
Variables such as `TAXI_OUT` and `DEP_DELAY` are already stored as
minutes, which is suitable for analysis. We also see that `CANCELLED`
and `DIVERTED` use bineary `0` and `1` coding rather than logical
`TRUE/FALSE`, which is fine as long as we handle it consistently.

These observations directly lead to the next steps. First we inspect
structure and missingness. Then we decide which datatype changes and
cleaning steps are actually needed for the pre-departure analysis.

To align this notebook more closely with the workbook on exploration and
preparation, we explicitly separate three questions:

-   What is the type and scale of each variable?

-   Which variables show data quality issues or unusual structure?

-   Which of those issues require action before modelling?

The next sections answer these questions step by step. We focus on
variables that are most relevant for a pre-departure delay analysis,
rather than producing every possible plot for all 28 columns.

```{r selected_numeric_diagnostics}
numeric_vars <- c("DISTANCE", "CRS_ELAPSED_TIME", "DEP_DELAY", "ARR_DELAY")
numeric_vars <- intersect(numeric_vars, names(flights))

numeric_diagnostics <- data.frame(
  variable = numeric_vars,
  mean = sapply(flights[numeric_vars], function(x) round(mean(x, na.rm = TRUE), 2)),
  median = sapply(flights[numeric_vars], function(x) round(median(x, na.rm = TRUE), 2)),
  sd = sapply(flights[numeric_vars], function(x) round(sd(x, na.rm = TRUE), 2)),
  min = sapply(flights[numeric_vars], function(x) round(min(x, na.rm = TRUE), 2)),
  max = sapply(flights[numeric_vars], function(x) round(max(x, na.rm = TRUE), 2))
)

numeric_diagnostics
```

This table functions as our first univariate numeric exploration step.
In workbook terms, we compare central tendency and spread before we look
at plots. The main thing to watch for is a large gap between mean and
median, or an extreme maximum relative to the bulk of the data, because
that signals skewness or potentially influential observations.

```{r selected_numeric_plots}
set.seed(46748717)
idx_diag <- sample(seq_len(nrow(flights)), min(200000, nrow(flights)))
flights_diag <- flights[idx_diag, ]

par(mfrow = c(2, 2))
hist(flights_diag$DISTANCE, breaks = 40, main = "Distance", xlab = "Miles", col = "grey80")
boxplot(flights_diag$DISTANCE, horizontal = TRUE, main = "Distance boxplot", col = "grey80")
hist(flights_diag$ARR_DELAY, breaks = 40, main = "Arrival delay", xlab = "Minutes", col = "grey80")
boxplot(flights_diag$ARR_DELAY, horizontal = TRUE, main = "Arrival delay boxplot", col = "grey80")
par(mfrow = c(1, 1))
```

We keep this visual check selective. `DISTANCE` is a core pre-departure
predictor and should be inspected as a regular numeric feature.
`ARR_DELAY` is not a predictor in the pre-departure setting, but it is
the outcome we eventually simplify into a binary target. Inspecting its
raw distribution helps justify why a classification target is more
practical here than modelling raw delay minutes directly: the delay
distribution is highly skewed and dominated by a long right tail.

## 4.2 Missing values overview and discussion

We now inspect missing values. The point is not only to count `NA`s, but
also to understand why they occur. In operational data, missingness can
be structural rather than accidental, so the next cleaning step should
depend on the reason for the missing value and not just on the count.

```{r missingness}
na_count <- sapply(flights, function(x) sum(is.na(x)))
na_pct   <- round(100 * na_count / nrow(flights), 2)

na_table <- data.frame(
  na_count = as.integer(na_count),
  na_pct = na_pct
)

na_table <- na_table[order(-na_table$na_pct), ]
na_table
```

The table already suggests three different missingness patterns. First,
`Unnamed..27` contains only missing values, so it does not carry useful
information and can be dropped. Second, several operational variables
such as `DEP_TIME`, `ARR_TIME`, `AIR_TIME`, `WHEELS_ON`, and
`WHEELS_OFF` are missing in roughly the same small share of rows. That
often indicates a shared operational reason rather than random data
entry errors. Third, the delay-cause columns also seem to be jointly
missing or jointly present.

That leads to the next check. Before removing anything, we verify
whether the delay-cause columns behave as structurally missing fields:
they should only be filled when a substantial realised delay exists. If
that is true, we should not treat those `NA`s as a classic data-quality
problem.

```{r delay-na-same}
cause_cols <- intersect(c("CARRIER_DELAY","WEATHER_DELAY",
                    "NAS_DELAY","SECURITY_DELAY","LATE_AIRCRAFT_DELAY"), names(flights))
delay15 <- !is.na(flights$ARR_DELAY) & flights$ARR_DELAY >= 15
cause_present <- !is.na(flights[[cause_cols[1]]])

miss <- is.na(flights[, cause_cols, drop=FALSE])
same_na_rows <- all(apply(miss, 2, function(x) identical(which(x), which(miss[,1]))))

cat("All the delay causes with NA values are for the same row, is", same_na_rows, "\n")
cat("Causes are present but there was no 15 min delay:", sum(cause_present & !delay15), "\n")
cat("More than 15 min delay but cause is NA value:", sum(delay15 & !cause_present))
```

Next we check whether a delayed flight can have multiple recorded delay
causes. This is useful context for interpretation later on, because it
tells us whether delay causes behave like a single-label explanation or
whether multiple operational problems can happen on the same flight.

```{r cause-compact}
k <- rowSums(flights[cause_present, cause_cols, drop=FALSE] > 0)
tab <- table(k)

multi_n <- sum(k > 1)
total_n <- length(k)
multi_pct <- round(100 * multi_n / total_n, 2)

cat("Of the flights with delay causes,", multi_pct, "% have multiple causes for delay.")
```

We now know that the delay-cause columns are structurally complete for
their purpose. The next step is to check the missing values in realised
operational fields such as `ARR_DELAY` and `ARR_TIME`. Here the working
assumption is that many of those missing values belong to cancelled or
diverted flights. If that is true, the missingness is again explained by
the flight outcome rather than by poor data quality.

```{r non_cancelled_missingness_check}
flights_wo_cancelled <- which(flights$CANCELLED == 0 & flights$DIVERTED == 0)
flights_wo_cancelled_df <- flights[flights_wo_cancelled, ]

na_count_wo_cd <- sapply(flights_wo_cancelled_df, function(x) sum(is.na(x)))
na_pct_wo_cd   <- round(100 * na_count_wo_cd / nrow(flights_wo_cancelled_df), 2)

na_table_wo_cd <- data.frame(
  variable = names(flights_wo_cancelled_df),
  na_count_wo_cd = as.integer(na_count_wo_cd),
  na_pct_wo_cd = na_pct_wo_cd
)

na_table_wo_cd <- na_table_wo_cd[order(-na_table_wo_cd$na_pct_wo_cd), ]
na_table_wo_cd
```

This table shows that cancelled or diverted flights explain almost all
of the missingness in realised performance fields. After removing those
rows, the only important remaining gaps are in `ARR_DELAY` and
`DEP_DELAY`. That matters because `ARR_DELAY` is used to define our
target, so we need a clear rule for these remaining cases before moving
on.

```{r check-missing-mostly-delays}
dep_delays_missing <- which(is.na(flights_wo_cancelled_df$DEP_DELAY)) 
arr_delays_missing <- which(is.na(flights_wo_cancelled_df$ARR_DELAY))
missing_delays <- flights_wo_cancelled_df[c(dep_delays_missing, arr_delays_missing), ]
head(missing_delays)
```

We now inspect those remaining missing delay values directly. Our
working assumption is that in this dataset an `NA` delay sometimes means
the scheduled and realised times are exactly equal, so the delay should
effectively be `0`. If that is true, these are repairable missing values
rather than rows we need to drop. We therefore test that assumption
explicitly before imputing anything.

```{r fill_zero_delays}
dep_na <- !is.na(missing_delays$CRS_DEP_TIME) & !is.na(missing_delays$DEP_TIME) &
                 is.na(missing_delays$DEP_DELAY) &
                 missing_delays$CRS_DEP_TIME == missing_delays$DEP_TIME

arr_na <- !is.na(missing_delays$CRS_ARR_TIME) & !is.na(missing_delays$ARR_TIME) &
                 is.na(missing_delays$ARR_DELAY) &
                 missing_delays$CRS_ARR_TIME == missing_delays$ARR_TIME

cat("DEP_DELAY NA but not CRS_DEP_TIME==DEP_TIME:", sum(dep_na), "\n")
cat("ARR_DELAY NA but not CRS_ARR_TIME==ARR_TIME:", sum(arr_na))
```

At this point we have explained the main missing-value patterns, so we
can move into cleaning with a clearer rationale.

The workbook logic is visible here: we do not immediately delete all
rows with missing values, because in operational data missingness is
often informative. We first distinguish between:

-   structurally missing values, for example delay causes that are only
    recorded after substantial delays;

-   outcome-related missing values, for example flights that were
    cancelled or diverted and therefore have no realised arrival
    metrics;

-   repairable missing values, for example cases where scheduled time
    and realised time are equal and the delay can logically be set to 0.

This distinction matters because it prevents us from throwing away valid
flights too early and helps us explain each cleaning choice.

## 4.3 Cleaning and datatypes

Before cleaning, we do a few final sanity checks. The purpose here is
simple: if values are clearly impossible, we need to deal with that
before we start deriving new variables from them.

```{r sanity_checks}
cat("There are", sum(flights$DISTANCE <= 0), "negative distances.", "\n")
cat("There are", sum(flights$AIR_TIME <= 0, na.rm = TRUE) , "negative airtimes.", "\n")
print("Departure delay summary:")
summary(flights$DEP_DELAY) 
print("Arrival delay summary:")
summary(flights$ARR_DELAY) 
```

We create a separate `flights_clean` object so the raw import remains
available for reference.

```{r create_flights_clean}
flights_clean <- flights
```

The first cleaning step follows directly from the structural inspection
above. We convert `FL_DATE` to a real date and drop the unnamed column
that contains no information.

```{r clean_types}
drop_cols <- grep("^Unnamed", names(flights_clean), value = TRUE)
flights_clean[drop_cols] <- NULL

flights_clean$FL_DATE <- as.Date(flights_clean$FL_DATE)

head(flights_clean, 4)
```

Next we deal with the time variables. Because they are stored as `HHMM`
integers, we convert them to proper datetime values so that time-based
comparisons and derived hour variables become more reliable.

```{r datetime_fix_overwrite}
hhmm_to_hhmm_str <- function(x) {
  x <- as.integer(x)
  hh <- x %/% 100
  mm <- x %% 100
  out <- sprintf("%02d:%02d", hh, mm)
  out[is.na(x) | mm > 59 | hh < 0 | hh > 29] <- NA
  out
}

time_cols <- intersect(c("WHEELS_OFF","WHEELS_ON","DEP_TIME","ARR_TIME","CRS_DEP_TIME","CRS_ARR_TIME"), names(flights_clean))

# Replace columns to POSIXct
for (cn in time_cols) {
  tmp_str <- hhmm_to_hhmm_str(flights_clean[[cn]])
  flights_clean[[cn]] <- as.POSIXct(
    paste(as.character(flights_clean$FL_DATE), tmp_str),
    format = "%Y-%m-%d %H:%M",
    tz = "UTC"
  )
}

# Overnight: is ARR_TIME < DEP_TIME + one day
if (all(c("DEP_TIME","ARR_TIME") %in% names(flights_clean))) {
  overnight <- !is.na(flights_clean$DEP_TIME) &
               !is.na(flights_clean$ARR_TIME) &
               (flights_clean$ARR_TIME < flights_clean$DEP_TIME)

  arrival_side <- intersect(c("ARR_TIME","WHEELS_ON","CRS_ARR_TIME"), time_cols)
  for (cn in arrival_side) {
    flights_clean[[cn]][overnight] <- flights_clean[[cn]][overnight] + 24*3600
  }
}

head(flights_clean[, time_cols, drop=FALSE], 5)
```

We now apply the repair rule discussed above: when the scheduled and
actual times are identical and the delay is missing, we set the delay to
`0`.

```{r impute_delays_posixct}
idx_dep0 <- which(
  is.na(flights_clean$DEP_DELAY) &
  !is.na(flights_clean$DEP_TIME) & !is.na(flights_clean$CRS_DEP_TIME) &
  flights_clean$DEP_TIME == flights_clean$CRS_DEP_TIME
)

idx_arr0 <- which(
  is.na(flights_clean$ARR_DELAY) &
  !is.na(flights_clean$ARR_TIME) & !is.na(flights_clean$CRS_ARR_TIME) &
  flights_clean$ARR_TIME == flights_clean$CRS_ARR_TIME
)

# Fill delay values with 0 only in the specific cases supported by the
# equality check above.
flights_clean$DEP_DELAY[idx_dep0] <- 0
flights_clean$ARR_DELAY[idx_arr0] <- 0

# check
head(flights_clean[idx_dep0, c("CRS_DEP_TIME","DEP_TIME","DEP_DELAY")])
head(flights_clean[idx_arr0, c("CRS_ARR_TIME","ARR_TIME","ARR_DELAY")])
```

With the datatype fixes and repairable missing values handled, we can
now define the modelling sample more clearly. We remove cancelled and
diverted flights because the main task is to compare completed flights
on meaningful arrival delay.

```{r filter_cancelled_diverted}
keep <- (is.na(flights_clean$CANCELLED) | flights_clean$CANCELLED == 0) &
        (is.na(flights_clean$DIVERTED)  | flights_clean$DIVERTED  == 0)

flights_clean <- flights_clean[keep, , drop = FALSE]

cat("Removed:", sum(!keep), "\n")
cat("Rows after:", nrow(flights_clean))
```

Once the analysis sample is fixed, the next step is feature creation. We
derive a binary delay indicator for the main target, and we add calendar
and scheduled-hour variables that are still known before departure and
therefore fit the business setting.

```{r create_analysis_features}
flights_clean$IS_DELAYED <- ifelse(flights_clean$ARR_DELAY >= 15, TRUE, FALSE)
flights_clean$WEEKDAY <- weekdays(flights_clean$FL_DATE)
flights_clean$MONTH <- as.integer(format(flights_clean$FL_DATE, "%m"))
flights_clean$SCHED_DEP_HOUR <- as.numeric(format(flights_clean$CRS_DEP_TIME, "%H")) +
  as.numeric(format(flights_clean$CRS_DEP_TIME, "%M")) / 60
flights_clean$SCHED_ARR_HOUR <- as.numeric(format(flights_clean$CRS_ARR_TIME, "%H")) +
  as.numeric(format(flights_clean$CRS_ARR_TIME, "%M")) / 60
```

The target is now based on `ARR_DELAY >= 15`. This is the cleanest
definition for the business question because it captures whether the
flight ultimately arrived with a meaningful disruption. We remove
cancelled and diverted flights before defining this target, so the class
labels are only assigned to completed flights.

```{r cleaning_summary}
cleaning_summary <- data.frame(
  step = c(
    "Original rows",
    "After removing cancelled/diverted flights",
    "Rows with defined delay target"
  ),
  n_rows = c(
    nrow(flights),
    nrow(flights_clean),
    sum(!is.na(flights_clean$IS_DELAYED))
  )
)

cleaning_summary$share_of_original <- round(
  100 * cleaning_summary$n_rows / cleaning_summary$n_rows[1], 2
)

cleaning_summary
```

The table above makes the cleaning impact transparent. This is useful in
the workbook context because it shows that our later analysis is based
on a clearly defined subset of operationally comparable flights, rather
than on an unclear mix of completed, cancelled and diverted flights.

## 4.4 Exploration log and preparation decisions

The workbook stresses that exploration should lead to explicit
preparation choices. We therefore summarize our main findings in a short
table.

| Variable or group | Observation | Action |
|----|----|----|
| Unnamed columns | Unnamed column contains only missing values | Drop |
| Cancelled/diverted flights | Operational outcome fields are often missing for these flights | Exclude from completed-flight modelling sample |
| Departure/arrival times | Stored as HHMM integers rather than usable times | Convert to POSIXct and derive scheduled hour features |
| Delay fields | Some zero-delay cases are stored as `NA` when scheduled and actual times match | Impute those cases with `0` when the equality check supports it |
| Carrier/origin categories | High-cardinality categories with uneven group sizes | Use volume filters in visual ranking and factor encoding in models |
| Distance and elapsed time | Numeric variables look plausible but can contain influential extremes | Keep, but inspect visually and avoid unnecessary transformation unless the analysis needs it |

```{r duplicate_check}
dup_key <- c("FL_DATE", "OP_CARRIER", "OP_CARRIER_FL_NUM", "ORIGIN", "DEST", "CRS_DEP_TIME")
dup_key <- intersect(dup_key, names(flights_clean))

duplicate_rows <- duplicated(flights_clean[, dup_key])

data.frame(
  key_columns_checked = paste(dup_key, collapse = ", "),
  n_duplicate_rows = sum(duplicate_rows),
  pct_duplicate_rows = round(100 * mean(duplicate_rows), 4)
)
```

This duplicate check follows the workbook logic that cleaning is not
only about missing values. In this data set, the practical question is
whether two rows appear to describe the same scheduled flight. If this
count is small or zero, that supports the assumption that the unit of
analysis is one row per flight. If it is not, the duplicates need manual
inspection before modelling.

Some workbook preparation steps are not needed here. Data integration is
not relevant yet because we currently work with one main flight file.
Likewise, we do not apply global standardization at this stage because
our current models are tree-based and therefore do not require all
predictors to be on the same scale. Where scaling does matter, such as
in PCA, we explicitly do it there.

# 5. Exploratory visualisation

We start the visual exploration with the base rate: the share of
completed flights that arrive 15 minutes late or more. This is the
natural first plot, because it tells us how common the target event is
before we interpret any subgroup patterns.

```{r mean_delay_rate}
delay_rate <- mean(flights_clean$IS_DELAYED)
delay_rate 
```

We first show a simple bar chart with delayed versus on-time flights.
This gives a direct visual check of the class balance.

```{r class_distribution_plot}
df_target <- data.frame(
  class = factor(
    ifelse(flights_clean$IS_DELAYED, "Delayed (15 min)", "On time"),
    levels = c("On time", "Delayed (15 min)")
  )
)

ggplot(df_target, aes(x = class)) +
  geom_bar() +
  labs(title = "Class distribution: delayed vs on-time", x = "", y = "Number of flights") +
  coord_flip()
```

This class balance matters for the rest of the notebook. A delay rate of
roughly 18% means the data is not perfectly balanced, but the delayed
class is still common enough to model without the extreme rarity issues
that would occur for cancellations. At the same time, the graph makes
clear that "On time" is still the dominant class. That matters when we
evaluate models later, because a model can look acceptable on accuracy
while still doing a poor job on the delayed flights we care about most.

```{r numeric_vs_target_boxplots}
set.seed(46748717)
idx_box <- sample(seq_len(nrow(flights_clean)), min(150000, nrow(flights_clean)))
flights_box <- flights_clean[idx_box, ]

par(mfrow = c(1, 2))
boxplot(SCHED_DEP_HOUR ~ IS_DELAYED,
        data = flights_box,
        outline = FALSE,
        names = c("On time", "Delayed"),
        main = "Scheduled departure hour by delay status",
        ylab = "Scheduled departure hour",
        col = c("grey80", "tomato"))

boxplot(CRS_ELAPSED_TIME ~ IS_DELAYED,
        data = flights_box,
        outline = FALSE,
        names = c("On time", "Delayed"),
        main = "Scheduled elapsed time by delay status",
        ylab = "Scheduled elapsed time (minutes)",
        col = c("grey80", "tomato"))
par(mfrow = c(1, 1))
```

This figure adds the workbook perspective of comparing a numerical
variable against a categorical one. It helps us see whether delayed and
on-time flights differ in their planned characteristics before we start
modelling. We suppress plotted outlier points here because the sample is
very large; otherwise the boxes become unreadable. The distribution
itself remains visible through the median and spread.

What we mainly want to read from these boxplots is whether the center
and spread differ between the two groups. If the delayed flights tend to
leave later in the day, or if their planned elapsed time looks
different, that suggests these variables contain predictive signal even
before departure. The goal here is not to prove causality, but to check
whether the delayed and on-time groups already look different in simple,
planned characteristics.

These two plots show delay rate per month and the number of flights per
month. The main thing we see is that delays go up in the summer months.
There are also a bit more flights in that period, but the rise in delay
rate is larger than the rise in volume, so seasonality seems to matter.
That is an important distinction. If the delay pattern were only driven
by more flights, the rate would stay much flatter. Because the rate
itself changes, month looks like a useful explanatory variable rather
than just a background descriptor.

```{r monthly_delay_patterns}
df_month_rate <- aggregate(IS_DELAYED ~ MONTH, flights_clean, function(x) mean(x, na.rm = TRUE))
df_month_n    <- aggregate(IS_DELAYED ~ MONTH, flights_clean, length)
names(df_month_n)[2] <- "n"

df_month <- merge(df_month_rate, df_month_n, by = "MONTH")
names(df_month)[2] <- "delay_rate"

pct_label <- function(x) paste0(round(100*x, 1), "%")

ggplot(df_month, aes(x = MONTH, y = delay_rate)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(labels = pct_label) +
  coord_cartesian(ylim = c(0, 0.30)) +
  labs(title = "Delay rate per month", x = "Month", y = "Delay rate")

ggplot(df_month, aes(x = MONTH, y = n)) +
  geom_col() +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Amount of flights per month", x = "Month", y = "#")
```

The weekday pattern is fairly stable. Most days sit close together, with
slightly higher delay rates on Thursday and Friday and a lower rate on
Saturday. So weekday does matter a bit, but not as strongly as some
other variables. That suggests it makes sense to keep weekday in the
analysis, but not to expect it to explain delays on its own. In other
words, weekday seems more like a supporting variable than one of the
strongest drivers in the notebook.

```{r weekday_delay_rates}
weekday_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

df_wday_rate <- aggregate(IS_DELAYED ~ WEEKDAY, flights_clean, mean)
names(df_wday_rate)[2] <- "delay_rate"
df_wday_rate$WEEKDAY <- factor(df_wday_rate$WEEKDAY, levels = weekday_levels)
df_wday_rate <- df_wday_rate[order(df_wday_rate$WEEKDAY), ]

ggplot(df_wday_rate, aes(x = WEEKDAY, y = delay_rate)) +
  geom_col() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Delay rate per weekday", x = "Day of the week", y = "Delay rate")

```

```{r weekday_hour_heatmap}
df_hw <- flights_clean[, c("WEEKDAY", "SCHED_DEP_HOUR", "IS_DELAYED")]
df_hw$DEP_HOUR_BIN <- floor(df_hw$SCHED_DEP_HOUR)
df_hw <- aggregate(IS_DELAYED ~ WEEKDAY + DEP_HOUR_BIN, df_hw, mean)
names(df_hw)[3] <- "delay_rate"
df_hw$WEEKDAY <- factor(df_hw$WEEKDAY, levels = weekday_levels)

ggplot(df_hw, aes(x = DEP_HOUR_BIN, y = WEEKDAY, fill = delay_rate)) +
  geom_tile(color = "white") +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  scale_fill_gradient(low = "white", high = "steelblue", labels = pct_label, limits = c(0, 0.4)) +
  labs(title = "Delay rate by weekday and departure hour",
       x = "Planned departure hour", y = "Day of the week", fill = "Delay rate")

```

This heatmap is more useful than the weekday bar chart on its own,
because it shows when delay risk builds up during the day. If later
flights are more often delayed, that can point to delays stacking up as
the day goes on. That is exactly the kind of pattern that a simple
weekday average would hide. So the main conclusion here is that time of
day matters more than weekday alone, and that it makes sense to keep a
departure-time feature in the later models.

```{r carrier_delay_rates}
MIN_N_CARRIER <- 5000

carrier_rate <- aggregate(IS_DELAYED ~ OP_CARRIER, flights_clean, mean)
carrier_n    <- aggregate(IS_DELAYED ~ OP_CARRIER, flights_clean, length)
names(carrier_n)[2] <- "n"

carrier_summary <- merge(carrier_rate, carrier_n, by = "OP_CARRIER")
names(carrier_summary)[2] <- "delay_rate"

carrier_summary <- subset(carrier_summary, n >= MIN_N_CARRIER)
carrier_summary <- carrier_summary[order(-carrier_summary$delay_rate), ]

top_carriers <- head(carrier_summary, 10)

ggplot(top_carriers, aes(x = reorder(OP_CARRIER, delay_rate), y = delay_rate)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Delay rate (>=15m) per airline (top 10, min volume filter)",
       x = "Airline", y = "Delay rate")
```

We use a minimum-volume filter before ranking carriers and airports.
Without that, very small groups can end up at the top just because they
have only a few flights. For this analysis we want stable patterns, not
coincidental extremes.

This is also an explicit assumption in our analysis: we assume that
groups with at least `r MIN_N_CARRIER` flights give a more reliable
picture than very small groups. We choose this cutoff for stability, not
because `r MIN_N_CARRIER` is a universal rule.

The airline chart should therefore be read as a volume-aware comparison.
If some carriers still stand out after the minimum-volume filter, that
is more likely to reflect a real operational pattern rather than noise.
We do not interpret that as a causal airline effect by itself, but it
does justify keeping carrier information in the later supervised models.

```{r origin_delay_rates}
MIN_N_AIRPORT <- 20000

origin_rate <- aggregate(IS_DELAYED ~ ORIGIN, flights_clean, mean)
origin_n    <- aggregate(IS_DELAYED ~ ORIGIN, flights_clean, length)
names(origin_n)[2] <- "n"

origin_summary <- merge(origin_rate, origin_n, by = "ORIGIN")
names(origin_summary)[2] <- "delay_rate"

origin_summary <- subset(origin_summary, n >= MIN_N_AIRPORT)
origin_summary <- origin_summary[order(-origin_summary$delay_rate), ]

top_origins <- head(origin_summary, 15)

ggplot(top_origins, aes(x = reorder(ORIGIN, delay_rate), y = delay_rate)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Delay rate (>=15m) per origin airport (top 15, min volume filter)",
       x = "Origin airport", y = "Delay rate")
```

```{r origin_volume_vs_delay}
origin_top50_vol <- origin_summary[order(-origin_summary$n), ][1:50, ]

ggplot(origin_top50_vol, aes(x = n, y = delay_rate)) +
  geom_point() +
  scale_x_log10() +
  scale_y_continuous(labels = pct_label) +
  labs(title = "Origin airports: volume vs delay rate (top 50 volume)",
       x = "Number of flights (log scale)", y = "Delay rate")
```

This scatterplot helps us separate volume from risk. A large airport
with a high delay rate matters more than a very small airport with a
high but unstable rate, because it affects many more flights. So for the
business side, the most relevant airports are not only those with high
rates, but especially those with both high rates and high traffic.

Again, this rests on an assumption: we treat airports with at least
`r MIN_N_AIRPORT` flights as more useful for comparison because their
rates are less noisy. That cutoff is a practical choice for this
notebook, not a fixed industry standard.

Next we look at two simple relationships: distance and planned departure
time.

```{r distance_hour_delay_risk}
set.seed(46748717)
N_VIS <- min(200000, nrow(flights_clean))
idx_vis <- sample(seq_len(nrow(flights_clean)), N_VIS)
flights_vis <- flights_clean[idx_vis, ]

# First we bin distance so we can see the average delay risk in a simple
# and readable way instead of plotting millions of raw points.
distance_breaks <- pretty(flights_vis$DISTANCE, n = 12)
flights_vis$DISTANCE_BIN <- cut(
  flights_vis$DISTANCE,
  breaks = unique(distance_breaks),
  include.lowest = TRUE
)

# For each distance bin we calculate both the average delay rate and the
# number of flights, so we can ignore bins with almost no observations.
distance_risk <- aggregate(IS_DELAYED ~ DISTANCE_BIN, flights_vis, mean)
distance_mid  <- aggregate(DISTANCE ~ DISTANCE_BIN, flights_vis, mean)
distance_n    <- aggregate(IS_DELAYED ~ DISTANCE_BIN, flights_vis, length)

names(distance_risk)[2] <- "delay_rate"
names(distance_mid)[2]  <- "distance_mid"
names(distance_n)[2]    <- "n"

distance_risk <- merge(distance_risk, distance_mid, by = "DISTANCE_BIN")
distance_risk <- merge(distance_risk, distance_n, by = "DISTANCE_BIN")
distance_risk <- subset(distance_risk, n >= 100)
distance_risk <- distance_risk[order(distance_risk$distance_mid), ]

print(
  ggplot(distance_risk, aes(x = distance_mid, y = delay_rate)) +
    geom_line() +
    geom_point() +
    scale_y_continuous(labels = pct_label, limits = c(0, 1)) +
    labs(title = "Delay risk vs distance",
         x = "Distance (miles)", y = "Delay risk")
)

# We do the same idea for planned departure hour, but now on whole hours
# because that is easier to read than the raw scheduled time.
flights_vis$DEP_HOUR_BIN <- floor(flights_vis$SCHED_DEP_HOUR)

hour_risk <- aggregate(IS_DELAYED ~ DEP_HOUR_BIN, flights_vis, mean)
hour_n    <- aggregate(IS_DELAYED ~ DEP_HOUR_BIN, flights_vis, length)

names(hour_risk)[2] <- "delay_rate"
names(hour_n)[2]    <- "n"

hour_risk <- merge(hour_risk, hour_n, by = "DEP_HOUR_BIN")
hour_risk <- subset(hour_risk, n >= 100)
hour_risk <- hour_risk[order(hour_risk$DEP_HOUR_BIN), ]

print(
  ggplot(hour_risk, aes(x = DEP_HOUR_BIN, y = delay_rate)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = 0:23) +
    scale_y_continuous(labels = pct_label, limits = c(0, 1)) +
    labs(title = "Delay risk vs planned departure time",
         x = "Scheduled departure hour", y = "Delay risk")
)
```

These plots are just meant to explore the pattern. They help us decide
which variables are worth keeping and how to code them. Scheduled
departure time clearly works better as an hour variable than as the raw
HHMM value.

For distance, we mainly look for broad non-linear patterns rather than a
perfect straight-line relationship. For departure hour, we mainly want
to see whether delay risk changes over the day in a structured way. If
that pattern is visible, then the variable deserves a place in the
feature set. So even without a causal claim, these plots are useful
because they show that both route length and time of day capture real
variation in delay risk.

There is also a small plotting assumption here. We only keep bins with
at least 100 flights, because otherwise the average rate per bin becomes
too unstable to read. That choice is meant to make the plot more
reliable, not to change the underlying pattern.

```{r route_risk_summary}
route_summary <- aggregate(IS_DELAYED ~ ORIGIN + DEST, flights_clean, mean)
route_n <- aggregate(IS_DELAYED ~ ORIGIN + DEST, flights_clean, length)
names(route_summary)[3] <- "delay_rate"
names(route_n)[3] <- "n"

route_summary <- merge(route_summary, route_n, by = c("ORIGIN", "DEST"))
route_summary <- subset(route_summary, n >= 3000)
route_summary <- route_summary[order(-route_summary$delay_rate), ]

head(route_summary, 10)
```

This route table adds one more useful view. It shows whether high delay
rates mainly belong to certain airports in general, or to specific
origin-destination combinations. If some routes stay risky even after we
have already seen airport-level differences, that suggests route-level
network structure also matters.

Here too we make a practical assumption by keeping only routes with at
least 3000 flights. The idea is the same as above: we prefer stable
patterns over routes that look extreme only because they are rare.

# 6. Dimension reduction and clustering

PCA and clustering are exploratory steps. We only use them if we can
build a sensible feature set with variables that are known before
departure. If we put in raw times, airport labels, or post-event
variables, the output becomes much less useful.

## 6.1 PCA

We use PCA here to get a quick view of the structure in the data. The
first step is not choosing components yet, but deciding which variables
should even go into PCA and clustering.

| Variable | PCA role | Clustering role | Representation | Why |
|----|----|----|----|----|
| `FL_DATE` | Transform | Transform | Do not use raw date; derive month/weekday information instead | Raw dates create arbitrary scale differences |
| `OP_CARRIER` | Post hoc only | Post hoc only | No dummy expansion in unsupervised step | One-hot encoding carriers would add dimensions and dominate structure |
| `OP_CARRIER_FL_NUM` | Exclude | Exclude | Identifier-like flight number; not analytically meaningful | Mostly identifies a flight record rather than a latent trait |
| `ORIGIN` | Post hoc only | Post hoc only | Retain for post hoc interpretation, not as dummy set | Airport dummies would mainly recover airport labels rather than shared structure |
| `DEST` | Post hoc only | Post hoc only | Retain for post hoc interpretation, not as dummy set | Airport dummies would mainly recover airport labels rather than shared structure |
| `DISTANCE` | Keep | Keep | Use as numeric route-length feature | Operationally meaningful and known before departure |
| `CRS_DEP_TIME` | Transform | Transform | Convert to cyclical sin/cos features | Raw HHMM values are not valid linear distances |
| `CRS_ARR_TIME` | Transform | Transform | Convert to cyclical sin/cos features | Raw HHMM values are not valid linear distances |
| `CRS_ELAPSED_TIME` | Keep | Watch for redundancy | Use as numeric schedule-length feature | Operationally meaningful and known before departure |
| `DEP_TIME / ARR_TIME / WHEELS_*` | Exclude | Exclude | Known only after departure/arrival | Post-event information must be excluded |
| `DEP_DELAY / ARR_DELAY` | Exclude | Exclude | Leakage / target-adjacent outcome information | Would force PCA/clustering to recover realised delay instead of latent structure |
| Delay-cause columns | Exclude | Exclude | Recorded after the event and therefore leakage | Pure leakage for pre-departure analysis |
| `CANCELLED / DIVERTED` | Exclude | Exclude | Used only as sample filter, not as features | These variables already define which rows remain in the analysis |
| `MONTH` | Transform | Transform | Convert to cyclical sin/cos features | Month 12 is adjacent to month 1, so linear coding is inappropriate |
| `WEEKDAY` | Transform | Transform | Represent coarsely as weekend dummy | A low-dimensional weekend indicator is preferable to seven weekday dummies |

This screening leads to two main choices. First, we do not put raw
carrier and airport labels into PCA or clustering, because they would
mostly make the analysis split on labels. Second, time variables need to
be transformed first, because time is cyclical and not linear.

In this step we prepare the variables for PCA and clustering. We convert
time variables into a cyclical form so that times such as 23:00 and
00:00 stay close together. This matters because raw clock times do not
work well here as simple numeric values. We also convert `WEEKDAY` into
a simple weekend indicator so the feature space stays manageable.

The table below then summarizes what we did for each variable and why.

```{r pca_cluster_preparation, echo=FALSE}
flights_structure <- flights_clean

# Time variables are cyclical, so we convert them to sine/cosine pairs
# before PCA and clustering. This keeps 23:00 and 00:00 close together.
flights_structure$DEP_HOUR_SIN <- sin(2 * pi * flights_structure$SCHED_DEP_HOUR / 24)
flights_structure$DEP_HOUR_COS <- cos(2 * pi * flights_structure$SCHED_DEP_HOUR / 24)
flights_structure$ARR_HOUR_SIN <- sin(2 * pi * flights_structure$SCHED_ARR_HOUR / 24)
flights_structure$ARR_HOUR_COS <- cos(2 * pi * flights_structure$SCHED_ARR_HOUR / 24)
flights_structure$MONTH_SIN <- sin(2 * pi * flights_structure$MONTH / 12)
flights_structure$MONTH_COS <- cos(2 * pi * flights_structure$MONTH / 12)

# Weekday is simplified to weekend versus non-weekend to keep the
# unsupervised feature space small and interpretable.
flights_structure$IS_WEEKEND <- ifelse(
  flights_structure$WEEKDAY %in% c("Saturday", "Sunday"), 1, 0
)
```

| Feature | Preparation | Why |
|----|----|----|
| `DISTANCE` | Z-score scaling only | This is already a meaningful numeric variable, so we only need to standardize the scale |
| `CRS_ELAPSED_TIME` | Z-score scaling only | This is also a meaningful numeric variable, but we keep an eye on overlap with distance |
| `SCHED_DEP_HOUR` | Cyclical encoding (`sin`/`cos`) + scaling | Raw HHMM values would create false distances between times |
| `SCHED_ARR_HOUR` | Cyclical encoding (`sin`/`cos`) + scaling | Raw HHMM values would create false distances between times |
| `MONTH` | Cyclical encoding (`sin`/`cos`) + scaling | Month 12 and month 1 should be close together, so linear coding is not ideal |
| `WEEKDAY` | Weekend dummy + scaling | A simple weekend indicator keeps the feature space small and clear |

We use Z-score standardization here because PCA and distance-based
clustering are sensitive to scale. This puts the variables on a more
equal footing. We do not add an extra log transformation for `DISTANCE`
or `CRS_ELAPSED_TIME`, because we want to keep those variables easy to
interpret. This is also an explicit assumption: we assume that keeping
the variables easy to interpret is more useful here than adding another
transformation layer just to chase a slightly neater distribution.

```{r pca_cluster_feature_space}
# Here we define the variables that are allowed to shape the PCA and
# clustering space, plus a few label variables kept only for later interpretation.
representation_vars <- c(
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "DEP_HOUR_SIN", "DEP_HOUR_COS",
  "ARR_HOUR_SIN", "ARR_HOUR_COS",
  "MONTH_SIN", "MONTH_COS",
  "IS_WEEKEND"
)

post_hoc_vars <- c("IS_DELAYED", "OP_CARRIER", "ORIGIN", "DEST")

structure_df <- flights_structure[, c(representation_vars, post_hoc_vars)]
structure_df <- na.omit(structure_df)

# We sample a manageable subset and standardize the feature space so no
# variable dominates only because of its scale.
set.seed(46748717)
N_STRUCTURE <- min(50000, nrow(structure_df))
idx_structure <- sample(seq_len(nrow(structure_df)), N_STRUCTURE)

structure_sample <- structure_df[idx_structure, ]
structure_z <- as.data.frame(scale(structure_sample[, representation_vars]))

dim(structure_sample)
summary(structure_z)
```

## 6.2 Correlation heatmap

```{r correlation_heatmap}
cor_mat <- round(cor(structure_z), 2)
cor_mat

reshape2::melt(cor_mat) |>
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "firebrick", mid = "white", high = "steelblue", midpoint = 0) +
  labs(title = "Correlation heatmap of PCA/clustering feature space",
       x = "", y = "", fill = "Correlation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r correlation_pairs}
high_corr_idx <- which(abs(cor_mat) >= 0.75 & lower.tri(cor_mat), arr.ind = TRUE)

if (nrow(high_corr_idx) == 0) {
  high_corr_pairs <- data.frame(
    variable_1 = character(),
    variable_2 = character(),
    correlation = numeric()
  )
} else {
  high_corr_pairs <- data.frame(
    variable_1 = rownames(cor_mat)[high_corr_idx[, 1]],
    variable_2 = colnames(cor_mat)[high_corr_idx[, 2]],
    correlation = cor_mat[high_corr_idx],
    row.names = NULL
  )
}

high_corr_pairs
```

This heatmap is a quick check before PCA and clustering. If variables
are strongly correlated, they can capture the same signal twice. That is
especially relevant for clustering, where this can distort the
distances. The next step follows directly from this: once we know which
variables overlap strongly, we can judge more clearly whether PCA is
useful and whether clustering may be double-counting some structure.

The cutoff of `|correlation| >= 0.75` in the table below is an explicit
assumption. We use it as a practical signal for "strong enough to pay
attention", not as a hard statistical law.

## 6.3 Principal component analysis

```{r pca_run}
pca_results <- prcomp(structure_z, center = FALSE, scale. = FALSE)

pca_var_explained <- (pca_results$sdev)^2 / sum((pca_results$sdev)^2)
pca_var_table <- data.frame(
  PC_num = seq_along(pca_var_explained),
  PC = paste0("PC", seq_along(pca_var_explained)),
  Variance_Explained = pca_var_explained,
  Cumulative = cumsum(pca_var_explained)
)

pca_var_table
```

```{r pca_variance_plots}
ggplot(pca_var_table, aes(x = PC, y = Variance_Explained)) +
  geom_col(fill = "steelblue", width = 0.7) +
  geom_text(aes(label = paste0(round(100 * Variance_Explained, 1), "%")),
            vjust = -0.4, size = 3) +
  ylim(0, max(pca_var_table$Variance_Explained) * 1.15) +
  labs(title = "Scree plot",
       x = "Principal component", y = "Variance explained")

ggplot(pca_var_table, aes(x = PC_num, y = Cumulative)) +
  geom_line(color = "steelblue") +
  geom_point(size = 2, color = "steelblue") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey50") +
  scale_x_continuous(breaks = pca_var_table$PC_num, labels = pca_var_table$PC) +
  scale_y_continuous(labels = pct_label, limits = c(0, 1)) +
  labs(title = "Cumulative explained variance",
       x = "Principal component", y = "Cumulative variance explained")
```

```{r pca_loadings}
top_loadings <- function(pca_obj, pc = 1, n = 5) {
  out <- data.frame(
    variable = names(pca_obj$rotation[, pc]),
    loading = pca_obj$rotation[, pc]
  )
  out$abs_loading <- abs(out$loading)
  out <- out[order(-out$abs_loading), ]
  head(out, n)
}

top_loadings(pca_results, pc = 1, n = 8)
top_loadings(pca_results, pc = 2, n = 8)
top_loadings(pca_results, pc = 3, n = 8)

plot_top_loadings <- function(pca_obj, pc = 1, n = 8) {
  out <- top_loadings(pca_obj, pc = pc, n = n)
  out$variable <- factor(out$variable, levels = rev(out$variable))

  ggplot(out, aes(x = variable, y = loading, fill = loading > 0)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    labs(title = paste("Main loadings for PC", pc),
         x = "Variable", y = "Loading")
}

print(plot_top_loadings(pca_results, pc = 1, n = 8))
print(plot_top_loadings(pca_results, pc = 2, n = 8))
```

The point of PCA here is not to replace the original variables straight
away. It is mainly a way to see whether a few components already capture
most of the structure, and which variables drive those components. The
scree plot shows how much variance each component explains, and the
loading plots show which variables matter most inside each component.

So the scree plot is mainly about compression: do we need many
dimensions, or do a few already summarize most of the structure? The
loading plots are mainly about interpretation: is a component mostly
capturing route length, schedule timing or calendar effects? Together,
these graphs help us decide whether there is strong overlap between
variables and whether PCA teaches us something useful about the feature
space.

The dashed line at 80% cumulative explained variance is also a chosen
baseline. We use 0.80 because it is a common rule of thumb in PCA: if a
small number of components explains around 80% of the variance, that is
often seen as a reasonably compact summary. We do not treat it as a hard
pass/fail rule. It is simply a practical benchmark that helps us decide
whether the PCA looks compact enough to be useful. The next step then
follows naturally: if the variance is not captured compactly enough, we
keep the original variables as the main representation.

```{r pca_followup_decisions, include=FALSE}
cum4 <- pca_var_table$Cumulative[min(4, nrow(pca_var_table))]
distance_elapsed_cor <- cor(structure_z$DISTANCE, structure_z$CRS_ELAPSED_TIME)
```

The first four principal components explain `r round(100 * cum4, 1)`% of
the variance. That is helpful for understanding the data, but not enough
to drop the original variables. For the later models, the original
variables are still easier to explain.

The correlation between `DISTANCE` and `CRS_ELAPSED_TIME` is
`r round(distance_elapsed_cor, 2)`.
`r if (abs(distance_elapsed_cor) >= 0.85) "That overlap is quite strong, so we do not want both variables to weigh too heavily in the clustering step." else "That overlap is not extreme, so we can keep both for now, but we still read the clustering results with that overlap in mind."`

We also leave carrier and airport dummies out of PCA and clustering,
because they would mainly separate labels instead of showing broader
patterns.

The `0.85` cutoff here is also an assumption. We use it as a practical
line for "very strong overlap" when deciding whether two variables may
dominate the clustering distance too much. It is not an exact rule, but
it makes the screening decision transparent.

## 6.4 Clustering

Clustering answers a different question than PCA. Here we use it to see
whether flights fall into a few recurring types based on route length,
timing and simple calendar structure. These clusters are not delay
classes. They are better seen as flight profiles.

```{r clustering_input}
cluster_vars <- c(
  "DISTANCE",
  "DEP_HOUR_SIN", "DEP_HOUR_COS",
  "ARR_HOUR_SIN", "ARR_HOUR_COS",
  "MONTH_SIN", "MONTH_COS",
  "IS_WEEKEND"
)

if (abs(distance_elapsed_cor) < 0.85) {
  cluster_vars <- c("DISTANCE", "CRS_ELAPSED_TIME", setdiff(cluster_vars, "DISTANCE"))
}

set.seed(46748717)
N_CLUSTER <- min(20000, nrow(structure_sample))
idx_cluster <- sample(seq_len(nrow(structure_sample)), N_CLUSTER)

cluster_data_z <- structure_z[idx_cluster, cluster_vars, drop = FALSE]
cluster_data_original <- flights_structure[idx_structure[idx_cluster], c(
  "DISTANCE", "CRS_ELAPSED_TIME", "SCHED_DEP_HOUR", "SCHED_ARR_HOUR",
  "IS_WEEKEND", "IS_DELAYED", "OP_CARRIER", "ORIGIN", "DEST"
)]
```

```{r elbow_plot}
wss <- sapply(2:6, function(k) {
  kmeans(cluster_data_z, centers = k, nstart = 20)$tot.withinss
})

elbow_df <- data.frame(k = 2:6, tot_withinss = wss)

ggplot(elbow_df, aes(x = k, y = tot_withinss)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 2:6) +
  labs(title = "Elbow plot for k-means", x = "Number of clusters", y = "Total within-cluster sum of squares")
```

We use the elbow plot only as a guide. There is no single correct number
of clusters, so we choose a value that is still easy to interpret. The
main thing to look for is whether the curve starts flattening out. If it
does, adding more clusters gives less extra structure and mostly adds
complexity. So the next step follows from that logic: once the gain
starts flattening out, it makes more sense to pick a moderate number of
clusters than to force many small groups.

```{r kmeans_clustering}
set.seed(46748717)
K_CHOSEN <- 4
kmeans_result <- kmeans(cluster_data_z, centers = K_CHOSEN, nstart = 25)
cluster_profile_df <- cluster_data_original
cluster_profile_df$cluster <- factor(kmeans_result$cluster)
cluster_profile_df$IS_DELAYED <- as.numeric(cluster_profile_df$IS_DELAYED)
```

The earlier projection on the first two dimensions looked harder to read
than it was worth. That shape is partly caused by the cyclical encoding
of time variables with sine and cosine. That is common in this kind of
analysis, because 23:00 and 00:00 should be treated as close together.
For the clustering itself that is useful, but for explaining the
clusters it is often clearer to go back to the original variables.

So for our report we do not use that projection as the main cluster
plot. Even though the cyclical encoding is the better choice for the
actual clustering, the projected shape is hard to read. You cannot
easily see from that figure what a cluster really means in practice.

That is why we switched to more recognizable cluster plots in the
original variables. These make it easier to see whether a cluster mainly
contains shorter or longer flights, and whether it is more focused on
morning, afternoon or evening departures. For this project that is a
clearer way to explain the clusters.

The choice of `K_CHOSEN = 4` is an explicit assumption as well. We pick
four clusters because it looks like a reasonable balance between some
separation in the elbow plot and still being easy to explain in the
report. We do not claim that four is the single true number of flight
types in the data.

```{r cluster_profiles}

cluster_sizes <- data.frame(
  cluster = factor(names(table(cluster_profile_df$cluster))),
  n = as.integer(table(cluster_profile_df$cluster))
)

cluster_sizes

ggplot(cluster_sizes, aes(x = cluster, y = n)) +
  geom_col() +
  labs(title = "Number of flights per cluster",
       x = "Cluster", y = "Number of flights")

cluster_summary <- aggregate(
  cbind(DISTANCE, CRS_ELAPSED_TIME, SCHED_DEP_HOUR, SCHED_ARR_HOUR, IS_WEEKEND, IS_DELAYED) ~ cluster,
  data = cluster_profile_df,
  FUN = mean
)

cluster_summary

ggplot(cluster_profile_df, aes(x = cluster, y = DISTANCE, fill = cluster)) +
  geom_boxplot(show.legend = FALSE, outlier.shape = NA) +
  labs(title = "Distance per cluster",
       x = "Cluster", y = "Distance (miles)")

ggplot(cluster_profile_df, aes(x = cluster, y = SCHED_DEP_HOUR, fill = cluster)) +
  geom_boxplot(show.legend = FALSE, outlier.shape = NA) +
  labs(title = "Scheduled departure hour per cluster",
       x = "Cluster", y = "Scheduled departure hour")
```

The cluster summaries are mainly descriptive. If the clusters differ a
lot in route length and timing, but only a little in delay rate, then
they are telling us more about flight types than about delay risk. The
cluster size plot also matters here, because it tells us whether we are
looking at broad, meaningful groups or very small niche groups. The
boxplots help with the interpretation: they show in plain terms whether
a cluster contains shorter or longer flights and whether it is more of a
morning or later-day cluster.

Because the cluster numbers themselves have no direct meaning, we also
translate them into simple archetypes. This is an interpretation step,
so it is based on the average values in each cluster rather than on a
hard external label. The goal is simply to make the clusters easier to
talk about in the report.

```{r cluster_archetypes, include=FALSE}
describe_dep_period <- function(x) {
  if (x < 10) {
    "morning"
  } else if (x < 14) {
    "midday"
  } else if (x < 18) {
    "afternoon"
  } else {
    "evening"
  }
}

distance_breaks <- quantile(
  cluster_summary$DISTANCE,
  probs = c(0, 1/3, 2/3, 1),
  na.rm = TRUE
)
distance_breaks <- unique(distance_breaks)

if (length(distance_breaks) < 4) {
  distance_label <- rep("mixed-distance", nrow(cluster_summary))
} else {
  distance_label <- cut(
    cluster_summary$DISTANCE,
    breaks = distance_breaks,
    include.lowest = TRUE,
    labels = c("short-distance", "medium-distance", "long-distance")
  )
}

delay_label <- ifelse(
  cluster_summary$IS_DELAYED >= median(cluster_summary$IS_DELAYED),
  "slightly higher delay risk",
  "slightly lower delay risk"
)

weekend_label <- ifelse(
  cluster_summary$IS_WEEKEND >= median(cluster_summary$IS_WEEKEND),
  "with a relatively larger weekend share",
  "with a relatively larger weekday share"
)

cluster_archetypes <- data.frame(
  cluster = as.character(cluster_summary$cluster),
  archetype = paste(as.character(distance_label),
                    sapply(cluster_summary$SCHED_DEP_HOUR, describe_dep_period),
                    "flight profile"),
  description = paste(
    "This cluster mainly groups",
    as.character(distance_label),
    "flights that depart mostly in the",
    sapply(cluster_summary$SCHED_DEP_HOUR, describe_dep_period),
    "period,",
    weekend_label,
    "and with",
    delay_label,
    "than the other clusters."
  ),
  stringsAsFactors = FALSE
)

cluster_archetype_text <- paste0(
  "Cluster ", cluster_archetypes$cluster, " looks most like a ",
  cluster_archetypes$archetype, ". ",
  cluster_archetypes$description
)
```

We do not treat these clusters as hard labels. They are just a practical
way to describe the main flight types that appear in the sample. That is
enough for this workbook, because the goal of clustering here is
interpretation rather than building a final segmentation system.

-   `r cluster_archetype_text[1]`

-   `r cluster_archetype_text[2]`

-   `r cluster_archetype_text[3]`

-   `r cluster_archetype_text[4]`

These short archetypes should be read as summaries, not as exact rules.
They help us explain what each cluster mainly captures in terms of
distance, departure timing, weekday versus weekend mix, and relative
delay risk.

```{r cluster_density_plots}
cluster_density_df <- cluster_profile_df

density_vars <- c("DISTANCE", "CRS_ELAPSED_TIME", "SCHED_DEP_HOUR", "SCHED_ARR_HOUR")
density_vars <- intersect(density_vars, names(cluster_density_df))

cluster_density_long <- reshape2::melt(
  cluster_density_df[, c("cluster", density_vars)],
  id.vars = "cluster"
)

ggplot(cluster_density_long, aes(x = value, color = cluster)) +
  geom_density() +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Cluster distributions for key continuous variables",
       x = "Value", y = "Density", color = "Cluster")
```

These density plots help us check whether the clusters really differ in
their underlying distributions and not only in their averages. We only
use them for continuous variables, because they are not very useful for
binary ones like `IS_WEEKEND`. If two clusters have very similar density
curves, that is a sign that the clustering may not add much practical
insight for that variable. If the curves separate clearly, the clusters
are easier to explain and defend.

```{r hierarchical_clustering}
set.seed(46748717)
N_HCLUST <- min(150, nrow(cluster_data_z))
idx_hclust <- sample(seq_len(nrow(cluster_data_z)), N_HCLUST)

hclust_input <- cluster_data_z[idx_hclust, , drop = FALSE]
hclust_dist <- dist(hclust_input, method = "euclidean")
hclust_result <- hclust(hclust_dist, method = "ward.D2")

plot(hclust_result, labels = FALSE, hang = -1,
     main = "Hierarchical clustering dendrogram (sample)",
     xlab = "", sub = "", ylab = "Height")
```

Hierarchical clustering is only used on a small sample. We do not choose
it as the main clustering method for three reasons. First, on a dataset
this size it becomes too heavy, because it needs all pairwise distances.
Second, the final dendrogram is harder to translate into a small number
of clear flight types than the k-means output. Third, the result depends
strongly on the chosen sample and representation, so it works better
here as a visual check than as the main segmentation.

So our clustering conclusion is fairly simple. The clusters are useful
to describe flight types, but we do not use them directly in the main
prediction models. For that, the original engineered variables are still
the better choice.

# 7. Modelling

In this section we start to model our data. The goal is to compare
models that are clearly useful for our delay problem. The focus will be
mainly on the delayed class, this is because missing flagging a flight
that has a real delay is worse than giving a flagging a flight and it
eventually not having a delay. It must be noted that if this is the case
it could seem that it is better to always install extra crew and be more
prepared, however this is a trade-off for costs it is important to weigh
these and find an optimum. This is hard to do without the data for these
costs that is why it is chosen to punish take missing high risk a bit
more but not unrealistically much.

## 7.1 Classification setup

We prepare a modelling data set by splitting our data set into training
and testing data. The training data is as always used to estimate the
classification rules, while the testing data is held back to inspect how
the model behaves on unseen flights.

## 7.2 Preparing the modelling data

```{r prepare_modelling_data}
flights_model <- flights_clean

flights_model$IS_DELAYED_15 <- ifelse(flights_model$ARR_DELAY >= 15, "Delayed", "On_time")
flights_model$IS_DELAYED_15 <- factor(flights_model$IS_DELAYED_15)

flights_model$DEP_HOUR <- as.numeric(format(flights_model$CRS_DEP_TIME, "%H")) +
                          as.numeric(format(flights_model$CRS_DEP_TIME, "%M")) / 60

flights_model$ARR_HOUR <- as.numeric(format(flights_model$CRS_ARR_TIME, "%H")) +
                          as.numeric(format(flights_model$CRS_ARR_TIME, "%M")) / 60

factor_vars <- c("OP_CARRIER", "ORIGIN", "DEST", "WEEKDAY")

for (v in factor_vars) {
  if (v %in% names(flights_model)) {
    flights_model[[v]] <- factor(flights_model[[v]])
  }
}

model_vars <- c(
  "IS_DELAYED_15",
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "DEP_HOUR",
  "ARR_HOUR",
  "MONTH",
  "WEEKDAY",
  "OP_CARRIER",
  "ORIGIN",
  "DEST"
)

model_df <- flights_model[, model_vars]
model_df <- na.omit(model_df)

str(model_df)
summary(model_df$IS_DELAYED_15)

set.seed(46748717)

N_MODEL <- min(300000, nrow(model_df))   # adjust if needed
idx <- sample(seq_len(nrow(model_df)), N_MODEL)

model_df_small <- model_df[idx, ]
dim(model_df_small)

set.seed(46748717)

split <- rsample::initial_split(model_df_small, prop = 0.7)
training <- training(split)
testing  <- testing(split)

dim(training)
dim(testing)
```

We keep the feature set simple and close to the workbook. All variables
used here are known before departure and are easy to explain. The next
step follows directly from the earlier exploration: because the EDA
showed useful patterns for time, distance, carrier and calendar
variables, we carry those into the supervised models.

There are also explicit assumptions here. We cap the modelling sample at
`r N_MODEL` rows to keep the notebook workable, and we use a 70/30
train-test split as a standard evaluation setup. Neither choice is the
only possible one, but both are common and easy to explain.

```{r training_testing_comparison}
split_compare <- rbind(
  data.frame(split = "Training", training[, c("IS_DELAYED_15", "DISTANCE", "DEP_HOUR")]),
  data.frame(split = "Testing", testing[, c("IS_DELAYED_15", "DISTANCE", "DEP_HOUR")])
)

ggplot(split_compare, aes(x = IS_DELAYED_15, fill = split)) +
  geom_bar(position = "dodge") +
  labs(title = "Class balance in training and testing data",
       x = "Delay class", y = "Number of flights")

ggplot(split_compare, aes(x = DISTANCE, fill = split)) +
  geom_histogram(bins = 30, color = "white") +
  facet_wrap(~ split, ncol = 1) +
  labs(title = "Distance distribution: training vs testing",
       x = "Distance (miles)", y = "Count")

ggplot(split_compare, aes(x = DEP_HOUR, fill = split)) +
  geom_histogram(bins = 24, color = "white") +
  facet_wrap(~ split, ncol = 1) +
  labs(title = "Departure-hour distribution: training vs testing",
       x = "Scheduled departure hour", y = "Count")
```

These checks are simple but important. If training and testing look very
different, then the model comparison is less reliable. Here we mainly
want to see whether the split produced two similar samples.

```{r model_eval_helpers, include=FALSE}
# This helper function lets us calculate the same evaluation measures in
# exactly the same way for every model we compare later on.
metric_summary <- function(actual, pred) {
  actual <- factor(actual, levels = c("On_time", "Delayed"))
  pred   <- factor(pred, levels = c("On_time", "Delayed"))

  tp <- sum(actual == "Delayed" & pred == "Delayed")
  tn <- sum(actual == "On_time" & pred == "On_time")
  fp <- sum(actual == "On_time" & pred == "Delayed")
  fn <- sum(actual == "Delayed" & pred == "On_time")

  accuracy  <- (tp + tn) / (tp + tn + fp + fn)
  recall    <- ifelse(tp + fn == 0, NA, tp / (tp + fn))
  precision <- ifelse(tp + fp == 0, NA, tp / (tp + fp))
  f1        <- ifelse(
    is.na(recall) | is.na(precision) | (recall + precision) == 0,
    NA,
    2 * recall * precision / (recall + precision)
  )

  data.frame(
    accuracy = accuracy,
    recall_delayed = recall,
    precision_delayed = precision,
    f1_delayed = f1,
    tp = tp, tn = tn, fp = fp, fn = fn
  )
}

# Hard voting means each model gets one vote and the majority decides.
hard_vote <- function(pred_df) {
  delayed_votes <- rowSums(pred_df == "Delayed")
  factor(
    ifelse(delayed_votes >= ceiling(ncol(pred_df) / 2), "Delayed", "On_time"),
    levels = c("On_time", "Delayed")
  )
}
```

## 7.3 Model 1: C5.0 decision tree

```{r c50_tree_model}

model_tree <- C50::C5.0(IS_DELAYED_15 ~ ., data = training)

summary(model_tree)
```

The decision tree is our first baseline because it is easy to explain.
We can see which variables are used first and how the data gets split.

```{r tree_conditions_plot}
tree_plot_model <- rpart::rpart(
  IS_DELAYED_15 ~ .,
  data = training,
  method = "class",
  control = rpart::rpart.control(maxdepth = 3, minsplit = 3000, cp = 0.001)
)

if (nrow(tree_plot_model$frame) > 1 &&
    any(tree_plot_model$frame$var != "<leaf>")) {
  plot(tree_plot_model, uniform = TRUE, margin = 0.1)
  text(tree_plot_model, use.n = TRUE, all = TRUE, cex = 0.7)
} else {
  cat("The shallow rpart check also stayed at the root node, so there are no split conditions to plot here.\n")
}
```

The C5.0 summary is still the main output, but this extra shallow tree
makes the first split rules easier to read.

One thing that can happen with this baseline tree is that it stays very
small, or even collapses to a single root node. If that happens, the
reason is usually not that the code failed, but that the model does not
see any split that improves enough on the majority-class prediction. In
our data, the delayed class is clearly smaller than the on-time class,
so a very conservative tree can reduce a lot of training error by simply
predicting `On_time` for almost everything.

That is an important result in itself. It tells us that a plain baseline
tree is too conservative for this problem and too focused on overall
error. This fits the business setting only partly, because for us a
missed real delay matters more than a false alarm. That is exactly why
the next steps follow: after this baseline, we test approaches such as
boosting and cost-sensitive classification that can push the model to
pay more attention to delayed flights.

## 7.4 Inspecting tree behaviour on unseen data

```{r c50_tree_test_evaluation}

pred_test <- predict(model_tree, testing)

gmodels::CrossTable(
  testing$IS_DELAYED_15,
  pred_test,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

# Accuracy
mean(pred_test == testing$IS_DELAYED_15)

actual <- testing$IS_DELAYED_15
pred   <- pred_test

tp <- sum(actual == "Delayed" & pred == "Delayed")
tn <- sum(actual == "On_time" & pred == "On_time")
fp <- sum(actual == "On_time" & pred == "Delayed")
fn <- sum(actual == "Delayed" & pred == "On_time")

accuracy  <- (tp + tn) / (tp + tn + fp + fn)
recall    <- tp / (tp + fn)
precision <- tp / (tp + fp)
f1_tree   <- 2 * recall * precision / (recall + precision)

data.frame(
  accuracy = accuracy,
  recall_delayed = recall,
  precision_delayed = precision,
  f1_delayed = f1_tree,
  tp = tp, tn = tn, fp = fp, fn = fn
)
```

The confusion matrix matters because accuracy alone is not enough. For
this problem, false negatives matter most: flights predicted as on time
that are actually delayed. So recall is especially important here. This
is why the next model steps follow: after the baseline tree, we
deliberately try approaches that may change that recall/precision
balance.

## 7.5 Altering tree behaviour: boosting

```{r boosted_c50_evaluation}
model_tree_boost <- C50::C5.0(IS_DELAYED_15 ~ ., data = training, trials = 10)

pred_boost <- predict(model_tree_boost, testing)

gmodels::CrossTable(
  testing$IS_DELAYED_15,
  pred_boost,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

tp_boost <- sum(actual == "Delayed" & pred_boost == "Delayed")
tn_boost <- sum(actual == "On_time" & pred_boost == "On_time")
fp_boost <- sum(actual == "On_time" & pred_boost == "Delayed")
fn_boost <- sum(actual == "Delayed" & pred_boost == "On_time")

data.frame(
  model = "Boosted C5.0",
  accuracy = (tp_boost + tn_boost) / (tp_boost + tn_boost + fp_boost + fn_boost),
  recall_delayed = tp_boost / (tp_boost + fn_boost),
  precision_delayed = tp_boost / (tp_boost + fp_boost),
  f1_delayed = 2 * (tp_boost / (tp_boost + fn_boost)) * (tp_boost / (tp_boost + fp_boost)) /
    ((tp_boost / (tp_boost + fn_boost)) + (tp_boost / (tp_boost + fp_boost)))
)
```

Boosting is useful here because it lets the model correct some mistakes
from earlier trees. We keep the number of trials low so the workflow
stays simple. The choice of 10 trials is an explicit assumption: enough
to see whether repeated trees help, but still simple and workbook-like.

## 7.6 Altering tree behaviour: cost-sensitive classification

```{r cost_sensitive_c50_evaluation}
cost_matrix <- matrix(c(NA, 1,
                        2, NA),
                      nrow = 2,
                      ncol = 2,
                      byrow = FALSE)
rownames(cost_matrix) <- colnames(cost_matrix) <- c("Delayed", "On_time")

model_tree_cost <- C50::C5.0(IS_DELAYED_15 ~ ., data = training, costs = cost_matrix)

pred_cost <- predict(model_tree_cost, testing)

gmodels::CrossTable(
  testing$IS_DELAYED_15,
  pred_cost,
  prop.chisq = FALSE,
  prop.c = FALSE,
  prop.r = FALSE,
  prop.t = FALSE,
  dnn = c("Actual delay status", "Predicted delay status")
)

tp_cost <- sum(actual == "Delayed" & pred_cost == "Delayed")
tn_cost <- sum(actual == "On_time" & pred_cost == "On_time")
fp_cost <- sum(actual == "On_time" & pred_cost == "Delayed")
fn_cost <- sum(actual == "Delayed" & pred_cost == "On_time")

data.frame(
  model = "Cost-sensitive C5.0",
  accuracy = (tp_cost + tn_cost) / (tp_cost + tn_cost + fp_cost + fn_cost),
  recall_delayed = tp_cost / (tp_cost + fn_cost),
  precision_delayed = tp_cost / (tp_cost + fp_cost),
  f1_delayed = 2 * (tp_cost / (tp_cost + fn_cost)) * (tp_cost / (tp_cost + fp_cost)) /
    ((tp_cost / (tp_cost + fn_cost)) + (tp_cost / (tp_cost + fp_cost)))
)
```

This step fits the business problem well. We explicitly say that missing
a real delay is worse than a false alarm. That usually increases recall,
even if precision or accuracy drops a bit.

## 7.7 Model 2: Random Forest

```{r random_forest_evaluation}
rf_vars <- c(
  "IS_DELAYED_15",
  "DISTANCE",
  "CRS_ELAPSED_TIME",
  "DEP_HOUR",
  "ARR_HOUR",
  "MONTH",
  "WEEKDAY",
  "OP_CARRIER"
)

training_rf <- training[, rf_vars]
testing_rf  <- testing[, rf_vars]

set.seed(46748717)
model_rf <- randomForest::randomForest(IS_DELAYED_15 ~ ., data = training_rf, ntree = 200)

pred_rf <- predict(model_rf, testing_rf)

tp_rf <- sum(actual == "Delayed" & pred_rf == "Delayed")
tn_rf <- sum(actual == "On_time" & pred_rf == "On_time")
fp_rf <- sum(actual == "On_time" & pred_rf == "Delayed")
fn_rf <- sum(actual == "Delayed" & pred_rf == "On_time")

data.frame(
  model = "Random forest",
  accuracy = (tp_rf + tn_rf) / (tp_rf + tn_rf + fp_rf + fn_rf),
  recall_delayed = tp_rf / (tp_rf + fn_rf),
  precision_delayed = tp_rf / (tp_rf + fp_rf),
  f1_delayed = 2 * (tp_rf / (tp_rf + fn_rf)) * (tp_rf / (tp_rf + fp_rf)) /
    ((tp_rf / (tp_rf + fn_rf)) + (tp_rf / (tp_rf + fp_rf)))
)

importance(model_rf)
varImpPlot(model_rf, main = "Random forest variable importance")
```

The random forest is a second benchmark next to the single tree. Here we
use a slightly simpler feature set, because this `randomForest`
implementation cannot handle predictors with too many categories, such
as `ORIGIN` and `DEST`. We also use 200 trees as a practical benchmark:
large enough to stabilize the forest somewhat, but still light enough
for this notebook.

## 7.8 Comparison

```{r model_comparison_table}
comparison_table <- data.frame(
  model = c("C5.0 tree", "Boosted C5.0", "Cost-sensitive C5.0", "Random forest"),
  accuracy = c(
    accuracy,
    (tp_boost + tn_boost) / (tp_boost + tn_boost + fp_boost + fn_boost),
    (tp_cost + tn_cost) / (tp_cost + tn_cost + fp_cost + fn_cost),
    (tp_rf + tn_rf) / (tp_rf + tn_rf + fp_rf + fn_rf)
  ),
  recall_delayed = c(
    recall,
    tp_boost / (tp_boost + fn_boost),
    tp_cost / (tp_cost + fn_cost),
    tp_rf / (tp_rf + fn_rf)
  ),
  precision_delayed = c(
    precision,
    tp_boost / (tp_boost + fp_boost),
    tp_cost / (tp_cost + fp_cost),
    tp_rf / (tp_rf + fp_rf)
  ),
  f1_delayed = c(
    f1_tree,
    2 * (tp_boost / (tp_boost + fn_boost)) * (tp_boost / (tp_boost + fp_boost)) /
      ((tp_boost / (tp_boost + fn_boost)) + (tp_boost / (tp_boost + fp_boost))),
    2 * (tp_cost / (tp_cost + fn_cost)) * (tp_cost / (tp_cost + fp_cost)) /
      ((tp_cost / (tp_cost + fn_cost)) + (tp_cost / (tp_cost + fp_cost))),
    2 * (tp_rf / (tp_rf + fn_rf)) * (tp_rf / (tp_rf + fp_rf)) /
      ((tp_rf / (tp_rf + fn_rf)) + (tp_rf / (tp_rf + fp_rf)))
  )
)

comparison_table
```

For this case, recall on delayed flights is still the most important
metric. A model with slightly lower accuracy can still be better if it
finds more real delays.

The extra F1 column is useful because it summarizes the balance between
precision and recall in one number. We still do not treat it as the main
metric, but it helps us see when a model improves one of the two at the
expense of the other.

## 7.9 Why one split is not enough

One train-test split gives only one answer. To see whether our
conclusions are stable, we repeat the same evaluation several times on a
smaller fixed sample. We keep this step simple and limit it to three
different model behaviours: a plain tree, a cost-sensitive tree and the
random forest. We do not repeat every model from above, because that
would mainly add runtime without adding much new insight.

```{r repeated_evaluation}
set.seed(46748717)
N_EVAL <- min(60000, nrow(model_df_small))
idx_eval <- sample(seq_len(nrow(model_df_small)), N_EVAL)
eval_df <- model_df_small[idx_eval, ]

# We repeat the same evaluation several times on the same smaller sample
# so we can see how much results move around across random splits.
N_REPEATS <- 8
repeated_results <- data.frame()

for (i in seq_len(N_REPEATS)) {
  set.seed(46748717 + i)
  split_eval <- rsample::initial_split(eval_df, prop = 0.7, strata = IS_DELAYED_15)
  training_eval <- rsample::training(split_eval)
  testing_eval  <- rsample::testing(split_eval)

  # Plain decision tree
  tree_eval <- C50::C5.0(IS_DELAYED_15 ~ ., data = training_eval)
  pred_tree_eval <- predict(tree_eval, testing_eval)
  metrics_tree_eval <- metric_summary(testing_eval$IS_DELAYED_15, pred_tree_eval)
  metrics_tree_eval$model <- "C5.0 tree"
  metrics_tree_eval$iteration <- i

  # Cost-sensitive tree
  cost_eval <- C50::C5.0(IS_DELAYED_15 ~ ., data = training_eval, costs = cost_matrix)
  pred_cost_eval <- predict(cost_eval, testing_eval)
  metrics_cost_eval <- metric_summary(testing_eval$IS_DELAYED_15, pred_cost_eval)
  metrics_cost_eval$model <- "Cost-sensitive C5.0"
  metrics_cost_eval$iteration <- i

  # Random forest on the reduced feature set
  training_eval_rf <- training_eval[, rf_vars]
  testing_eval_rf  <- testing_eval[, rf_vars]
  rf_eval <- randomForest::randomForest(IS_DELAYED_15 ~ ., data = training_eval_rf, ntree = 100)
  pred_rf_eval <- predict(rf_eval, testing_eval_rf)
  metrics_rf_eval <- metric_summary(testing_eval_rf$IS_DELAYED_15, pred_rf_eval)
  metrics_rf_eval$model <- "Random forest"
  metrics_rf_eval$iteration <- i

  # Simple hard-vote ensemble based on the three models above
  pred_ensemble_eval <- hard_vote(data.frame(
    tree = pred_tree_eval,
    cost_sensitive = pred_cost_eval,
    rf = pred_rf_eval
  ))
  metrics_ensemble_eval <- metric_summary(testing_eval$IS_DELAYED_15, pred_ensemble_eval)
  metrics_ensemble_eval$model <- "Hard vote ensemble"
  metrics_ensemble_eval$iteration <- i

  repeated_results <- rbind(
    repeated_results,
    metrics_tree_eval,
    metrics_cost_eval,
    metrics_rf_eval,
    metrics_ensemble_eval
  )
}

# We compare the spread in accuracy and recall, not just the average.
ggplot(repeated_results, aes(x = model, y = accuracy)) +
  geom_boxplot() +
  labs(title = "Accuracy across repeated train-test splits",
       x = "Model", y = "Accuracy") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

ggplot(repeated_results, aes(x = model, y = recall_delayed)) +
  geom_boxplot() +
  labs(title = "Recall across repeated train-test splits",
       x = "Model", y = "Recall for delayed flights") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

aggregate(
  cbind(accuracy, recall_delayed, precision_delayed, f1_delayed) ~ model,
  data = repeated_results,
  FUN = mean
)
```

This step matters because model performance is not just one number. Some
models look more stable across splits, while others move around more.
That is especially important here, because we want a model that behaves
reasonably well and not only on one lucky test set. So when reading
these boxplots, the spread matters almost as much as the center. A model
with slightly lower average performance but much less variation can
still be more trustworthy in practice.

There are explicit assumptions here too: we repeat the evaluation 8
times on a subset of `r N_EVAL` rows. That is not meant as a definitive
validation study, but as a practical way to make instability visible
without making the notebook too heavy.

## 7.10 Looking beyond one threshold

So far we mostly looked at class labels such as "Delayed" or "On_time".
ROC curves and AUC add another view: they show how well the model ranks
flights from lower to higher delay risk across many possible thresholds.
We only include this for models that return usable class probabilities.
That is an important restriction here: the cost-sensitive C5.0 model is
built to change the classification decision under asymmetric costs, so
its output is not treated as an ordinary probability ranking for ROC/AUC
purposes.

```{r roc_auc}
# For ROC/AUC we need predicted probabilities instead of final class labels.
tree_prob <- predict(model_tree, testing, type = "prob")[, "Delayed"]
rf_prob   <- predict(model_rf, testing_rf, type = "prob")[, "Delayed"]

# Each ROC curve shows how the trade-off changes when we move the decision threshold.
roc_tree <- pROC::roc(testing$IS_DELAYED_15, tree_prob,
                      levels = c("On_time", "Delayed"), quiet = TRUE)
roc_rf   <- pROC::roc(testing_rf$IS_DELAYED_15, rf_prob,
                      levels = c("On_time", "Delayed"), quiet = TRUE)

plot(roc_tree, col = "steelblue", main = "ROC curves on the testing data")
lines(roc_rf, col = "darkgreen")
legend("bottomright",
       legend = c("C5.0 tree", "Random forest"),
       col = c("steelblue", "darkgreen"),
       lwd = 2, bty = "n")

# The AUC table gives one compact summary per model.
auc_table <- data.frame(
  model = c("C5.0 tree", "Random forest"),
  auc = c(
    as.numeric(pROC::auc(roc_tree)),
    as.numeric(pROC::auc(roc_rf))
  )
)

auc_table
```

AUC is useful here because it is not tied to one single threshold. We
still keep the confusion matrix and recall as our main practical view,
because the airline eventually needs a yes/no decision. But AUC helps us
check whether a model is at least ranking higher-risk flights above
lower-risk ones. In the ROC plot, curves that sit higher and more to the
left indicate better ranking performance. The AUC table then summarizes
that graph in one value, but the graph itself is useful because it shows
whether model differences are large or only modest.

The cost-sensitive tree is left out of this section on purpose. That is
not because it is unimportant, but because its output is calibrated to a
cost rule rather than to a neutral probability scale. For that model,
the confusion matrix, recall and precision are the more appropriate
evaluation tools.

## 7.11 Model disagreement and simple voting

If models disagree on the same flights, that is useful information. It
usually means those flights are harder to classify. We therefore first
look at how often the main models agree, and then test one simple hard
voting ensemble.

```{r model_disagreement}
# We first put the three main model predictions next to each other to
# see how often they make the same call on the same flight.
pred_patterns <- data.frame(
  tree = pred_test,
  cost_sensitive = pred_cost,
  rf = pred_rf
)

# This summary tells us how much agreement there is between the models.
agreement_summary <- data.frame(
  outcome = c("All three agree", "At least one model disagrees"),
  share = c(
    mean(pred_patterns$tree == pred_patterns$cost_sensitive &
           pred_patterns$cost_sensitive == pred_patterns$rf),
    mean(!(pred_patterns$tree == pred_patterns$cost_sensitive &
             pred_patterns$cost_sensitive == pred_patterns$rf))
  )
)

ggplot(agreement_summary, aes(x = outcome, y = share)) +
  geom_col() +
  scale_y_continuous(labels = pct_label, limits = c(0, 1)) +
  labs(title = "How often do the main models agree?",
       x = "", y = "Share of testing flights")

# Then we create the simple voting ensemble and compare it with the
# individual models in the same metric table.
pred_ensemble <- hard_vote(pred_patterns)
ensemble_metrics <- metric_summary(testing$IS_DELAYED_15, pred_ensemble)

ensemble_metrics

comparison_with_ensemble <- rbind(
  comparison_table,
  data.frame(
    model = "Hard vote ensemble",
    accuracy = ensemble_metrics$accuracy,
    recall_delayed = ensemble_metrics$recall_delayed,
    precision_delayed = ensemble_metrics$precision_delayed,
    f1_delayed = ensemble_metrics$f1_delayed
  )
)

comparison_with_ensemble
```

We keep the ensemble very simple on purpose. Hard voting is easy to
explain: each model gets one vote and the majority decides. That makes
it a good workbook-level ensemble.

The agreement plot should be read as an uncertainty signal. If the
models disagree often, that means many flights are borderline cases and
there is not one obvious answer. The ensemble then becomes interesting
because it may smooth out some of that disagreement, even if it does not
become the single best model on every metric. So the next step follows
logically: once we see disagreement between models, it makes sense to
check whether a simple voting rule gives a more stable compromise.

We do not add weighted voting, stacking or another optimized ensemble
layer here. Those methods would add extra tuning and make it harder to
see whether any improvement comes from the ensemble idea itself or from
extra optimization. We also do not use soft voting, because then we
would need to treat the probabilities from different models as directly
comparable, which is a stronger assumption than we need here.

So the point of this section is not to build the most advanced ensemble.
It is just to check whether a very simple combination changes the
stability or the error balance in a useful way.

```{r conclusion_helpers, include=FALSE}
best_recall_idx <- which.max(comparison_with_ensemble$recall_delayed)[1]
best_auc_idx <- which.max(auc_table$auc)[1]

best_recall_model <- comparison_with_ensemble$model[best_recall_idx]
best_recall_value <- comparison_with_ensemble$recall_delayed[best_recall_idx]
best_precision_value <- comparison_with_ensemble$precision_delayed[best_recall_idx]
best_auc_model <- auc_table$model[best_auc_idx]
best_auc_value <- auc_table$auc[best_auc_idx]
```

# 8. Conclusions and recommendations

## 8.1 Operational recommendations

Our main question was whether we can estimate before departure if a
scheduled flight is likely to be delayed by 15 minutes or more. Based on
this notebook, the answer is yes, but mainly as a risk signal and not as
a perfect prediction.

The data exploration already showed that delay risk is not random. It
changes by month, planned departure time, airline, airport and route.
That means there is enough structure in the pre-departure data to make
useful distinctions between lower-risk and higher-risk flights.

The modelling results support that conclusion. Different models do not
behave exactly the same, but they all show that delayed flights can be
identified better than chance. In our current results, the highest
recall is reached by `r best_recall_model`, with a recall of
`r round(100 * best_recall_value, 1)`%. That matters most for our case,
because missing a real delay is more costly than raising a false alarm.
At the same time, the precision of that model is
`r round(100 * best_precision_value, 1)`%, so this should still be seen
as an early warning tool and not as an automatic decision system.

The ROC/AUC results point in the same direction. The best ranking
performance in our current test results comes from `r best_auc_model`
with an AUC of `r round(best_auc_value, 3)`. This means the model is not
just making one fixed yes/no decision, but is also reasonably able to
put higher-risk flights above lower-risk flights.

So the practical answer to the business question is: with the variables
we know before departure, we can already build a useful delay risk
indicator. That indicator is good enough to help operations teams decide
where to pay extra attention, prepare standby capacity or communicate
earlier with passengers. It is not good enough to replace human
judgment, but it is useful as decision support.

## 8.2 Limitations & next steps

This notebook still has clear limits. We only use historical operational
variables and a first set of engineered features. Weather, network
effects, inbound aircraft information and congestion measures could be
good next steps.

## 8.3 Assignment check

To make it easier to check the assignment, we summarize the main parts
that are already covered in the workbook:

-   Business context, analytics task and success criteria are stated
    explicitly.

-   Data exploration is documented with interpretation rather than plots
    only.

-   Data preparation choices are motivated, including missing values,
    datatype fixes and derived variables.

-   PCA and clustering are only used after variable screening and only
    on a clear pre-departure feature space.

-   Classification models are evaluated with a train-test split and a
    confusion matrix, with explicit attention to recall versus
    precision.

-   Model evaluation goes beyond one split by checking repeated
    evaluation, ROC/AUC and a simple voting ensemble.

-   Operational recommendations, limitations and next steps are stated
    separately.

So overall, the workbook now covers the main analytical parts that would
normally be expected. The last important step is still to knit the file
and check whether the written interpretation matches the final output.
